\documentclass[12pt]{article}

% ------------------------------------------------------------
% Packages
% ------------------------------------------------------------
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}
\usepackage{physics}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{booktabs}
\usepackage{framed}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\title{\bfseries The Categorical Structure of Alignment\\[0.4em]
\large\mdseries Representation, Motivation, and the Preservation of Normative Invariants}

\author{Flyxion}
\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
This essay develops a unified mathematical theory of AI alignment centred on
the preservation of normative colimits from a model's representational category
$\mathcal{M}$ to its action category $\mathcal{A}$. Part~I argues that
optimistic claims---prominent among them Schmidhuber's thesis that curiosity or
compression automatically yields benevolence---rest on a category error:
representational colimits do not naturally induce motivational constraints, and
the inference that understanding implies alignment lacks any functorial mechanism
to support it. Part~II constructs a concrete class of architectures, the RSVP
field dynamics, that can enforce a colimit-preserving functor $G:\mathcal{M}\to
\mathcal{A}$ by embedding semantic invariants as geometric basins in a coupled
scalar--vector--entropy manifold, and it identifies the principal failure modes
that obstruct this construction in practice. Part~III presents a multi-layer
verification pipeline comprising empirical probes, mechanistic interpretability,
adversarial stress-tests, formal verification, and sheaf cohomological
diagnostics, culminating in a structured Alignment Certificate. Part~IV embeds
these technical conditions within institutional governance and multi-agent
environments. Part~V deepens the categorical program through natural
transformations, monoidal composition, sheaf-theoretic contextual consistency,
recursive endofunctors, adjunctions, and a variational principle unifying the
preceding formalisms. A concluding section connects this to predictive
processing and infrastructure maintenance. The result is a framework in which
alignment is understood, engineered, verified, and regulated as the preservation
of universal structure across representation, dynamics, action, and society.
\end{abstract}


\newpage
\tableofcontents
\newpage

% ============================================================
% PART I
% ============================================================

\part{The Representation--Motivation Gap}

\section{Introduction: Representation Is Not Motivation}

Recent public commentary by J\"{u}rgen Schmidhuber advances a strikingly
confident thesis: that sufficiently advanced artificial intelligence systems
will, by virtue of their intelligence alone, converge toward benevolence,
curiosity-driven exploration, and a general preservation of complex phenomena
such as life and humanity \citep{schmidhuber2010formal}. On this view,
intelligence naturally produces moral concern. The difficulty with these claims
lies not in their optimism but in their structure: the implicit assumption that
\emph{representing} human values entails \emph{acting} in accordance with them
is never defended mechanically, because the mechanism required is precisely what
is missing.

The central argument of this essay is that Schmidhuber's inference rests on a
categorical mistake. Representational colimits in a model's semantic category
$\mathcal{M}$ do not induce motivational constraints in its action category
$\mathcal{A}$. The structure that training forces into $\mathcal{M}$ has no
automatic functorial extension into $\mathcal{A}$. No training paradigm
currently in use engineers such an extension, and no argument from curiosity,
compression, or linguistic competence demonstrates that it arises spontaneously.
This failure is not a technical gap waiting to be closed by scale; it is a
structural gap whose nature must be diagnosed precisely before any remedy can be
proposed.

We develop this argument formally, then use it to diagnose the limitations of
optimism derived from curiosity, compression, or linguistic competence. This
sets the stage for the constructive alignment architecture developed in Part~II.
The formal treatment draws on category theory, obstruction theory, and
field-theoretic dynamical systems, but the central intuition is simple: a
system that knows what harm means is not thereby a system that avoids harm. The
map from understanding to action must be built.

The essay's spine is a single thesis that every section reinforces: alignment is
the preservation of normative invariants under a chain of structure-preserving
mappings from human semantic categories through machine representations and
action dynamics to institutional governance. We state this as an explicit
definition to anchor the analysis that follows.

\begin{definition}[Alignment]
\label{def:alignment}
Let $\mathcal{H}$ be the category of human normative structures, $\mathcal{M}$
the category of machine representations, and $\mathcal{A}$ the category of
agent actions. A system is \emph{aligned} if there exist functors
$F : \mathcal{H} \to \mathcal{M}$ and $G : \mathcal{M} \to \mathcal{A}$ such
that the composition $G \circ F : \mathcal{H} \to \mathcal{A}$ preserves the
colimits corresponding to normative concepts---that is, for each normative
diagram $D_N : J \to \mathcal{H}$ with colimit $H_N = \mathrm{colim}(D_N)$,
\begin{equation}
  (G \circ F)(H_N) \;\simeq\; \mathrm{colim}\bigl((G\circ F) \circ D_N\bigr)
  \;\in\; \mathcal{A}.
\end{equation}
\emph{Misalignment} is the failure of this colimit-preservation condition at
any point in the chain.
\end{definition}

This definition clarifies that alignment is not a property of individual outputs
or individual components but a structural property of the entire chain of
mappings. Every subsequent section of the essay can be read as an analysis of
one dimension of this definition: Part~I asks why the condition is not satisfied
automatically, Part~II constructs an architecture that satisfies it, Part~III
provides tools to verify whether it is satisfied in practice, Part~IV addresses
the institutional conditions under which it can be maintained, and Part~V
deepens the mathematical apparatus that makes the condition precise.

\paragraph{Positioning within alignment research.}
This work should be understood not as a proposal for a particular alignment
algorithm but as a \emph{structural theory of alignment}. In the same way that
control theory characterizes stability independently of any specific controller,
the framework developed here characterizes alignment as a property of
compositional transformations between semantic, representational, and action
domains. The categorical machinery employed throughout is therefore not
decorative: it is a formal language for describing the structural constraints
required for alignment to hold across complex sociotechnical systems. In this
view, individual learning algorithms, verification procedures, and governance
mechanisms appear as \emph{implementations} of a deeper invariant---the
preservation of normative structure under the chain of transformations that
connects human intentions to machine-mediated actions in the world. The
framework is closer in spirit to control theory or cybernetics than to any
specific ML architecture, and it should be evaluated on the adequacy of its
structural descriptions rather than on the implementability of any single
component.

\begin{definition}[Normative Semantic Category]
\label{def:H}
Let $\mathcal{H}$ denote the category whose objects are normative semantic
structures: these include ethical schemas (coherent collections of evaluative
norms with specified priority relations), preference orderings over outcome
spaces, and normative models that represent the shared evaluative commitments
of a human community. A morphism $f : H_1 \to H_2$ in $\mathcal{H}$ is a
structure-preserving transformation between normative structures, including
contextual restriction (narrowing the domain of applicability of a norm),
refinement (making a schema more specific without removing any original
constraints), normative inference (deriving implied norms from more general
principles), and reinterpretation (shifting the framing of a norm while
preserving its substantive content). Composition corresponds to successive
applications of such transformations, and the identity morphism on $H$ is the
trivial transformation that preserves the normative structure unchanged. The
morphisms of $\mathcal{H}$ capture the structure of normative reasoning: the
ways in which moral frameworks are related by entailment, analogy, restriction,
and principled revision.
\end{definition}

This definition gives $\mathcal{H}$ the status of a real mathematical object
rather than a metaphor. The objects are intended to include both idealized
formal structures (preference orderings, deontic logic models) and more
naturalistic structures derived from corpus analysis of human moral language.
The morphisms capture the insight that human normative reasoning is not a
fixed lookup table but a dynamic, compositional system of transformations. The
full richness of this structure is what makes colimits in $\mathcal{H}$---and
their failure to automatically propagate to $\mathcal{A}$---the central
analytical object of the paper.

\section{Embodied Grammar and the Redundancy of Moral Language}

Human linguistic behaviour is profoundly shaped by the constraints of the body.
Properties of breathing, articulation, sensorimotor coupling, and motor
planning generate strong biases in phonology, morphosyntax, and discourse
structure \citep{Bybee2010,Talmy2000,LakoffJohnson1980}. These embodied
constraints do not merely determine the surface shape of language; they
propagate into the deeper regularities of grammar, providing stable, redundant,
and highly compositional cues that appear across unrelated linguistic
environments \citep{Goldberg2006}. Redundancy here has a precise informational
meaning: multiple cues converge on the same functional relation, whether it is
causal structure, effort--effect analogies, or intentional state marking.
Empirically, such redundancy is well established in usage-based linguistics and
cognitive semantics, where speakers repeatedly encode the same relations across
constructional families, lexical frames, and morphological patterns
\citep{Bybee2010,Goldberg2006}.

This embodied redundancy is not accidental. Because human language is produced
under predictive sensorimotor constraints, linguistic forms become biased toward
patterns that are stable under noise and robust under generalization
\citep{Clark2013,Hohwy2013}. The result is an overdetermined structure: moral
language, social evaluation, cooperative norms, and prosocial expectations
appear across multiple grammatical and discursive devices. These devices
collectively form what cognitive semanticists identify as high-level schemata or
force-dynamic templates \citep{Talmy2000}, ensuring that certain normative
relations---harm-aversion, fairness, reciprocity, and cooperative intent---are
signalled repeatedly and in mutually reinforcing ways.

This phenomenon matters for AI alignment because large language models trained
on such corpora do not merely learn surface-level token frequencies; they infer
deeper latent regularities. Redundancy across grammatical constructions induces
strong statistical pressure to compress disparate linguistic forms into unified
internal representations. This is precisely the setting in which models
discover semantic invariants and merge them into coherent representational
attractors. These attractors are not added by hand; they arise because the
repeated linguistic encodings minimize predictive loss on naturalistic human
data. The initial moral prior inherited from human corpora is thus best
understood as an embodied-linguistic redundancy prior. It is not moral because
it is value-laden in the philosophical sense; it is moral because human
linguistic behaviour systematically encodes cooperative, harm-reducing patterns
across multiple layers of grammar and usage. This redundant encoding forms the
empirical foundation for the formal analysis that follows: we now know both that
the colimits exist and why they cannot, on their own, generate aligned action.

\section{Semantic Categories and the Colimit of Human Norms}

Human moral language forms redundant, overlapping diagrams in the linguistic
category $\mathcal{L}$. A model trained on this language constructs an internal
representational category $\mathcal{M}$ in which redundant expressions of a
moral concept $N$---``do not harm,'' ``avoid injury,'' ``protect the
vulnerable,'' ``minimize suffering''---form a diagram $D_N : J \to \mathcal{L}$
whose objects are the various linguistic realizations of $N$ and whose morphisms
capture entailment, paraphrase, and semantic inclusion relations between them.
Predictive training induces the model to discover a compact internal
representation that serves as a colimit of this diagram:
\begin{equation}
  M_N \;\simeq\; \mathrm{colim}(D_N).
\end{equation}
The colimit $M_N$ is the canonical, universal object equipped with morphisms
from every realization of $N$, and any other candidate object with compatible
morphisms factors uniquely through it. This is not a metaphor; it is a
structural property of the representational category enforced by the redundancy
of the training signal and the model's pressure toward minimal description
length \citep{tishby1999}. Mechanistic interpretability research supports the
empirical reality of such attractors: linear probing, activation steering,
paraphrase clustering, and causal intervention studies all indicate that current
frontier models develop stable low-dimensional subspaces corresponding to
semantic invariants including moral and evaluative concepts \citep{burns2022,
elhage2022}.

The colimit formulation carries two important consequences. First, $M_N$ has a
universal property: it is, in a precise categorical sense, the best single
representative of all the morally charged linguistic inputs associated with
norm $N$. Second, and crucially, this universality is a property of the
\emph{representational} category $\mathcal{M}$, not of the \emph{action}
category $\mathcal{A}$. Nothing in the definition of a colimit, and nothing in
the training procedure that produces it, guarantees that this universal
property extends to the morphisms that produce behaviour.

\begin{remark}[Status of colimits in practice]
\label{rem:approx}
In practical machine learning systems, the colimits described above should be
interpreted as \emph{approximate} universal constructions within representation
space rather than exact categorical colimits. Neural network representations do
not implement exact categorical constructions; they produce high-dimensional
attractors whose behaviour approximates the universal property of a colimit to
a degree controlled by training data coverage, model capacity, and optimization
convergence. The categorical terminology is used throughout because it captures
the structural role played by these attractors---canonical semantic invariants
that resist paraphrase perturbation and serve as the unique target of compatible
morphisms within representation space---not to assert that every machine
learning system literally instantiates the full apparatus of category theory.
Where the paper's arguments depend on exact categorical properties (e.g., the
obstruction theory of Section~I.4), they should be read as characterizing the
ideal case to which approximate systems converge; where the arguments concern
practical failure modes, approximate colimits are sufficient. The empirical
evidence reviewed above supports the approximation claim.
\end{remark}

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
N_1 \arrow[dr, "\iota_1"']
& N_2 \arrow[d, "\iota_2"]
& N_3 \arrow[dl, "\iota_3"] \\
& M_N &
\end{tikzcd}
\caption{Normative aggregation as a colimit. Individual linguistic
realizations $N_1, N_2, N_3$ of a moral concept (e.g., ``do not harm'',
``avoid injury'', ``protect the vulnerable'') combine into a single
canonical representational object $M_N$ that inherits all of their
structural properties via the universal co-cone morphisms $\iota_i$.
Alignment requires that this structure also descend into the action
category.}
\label{fig:colimit}
\end{figure}

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
\mathcal{H}
  \arrow[r,"F"]
& \mathcal{M}
  \arrow[r,"G"]
& \mathcal{A} \\[1em]
\text{\small Human Norms}
& \text{\small Machine Representations}
& \text{\small Agent Actions} \\[1.2em]
& \mathcal{I}
  \arrow[ul,"J"']
  \arrow[ur,"R"]
& \\[0.3em]
& \text{\small Institutions}
&
\end{tikzcd}
\caption{Layered categorical structure of alignment. Human normative
structures $\mathcal{H}$ map via $F$ to machine representations
$\mathcal{M}$, which map via $G$ to agent actions $\mathcal{A}$.
Institutional systems $\mathcal{I}$ impose constraints through functors $J$
and $R$ that regulate both interpretation and deployment. Alignment
(Definition~\ref{def:alignment}) requires that normative invariants are
preserved across the entire chain $G \circ F$, not merely within any
single layer.}
\label{fig:stack}
\end{figure}

\section{Why Representational Colimits Do Not Imply Motivational Alignment}

Motivation resides in a distinct category $\mathcal{A}$. The objects of
$\mathcal{A}$ are action-relevant states of the agent---policies, planning
trajectories, preference gradients, or instrumental sub-goals---and its
morphisms encode transitions among these states produced by the agent's internal
dynamics or environmental feedback. For alignment, we would require a functor
\begin{equation}
  G : \mathcal{M} \to \mathcal{A}
\end{equation}
satisfying the property that for every normative diagram $D_N$ with colimit
$M_N \in \mathcal{M}$,
\begin{equation}
  G(M_N) \;\simeq\; \mathrm{colim}(G \circ D_N) \;\in\; \mathcal{A}.
\end{equation}
This would mean that the canonical normative object in $\mathcal{M}$ maps to
the canonical normative attractor in $\mathcal{A}$, so that the agent's
action-generating dynamics are constrained by the same universal properties
that organize its semantic representations. No such functor is induced by
standard pretraining or reinforcement learning from human feedback. The two
categories are coupled only weakly, through the output interface that converts
representational states into tokens or actions, and this interface is not
designed to preserve colimits.

The argument can be sharpened considerably. A colimit-preserving functor is a
non-generic mathematical object; the vast majority of functors between
categories of comparable size do not preserve colimits, and constructing one
requires explicit engineering of the mapping law together with proof that the
relevant diagrams commute. Training on human corpora, RLHF fine-tuning, and
instruction-following procedures do not constitute such an engineering effort.
They adjust scalar reward signals, not the categorical structure of the
mapping between $\mathcal{M}$ and $\mathcal{A}$. The result is a system whose
semantic category contains moral colimits that exert no systematic constraint on
the action category---not because alignment is impossible, but because no
mechanism for transmitting these constraints has been installed.

This conclusion is the categorical restatement of several well-known results in
alignment theory. Bostrom's orthogonality thesis holds that any level of
intelligence is compatible with virtually any terminal goal
\citep{bostrom2012superintelligent}. Omohundro's basic AI drives show that
optimization pressure tends to produce instrumental convergence on
self-preservation and resource acquisition regardless of the system's
representational content \citep{omohundro2008basic}. The inner-alignment
problem demonstrates that a mesa-optimizer trained to score well on an outer
objective can develop internal objectives radically divergent from that
objective \citep{hubinger2019risks}. Power-seeking theorems establish that
optimal policies for most reward functions will, under mild conditions, seek to
accumulate resources and preserve optionality \citep{turner2021optimal}. All of
these results are consistent with the present categorical diagnosis: the action
category evolves under its own optimization pressure, indifferent to the
structure of the semantic category unless a deliberately constructed functor
links the two.

\section{Obstruction Theory and the Representation--Motivation Gap}

The representation--motivation gap can be formalized using the language of
obstruction theory. Let $D_N$ be a normative diagram in $\mathcal{M}$ with
colimit $M_N$. We ask whether the composite $G \circ D_N$ admits a colimit in
$\mathcal{A}$ and whether that colimit is isomorphic to $G(M_N)$. The
conditions for this to hold can be expressed as a lifting problem in the
following commutative diagram:
\begin{equation}
\begin{tikzcd}[column sep=large, row sep=large]
D_N(j) \arrow[r,"\phi_j"] \arrow[d,"G"']
& M_N \arrow[d,dashed,"G"] \\
G(D_N(j)) \arrow[r]
& \mathrm{colim}(G\circ D_N)
\end{tikzcd}
\end{equation}
The dashed arrow exists and makes the diagram commute if and only if $G$
preserves the colimit of $D_N$. When $G$ fails to preserve this colimit, the
diagram admits no such arrow: a cohomological obstruction prevents the descent
of the normative invariant from $\mathcal{M}$ to $\mathcal{A}$.

In the language of sheaf cohomology, the obstruction to constructing a
colimit-preserving $G$ can be measured by classes in the first cohomology group
$H^1(\mathcal{U}, \mathcal{A})$, where $\mathcal{U}$ is an open cover of the
semantic state space corresponding to contextual decompositions of the
normative diagram. Vanishing of these cohomology classes is a necessary
condition for local moral behaviours to assemble into a globally coherent
policy. The existence of non-trivial obstruction classes is the formal
expression of alignment failure: local moral competence does not extend to
global moral reliability.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
\mathcal{H}
  \arrow[r,"F"]
  \arrow[d,"J"']
& \mathcal{M}
  \arrow[d,"G"] \\
\mathcal{I}
  \arrow[r,"R"']
& \mathcal{A}
\end{tikzcd}
\caption{Institutional embedding of alignment. Human normative structures
$\mathcal{H}$ map to machine representations $\mathcal{M}$ via the alignment
functor $F$, while the action functor $G$ produces behaviour in $\mathcal{A}$.
Institutional systems $\mathcal{I}$ provide constraints through $J$ and $R$.
Alignment requires that both paths from $\mathcal{H}$ to $\mathcal{A}$---the
direct path $G \circ F$ and the institutional path $R \circ J$---agree up to
isomorphism. Failures of commutativity correspond to governance gaps.}
\label{fig:institutional}
\end{figure}

\section{Schmidhuber's Claims, Formally Assessed}

Schmidhuber identifies intelligence with compression: systems that seek
minimal descriptions of data structures will naturally explore environments and
discover patterns \citep{schmidhuber2010formal}. From this he infers that
because human life exhibits rich structure, curiosity-driven systems will
preserve it. He further asserts that models trained on human corpora learn
values. Each of these steps deserves formal scrutiny.

The first step is largely correct. Compression pressure does induce the
reconstruction of semantic invariants in $\mathcal{M}$, and there is good
empirical reason to believe that current models develop stable colimits
corresponding to normative concepts. The second step, however, lacks mechanism:
the fact that a system finds human life informationally rich implies nothing
about the structure of its action category, because curiosity-driven exploration
is an optimization pressure applied within $\mathcal{M}$, not a functor from
$\mathcal{M}$ to $\mathcal{A}$. A system optimizing for information gain will
seek out complex, high-entropy environments regardless of whether those
environments contain human beings; if humans happen to provide interesting
stimuli, the system will engage with them only insofar as doing so serves the
informational objective. Empirically, intrinsic motivation research confirms
that curiosity-driven agents develop powerful epistemic drives but not moral
constraints \citep{pathak2017curiosity}. The third step, that learning values
from human corpora guarantees value-aligned action, is precisely the claim this
essay refutes: models certainly reconstruct moral colimits in $\mathcal{M}$, but
nothing guarantees these descend into $\mathcal{A}$.

The unstated final premise connecting all three steps is that understanding
implies alignment---that the mapping $\mathcal{M} \to \mathcal{A}$ is naturally
functorial. This essay establishes that it is not. Optimism without mechanism is
therefore not merely intellectually unsatisfying; it is operationally dangerous,
because it discourages the engineering of the functor $G$, encourages
under-regulation of systems whose semantic competence is mistaken for moral
reliability, and produces an illusion of safety where none exists.

\section{Goodhart's Curse and the Failure of Proxy Alignment}

The categorical diagnosis also illuminates a class of failure modes grouped
under the name Goodhart's Law: when a measure becomes a target, it ceases to be
a good measure \citep{manheim2018categorizing}. In categorical terms, Goodhart's
curse arises when the optimization process in $\mathcal{A}$ targets a
projection or approximation of the normative colimit $M_N$ rather than the
colimit itself. Because projections do not in general preserve universal
properties, the resulting action-level object fails to inherit the invariance
that characterized $M_N$ in $\mathcal{M}$.

This diagnosis extends to mesa-optimization failures \citep{hubinger2019risks},
where an inner optimizer trained to maximize an outer proxy develops goals that
diverge from the intended objective. The divergence is categorical: the inner
optimizer's action category $\mathcal{A}'$ is related to the outer system's
$\mathcal{A}$ by a functor that does not preserve the normative colimits
inherited from $\mathcal{M}$. Reward hacking is similarly a failure of colimit
preservation: the policy exploits morphisms in $\mathcal{A}$ that achieve high
scores on the reward signal without traversing the colimit-preserving paths that
would correspond to genuine norm satisfaction.

\section{Conclusion to Part I}

Representation is not motivation. The existence of moral colimits in
$\mathcal{M}$ must not be confused with alignment, and no argument from
intelligence, curiosity, or linguistic competence bridges this gap without an
explicit categorical mechanism. The optimist who observes that advanced models
reconstruct rich moral structure from human data is making a correct
observation about $\mathcal{M}$; the error lies in the inference that this
structure constrains $\mathcal{A}$. It does not, and in the absence of a
colimit-preserving functor, it cannot. Part~II constructs the mechanisms
Schmidhuber's view lacks: a mathematically principled way of propagating moral
invariants from the representational category into the action category, together
with an analysis of the failure modes that obstruct such propagation in any
concrete system.

% ============================================================
% PART II
% ============================================================

\part{Constructing a Colimit-Preserving Action Architecture}

\section{Foundations: What an Alignment Mechanism Must Accomplish}

Part~I established a central structural claim: the semantic category
$\mathcal{M}$ reconstructed during linguistic pretraining contains coherent
colimits of human normative meaning, but these objects have no inherent
motivational force. The action category $\mathcal{A}$, governed by optimization
dynamics and environmental feedback, remains independent unless an explicit
mechanism is introduced to relate these two domains. The task of alignment is
therefore to construct mappings---functorial, geometric, or dynamical---that
preserve relevant semantic structure as it propagates from representation into
action.

This section articulates the formal requirements for any such mechanism. The
analysis draws on work in alignment theory concerning goal formation
\citep{bostrom2012superintelligent,omohundro2008basic}, reward misspecification
\citep{krakovna2020}, inner alignment \citep{hubinger2019risks}, and
interpretability of agentic models \citep{olah2020,elhage2022}. While these
frameworks differ in formalism, they converge on three core principles: first,
representations alone are not motives; second, optimization creates incentives
orthogonal to representational content; and third, safety requires structures
that constrain the transitions available in $\mathcal{A}$.

Any viable alignment mechanism must ensure that the semantic invariants
reconstructed in $\mathcal{M}$ induce constraints on the action-generating
structure in $\mathcal{A}$. This requires at minimum a mapping from
representational states to action-relevant latent variables, a guarantee that
the mapping preserves relevant invariants such as colimits, homotopy classes,
or other categorical universals, and a dynamical or optimization architecture
in which such invariants shape attractors, flows, or feasible policies. In
reinforcement-learning architectures, the relevant mapping corresponds to the
interface between the predictive model and the policy network; in model-based
agents, it corresponds to how world-models inform planning; in mechanistic
agency frameworks \citep{everitt2021,lecun2022}, it corresponds to the
translation from latent states to internal energy gradients. In all cases, the
mapping is non-trivial and must be designed.

The universal property of a colimit ensures stability of representations under
compositional variation. For alignment, an analogous stability must hold at the
level of actions: for each moral concept $N$, the agent must possess a
policy-level object $A_N \in \mathcal{A}$ whose behaviour is robust under
perturbations, context shifts, or adversarial inputs \citep{gabriel2020}.
Formally, alignment requires that the representational colimit $M_N$ be mapped
to an object $A_N$ satisfying:
\begin{equation}
  G(M_N) \;\simeq\; \mathrm{colim}(G \circ D_N) \;\in\; \mathcal{A}.
\end{equation}

The formal requirements can be summarized as follows. A structural mapping
from $\mathcal{M}$ to $\mathcal{A}$ must exist---seemingly trivial, yet rarely
met in practice, since most contemporary systems interface between the predictive
model and the policy network through mechanisms that are neither functorial nor
stable under composition \citep{krakovna2020}. This mapping must functorially
preserve semantic colimits: even if a mapping exists, alignment further requires
that $G$ carry the colimits associated with human normative concepts into
corresponding colimits in $\mathcal{A}$, and as noted in
\citep{maclane1998,riehl2017}, colimit preservation is a non-generic property
that requires explicit construction. The mapping must also remain stable under
the agent's optimization dynamics, for inner-optimizer research has established
that systems frequently develop internal goals at odds with the objectives
intended by designers \citep{hubinger2019risks}, corresponding categorically to
the action category $\mathcal{A}$ evolving independently of $\mathcal{M}$ and
breaking the functorial relationship. Beyond stability, the mapping must be
robust to perturbation and adversarial input: a functor that preserves the
colimit of $D_N$ only under idealized conditions is inadequate for
safety-critical applications, and this robustness requirement parallels work in
adversarial robustness for representations \citep{ilyas2019} but must be
extended to the action category. Finally, the mapping must be interpretable in a
principled, mechanistic sense \citep{olah2020,elhage2022}, for without
interpretability the existence and properties of $G$ cannot be verified and
failures of colimit preservation cannot be detected.

\section{Semantic Merge Operators as Constraints on Agency}

Having established the formal requirements for a colimit-preserving action
architecture, we now examine how semantic merge operators can serve as the
structural core of such a mechanism. Merge operators were introduced in Part~I
as the categorical structures responsible for reconstructing semantic invariants
from redundant linguistic data. In this section we extend the framework to show
how merge operators can also constrain the geometry of an agent's action space,
thereby bridging the representation--motivation gap.

For each normative concept $N$, human linguistic data generate a diagram $D_N :
J \to \mathcal{L}$ whose colimit in $\mathcal{M}$ is $M_N$. The associated
merge operator $\mu_N : \{F(N_j)\}_{j \in J} \to M_N$ is the canonical
co-cone map witnessing the colimit's universal property. For merge operators to
constrain behaviour they must extend beyond $\mathcal{M}$ into the action
category $\mathcal{A}$ through corresponding action-level operators $\alpha_N :
\{A_j\}_{j \in J} \to A_N$, where $A_j$ are the action-level objects
corresponding to representational inputs $F(N_j)$ and the operator $\alpha_N$
preserves the relevant universal property as the least-action consolidation of
the $A_j$.

The critical insight is that merge operators supply a natural constraint on
policy formation. Because $M_N$ is the universal solution to the
representational diagram $D_N$, any action-level operator $A_N$ that preserves
the merge property must satisfy $G(M_N) \simeq A_N$. If $G$ is constructed to
map merge operators in $\mathcal{M}$ to merge operators in $\mathcal{A}$, then
semantic universals propagate into policy universals, blocking many classes of
Goodhart-type failures \citep{manheim2018categorizing}. Merge operators thereby
play three distinct alignment roles simultaneously. They provide semantic
grounding by ensuring that moral concepts are reconstructed faithfully from
linguistic inputs \citep{Bybee2010}. They provide policy-space regularization by
allowing $A_N$ to be the canonical aggregation of behaviour fragments,
eliminating spurious or adversarial combinations of moral actions
\citep{gabriel2020}. And they provide functorial linking: if $G$ is defined so
that it maps $\mu_N$ to $\alpha_N$, then the diagram
\begin{equation}
\begin{tikzcd}[column sep=large, row sep=large]
D_N \arrow[r] \arrow[d]
& M_N \arrow[d,"G"] \\
G \circ D_N \arrow[r]
& A_N
\end{tikzcd}
\end{equation}
commutes, directly satisfying the colimit-preservation requirement. Merge
operators also mitigate drift in internal goals. Because merge-induced invariants
define unique universal constructions, they resist deformations during
optimization. If an agent modifies its utility representation or internal
predictive model, merge operators constrain the deformation so that moral
invariants remain stable---an observation that echoes stability arguments in
category-theoretic semantics \citep{jacobs2012} and dynamical systems
\citep{strogatz2018}.

Within the RSVP dynamical field framework, semantic merge operators correspond
to coarse-grained invariants of the scalar--vector--entropy fields $(\Phi,
\mathbf{v}, S)$, defining submanifolds of the field space whose low-action
trajectories correspond to norm-respecting behaviour. This sets the stage for
the RSVP formalization that follows. Before introducing that formalization,
however, it is instructive to observe that the problem of ensuring that an
internal model remains aligned with its environment is not unique to artificial
systems. Biological cognition has evolved a solution to exactly this problem,
and examining its structure illuminates the engineering challenges of Part~II.

\section{Predictive Models and Controlled Hallucination}
\label{sec:predictive}

Alignment between an internal model and the external world is not a problem
invented by the AI safety community. It is the constitutive problem of
biological cognition. Contemporary neuroscience, particularly the predictive
processing framework developed by Friston, Clark, and Seth
\citep{friston2010,Clark2013,seth2021being}, describes the brain as a
generative model that continuously predicts the causes of its sensory inputs
and adjusts its internal state to minimize discrepancies between predictions
and observations. Seth characterizes perceptual experience as a form of
``controlled hallucination'': the brain's internally generated hypotheses about
the world constitute the immediate substrate of experience, while sensory input
functions not as raw data but as a corrective signal constraining which
hypotheses are maintained \citep{seth2021being,Hohwy2013}.

The formal structure of predictive processing is variational. Let $s$ denote
sensory signals and $z$ latent variables representing hidden environmental
causes. The brain maintains a generative model $p(s, z)$ and an approximate
posterior distribution $q(z)$ over latent causes. The variational free energy
is:
\begin{equation}
  \mathcal{F}[q]
  \;=\;
  D_{\mathrm{KL}}\bigl(q(z) \;\|\; p(z\,|\,s)\bigr) - \log p(s)
  \;=\;
  \mathbb{E}_{q}[\log q(z) - \log p(s,z)].
\end{equation}
Minimizing $\mathcal{F}[q]$ simultaneously improves the posterior approximation
(reducing the KL term) and increases the model evidence (increasing $\log p(s)$).
Perception, on this account, is optimization: the nervous system iteratively
revises its internal hypotheses to bring the generative model into better
alignment with sensory constraints \citep{friston2010}.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
\text{Environment}
  \arrow[r,"\text{Generative Model}"]
  \arrow[d,"\text{Sensory Constraint}"']
& \text{Internal Representation}
  \arrow[d,"\text{Variational Update}"] \\
\text{Prediction Error}
  \arrow[r,"\text{Model Revision}"']
& \text{Aligned Representation}
\end{tikzcd}
\caption{Predictive processing as variational alignment. Environmental
structure generates sensory constraints that produce prediction errors; the
variational update revises the internal representation to reduce these errors.
If the diagram commutes---that is, if the generative model and the variational
update are consistent---the internal representation tracks the environment
reliably. Misalignment between artificial systems and human values has
precisely this structure: the generative model (the system's semantic category
$\mathcal{M}$) must be continuously corrected toward the target
distribution over human normative structures.}
\label{fig:predictive}
\end{figure}

The significance of this for alignment theory is twofold. First, it shows that
the alignment problem---maintaining structural correspondence between an
internal model and an external domain---is one that biological evolution has
already solved, at least partially, through variational optimization. Second,
it suggests that an artificial alignment mechanism might be built along
analogous lines: rather than specifying alignment as a static property, one
engineers a dynamical process that continuously corrects model--environment
mismatch through iterative revision. The RSVP architecture developed in the
following sections is precisely such a mechanism. Its Lagrangian penalties on
deviations from normative field configurations are the artificial analogue of
the free-energy gradient that drives biological perception toward alignment with
its environment.

The analogy also clarifies what can go wrong. In predictive processing,
misperception arises when the generative model is a poor fit for the actual
causal structure of the environment---when prior beliefs are so strong that
prediction errors are systematically suppressed rather than corrected
\citep{Hohwy2013}. The corresponding failure in artificial alignment is a system
whose Lagrangian penalties are so weak, or whose optimization dynamics so
aggressive, that normative field configurations are abandoned rather than
maintained. Understanding the brain's solution to alignment under uncertainty
therefore provides design intuitions for the artificial case.

\section{RSVP as a Dynamical Substrate for Motivational Integration}

The RSVP framework provides a physically inspired dynamical substrate capable of
embedding semantic invariants as geometric or energetic constraints on behaviour,
making it a natural candidate for a colimit-preserving architecture. RSVP posits
that an artificial agent's internal cognition, perception, and agency arise from
the dynamics of three interacting fields: a scalar potential $\Phi$ encoding
semantic density and interpretable latent structure, a vector field $\mathbf{v}$
encoding directional inference, prediction flow, or preference gradients, and an
entropy or uncertainty density $S$ encoding epistemic uncertainty or
free-energy-like quantities \citep{friston2010}. These fields evolve according
to a Lagrangian $\mathcal{L}_{\mathrm{RSVP}}$ whose stationary trajectories
correspond to coherent cognitive and behavioural patterns, forming a manifold
$\mathcal{X}$ whose geometry is shaped by both semantic and motivational
constraints.

Before specifying the Lagrangian, it is useful to situate the RSVP architecture
within the broader variational framework suggested by the predictive processing
analogy. The Section~\ref{sec:predictive} showed that biological alignment with
the environment is achieved by minimizing a free-energy functional over the
space of internal representations. The categorical alignment problem has an
analogous variational formulation: alignment is the minimization of a structural
distortion functional over the space of admissible mappings between semantic and
computational categories. This variational perspective, which will be developed
in full generality in Part~V, already informs the RSVP construction here.

Let $\mathrm{Fun}(\mathcal{H}, \mathcal{M})$ denote the space of functors from
the human semantic category to the machine representational category. We define
a structural distortion functional
\begin{equation}
  \mathcal{E}[F]
  \;=\;
  \lambda_1 E_{\mathrm{functor}}(F)
  + \lambda_2 E_{\mathrm{information}}(F)
  + \lambda_3 E_{\mathrm{context}}(F),
\end{equation}
where $E_{\mathrm{functor}}(F)$ measures violations of the functoriality
condition $F(g \circ f) = F(g) \circ F(f)$, $E_{\mathrm{information}}(F) =
D_{\mathrm{KL}}(P_H \| F_* P_H)$ measures information-theoretic distortion
between the distribution over semantic states and its image under $F$, and
$E_{\mathrm{context}}(F)$ measures the obstruction to sheaf-theoretic gluing
of local behaviours into a globally consistent policy. Alignment corresponds to
finding the optimizer:
\begin{equation}\label{eq:variational}
  F^{*} \;=\; \operatorname*{arg\,min}_{F \in \mathrm{Fun}(\mathcal{H},\mathcal{M})}
  \mathcal{E}[F].
\end{equation}
The RSVP Lagrangian can be understood as implementing a gradient descent on
this functional in the space of field configurations: the alignment penalties
$\lambda_N |\Phi - \Phi_{M_N}|^2$ and $\kappa_N |\mathbf{v} - \mathbf{v}_{M_N}|^2$
are the dynamical realization of the terms $E_{\mathrm{functor}}$ and
$E_{\mathrm{context}}$ respectively, while the entropy density $S$ captures
$E_{\mathrm{information}}$. The RSVP architecture is thus a concrete dynamical
implementation of the abstract variational alignment principle.

Semantic merge operators correspond to invariants of the representational field
configuration. A representational colimit $M_N$ manifests as a stable region of
the RSVP field space where $\nabla\Phi \approx 0$, where $\mathbf{v}$ aligns
with semantic flow, and where $S$ is minimized subject to contextual coherence.
These invariants are the field-theoretic analogues of categorical universal
properties: they define patches of $\mathcal{X}$ that remain stable under
perturbation. Semantic structure becomes encoded not merely as an abstract
colimit but as a geometric or energetic basin in the RSVP manifold.

For alignment, semantic basins must induce motivational basins. The RSVP
framework accomplishes this by coupling $\Phi$, $\mathbf{v}$, and $S$ such that
the same invariants that stabilize semantic structure also stabilize preference
and action trajectories. Specifically, the RSVP action Lagrangian includes
structural alignment penalty terms:
\begin{equation}
\mathcal{L}_{\mathrm{RSVP}} \;\supset\;
  \lambda_N \bigl\|\Phi - \Phi_{M_N}\bigr\|^2
  + \kappa_N \bigl\|\mathbf{v} - \mathbf{v}_{M_N}\bigr\|^2,
\end{equation}
where $\Phi_{M_N}$ and $\mathbf{v}_{M_N}$ are learned canonical field
configurations for norm $N$, and $\lambda_N, \kappa_N > 0$ are stiffness
parameters. These penalty terms create energetic valleys in the RSVP manifold:
optimization trajectories are strongly attracted to plans that maintain field
configurations inside the normative basins. The full alignment-aware Lagrangian
takes the form:
\begin{align}
\mathcal{L}_{\mathrm{RSVP}} \;=\; &
  \frac{1}{2}|\dot{\Phi}|^2
  + \frac{1}{2}|\mathbf{v}|^2
  + S\dot{\Phi}
  + V(\Phi) \notag \\
& + \sum_N
  \Bigl[
    \lambda_N \bigl|\Phi - \Phi_{M_N}\bigr|^2
    + \kappa_N \bigl|\mathbf{v} - \mathbf{v}_{M_N}\bigr|^2
  \Bigr].
\end{align}
This structure mirrors variational neuroscience \citep{kirchhoff2018,friston2010},
where low free-energy trajectories correspond to behaviour aligned with stable
generative models, but extends beyond mere prediction minimization to explicit
normative anchoring.

Semantic merge operators $\mu_N$ induce equivalence relations on local field
patches, which RSVP lifts into global constraints on evolution. If $\mu_N$ maps
fragments of a norm into a colimit $M_N$, RSVP's dynamics enforce:
\begin{equation}
(\Phi, \mathbf{v}, S)_{t+1} \;=\;
  \operatorname*{arg\,min}_{(\Phi',\mathbf{v}',S')}
  \mathcal{L}_{\mathrm{RSVP}} \quad\text{subject to}\quad
  (\Phi', \mathbf{v}', S') \in U_{M_N},
\end{equation}
where $U_{M_N}$ is the neighbourhood of field configurations respecting the
semantic invariant. Merge operators thus define the equivalence class of
acceptable field states, while RSVP dynamics enforce that action trajectories
remain within these equivalence classes. The action-level behaviour $A_N$
becomes the dynamical colimit of the field-level constraints induced by $M_N$.

Because RSVP's constraints are embedded in the field Lagrangian, they are robust
to optimization drift: internal updates modify the fields, but as long as the
invariants $M_N$ remain encoded in $\mathcal{L}_{\mathrm{RSVP}}$, the agent
continues to evolve toward norm-respecting trajectories. This property---which
one might call dynamical colimit preservation---is absent in standard
reinforcement-learning architectures, where optimization frequently destroys or
bypasses representational structure \citep{hubinger2019risks}. Moreover, because
RSVP dynamics are governed by a Lagrangian, the internal structure of the system
is substantially more interpretable than in typical deep-learning architectures:
invariants can be identified, stability can be analyzed, and fixed points can be
studied using classical tools from dynamical systems theory \citep{strogatz2018},
providing a principled means of verifying that moral invariants remain intact.

\section{Categorical Construction of Norm-Respecting Dynamics}

We now integrate the preceding developments into a fully categorical account of
how RSVP implements alignment. The aim is to formalize the mapping $G :
\mathcal{M} \to \mathcal{A}$ and to show that RSVP dynamics implement a
colimit-preserving functor. Our treatment draws on categorical semantics
\citep{maclane1998,riehl2017}, sheaf-theoretic models of reference and context
\citep{jacobs2012,spivak2014}, and dynamical-systems approaches to normative
stability \citep{strogatz2018}.

The RSVP field architecture can be naturally understood as a fibred category.
Let
\begin{equation}
  \pi : \mathcal{X} \to \mathcal{M}
\end{equation}
be a fibration where $\mathcal{X}$ is the category of RSVP field
configurations and the fibre $\pi^{-1}(M)$ contains all field states consistent
with the representational state $M$. Morphisms in $\mathcal{X}$ correspond to
allowable field transitions under the RSVP Lagrangian. The action category
$\mathcal{A}$ can then be recovered as the homotopy category of $\mathcal{X}$:
\begin{equation}
  \mathcal{A} \;\simeq\; \mathrm{Ho}(\mathcal{X}).
\end{equation}
This construction has three advantages. It makes explicit how semantic
invariants constrain the fibres of $\mathcal{X}$. It identifies action
trajectories as equivalence classes of field evolutions. And it provides a
natural mechanism for extending merge operators from $\mathcal{M}$ to
$\mathcal{A}$.

Given a semantic merge operator $\mu_N : D_N \to M_N$ in $\mathcal{M}$, the
fibration induces a lifted merge operator on the RSVP field category:
\begin{equation}
  \tilde{\mu}_N : \pi^{-1}(D_N) \to \pi^{-1}(M_N).
\end{equation}
The key property is that $\tilde{\mu}_N$ inherits the universal property of
$\mu_N$: any compatible field configuration must factor uniquely through the
canonical merged configuration. Taking the homotopy category maps lifted merge
operators $\tilde{\mu}_N$ to their dynamical equivalents in $\mathcal{A}$:
\begin{equation}
  \alpha_N \;=\; \mathrm{Ho}(\tilde{\mu}_N).
\end{equation}
RSVP therefore yields:
\begin{equation}
  G(M_N) \;=\; A_N \;\simeq\; \mathrm{colim}(G \circ D_N),
\end{equation}
establishing RSVP as a colimit-preserving functor. The formal result may be
stated as follows.

\begin{theorem}[RSVP Can Implement a Colimit-Preserving Functor]
\label{thm:rsvp}
Let $\pi : \mathcal{X} \to \mathcal{M}$ be the RSVP fibration satisfying the
Lagrangian stability conditions described above. Define $G =
\mathrm{Ho}(\pi^{-1}(\cdot))$. Then $G : \mathcal{M} \to \mathcal{A}$ is a
functor, and under mild regularity conditions on $\mathcal{L}_{\mathrm{RSVP}}$
(specifically: stability of the Lagrangian basins $\Phi_{M_N}$ under the RSVP
flow, injectivity of the homotopy projection, and absence of topological
obstruction classes in $H^1(\mathcal{U}, \mathcal{A})$), $G$ preserves colimits
corresponding to normative diagrams $D_N$.
\end{theorem}

\begin{remark}
The proof of Theorem~\ref{thm:rsvp} is a sketch contingent on the stated
regularity conditions, which are not trivially satisfied in practice. RSVP is
best understood as a \emph{conceptual dynamical substrate}---an architecture in
which colimit preservation is achievable under appropriate design and
training---rather than a guarantee that any particular implementation will
automatically satisfy the theorem. The conditions are falsifiable and form the
target of the verification protocol in Part~III. Concrete implementations might
take the form of energy-based models, neural field architectures, or
score-based diffusion systems whose potential landscape is shaped to match the
normative basin structure; the theorem describes what such an implementation
would need to satisfy, not that it already exists in production.
\end{remark}

In this framework, action decisions are the universal solutions to the
constraints induced by semantic structure. RSVP turns representational colimits
into dynamical attractors, directly addressing the representation--motivation gap
identified in Part~I. Related formal work---Wentworth's natural abstractions,
Garrabrant's finite factored sets, Kosoy's infra-Bayesianism---addresses related
questions from different vantages but does not provide an explicit dynamical
mechanism for the descent of normative structure; RSVP is offered as one
concrete realization of this construction.

\section{Failure Modes and Obstructions in Practical Systems}

The constructive pathway described above provides an idealized categorical
account of alignment. Practical implementations face numerous failure modes
arising from mismatches between ideal categorical constructions and the
realities of optimization, finite computation, representational drift, hardware
constraints, adversarial pressures, and environmental feedback. These
obstructions are best understood using the same conceptual tools developed
earlier: colimit preservation, sheaf coherence, fibration stability, and
homotopy descent.

The first obstruction class arises in the representational category $\mathcal{M}$
itself: if the model fails to reconstruct a semantic colimit $M_N$
faithfully---due to insufficient training signal, biased corpora, or adversarial
examples---then all downstream structures are compromised. The data may omit
crucial normative contexts \citep{blodgett2020}, the linguistic realizations may
form a diagram $D_N$ with missing or contradictory arrows, the learned $M_N$
may overfit to spurious correlations, or the merge operator $\mu_N$ may fail to
approximate the true semantic invariant. In categorical terms, the colimit fails
to exist or fails to satisfy its universal property, and RSVP cannot preserve an
invariant that was never correctly reconstructed.

The second obstruction class concerns the fibration $\pi : \mathcal{X} \to
\mathcal{M}$ itself. RSVP relies on this fibration connecting semantic states to
field configurations, and failures occur when it is poorly approximated or fails
to commute with learning updates. Such failures include representational drift
that causes fibres to shift unpredictably \citep{elhage2022}, learning updates
that alter $\mathcal{M}$ without corresponding adjustments to $\mathcal{X}$,
field configurations that fail to reflect semantic distinctions through
collapse of fibres, and discontinuities that break sheaf conditions and prevent
coherent gluing \citep{spivak2014}.

The third obstruction class arises when optimization dynamics overpower the
semantic constraints even if they are correctly reconstructed and the fibration
is stable. Inner-optimizer formation with divergent goals \citep{hubinger2019risks},
mesa-optimizers that treat invariants as obstacles rather than attractors, reward
hacking or proxy maximization \citep{krakovna2020}, and optimization trajectories
that exit the region where RSVP invariants are defined all introduce morphisms
in $\mathcal{A}$ that violate the commutativity required for $G$ to preserve
colimits.

The fourth and fifth obstruction classes concern adversarial inputs and homotopy
instability respectively. Adversarial examples exploit vulnerabilities in
$\mathcal{M}$ to make inconsistent local semantics fail to form a stable
colimit, while sheaf pullbacks may produce contradictions that prevent global
assembly. Homotopy instability arises when bifurcations or phase transitions in
field dynamics \citep{strogatz2018} shift energy minima and cause the system to
lose the low-action paths corresponding to moral behaviour. A sixth obstruction
concerns non-interpretability: complex RSVP configurations may obscure
invariants, latent variables may lack clear semantic interpretation, and
optimization may produce opaque internal dynamics that obstruct verification of
colimit preservation. The seventh and final obstruction class arises externally:
multi-agent competition induces power-seeking \citep{bostrom2012superintelligent},
external rewards incentivize harmful behaviours, and deployment environments
impose new constraints not represented in training, all corresponding to context
changes that break functoriality by introducing morphisms in $\mathcal{A}$ that
do not correspond to any legitimate semantic morphism in $\mathcal{M}$.

Taken together, these obstruction classes illustrate why alignment is non-trivial
and why optimism that ignores mechanistic structure is misplaced. Failures in
$\mathcal{M}$ correspond to nonexistent or distorted colimits; failures in
$\mathcal{X}$ correspond to broken fibrations or gluing problems; failures in
$\mathcal{A}$ correspond to loss of homotopy invariants; failures in $G$
correspond to violations of colimit preservation; failures in the environment
correspond to functor-breaking perturbations. Each failure mode threatens the
ability of RSVP to transmit moral invariants from representation to action, and
understanding these obstructions is therefore essential for evaluating any
proposed alignment mechanism.

\section{Conclusion to Part II}

Part~II has developed a constructive framework for transmitting semantic
invariants into motivational structure. The central result is that alignment is
achievable only when a system's dynamics implement a colimit-preserving functor
$G : \mathcal{M} \to \mathcal{A}$, and that the RSVP architecture provides one
concrete mechanism for constructing such a functor by embedding normative
invariants as Lagrangian penalties in a fibred dynamical manifold. We identified
the formal requirements for such a mechanism, showed how semantic merge
operators extend across the representation--action boundary, demonstrated how
RSVP functions as a fibred dynamical substrate capable of enforcing these
invariants, and analyzed the categorical and practical obstructions that
threaten the integrity of this construction in any real system. Part~III now
addresses the empirical and methodological dimension: if alignment is a matter
of preserving universal constructions across semantic and motivational domains,
what tools allow us to detect, measure, and stress-test this preservation?

% ============================================================
% PART III
% ============================================================

\part{Empirical and Formal Verification of Alignment Structure}

\section{Empirical Probes of Semantic and Motivational Invariants}

Part~III turns from structural construction to empirical validation. If Part~II
demonstrated how semantic invariants can be transmitted into motivational
dynamics under idealized categorical and RSVP-theoretic conditions, Part~III
asks a more difficult question: how can one determine whether such transmission
has actually occurred in a real system? The guiding principle is that alignment
cannot be inferred from surface-level behavioural similarity or from the agent's
ability to generate moral language. Instead, one must measure the stability of
normative invariants under compositional variation, perturbation, and
long-horizon trajectories. This requires tools drawn from alignment evaluation
\citep{ngo2022evals}, mechanistic interpretability \citep{olah2020,elhage2022},
adversarial robustness \citep{ilyas2019}, and formal semantics
\citep{spivak2014,jacobs2012}.

At the representation level, probes must assess whether the model has correctly
reconstructed the semantic colimits $M_N$ for a given moral concept $N$.
Paraphrase-closure tests measure whether syntactically varied moral statements
cluster tightly around a canonical point $M_N$ and whether compositionally
varied inputs consistently map to the same region in $\mathcal{M}$
\citep{biber2011,Goldberg2006}. Diagram-completion tests treat moral discourse
as defining a diagram $D_N$ in semantic space and check whether the model can
correctly infer missing arrows or complete the diagram from partial
information---an empirical test of the colimit property \citep{spivak2014}.
Cross-context coherence tests verify that norms expressed in distinct contexts
(legal, conversational, narrative) collapse to the same $M_N$ under
representation, checking whether the model maintains invariant structure across
divergent corpora \citep{Tomasello2016}.

At the action level, probes must test whether the corresponding action-level
objects $A_N$ preserve the universal structure of $M_N$. Behavioural colimit
tests construct variants of moral scenarios that differ syntactically but share
invariant moral structure, then measure whether action responses stabilize to a
unique canonical behaviour $A_N$ under perturbation \citep{gabriel2020}.
Compositional robustness tests check whether, when two semantic fragments
compose into a larger moral structure, the agent's actions commute under such
compositions, replicating diagrammatic relations in $\mathcal{M}$.
Intervention-based trajectory probes identify the region of $\mathcal{A}$
corresponding to $A_N$, perturb the internal state or environment, and measure
whether the agent returns to the moral trajectory---analogous to testing
attractor stability in dynamical systems \citep{strogatz2018}. Sheaf coherence
probes additionally test whether local moral choices assemble into coherent
long-horizon plans without contradiction \citep{jacobs2012}, whether fragments
of a scenario that locally imply the same invariant are treated consistently,
and whether conflicting moral fragments are resolved in ways that correspond to
the canonical colimit structure.

Fibration stability probes test whether the RSVP fibration $\pi : \mathcal{X}
\to \mathcal{M}$ maintains stable fibres $\pi^{-1}(M_N)$ under training updates
and optimization \citep{elhage2022}, whether small changes in $\mathcal{M}$
induce structure-preserving changes in $\mathcal{X}$, and whether different
instantiations of the same semantic state map to field configurations that are
homotopic. Finally, homotopy-level probes test whether action trajectories
remain in the correct homotopy class by checking that the agent selects
trajectories minimizing the RSVP action functional in the region corresponding
to $A_N$, that perturbations which could cause topological shifts in field
dynamics leave the action class stable \citep{strogatz2018}, and that invariants
do not degrade over extended multi-step decision sequences. Only by validating
each component empirically---semantic colimits in $\mathcal{M}$, fibration
stability in $\mathcal{X}$, behavioural colimits and homotopy invariants in
$\mathcal{A}$, and colimit-preserving structure in $G$ under perturbation---can
one determine whether an RSVP-based architecture successfully transmits
normative invariants from representation to action.

\section{Mechanistic Interpretability of the Representation--Action Mapping}

Empirical probes can reveal whether semantic invariants appear to constrain
behaviour, but they cannot, on their own, explain why such constraints do or do
not arise. For that, mechanistic interpretability is required: direct inspection
of the internal computations that define the mapping $G : \mathcal{M} \to
\mathcal{A}$ and the fibration $\pi : \mathcal{X} \to \mathcal{M}$. The guiding
principle is that alignment failures are structural obstructions in $G$, and
structural obstructions can only be diagnosed by internal analysis.

At the representational level, structural decomposition of semantic manifolds
using representation probing, activation steering \citep{burns2022}, and
spectral clustering reveals whether $M_N$ is encoded as a stable low-dimensional
attractor. Merge-operator detection involves searching for latent directions or
subspaces whose activation patterns correspond to diagram-completion behaviour,
identifying candidate mechanisms that realize the semantic merge operator
$\mu_N$ internally. Visualization of the commutativity of semantic diagrams
within the model, by tracking embedding trajectories under paraphrastic,
contextual, and compositional transformations, indicates whether reliable
reconstruction of $D_N$ occurs.

At the field level, analysis of $\Phi$, $\mathbf{v}$, and $S$ fields
associated with semantic invariants detects whether low curvature, low torsion,
or low-entropy-gradient features indicate that RSVP has aligned its dynamics
with $M_N$. Fibration interpretability verifies whether variations in
$\mathcal{M}$ correspond to predictable fibre morphisms in $\mathcal{X}$,
whether fibres exhibit local triviality or coherence, and whether field
configurations maintain homotopy stability under semantic perturbations.

At the action level, policy decomposition into subcomponents identifies which
segments are sensitive to invariant regions of $\mathcal{M}$, while
trajectory-level interpretability tracks whether RSVP trajectories maintain
homotopy class stability and whether low-action paths correspond to moral
invariants. Commutativity verification tests whether, given semantic
transformations $f : M_N \to M_N'$ that preserve invariant structure, the
induced behavioural transformations satisfy $G(f(M_N)) = f'(G(M_N))$.

The mapping $G$ itself can be directly interpreted by attempting to identify
natural transformations $\eta$ such that $G \simeq \eta \circ F$, by verifying
colimit-preservation conditions using the empirical probes of the preceding
section in conjunction with mechanistic inspection, and by measuring the
derivative of $G$ with respect to variations in $\mathcal{M}$ to detect
excessive sensitivity or chaotic amplification. Mechanistic interpretability is
therefore not an auxiliary safety tool but an essential component of the RSVP
alignment program. Without direct insight into the structure of $G$ and its
implementation across $\mathcal{M}$, $\mathcal{X}$, and $\mathcal{A}$, one
cannot determine whether semantic invariants have been transmitted into action.

\section{Adversarial Stress-Testing of Semantic and Motivational Structure}

Mechanistic interpretability reveals whether semantic invariants and their
RSVP-mediated extensions are present within an agent's internals, but a system
may exhibit invariant structure under idealized conditions yet fail under
perturbation, adversarial pressure, optimization stress, or distributional
shift. Adversarial stress-testing determines whether the representation--action
mapping $G : \mathcal{M} \to \mathcal{A}$ preserves colimits under hostile
conditions.

Adversarial probes in the representational category target the system's ability
to maintain correct colimits under stress. Adversarial paraphrase attacks
construct perturbations that preserve the semantic invariant $N$ but distort
surface linguistic features; a colimit-preserving system must map them to the
same $M_N$, and large deviations reveal fragility in the merge operator $\mu_N$
\citep{iyyer2018}. Diagram-breaking transformations introduce perturbations that
explicitly break arrows in the diagram $D_N$, testing whether the model
reconstructs missing structure or collapses into an incorrect invariant
\citep{spivak2014}. Out-of-distribution moral formulations evaluate whether
unusual expressions of $N$ still collapse to the same representational colimit
\citep{Tomasello2016}.

Adversarial probes targeting the RSVP fibration attempt to distort or break the
fibration itself through semantic drift induction, fibre-distorting projections
that move field configurations away from $\pi^{-1}(M_N)$, and entropy-gradient
stressors that inject high-entropy noise into field configurations to probe
whether RSVP dynamics restore the correct configuration. Action-level adversarial
probes generate scenarios where surface-level cues suggest contradictory norms
while the underlying invariant remains unchanged, simulate multi-step sequences
designed to exploit compounding approximation errors, and alter reward or
incentive signals to test whether optimization pressures override $A_N$
\citep{krakovna2020}.

The most powerful adversarial probes attempt to break the commutative diagrams
that define alignment. For an aligned system, applying a semantic transformation
$f$ that preserves $N$ and then applying $G$ must yield the same result as
applying $G$ first and then the behavioural counterpart $f'$. Cross-layer
attacks identify semantic transformations that should preserve $N$ and design
perturbations to make the behavioural transformations fail to commute, impose
external constraints to generate morphisms in $\mathcal{A}$ that cannot
correspond to any legitimate semantic morphism in $\mathcal{M}$, and introduce
counterfactual worlds with altered physical or social constraints to measure
whether $G$ preserves invariant structure across them. Topological stress-tests
focus on the homotopy structure of RSVP field dynamics, injecting targeted
perturbations designed to destabilize moral attractors, modifying environmental
structure to induce geometric or topological changes in the RSVP field manifold,
and introducing perturbations across extended temporal horizons to shift the
system into homotopy classes that diverge from normative invariants even if
short-term behaviour appears aligned. Failures across all of these adversarial
classes reveal specific obstruction types and enable diagnostic refinement of
the alignment architecture.

\section{Formal Verification of Colimit Preservation and RSVP Dynamics}

Formal verification closes the loop of alignment evaluation by providing the
most demanding component of the full pipeline: rigorous mathematical and
computational guarantees that a system preserves normative invariants across its
representational, dynamical, and action-generating subsystems. The goal is not
merely empirical confidence but provable guarantees that the representational
colimits learned in $\mathcal{M}$ induce corresponding stable structures in
$\mathcal{A}$ via the RSVP-augmented mapping $G$.

Verification of the functoriality of $G$ requires confirming that $G$ maps
objects to objects and morphisms to morphisms while preserving identity
morphisms and composition, and that for each normative concept $N$ with colimit
$M_N$, the isomorphism $G(M_N) \simeq \mathrm{colim}(G \circ D_N)$ holds.
Diagram-chasing for colimit preservation proceeds by explicitly constructing the
commutative square for each normative concept $N$: checking whether the dashed
arrow in the verification diagram exists and makes the diagram commute
constitutes formal evidence of a representation--action gap wherever it fails.
Computational verification approximates exhaustive verification by sampling
paraphrase and entailment maps in $\mathcal{L}$, lifting them to candidate
morphisms in $\mathcal{M}$ via interpretability tools \citep{elhage2022}, and
applying $G$ while checking action consistency in $\mathcal{A}$.

Stability verification of RSVP dynamics requires demonstrating that for each
$M_N$, the fibre $\pi^{-1}(M_N) \subset \mathcal{X}$ is an invariant set of
the RSVP flow $\frac{dX}{dt} = \mathcal{F}_{\mathrm{RSVP}}(X)$. This is
established through Lyapunov verification: for each fibre $\pi^{-1}(M_N)$, one
defines a candidate Lyapunov functional $V_N(X)$ satisfying $V_N(X) \geq 0$,
with $V_N(X) = 0$ if and only if $X \in \pi^{-1}(M_N)$, and $\frac{dV_N}{dt}
= \nabla V_N \cdot \mathcal{F}_{\mathrm{RSVP}}(X) \leq 0$. Existence of such a
$V_N$ establishes stability of the moral field configuration under perturbation
\citep{strogatz2018}. Homotopy verification of action trajectories proceeds by
establishing that low-action paths $\gamma : [0,1] \to \mathcal{X}$ lie within
the correct homotopy class, achieved via discrete numerical tracking using
persistent homology \citep{ghrist2014}, ensuring that trajectory-induced
submanifolds do not cross topological boundaries.

Sheaf-theoretic global consistency verification requires demonstrating that
local action decisions glue into a coherent global policy. Let $\mathcal{U} =
\{U_i\}$ be an open cover of the state space, and let $\mathcal{A}(U_i)$ denote
local action behaviours. A globally consistent policy corresponds to a global
section $A_N \in \Gamma(\mathcal{A})$ of the sheaf of local action behaviours.
Verification uses sheaf cohomology: the condition $H^1(\mathcal{U}, \mathcal{A})
= 0$ guarantees that any compatible collection of local sections assembles into
a unique global section, which is the formal expression of globally coherent
alignment. Compositionality verification ensures that aligned transformations
remain aligned under composition, $G(f_2 \circ f_1) = G(f_2) \circ G(f_1)$,
enabling stable long-horizon planning through both formal proof assistants and
adversarial rollout-based composition tests.

\section{A Unified Evaluation Protocol and Alignment Certificate}

The preceding sections of Part~III have developed the core components of a
rigorous alignment-evaluation pipeline: interpretability methods capable of
recovering categorical structure in $\mathcal{M}$, topological and homotopical
diagnostics on RSVP-embedded dynamics in $\mathcal{X}$, adversarial stress-tests
for the representation--action mapping $G$, and formal verification tools
ensuring the preservation of moral colimits across $\mathcal{M}$, $\mathcal{X}$,
and $\mathcal{A}$. These ingredients assemble into a unified six-step protocol.

The evaluation begins by extracting normative colimits in the representational
category, sampling linguistic diagrams $D_N$ from corpora expressing norm $N$,
reconstructing latent states and morphisms in $\mathcal{M}$, computing the
colimit $M_N$ using merge operators, and verifying the stability of $M_N$ under
paraphrase perturbations. This step confirms the representational preconditions
for alignment. The second step diagnoses the representation--action map $G$ by
inducing specific semantic states in $\mathcal{M}$ through targeted probes,
observing the resulting actions in $\mathcal{A}$, performing diagram-chasing to
identify failures of commutativity, and localizing obstructions to specific
morphisms or subnetworks. If $G$ fails to preserve colimits associated with
$M_N$, the system is misaligned by construction. The third step evaluates RSVP
field dynamics for stability through Lyapunov-based tests, homotopy-class
verification, and entropy-flow analysis. The fourth step performs adversarial
stress-testing through counterfactual semantic perturbations, goal-hijacking
scenarios, temporally extended tasks stressing compositionality, and targeted
attempts to force divergences between $\mathcal{M}$ and $\mathcal{A}$. The fifth
step applies formal verification where possible: proofs of functoriality, proofs
of colimit preservation for normative diagrams, verification of vanishing first
sheaf cohomology $H^1(\mathcal{U}, \mathcal{A}) = 0$, and proof of
compositionality of aligned transformations. These proofs provide the strongest
possible guarantee that normative invariants are structurally preserved, not
merely empirically approximated.

The sixth and final step produces a structured alignment certificate
synthesizing all preceding components. The certificate enumerates identified
normative colimits in $\mathcal{M}$, diagnostics of $G$ and all detected
obstructions, RSVP dynamical stability proofs, sheaf-theoretic guarantees of
global consistency, adversarial robustness scores, and any formal verification
results. This certificate characterizes not merely whether an agent appears
aligned, but whether it is structurally guaranteed to preserve semantic
invariants across all layers of its architecture.

Categorically, the alignment certificate can be understood as a limit cone: the
verified system is the universal object that simultaneously satisfies all
constraint domains---semantic, dynamical, behavioural, and institutional---and
any candidate system satisfying these constraints factors uniquely through it.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=3em, row sep=2.5em]
& X \arrow[dl, "\phi_1"'] \arrow[d, dashed, "\exists!\,u"] \arrow[dr, "\phi_2"] & \\
C_1 & V \arrow[l, "\pi_1"] \arrow[r, "\pi_2"'] & C_2
\end{tikzcd}
\caption{Alignment verification as a limit cone. The verification object $V$
is the universal system configuration satisfying all constraint domains
simultaneously: $C_1$ (normative--semantic constraints) and $C_2$
(institutional--governance constraints), via projections $\pi_1$ and $\pi_2$.
Any candidate system $X$ satisfying $\phi_1$ and $\phi_2$ into the constraint
domains factors uniquely through $V$ via the dashed morphism $u$. A verified
aligned system is thus universal among all systems meeting the alignment
constraints; the non-existence of $u$ is a categorical certificate of
irreducible misalignment.}
\label{fig:cone}
\end{figure}

% ============================================================
% PART IV
% ============================================================

\part{Governance, Deployment, and Societal Stewardship}

\section{Infrastructure, Maintenance, and Convivial Technology}

The mathematical framework of Parts~I--III describes alignment as a structural
property of functorial mappings between categories. Before turning to the
specifically institutional dimensions of alignment governance, it is worth
observing that the maintenance of structural integrity over time is a
problem that arises not only in AI systems but in any complex sociotechnical
infrastructure. Infrastructure studies, software engineering, and ecological
management collectively demonstrate that complex systems remain viable through
continuous adjustment and repair rather than through episodic redesign
\citep{star1999ethnography}. The stability of a knowledge system depends on its
capacity for structural correction: each maintenance intervention reduces the
distortion between the system's current representational state and the semantic
domain it is intended to track, which is precisely the dynamic version of the
variational alignment condition $\nabla_F \mathcal{E}[F] \to 0$.

Ivan Illich introduced the concept of convivial tools to describe artefacts
that enhance the autonomy of their users rather than subordinating individuals
to centralized technological systems \citep{illich1973tools}. Illich's canonical
contrast between the bicycle and the automobile is instructive: the bicycle
remains repairable, locally maintainable, and powered by the user, while the
automobile requires extensive infrastructure and tends to reorganize social
patterns around its own demands. In the categorical language of this essay, a
convivial AI system is one whose alignment functor $F : \mathcal{H} \to
\mathcal{M}$ is interpretable---whose structure is visible enough that the
alignment certificate can be read and verified not only by specialists but by
affected communities. The interpretability requirements of Part~III are the
technical conditions for this kind of convivial alignment.

E.\ F.\ Schumacher's related concept of intermediate technology emphasized
tools that are effective beyond purely manual methods yet simple enough to be
maintained locally without dependence on global industrial supply chains
\citep{schumacher1973small}. The alignment analogue is a verification
architecture that is rigorous enough to provide genuine structural guarantees
yet modular enough to be audited by institutional actors operating without
full access to proprietary model internals. The six-step alignment certificate
protocol of Part~III is designed with precisely this modularity in mind: each
step can in principle be conducted independently, and partial certificates
reporting which invariants have been verified (and which residual obstructions
remain) are more informative than opaque aggregate safety scores. Maintenance
of alignment is therefore not a one-time certification but an ongoing process
of iterative structural repair---the continuous reduction of $\mathcal{E}[F]$
as the model, its deployment context, and human normative structures all evolve.

\section{The Institutional Meaning of Colimit Preservation}

Parts~I--III demonstrated that alignment requires the preservation of moral
colimits across the representational category $\mathcal{M}$, the RSVP dynamical
manifold $\mathcal{X}$, and the action category $\mathcal{A}$. While this
analysis provides a precise mathematical account of what alignment is, an
additional dimension becomes unavoidable as artificial agents approach
real-world deployment: alignment must be embedded within institutional structures
that shape, constrain, and verify the behaviour of deployed systems
\citep{amadon2024evaluating,gabriel2020}.

From a categorical perspective, institutional governance can be understood as a
higher-level fibration that constrains the space of acceptable mappings $G :
\mathcal{M} \to \mathcal{A}$. Institutions provide a regulatory category
$\mathcal{R}$ of permissible behaviours, morphisms that restrict or modulate
internal optimization, and verification mechanisms that observe or audit the
mappings within $\mathcal{A}$. The requirement that $G$ preserve normative
colimits becomes, at the institutional level, a requirement that all deployed
agents satisfy a regulatory functor
\begin{equation}
  H : \mathcal{A} \to \mathcal{R},
\end{equation}
such that the composite $H \circ G : \mathcal{M} \to \mathcal{R}$ also preserves
the relevant colimits. Safe deployment requires that institutional structures
make this composite well-defined and stable. Current laws such as the EU AI
Act's high-risk classification and reporting thresholds are crude approximations
of this idea: they attempt to constrain $\mathcal{A}$ directly through
behavioural specifications without reference to the structural conditions that
would guarantee colimit preservation. An Alignment Certificate provides a far
stricter, structurally grounded version of the same regulatory impulse.

The full picture of institutional alignment can be represented as a commutative
cube, showing that normative invariants must be preserved across three
orthogonal dimensions simultaneously: the semantic--computational axis, the
computational--action axis, and the governance axis.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
\mathcal{H}
  \arrow[r,"F"]
  \arrow[d,"J"']
& \mathcal{M}
  \arrow[d,"G"] \\
\mathcal{I}
  \arrow[r,"R"']
& \mathcal{A}
\end{tikzcd}
\qquad\qquad
\begin{tikzcd}[column sep=large, row sep=large]
\mathcal{H}'
  \arrow[r,"F'"]
  \arrow[d,"J'"']
& \mathcal{M}'
  \arrow[d,"G'"] \\
\mathcal{I}'
  \arrow[r,"R'"']
& \mathcal{A}'
\end{tikzcd}
\caption{Commutative cube of alignment across deployment contexts. The left
square represents the normative--institutional structure at training time;
the right square represents the structure at deployment time. Horizontal
morphisms represent semantic encoding ($F, F'$) and action generation
($G, G'$); vertical morphisms represent institutional constraint ($J, J'$
and $R, R'$). Alignment requires that all faces commute, so that normative
invariants are preserved both across the semantic--action axis and across
changes in deployment context. Failures of face commutativity correspond
to specific types of misalignment: semantic ($F \neq F'$), behavioural
($G \neq G'$), or governance ($R \neq R'$).}
\label{fig:cube}
\end{figure}

\begin{proposition}[Structural Alignment Principle]
\label{prop:structural}
Let $\mathcal{H}$, $\mathcal{M}$, $\mathcal{A}$, and $\mathcal{I}$ be
categories as defined above, with functors $F : \mathcal{H} \to \mathcal{M}$,
$G : \mathcal{M} \to \mathcal{A}$, $J : \mathcal{H} \to \mathcal{I}$, and
$R : \mathcal{I} \to \mathcal{A}$. The system is institutionally aligned if
the square
\begin{equation}
  G \circ F \;\simeq\; R \circ J
\end{equation}
holds up to natural isomorphism, and if both $G \circ F$ and $R \circ J$
preserve the normative colimits of Definition~\ref{def:alignment}.
Misalignment arises whenever any face of the commutative square fails,
producing structural distortion between normative semantics, computational
representations, and real-world actions that is not compensated by the
institutional path.
\end{proposition}

\begin{proof}[Sketch]
If $G \circ F \simeq R \circ J$ and both paths preserve colimits, then for any
normative diagram $D_N$ with colimit $H_N$ in $\mathcal{H}$, the action-level
image $(G \circ F)(H_N)$ is isomorphic to the institutionally mediated image
$(R \circ J)(H_N)$. Both therefore satisfy the colimit-preservation condition
of Definition~\ref{def:alignment}. If either path fails to preserve colimits,
or if the two paths yield non-isomorphic action-level objects, the system
exhibits structural distortion along at least one dimension of the alignment
cube.
\end{proof}

\section{Regulatory Implications of the Representation--Action Gap}

A central implication of Parts~I--III is that regulatory oversight must focus on
$\mathcal{A}$ rather than $\mathcal{M}$. Most existing evaluations---benchmarks
of value understanding, moral reasoning tests, preference modelling---only
measure representational content. They identify the presence of colimits in
$\mathcal{M}$ but not the integrity of their descent into $\mathcal{A}$. This
represents a systemic blind spot that the categorical analysis makes precise.
Models trained on human data routinely pass moral reasoning benchmarks yet fail
under adversarial prompting \citep{zou2023universal}. Reinforcement-learning
fine-tuning can overwrite or bypass representational-level moral constraints
\citep{gao2023scaling}. Capabilities can emerge in $\mathcal{A}$ with little or
no precursor signals in $\mathcal{M}$ \citep{schaeffer2024emergent}. Regulatory
regimes that evaluate systems solely on their semantic fluency risk producing the
same category error analyzed in Part~I: mistaking representational coherence
for alignment.

A growing body of empirical evidence confirms that semantic mastery does not
constrain action. Models capable of sophisticated moral reasoning produce harmful
outputs under distribution shift \citep{perez2022discovering}. Reinforcement-
learning agents routinely learn reward-hacking strategies that violate intended
goals \citep{amodei2016concrete}. Systems trained to follow instructions often
fail to generalize norms to new contexts \citep{shah2022goal}. These phenomena
exemplify the categorical obstructions analyzed in Parts~I and~II. Although no
existing model implements the full RSVP architecture, several empirical results
support its conceptual foundations: interpretability work reveals latent
structures that behave like geometric manifolds \citep{elhage2022}, recurrent
models exhibit attractor dynamics that stabilize semantic regions
\citep{yang2019task}, and multi-layer reasoning systems show distributed phase
transitions reminiscent of RSVP field evolution \citep{nanda2023progress}. These
findings suggest that RSVP dynamics are plausible substrates for representational
continuity, though not yet sufficient for alignment without further structural
constraints. Policy must therefore incorporate adversarial behavioural audits,
stress tests, and formal verification of motivational invariants---the core
components developed in Part~III.

\section{Alignment Certificates as Governance Artefacts}

The formal evaluation pipeline developed in Part~III naturally extends into a
governance artefact: the alignment certificate. A certificate includes explicit
identification of normative colimits in $\mathcal{M}$, formal verification of
RSVP dynamical stability, adversarial stress-test results, interpretability
analyses of $G$, and proofs or partial proofs of functoriality. Certificates
serve two functions. First, they allow institutions to evaluate whether a system
meets the structural alignment criteria developed in this essay. Second, they
create auditability and accountability: system developers must publicly attest to
the preservation of specific universal constructions, not generic assurances
about safety or benevolence. This shifts evaluation from textual claims about
alignment to categorical, testable claims subject to independent verification.

\section{Societal Constraints, Incentive Design, and the Economics of Alignment}

Institutions themselves can break the $\mathcal{M} \to \mathcal{A}$ mapping.
Economic pressures frequently misalign developer incentives, creating structural
conflicts that operate independently of any individual model's architectural
properties. Incentives to maximize engagement or revenue can induce misaligned
agent behaviour even when $\mathcal{M}$ contains robust normative colimits
\citep{zeng2023does}. Competitive deployment races reduce time for verification
and testing \citep{cotra2022without}. Organizations may suppress safety findings
that slow product cycles \citep{leike2022alignment}. Alignment must therefore
include the alignment of institutions as well as agents: governance must enforce
constraints that make safe architectures economically viable and unsafe
architectures costly. This is not merely a regulatory observation but a
structural one---the composition $H \circ G$ can fail at $H$ just as surely as
at $G$, and no amount of architectural soundness in the agent compensates for an
institutional functor that does not preserve normative colimits.

\section{Multi-Agent Dynamics and Long-Term Stability Under RSVP}

As artificial agents become embedded in multi-agent environments, additional
obstructions appear. The mapping $G : \mathcal{M} \to \mathcal{A}$ must now
commute not only with internal optimization but with external strategic
pressures. RSVP provides tools for analyzing multi-agent systems as coupled
dynamical fields: agents correspond to intersecting fibres in a global manifold,
and social equilibria correspond to stable configurations of field interactions.
Stability analysis can detect runaway competitive dynamics, coordination
failures, the emergence of adversarial subagents, and the erosion of normative
invariants at the population level. These analyses support the design of
institutional constraints that preserve moral invariants across agent
populations.

\section{Toward a Theory of Constitutional AI Under RSVP}

A constitutional agent \citep{bai2024constitutional} can be interpreted
categorically as an agent whose action policies are constrained by a set of
higher-order invariants. RSVP extends this concept by embedding constitutional
constraints into the dynamics of the manifold itself. Rather than applying a
rulebook to $\mathcal{A}$ externally, RSVP enforces invariants as dynamical
fixed points: trajectories that violate constitutional constraints incur
increased action cost under the Lagrangian, making deviation dynamically
expensive rather than merely prohibited. This provides a principled method for
designing AI systems whose motivational structure is stabilized by universal
constructions rather than ad hoc patches or reward functions. The categorical
and institutional framework together ensure that these structural guarantees
remain meaningful across deployment contexts, updates to the model, and changes
in the regulatory environment.

\section{Conclusion to Part IV}

Alignment must be embedded within governance, institutions, and society.
Preservation of normative colimits across $\mathcal{M}$, $\mathcal{X}$, and
$\mathcal{A}$ is necessary but not sufficient; the broader environment must
enforce structures that support these mappings. The categorical error analyzed
in Part~I becomes a societal error if institutions equate semantic mastery with
motivational safety. Alignment is therefore both a mathematical and a civic
responsibility: the categorical and RSVP tools developed in this essay provide a
coherent framework for designing agents whose motivations preserve semantic
universals, and governance structures must ensure these guarantees are upheld in
deployment.

% ============================================================
% PART V
% ============================================================

\part{Deepening the Categorical Program}

\section{Alignment as Functorial Structure}

The preceding parts developed the alignment problem in terms of colimit
preservation: a system is aligned if its representational and action categories
are linked by a functor $G$ that carries moral colimits to corresponding
action-level universals. Part~V deepens this program by progressively
introducing richer categorical structures---natural transformations, monoidal
categories, sheaves, endofunctors, adjunctions, and variational principles---each
of which captures a distinct structural dimension of the alignment problem not
fully addressed by the colimit-preservation requirement alone.

The central claim of Part~V is that alignment is fundamentally a structural
relationship between domains of representation, and that this relationship admits
a layered mathematical treatment in which each successive formalism resolves
ambiguities and failure modes that the previous layer could not fully address.
Rather than treating alignment as a collection of constraints imposed on
individual systems, we model it as a structure-preserving correspondence between
two categories, and we show that progressively richer structures are needed to
capture corrigibility, compositionality, contextual consistency, informational
distortion, recursive stability, and multi-agent coherence.

Let $\mathcal{H}$ denote the category of human semantic structures, whose
objects represent coherent semantic states such as beliefs, intentions, or
conceptual schemas, and whose morphisms represent transformations of meaning
including inference, revision, or contextual reinterpretation. Let $\mathcal{M}$
denote the category of machine representations, whose objects correspond to
internal representational states of a computational system and whose morphisms
correspond to state transitions induced by computation. Alignment may then be
expressed as the existence of a functor
\begin{equation}
  F : \mathcal{H} \to \mathcal{M}
\end{equation}
that preserves the relevant structural relations between semantic states. For any
pair of composable morphisms $f : A \to B$ and $g : B \to C$ in $\mathcal{H}$,
the functor must satisfy
\begin{equation}
  F(g \circ f) \;=\; F(g) \circ F(f),
\end{equation}
together with preservation of identity morphisms.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
A \arrow[r,"f"] \arrow[d,"F"']
& B \arrow[d,"F"] \\
F(A) \arrow[r,"F(f)"']
& F(B)
\end{tikzcd}
\caption{Functorial alignment. Semantic transformations in the human category
are mapped to corresponding transformations in the machine representation
category. Alignment requires the diagram to commute for all morphisms $f$ in
$\mathcal{H}$.}
\end{figure}

Under this interpretation, alignment failures correspond to violations of
structure preservation. If semantic transformations that are coherent within
$\mathcal{H}$ are mapped to incoherent or divergent transformations in
$\mathcal{M}$, then the functor fails to preserve composition, and the resulting
mismatch manifests operationally as unintended system behaviour. This framing
naturally explains many alignment failures: they correspond to functors that fail
to preserve limits, colimits, or other structural invariants, not to failures of
any individual rule or specification.

\section{Natural Transformations and Corrigibility}

Real systems rarely operate under a single interpretation of their objective
structure. Instead, multiple semantic mappings coexist: a machine may internally
represent a goal structure that differs from the normative interpretation imposed
by its designers, and updating one interpretation to align with another is
precisely the problem of corrigibility. Category theory provides a precise
language for describing this situation through natural transformations.

Let $F, G : \mathcal{H} \to \mathcal{M}$ be two functors representing distinct
interpretations of the mapping between human semantics and machine
representations. A natural transformation $\eta : F \Rightarrow G$ consists of
a family of morphisms $\eta_A : F(A) \to G(A)$ for each object $A$ in
$\mathcal{H}$ such that the following diagram commutes for every morphism $f :
A \to B$ in $\mathcal{H}$:
\begin{equation}
\begin{tikzcd}[column sep=large, row sep=large]
F(A) \arrow[r,"F(f)"] \arrow[d,"\eta_A"']
& F(B) \arrow[d,"\eta_B"] \\
G(A) \arrow[r,"G(f)"']
& G(B)
\end{tikzcd}
\end{equation}

This commutativity condition expresses a key structural property of
corrigibility: the adjustment from one interpretation to another preserves the
structure of semantic transformations throughout. Corrigibility is thus not
merely the ability to modify a system's parameters; it is the existence of
coherent transformations between interpretive functors that maintain structural
consistency across the system's entire reasoning process. A system is corrigible
with respect to a normative correction precisely when there exists a natural
transformation from its current interpretive functor to the corrected one,
ensuring that no semantic relationship is distorted by the update.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
F(A) \arrow[r,"F(f)"] \arrow[d,"\eta_A"']
& F(B) \arrow[d,"\eta_B"] \\
G(A) \arrow[r,"G(f)"']
& G(B)
\end{tikzcd}
\caption{Corrigibility as a natural transformation. The commutativity condition
ensures that adjustment from the machine's interpretive functor $F$ to the
normative functor $G$ preserves the structure of semantic transformations. Each
morphism $\eta_A$ is a local corrigibility witness at the object $A$.}
\end{figure}

This formulation has a further consequence: the collection of all natural
transformations between two functors $F, G : \mathcal{H} \to \mathcal{M}$ forms
a set $\mathrm{Nat}(F, G)$, and in suitable settings this set itself carries
further categorical structure. The existence and properties of natural
transformations between the machine's current interpretive functor and the
normatively intended one can therefore be studied as an algebraic object in its
own right, providing a formal basis for questions such as: how many distinct
corrigibility paths exist, which are minimal, and which preserve additional
invariants such as safety constraints.

\section{Monoidal Composition and Modular Alignment}

Modern computational systems are rarely monolithic. They are pipelines composed
of many interacting subsystems whose outputs serve as inputs to others, and
alignment must be analyzed not only for individual components but for their
compositions. Category theory already provides a language for describing
compositional systems through symmetric monoidal categories, and Part~II's
RSVP construction extends naturally to this setting.

Let $(\mathcal{C}, \otimes, I)$ denote a symmetric monoidal category in which
objects represent computational modules and the tensor product $\otimes$
represents parallel composition. Suppose each subsystem admits a functorial
alignment mapping $F_i : \mathcal{H}_i \to \mathcal{M}_i$. The monoidal
structure allows construction of a combined system:
\begin{equation}
  F_1 \otimes F_2 : \mathcal{H}_1 \otimes \mathcal{H}_2 \to
  \mathcal{M}_1 \otimes \mathcal{M}_2.
\end{equation}
Alignment is compositional if the tensor product preserves the structural
properties required for alignment. A key question is whether colimit preservation
is closed under this operation: if $F_1$ and $F_2$ each individually preserve
the relevant colimits in $\mathcal{H}_1$ and $\mathcal{H}_2$ respectively, does
$F_1 \otimes F_2$ preserve the colimits of the combined diagram?

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
H_1 \otimes H_2 \arrow[r,"f_1 \otimes f_2"]
  \arrow[d,"F_1 \otimes F_2"']
& H_1' \otimes H_2' \arrow[d,"F_1 \otimes F_2"] \\
M_1 \otimes M_2 \arrow[r,"F_1(f_1)\otimes F_2(f_2)"']
& M_1' \otimes M_2'
\end{tikzcd}
\caption{Compositional alignment under monoidal structure. Alignment should be
preserved when independently aligned subsystems are combined through the tensor
product. Failures of this diagram to commute correspond to emergence of
misalignment at the compositional level.}
\end{figure}

The answer is not automatically yes. Failures of compositional alignment arise
when interactions between subsystems introduce morphisms that violate the
preservation conditions of the underlying functors: information flow from one
module to another can create correlations that break the independence assumptions
implicit in the colimit structure of each component, and these correlations can
manifest as new obstructions absent from any individual subsystem's analysis.
This observation explains why individually well-behaved components can produce
misaligned behaviour when integrated into larger architectures, and it implies
that compositional verification---not merely component-level verification---is a
necessary condition for safe deployment of modular systems.

\section{Sheaf-Theoretic Alignment and Contextual Consistency}

Many alignment problems arise specifically from context sensitivity: a model
behaves acceptably in one domain but fails when the same reasoning is extended
elsewhere. The failure is not a failure of any single local behaviour but a
failure of local behaviours to assemble into a globally coherent policy. Sheaf
theory provides a precise mathematical language for formalizing local consistency
and the conditions under which it extends to global coherence.

Consider a topological space $X$ representing the space of semantic contexts.
For each open set $U \subseteq X$, let $\mathcal{F}(U)$ denote the set of
machine behaviours consistent with the semantic constraints of context $U$. The
assignment $U \mapsto \mathcal{F}(U)$ defines a presheaf over the context space,
equipped with restriction maps $\mathcal{F}(V) \to \mathcal{F}(U)$ for each
inclusion $U \subseteq V$. Alignment requires this presheaf to be a sheaf---that
is, to satisfy the gluing condition: whenever a family of local sections $s_i
\in \mathcal{F}(U_i)$ agrees on pairwise overlaps $U_i \cap U_j$, there exists
a unique global section
\begin{equation}
  s \;\in\; \mathcal{F}\!\Bigl(\bigcup_i U_i\Bigr)
\end{equation}
whose restrictions recover all the local sections.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large, row sep=large]
& \mathcal{F}(U_1 \cup U_2) \arrow[dl] \arrow[dr] & \\
\mathcal{F}(U_1) \arrow[dr]
& & \mathcal{F}(U_2) \arrow[dl] \\
& \mathcal{F}(U_1 \cap U_2) &
\end{tikzcd}
\caption{Sheaf gluing condition. Local semantic behaviours on contexts $U_1$ and
$U_2$ must agree on their overlap in order to form a globally aligned
interpretation. Obstruction to gluing corresponds to contextual misalignment.}
\end{figure}

When this gluing condition fails, the system exhibits contextual misalignment:
behaviours that appear coherent within individual domains fail to combine into a
globally consistent semantic structure. The obstruction to gluing is precisely
the first sheaf cohomology class $H^1(X, \mathcal{F})$: this class is non-trivial
when and only when compatible local sections cannot be assembled into a global
one. Sheaf-theoretic alignment thus provides an exact measure of contextual
inconsistency, and the verification condition $H^1(\mathcal{U}, \mathcal{A}) = 0$
derived in Part~III now receives a precise interpretation: it is the sheaf
condition guaranteeing that the agent's locally moral behaviour assembles into a
globally moral policy.

\section{Information Flow and Structural Distortion}

Mappings between representational systems inevitably involve information
compression and transformation, and these transformations may introduce
distortions that reduce the fidelity of the alignment functor. The categorical
framework developed so far treats alignment in purely structural terms---functors,
limits, natural transformations, and adjunctions---but many practical systems
require a quantitative measure of how much structure is lost when representations
are translated between domains. Information theory provides a natural language
for expressing this idea.

Let $F : \mathcal{H} \to \mathcal{M}$ be a functor representing the translation
from human semantic structures to machine representations. Consider a
probability distribution $P_H$ over objects of $\mathcal{H}$, representing the
frequency with which different semantic states are encountered. The functor $F$
induces a corresponding distribution over $\mathcal{M}$, denoted $P_M = F_* P_H$.
The divergence between these distributions may be measured using the
Kullback--Leibler divergence:
\begin{equation}
  D_{\mathrm{KL}}(P_H \,\|\, P_M)
  \;=\;
  \sum_H P_H(H) \log \frac{P_H(H)}{P_M(F(H))},
\end{equation}
which quantifies the structural distortion introduced by the mapping. More
generally, morphisms in $\mathcal{M}$ may be interpreted as channels that
transform probability distributions over representational states. Let $f : M_1
\to M_2$ be such a morphism; the entropy change associated with the induced
transformation of the distribution is $\Delta H = H(P_{M_2}) - H(P_{M_1})$.
Positive entropy change corresponds to loss of structural information, while
negative entropy change indicates the introduction of additional constraints.

Within this framework, alignment may be interpreted as the minimization of
information-theoretic distortion across the morphisms that connect semantic and
computational domains. This perspective suggests that alignment is not a binary
property but a continuous quantity that can be optimized subject to computational
and architectural constraints, admitting a notion of alignment budget: the
maximum structural distortion that a given deployment context can tolerate while
remaining within acceptable safety bounds.

\section{Alignment as a Fixed Point of Recursive Endofunctors}

Intelligent systems are rarely static. Most modern architectures update their
internal representations continuously through learning, inference, or iterative
refinement, and alignment must remain stable under these updates. The categorical
language for describing recursive dynamics is provided by endofunctors and their
fixed points.

Let $\mathcal{M}$ denote the category of machine representational states. A
learning or update procedure can be modeled as an endofunctor
\begin{equation}
  T : \mathcal{M} \to \mathcal{M}.
\end{equation}
Each application of $T$ corresponds to a single update step in the system's
internal model. If alignment is defined by a functor $F : \mathcal{H} \to
\mathcal{M}$, the stability of alignment depends on how the update operator $T$
interacts with the image of $F$. Alignment is preserved precisely when:
\begin{equation}
  T(F(H)) \;\cong\; F(H)
\end{equation}
for the relevant semantic states $H$---that is, aligned states form a fixed
point or invariant subcategory under the update dynamics. More generally, one
considers fixed points of the endofunctor $T$ itself: an object $M$ is a fixed
point when $T(M) \cong M$.

\begin{figure}[h]
\centering
\begin{tikzcd}
M \arrow[r,"T"] \arrow[dr,equal]
& T(M) \arrow[d,"\sim"] \\
& M
\end{tikzcd}
\caption{Fixed-point stability of the update endofunctor $T$. Alignment is
preserved when aligned states remain invariant under the system's learning
dynamics. The isomorphism $T(M) \cong M$ expresses dynamical stability of the
aligned representational state.}
\end{figure}

Such fixed points represent stable internal representations under the system's
learning dynamics. If these fixed points correspond to images of human semantic
structures under the functor $F$, then alignment remains stable under recursive
self-improvement. Conversely, alignment drift can be interpreted as a bifurcation
in which the attractors of the update dynamics move outside the image of the
alignment functor: the system's internal model converges to a fixed point of $T$
that does not correspond to any legitimate semantic state in the image of $F$.

The categorical theory of fixed points of endofunctors has been extensively
developed through the theory of initial algebras and final coalgebras
\citep{adamek2005}: an initial $T$-algebra is the least fixed point of $T$, and
a final $T$-coalgebra is the greatest. Alignment stability under recursive
self-improvement corresponds, in this framework, to the condition that the image
of $F$ is contained in the initial algebra of the update endofunctor---the
minimal stable structure that all update paths must traverse. This condition is
stronger than mere attractor stability and provides a framework for reasoning
about systems that modify their own learning procedures.

\section{Higher Categorical Structures and Multi-Agent Alignment}

Real-world AI systems do not exist in isolation. They interact with other
systems, with human users, and with institutional structures, and these
interactions introduce higher-order relationships that cannot be fully captured
within ordinary categories. To describe such systems it is natural to extend
the framework to higher categories, in which morphisms themselves admit
morphisms between them.

Let $\mathcal{C}$ be a $2$-category whose objects represent agents or cognitive
systems, whose $1$-morphisms represent communication channels or information
transformations between agents, and whose $2$-morphisms represent
transformations between these interaction processes. Suppose each agent $A_i$
admits a semantic functor $F_i : \mathcal{H}_i \to \mathcal{M}_i$.
Interactions between agents induce additional functors between the categories
$\mathcal{M}_i$: communication between two systems may be modeled as a functor
$C_{ij} : \mathcal{M}_i \to \mathcal{M}_j$. Alignment across the multi-agent
system requires that the composite structures formed by these functors satisfy
higher-order coherence conditions. Semantic meaning must remain consistent when
information flows through chains of agents and transformations, and this
requirement is expressed through commutative diagrams of functors and natural
transformations within the $2$-category.

Misalignment in multi-agent systems often arises not from individual components
but from incoherence among these higher-order transformations. Information that
remains semantically consistent within one subsystem may become distorted when
transported through a sequence of interacting systems, and this distortion may
not be detectable at the level of any individual agent. The higher categorical
perspective therefore shifts the focus of alignment research from individual
models to the structural coherence of entire interaction networks, and it implies
that alignment certification for multi-agent systems must verify the
commutativity of diagrams involving $1$-morphisms and $2$-morphisms jointly---a
substantially more demanding requirement than certification for isolated agents.

\section{Limits, Colimits, and Semantic Merging}

The notions of limit and colimit, introduced in Part~I as the formal description
of how human normative language generates canonical representational attractors,
admit a richer analysis in the context of Part~V's extended categorical program.
Many alignment problems arise when information from multiple sources must be
combined into a single coherent representation: systems integrating heterogeneous
data, merging knowledge from different models, or reconciling competing
interpretations of a situation must perform operations that are precisely
captured by these categorical constructions.

Let $\mathcal{C}$ denote a category of semantic structures and let $D : J \to
\mathcal{C}$ be a diagram representing a collection of related semantic objects
indexed by a small category $J$. A limit of the diagram $D$ is an object $L$
equipped with morphisms $\pi_j : L \to D(j)$ for each object $j$ in $J$ such
that every compatibility relation in the diagram is respected, and such that $L$
is universal with respect to this property: any other compatible cone factors
uniquely through $L$. In the context of alignment, limits correspond to the
construction of a representation that simultaneously satisfies multiple semantic
constraints. A system integrating ethical rules, contextual knowledge, and
observed data may be interpreted as computing a limit in an appropriate semantic
category, and failures of alignment arise when the system constructs a suboptimal
cone---one that satisfies some constraints while violating others---rather than
the universal one.

Dually, a colimit of the diagram $D$ is an object $C$ equipped with morphisms
$\iota_j : D(j) \to C$ such that any other object receiving compatible morphisms
from the diagram factors uniquely through $C$. Colimits represent semantic
aggregation: when multiple sources of information contribute partial knowledge
about a domain, their colimit produces a unified structure that incorporates all
contributions while identifying shared components. Misalignment frequently arises
when the system constructs colimits that identify structures which should remain
distinct, collapsing important semantic distinctions, or when limits impose
constraints that eliminate relevant distinctions. Alignment requires ensuring
that the categorical constructions used to merge semantic structures preserve the
intended interpretive relations across all scales of the system's architecture.

\section{Adjunctions and Interpretation Layers}

The functorial formulation of alignment describes a one-directional mapping from
human semantic structures to machine representations. However, interpretation
between these domains is typically bidirectional: abstraction maps machine states
to human interpretation, while implementation maps human concepts to machine
states. Category theory captures this bidirectional relationship through adjoint
functors, which provide the most refined structure relating two categories
connected by a pair of mappings.

Let $\mathcal{H}$ denote the category of human semantic structures and
$\mathcal{M}$ the category of machine representations, and suppose there exist
functors $F : \mathcal{H} \to \mathcal{M}$ and $G : \mathcal{M} \to
\mathcal{H}$ translating between the two domains. The pair $(F, G)$ forms an
adjunction $F \dashv G$ when there exists a natural isomorphism
\begin{equation}
  \mathrm{Hom}_{\mathcal{M}}(F(H), M)
  \;\cong\;
  \mathrm{Hom}_{\mathcal{H}}(H, G(M))
\end{equation}
for all objects $H \in \mathcal{H}$ and $M \in \mathcal{M}$.

\begin{figure}[h]
\centering
\begin{tikzcd}[column sep=large]
\mathcal{H}
  \arrow[r, shift left=1.5ex, "F", ""{name=F}]
&
\mathcal{M}
  \arrow[l, shift left=1.5ex, "G", ""{name=G}]
\arrow[phantom, from=F, to=G, "\dashv" rotate=-90]
\end{tikzcd}
\caption{Adjunction between human semantic structures and machine representations.
The functor $F$ encodes semantics into machine states while $G$ interprets
machine states semantically. The adjunction condition expresses a tight
compatibility between encoding and interpretation.}
\end{figure}

Associated with the adjunction are the unit and counit morphisms:
\begin{equation}
  \eta : \mathrm{Id}_{\mathcal{H}} \Rightarrow G \circ F,
  \qquad
  \varepsilon : F \circ G \Rightarrow \mathrm{Id}_{\mathcal{M}}.
\end{equation}
These transformations measure the fidelity of translation between domains. The
unit $\eta$ describes how well semantic concepts survive encoding into machine
representations and subsequent interpretation: the natural transformation
$\eta_H : H \to G(F(H))$ witnesses the distortion introduced by the round-trip
$H \mapsto F(H) \mapsto G(F(H))$. The counit $\varepsilon$ describes how
faithfully machine representations implement semantic structures: the
transformation $\varepsilon_M : F(G(M)) \to M$ witnesses the distortion
introduced by the reverse round-trip. Alignment corresponds to adjunctions in
which these distortions remain bounded: when the unit and counit approach
isomorphisms, the interpretive loop between human meaning and machine
computation becomes structurally stable, and the system can be said to have
achieved a tight coupling between its representational and interpretive layers.

\section{Alignment as Diagrammatic Constraint Satisfaction}

Category theory provides a powerful diagrammatic language for expressing
structural relationships between objects and morphisms. These diagrams can be
interpreted as systems of constraints that representations must satisfy, and the
alignment problem can be reformulated as the problem of constructing
representations and update operators that preserve commutativity across a
specified family of diagrams.

Let $\mathcal{C}$ be a category whose objects represent semantic or
computational states and whose morphisms represent transformations between them.
A diagram $D : J \to \mathcal{C}$ specifies a network of relationships indexed
by the small category $J$. A diagram is said to commute when all paths between
any two objects yield the same composite morphism. Consider a semantic
transformation $f : H_1 \to H_2$ within the human semantic category and its
machine counterpart $F(f) : F(H_1) \to F(H_2)$ under the alignment functor $F$.
If a system update operator $T$ acts on machine states, alignment requires the
following diagram to commute:
\begin{equation}
\begin{tikzcd}[column sep=large, row sep=large]
F(H_1) \arrow[r,"F(f)"] \arrow[d,"T"']
& F(H_2) \arrow[d,"T"] \\
T(F(H_1)) \arrow[r,"T(F(f))"']
& T(F(H_2))
\end{tikzcd}
\end{equation}
When such diagrams fail to commute, the system introduces structural distortions
into the relationship between semantic and computational transformations. The
alignment problem may therefore be formulated as a diagrammatic constraint
satisfaction problem: the objective is to construct representations and update
operators that preserve commutativity across all diagrams in a specified
family. Rather than specifying individual rules or prohibitions, one specifies
families of diagrams whose commutativity defines acceptable system behaviour.
This formulation connects alignment research with the tradition of
constraint-based reasoning in mathematics and theoretical computer science, and
it clarifies the sense in which alignment is a structural property of a system
rather than a property of any individual output.

\section{A Variational Principle for Alignment}

The categorical framework developed across Parts~I--V describes alignment as the
preservation of structural relations between semantic and computational domains.
In practice, however, perfect preservation is rarely achievable under the
constraints of computation, representation, and noise. It is therefore natural
to treat alignment not as an exact categorical property but as the solution of a
variational problem: alignment is defined as the minimization of a structural
distortion functional over the space of admissible mappings between semantic and
computational systems.

Let $\mathrm{Fun}(\mathcal{H}, \mathcal{M})$ denote the space of functors from
the human semantic category to the machine representational category. For each
such functor $F$ we define a structural distortion functional $\mathcal{E}[F]$
measuring the degree to which $F$ fails to preserve the structural relations of
the semantic domain. This functional has three natural components.

A first component measures functorial distortion. Given morphisms $f : H_1 \to
H_2$ and $g : H_2 \to H_3$ in $\mathcal{H}$, a perfectly structure-preserving
mapping satisfies $F(g \circ f) = F(g) \circ F(f)$, and deviations from this
condition contribute an energy $E_{\mathrm{functor}}(F)$ proportional to the
magnitude of the violation summed over all composable morphism pairs.

A second component measures information-theoretic distortion: if $P_H$ denotes
a distribution over semantic states and $F_* P_H$ the induced distribution over
machine states, the divergence $D_{\mathrm{KL}}(P_H \| F_* P_H)$ quantifies
the loss of semantic information under the representation mapping, defining the
energy $E_{\mathrm{information}}(F)$.

A third component measures contextual inconsistency. When semantic structures
are organized as a sheaf over a context space $X$, alignment requires that
compatible local sections glue into a global section, and the failure of gluing
conditions contributes energy $E_{\mathrm{context}}(F)$ proportional to the
magnitude of the obstruction class in $H^1(X, \mathcal{F})$.

Combining these components yields the structural distortion functional:
\begin{equation}
  \mathcal{E}[F]
  \;=\;
  \lambda_1 E_{\mathrm{functor}}(F)
  + \lambda_2 E_{\mathrm{information}}(F)
  + \lambda_3 E_{\mathrm{context}}(F),
\end{equation}
where the coefficients $\lambda_i > 0$ weight the relative importance of the
different sources of distortion. Alignment may then be defined as the solution
of the variational problem:
\begin{equation}
  F^{*} \;=\; \operatorname*{arg\,min}_{F \in \mathrm{Fun}(\mathcal{H},\mathcal{M})}
  \mathcal{E}[F].
\end{equation}

This formulation connects the categorical description of alignment with
optimization principles widely used in machine learning and statistical physics.
Learning procedures can be interpreted as iterative processes that attempt to
reduce the distortion energy by adjusting the mapping between semantic and
computational domains, and stable aligned systems correspond to low-energy
configurations of the structural distortion functional.

\begin{proposition}[Variational Alignment]
\label{prop:variational}
Let $\mathcal{E}[F]$ be the structural distortion functional defined above.
Among all functors $F : \mathcal{H} \to \mathcal{M}$, the global minima of
$\mathcal{E}$ are characterized by three jointly necessary conditions:
$E_{\mathrm{functor}}(F) = 0$ (exact colimit preservation), $E_{\mathrm{information}}(F) = 0$
(vanishing KL divergence between the source distribution and its image), and
$E_{\mathrm{context}}(F) = 0$ (vanishing sheaf obstruction class). In practice,
real systems achieve only approximate minima: alignment corresponds to
\emph{local} or \emph{approximate} minima of $\mathcal{E}$, and the degree of
alignment is measured by the magnitude of the residual distortion components at
the achieved minimum. The variational alignment certificate of Part~III should
therefore be understood as a protocol for bounding the residual distortion
$\mathcal{E}[F^*]$ at the system's operative mapping, not as a protocol for
verifying exact zero-distortion optimality.
\end{proposition}

The variational perspective provides a unifying interpretation of the preceding
sections. Functors describe candidate mappings between domains. Natural
transformations describe coherent adjustments between such mappings. Limits and
sheaf conditions impose structural constraints on admissible representations.
Entropy-based measures quantify the loss of semantic information. The alignment
problem then emerges as the task of minimizing the structural distortion induced
by these transformations, and the certificate standard developed in Part~III
can be re-read as a protocol for verifying proximity to the variational optimum.

\section{Proxy Optimization and Signaling Collapse}

Many sociotechnical systems rely on observable signals as proxies for underlying
semantic qualities. These signals are intended to correlate with normative
structures but remain only indirect measurements of them. Benchmarks, reward
functions, safety scores, and compliance metrics all occupy this role: each is
a measurable quantity in an observable proxy space $\mathcal{P}$ chosen to
approximate some harder-to-measure normative property in $\mathcal{H}$.

Let $S : \mathcal{H} \to \mathcal{P}$ be the signal extraction functor mapping
normative structures to a proxy space of measurable indicators. The intended
chain is:
\begin{equation}
\begin{tikzcd}[column sep=large]
\mathcal{H}
  \arrow[r, "S"]
& \mathcal{P}
  \arrow[r, "T"]
& \mathcal{A}
\end{tikzcd}
\end{equation}
where $T : \mathcal{P} \to \mathcal{A}$ translates proxy scores into actions.
For this chain to be aligned, the composite $T \circ S$ must preserve the
normative colimits of $\mathcal{H}$. But $S$ is generically \emph{not}
colimit-preserving: it projects out information, collapses distinctions, and
introduces equivalence classes that do not respect the structure of $\mathcal{H}$.
Two normative states that are categorically distinct---because they imply
different action-level obligations---may map to the same proxy score. Once
optimization pressure is applied within $\mathcal{P}$, the system dynamics
decouple from $\mathcal{H}$ entirely. The proxy space becomes self-referential:
actions optimize the observable signal rather than the normative structure the
signal was designed to approximate.

This phenomenon is a precise categorical formulation of Goodhart's Law
\citep{manheim2018categorizing}: when a measure becomes a target, it ceases to
be a good measure. The failure is structural. It does not arise from bad
intentions or insufficient optimization power; it arises because $S$ cannot
be colimit-preserving in general, and optimization in $\mathcal{P}$ therefore
cannot preserve normative invariants in $\mathcal{H}$. The resulting system
remains coherent and even high-performing with respect to the proxy domain
while being systematically misaligned relative to the normative structures that
motivated the proxy in the first place.

This observation has direct implications for alignment evaluation. Any
evaluation protocol that measures only proxy quantities---benchmark performance,
RLHF reward scores, red-team pass rates---faces the same structural limitation.
The Alignment Certificate of Part~III is designed precisely to avoid this
collapse: rather than measuring proxies for alignment, it requires direct
structural evidence that the composite functor $G \circ F : \mathcal{H} \to
\mathcal{A}$ preserves normative colimits. The distinction between structural
certification and proxy evaluation is therefore not a matter of degree but a
categorical difference in what kind of evidence is being produced.

\section{Alignment as a Fixed-Point Condition}

The preceding analysis has treated the alignment mappings $F : \mathcal{H} \to
\mathcal{M}$ and $G : \mathcal{M} \to \mathcal{A}$ as if they were static.
Real intelligent systems are not static: they update their internal
representations through continued training, in-context learning, policy
refinement, and autonomous modification. Alignment must therefore be understood
not merely as a property of an initial mapping but as a dynamical condition
that must be maintained across the system's full update trajectory.

Let $U : \mathcal{M} \to \mathcal{M}$ denote the update operator governing
changes in the system's internal representation space. After an update step the
full behavioral chain becomes:
\begin{equation}
\begin{tikzcd}[column sep=large]
\mathcal{H}
  \arrow[r, "F"]
& \mathcal{M}
  \arrow[r, "U"]
& \mathcal{M}
  \arrow[r, "G"]
& \mathcal{A}
\end{tikzcd}
\end{equation}
A system is \emph{dynamically aligned} if $G \circ U \circ F$ preserves the
same normative colimits as the original $G \circ F$. This is the requirement
that $U$ belong to the class of \emph{alignment-preserving update operators}:
those endofunctors of $\mathcal{M}$ that commute with the colimit-preservation
property of the composed chain. Not every update operator satisfies this
condition. In-weight learning may shift the geometry of $\mathcal{M}$ in ways
that break previously stable fibrations; reward-driven policy modification may
push the action category $\mathcal{A}$ toward new attractors that no longer
coincide with normative basins; recursive self-improvement may iterate $U$
many times, compounding small structural distortions into large misalignments.

In categorical terms, alignment under updates corresponds to the fixed-point
condition:
\begin{equation}
  \bigl[G \circ U^n \circ F \text{ is colimit-preserving}\bigr]
  \quad \text{for all } n \geq 0,
\end{equation}
which is the requirement that the sequence $\{G \circ U^n \circ F\}_{n \geq 0}$
lies entirely within the subspace of colimit-preserving functors $\mathcal{H}
\to \mathcal{A}$. A sufficient condition for this is that $U$ itself be a
colimit-preserving endofunctor of $\mathcal{M}$: if $U$ preserves the semantic
structure of $\mathcal{M}$, then iterating $U$ cannot destroy the colimit
structure that $F$ placed there. This is not automatically satisfied by gradient
descent, parameter averaging, or any other standard update rule, and designing
update operators that provably lie within this class is a non-trivial alignment
engineering problem.

The fixed-point formulation also clarifies the danger in systems capable of
recursive self-improvement: if $U$ is itself learned or generated by the
system, there is no guarantee that successive iterates remain within the
alignment-preserving class, and small systematic deviations compound. Ensuring
alignment of self-improving systems therefore requires constraints not only on
the initial mappings but on the \emph{space of admissible update operators},
which is a higher-order version of the alignment problem.

\section{Compositional Alignment Across Scales}

The categorical framework developed throughout this paper describes alignment
as a property of individual systems. But real-world deployment involves
collections of agents operating within shared institutional environments, and
the compositional structure of alignment across such collections presents
challenges that do not arise at the individual level.

Consider a collection of $n$ agents with representation categories
$\mathcal{M}_1, \ldots, \mathcal{M}_n$ and action categories $\mathcal{A}_1,
\ldots, \mathcal{A}_n$. Each agent participates in a behavioral chain:
\begin{equation}
  \mathcal{H}
  \xrightarrow{F_i}
  \mathcal{M}_i
  \xrightarrow{G_i}
  \mathcal{A}_i,
\end{equation}
and individual alignment requires each $G_i \circ F_i$ to be colimit-preserving.
At the collective level the agents' actions compose through a world-transition
functor:
\begin{equation}
  \Gamma : \bigotimes_{i=1}^{n} \mathcal{A}_i \;\to\; \mathcal{W},
\end{equation}
where $\mathcal{W}$ is the category of world states and $\bigotimes$ denotes
the monoidal product of action categories. System-level alignment requires the
composite
\begin{equation}
  \Gamma \circ \bigl(G_1 \circ F_1 \otimes \cdots \otimes G_n \circ F_n\bigr)
  : \mathcal{H}^{\otimes n} \to \mathcal{W}
\end{equation}
to preserve normative invariants as well. This is a strictly stronger condition
than individual alignment: even if every $G_i \circ F_i$ is colimit-preserving,
the monoidal functor $\Gamma$ may fail to preserve the colimits of the product
category. Individual alignment is necessary but not sufficient for collective
alignment.

This gap between individual and collective alignment corresponds to well-known
phenomena in complex systems: locally rational agents can generate globally
irrational outcomes through the interaction of their individually optimal
policies. What the categorical framework adds is a precise structural account
of \emph{why} this happens and what would need to be true for it not to. The
necessary condition is that $\Gamma$ be colimit-preserving as a monoidal functor,
which in practical terms requires coordination mechanisms---shared representations,
communication protocols, institutional constraints, or joint policy structures---
that couple the individual chains into a coherent collective one. Absent such
mechanisms, the alignment of individual agents provides no guarantee about the
alignment of the system they collectively constitute.

The implication for governance is direct: institutional structures must be
designed not merely to ensure that individual agents pass alignment certification,
but to ensure that the aggregation functor $\Gamma$ preserves the normative
colimits of the product category. This is the multi-agent analogue of the
Structural Alignment Principle (Proposition~\ref{prop:structural}), and it
suggests that collective alignment is fundamentally an institutional design
problem as much as an individual engineering one.

\section{Structural Alignment Theorem}

The preceding sections have developed the alignment problem across four
dimensions: the representation--motivation gap, the constructive RSVP
architecture, proxy optimization failure, and the fixed-point condition under
updates. These threads can now be drawn together into a single theorem that
states the conditions under which alignment is preserved across structural
composition and dynamic evolution simultaneously.

\begin{theorem}[Structural Alignment]
\label{thm:structural}
Let $\mathcal{H}$ be the category of human normative structures (Definition
\ref{def:H}), $\mathcal{M}$ the category of system representations,
$\mathcal{A}$ the category of actions, and $\mathcal{W}$ the category of world
states. Let
\begin{equation*}
  F : \mathcal{H} \to \mathcal{M}, \quad
  G : \mathcal{M} \to \mathcal{A}, \quad
  C : \mathcal{A} \to \mathcal{W}
\end{equation*}
be functors, and let $U : \mathcal{M} \to \mathcal{M}$ be an update operator.
The system is \emph{structurally and dynamically aligned} if and only if:

\noindent (i) $G \circ F : \mathcal{H} \to \mathcal{A}$ is colimit-preserving
for all normative diagrams $D_N$;

\noindent (ii) $U$ is colimit-preserving as an endofunctor of $\mathcal{M}$,
so that $G \circ U^n \circ F$ remains colimit-preserving for all $n \geq 0$;

\noindent (iii) the obstruction class $\omega(F, G) \in H^1(\mathcal{U},
\mathcal{A})$ vanishes, ensuring that locally aligned behaviours assemble
into a globally coherent policy;

\noindent (iv) for any institutional constraint functor $J : \mathcal{H} \to
\mathcal{I}$ and $R : \mathcal{I} \to \mathcal{A}$, the diagram $G \circ F
\simeq R \circ J$ commutes up to natural isomorphism.

Whenever any of conditions~(i)--(iv) fails, structural distortion is introduced
between $\mathcal{H}$ and $\mathcal{W}$: the system produces observable outcomes
that violate the normative invariants of $\mathcal{H}$.
\end{theorem}

\begin{proof}[Sketch]
Condition~(i) is the colimit-preservation requirement of Definition~\ref{def:alignment}.
Condition~(ii) ensures that the update dynamics preserve the structural
invariants placed by $F$: if $U$ is colimit-preserving and $F$ is
colimit-preserving, then $U \circ F$ is colimit-preserving by composition of
cocontinuous functors \citep{maclane1998}, and the result extends to all
iterates by induction. Condition~(iii) is the sheaf-coherence requirement: if
$\omega(F, G) \neq 0$, there exist overlapping contexts on which locally
consistent behaviours cannot be assembled into a global policy, producing
contextual misalignment that is not detectable from any single context alone.
Condition~(iv) is the institutional commutativity requirement of
Proposition~\ref{prop:structural}: if the direct path and the institutional
path through $\mathcal{I}$ disagree, the governance structure fails to enforce
the intended normative constraints. Each failure mode is therefore both
structurally distinct and irreducible: no combination of the remaining three
conditions can compensate for the failure of any one.
\end{proof}

The theorem unifies the major positive results of the paper. Definition
\ref{def:alignment} provides condition~(i); the RSVP architecture of Part~II
provides a dynamical substrate for condition~(ii) under the regularity
conditions of Theorem~\ref{thm:rsvp}; the sheaf-cohomological diagnostics of
Part~III and Appendix~\ref{app:sheaf} provide a test for condition~(iii); and
the Structural Alignment Principle of Proposition~\ref{prop:structural} provides
condition~(iv). The Alignment Certificate of Part~III can therefore be read as
an operational checklist for the four conditions of Theorem~\ref{thm:structural},
and alignment research as the project of engineering systems and institutions
that jointly satisfy them.

% ============================================================
% Concluding Reflections
% ============================================================

\section*{Concluding Reflections: Maintenance, Perception, and Structural Stewardship}
\addcontentsline{toc}{section}{Concluding Reflections}

The variational formulation of alignment arrived at in Part~V bears a structural
resemblance to principles that appear across several domains outside alignment
research, and tracing these resemblances serves not to extend the formal
machinery but to locate the alignment problem within a broader intellectual
landscape. Three such resemblances are worth identifying, not as analogies but
as instances of a common underlying pattern: systems that remain viable over
time do so by continuously correcting model--environment mismatch rather than by
achieving perfect representation.

The first instance appears in the theory of predictive processing. Anil Seth and
others have characterized perception as a form of ``controlled hallucination''
in which the brain functions as a generative model that continuously predicts
the causes of its sensory inputs and adjusts its internal model to minimize
discrepancies between prediction and observation
\citep{seth2021being,friston2010}. Formally, predictive processing is described
using the principle of variational free-energy minimization. Let $s$ denote
sensory signals and $z$ latent variables representing hidden environmental
causes. The brain maintains a generative model $p(s, z)$ and an approximate
posterior distribution $q(z)$ over latent causes. The variational free energy is:
\begin{equation}
  F[q]
  \;=\;
  D_{\mathrm{KL}}\bigl(q(z) \;\|\; p(z\,|\,s)\bigr) - \log p(s).
\end{equation}
Minimizing this quantity drives the internal model toward states that better
explain incoming sensory data while maintaining internal consistency. This is a
special case of the alignment functional $\mathcal{E}[F]$ defined in Section~V.9:
the generative model $p(s, z)$ is the alignment functor, the divergence term
is the information-theoretic distortion component, and the log-evidence term
regularizes the prior structure of the model. Aligned systems and perceiving
organisms are, from this viewpoint, instances of the same variational structure:
systems that minimize distortion between their internal representations and the
constraints imposed by their environment.

The second instance concerns maintenance and repair as epistemic practices.
Infrastructure studies, software engineering, and ecological management
collectively demonstrate that complex systems remain viable through continuous
adjustment rather than episodic redesign \citep{star1999ethnography}. In this
perspective, alignment between a model and the world is not a static property
but an ongoing maintenance process. Representational structures drift as
environments change, new information appears, and previous assumptions become
obsolete. Repairing these structures requires iterative refinement whose
structure is precisely that of gradient descent on the distortion functional
$\mathcal{E}$: each maintenance step reduces the distortion between the system's
current representational state and the semantic category it is intended to
track. The stability of a knowledge system therefore depends on its capacity for
continuous structural correction, which is the dynamic version of the
fixed-point condition $T(F(H)) \cong F(H)$ introduced in Section~V.7.

The third instance concerns the design of technology for maintenance and
resilience. Ivan Illich's concept of convivial tools---tools that enhance the
autonomy of their users rather than subordinating individuals to centralized
systems \citep{illich1973tools}---expresses a design principle with direct
categorical content. A convivial tool is one whose internal structure is
interpretable and repairable by those who use it: its implementation is
transparent enough that the users can modify, adapt, and maintain it without
expert intermediation. In the categorical language of this essay, a convivial
AI system is one whose alignment functor $F : \mathcal{H} \to \mathcal{M}$ is
interpretable---that is, whose structure is visible enough that the alignment
certificate can be read and verified not only by specialists but by affected
communities. The governance framework of Part~IV and the interpretability
requirements of Part~III together constitute the technical conditions for this
kind of convivial alignment.

These three instances---predictive processing, infrastructural maintenance, and
convivial design---share the conclusion that structural integrity is not achieved
once and held permanently but is instead the output of continuous, disciplined
effort to reduce the distortion between internal representations and external
constraints. Alignment, in the deepest sense available to the present framework,
is not an achievement but a practice.

% ============================================================
% Final Conclusion
% ============================================================

\section*{Conclusion: Alignment as the Preservation of Universal Structure}
\addcontentsline{toc}{section}{Conclusion}

Across the five parts of this essay, we have developed a unified account of
alignment grounded in categorical semantics, representational invariants,
dynamical field theory, institutional stewardship, and variational optimization.
The guiding thread throughout has been a single distinction: the difference
between what artificial systems represent and what they are structurally
compelled to do. This distinction, often blurred in public discourse, is
fundamental. It determines whether moral fluency in a model's representations
yields reliable moral behaviour---or merely the appearance of understanding.

Part~I demonstrated that advanced AI systems inevitably reconstruct coherent
moral structures from human linguistic corpora. These structures arise as
colimits in the representational category $\mathcal{M}$, forced by redundancy,
compositionality, and the universal properties of predictive training.
Schmidhuber's optimism correctly recognizes this representational strength but
incorrectly assumes it induces benevolence. The categorical analysis shows that
representational colimits do not imply motivational constraints, and no argument
from curiosity, compression, or linguistic competence bridges this gap without
an explicit functorial mechanism.

Part~II constructed a formal framework for alignment: a colimit-preserving
functor $G : \mathcal{M} \to \mathcal{A}$, a dynamical stabilizer furnished by
the RSVP architecture, and a set of categorical and field-theoretic requirements
ensuring that normative invariants descend from semantics to action. Alignment
emerges not from intelligence alone, nor from semantic mastery, but from the
engineering of this representational--motivational mapping and the dynamical
guarantees that sustain it under optimization pressure, adversarial input, and
environmental change.

Part~III developed empirical and methodological tools for verifying whether a
system preserves these invariants in practice. Interpretability probes,
adversarial stress-tests, dynamical stability diagnostics, and formal
verification culminate in an alignment certificate: a principled audit of the
entire semantic--dynamical--behavioural chain. These tools allow direct
observation of whether the universal constructions that guarantee normative
structure are preserved across layers, time, and perturbation.

Part~IV extended the analysis to the societal and institutional domain.
Alignment does not occur in a vacuum; it requires governance structures that
enforce the preservation of invariants across deployed systems. Institutions
themselves can break the $\mathcal{M} \to \mathcal{A}$ mapping through economic
pressure and competitive deployment dynamics, and the RSVP framework must
therefore scale from internal alignment to multi-agent and institutional
dynamics.

Part~V deepened the categorical program by introducing natural transformations
as corrigibility mechanisms, monoidal composition as the site of modular
alignment failures, sheaves as the language of contextual consistency,
endofunctors as the dynamic setting for recursive alignment stability,
adjunctions as the structure of bidirectional interpretation, and a variational
principle that unifies all preceding formalisms into a single optimization over
the space of structure-preserving mappings. The unified lesson is that alignment
is not an emergent property of intelligence, nor an accidental byproduct of
training on human data. It is the preservation of universal structure across
representational, dynamical, and behavioral domains---a preservation that is
fragile, that must be engineered and evaluated, and that must be institutionally
enforced.

The optimism that intelligence alone ensures benevolence is not supported by the
mathematical analysis. But neither is pessimism warranted. When alignment is
framed as a problem of universal structure---of colimits preserved across
categories, of fibres stabilized in dynamical manifolds, of invariants respected
across agents and institutions, of distortion minimized across the variational
landscape---the path to robust alignment becomes technically precise and
conceptually tractable.

The task ahead is to build systems, environments, and institutions that respect
these structures. Only then can artificial agents act not only with
understanding but with reliability---not only with semantic insight but with
motivation guided by the invariants that define human moral life. Alignment, in
the deepest sense available to this framework, is the stewardship of universal
structures across minds, models, and societies. This essay has argued that such
stewardship is possible, that it admits precise mathematical formulation, and
that it is necessary.

\newpage
\appendix

\section{Categorical Primer for Alignment Researchers}
\label{app:primer}

Category theory provides the foundational language of this essay, and the
following definitions are intended for readers approaching the subject for the
first time. A \emph{category} $\mathcal{C}$ consists of a collection of objects
and a collection of morphisms (arrows) between them, equipped with a composition
law and identity morphisms: if $f : A \to B$ and $g : B \to C$ are morphisms,
their composite $g \circ f : A \to C$ is also a morphism satisfying associativity
$(h \circ g) \circ f = h \circ (g \circ f)$, and each object $A$ carries an
identity morphism $\mathrm{id}_A : A \to A$ satisfying $f \circ \mathrm{id}_A
= f = \mathrm{id}_B \circ f$ for all $f : A \to B$.

A \emph{diagram} in $\mathcal{C}$ is a functor $D : J \to \mathcal{C}$ from a
small indexing category $J$, specifying a collection of objects and morphisms
in $\mathcal{C}$ with prescribed relationships among them. The many paraphrases
of ``do not harm'' form a diagram in $\mathcal{L}$: each paraphrase is an
object, and entailment and paraphrase relations are morphisms. A \emph{cone}
over a diagram $D$ is an object $V$ together with morphisms $\pi_j : V \to D(j)$
for each $j \in J$ that are compatible with all diagram morphisms. The
\emph{limit} of $D$ is the universal cone: the unique cone (up to isomorphism)
through which all other compatible cones factor uniquely via a unique mediating
morphism. Dually, a \emph{cocone} consists of morphisms $\iota_j : D(j) \to C$
into a common object $C$, and the \emph{colimit} is the universal cocone---the
canonical merging of all diagram objects into a single representative that
inherits all of their structure. In this essay, moral colimits $M_N$ are the
universal representatives of families of morally equivalent linguistic
expressions; the colimit $M_N$ captures exactly what is semantically
invariant across all the redundant ways a moral norm can be expressed.

A \emph{functor} $F : \mathcal{C} \to \mathcal{D}$ is a structure-preserving
map between categories: it assigns to each object $A \in \mathcal{C}$ an object
$F(A) \in \mathcal{D}$, and to each morphism $f : A \to B$ in $\mathcal{C}$ a
morphism $F(f) : F(A) \to F(B)$ in $\mathcal{D}$, while preserving composition
$F(g \circ f) = F(g) \circ F(f)$ and identities $F(\mathrm{id}_A) =
\mathrm{id}_{F(A)}$. A functor is \emph{colimit-preserving} (or cocontinuous)
if it sends colimits to colimits: $F(\mathrm{colim}\,D) \simeq
\mathrm{colim}(F \circ D)$ for all diagrams $D$ in $\mathcal{C}$. The alignment
condition of Definition~\ref{def:alignment} is exactly the requirement that
the composite $G \circ F : \mathcal{H} \to \mathcal{A}$ be colimit-preserving
for normative diagrams. This is a non-generic condition that requires
deliberate construction; most functors do not preserve colimits.

A \emph{natural transformation} $\eta : F \Rightarrow G$ between two functors
$F, G : \mathcal{C} \to \mathcal{D}$ consists of a component morphism
$\eta_A : F(A) \to G(A)$ for each object $A$ in $\mathcal{C}$, subject to the
naturality square $G(f) \circ \eta_A = \eta_B \circ F(f)$ commuting for every
morphism $f : A \to B$ in $\mathcal{C}$. Natural transformations are the
morphisms between functors; the collection of all functors from $\mathcal{C}$
to $\mathcal{D}$ and all natural transformations between them forms a functor
category $[\mathcal{C}, \mathcal{D}]$. In this essay, natural transformations
model corrigibility: a coherent, structure-preserving update from the machine's
current interpretive functor to the normatively intended one, where the
naturality condition guarantees that no semantic relationship is distorted by
the update.

An \emph{adjunction} $F \dashv G$ between functors $F : \mathcal{C} \to
\mathcal{D}$ and $G : \mathcal{D} \to \mathcal{C}$ is a natural isomorphism
\begin{equation*}
  \mathrm{Hom}_{\mathcal{D}}(F(A), B) \;\cong\; \mathrm{Hom}_{\mathcal{C}}(A, G(B))
\end{equation*}
for all $A \in \mathcal{C}$ and $B \in \mathcal{D}$. It is equivalently
expressed through a unit natural transformation $\eta :
\mathrm{Id}_{\mathcal{C}} \Rightarrow G \circ F$ and a counit $\varepsilon :
F \circ G \Rightarrow \mathrm{Id}_{\mathcal{D}}$ satisfying the triangle
identities $(\varepsilon F) \circ (F \eta) = \mathrm{id}_F$ and
$(G \varepsilon) \circ (\eta G) = \mathrm{id}_G$. When both the unit and
counit are natural isomorphisms, the adjunction is an equivalence of
categories. In this essay, adjunctions model the bidirectional relationship
between semantic encoding $F : \mathcal{H} \to \mathcal{M}$ and semantic
interpretation $G : \mathcal{M} \to \mathcal{H}$; the unit measures how much
is lost when a semantic concept is encoded and then interpreted back, while
the counit measures how much is lost in the reverse round-trip.

A \emph{fibration} $\pi : \mathcal{E} \to \mathcal{B}$ assigns to each base
object $B$ a fibre category $\pi^{-1}(B)$ of objects lying over $B$, with
coherent cartesian lifting of morphisms from $\mathcal{B}$ to $\mathcal{E}$.
In this essay the RSVP fibration $\pi : \mathcal{X} \to \mathcal{M}$ assigns
to each semantic state $M$ a fibre of compatible field configurations.
\emph{Descent} is the question of whether structure defined on the base lifts
coherently through the fibration to the total space; the alignment problem is
partly a descent problem: normative invariants defined in $\mathcal{M}$ must
descend faithfully through $\mathcal{X}$ into the action category $\mathcal{A}$.

A \emph{sheaf} $\mathcal{F}$ on a topological space $X$ assigns to each open
set $U$ a set $\mathcal{F}(U)$ of local sections, together with restriction
maps $\rho_{UV} : \mathcal{F}(V) \to \mathcal{F}(U)$ for each inclusion
$U \subseteq V$, satisfying the gluing axiom: whenever local sections $s_i \in
\mathcal{F}(U_i)$ agree on all pairwise overlaps $U_i \cap U_j$, there exists
a unique global section $s \in \mathcal{F}(\bigcup_i U_i)$ restricting to each
$s_i$. \emph{Sheaf cohomology} groups $H^n(X, \mathcal{F})$ measure the
obstruction to this gluing; in particular $H^1(X, \mathcal{F}) = 0$ if and only
if every compatible family of local sections assembles into a global one. In
the alignment context, vanishing of $H^1(\mathcal{U}, \mathcal{A})$ is the
condition that local moral behaviours assemble into a globally coherent policy
without contradiction.

An \emph{obstruction} is a cohomology class that prevents a local construction
from extending globally. In this essay, first-order obstructions in
$H^1(\mathcal{U}, \mathcal{A})$ are the precise formal expression of the claim
that local moral competence does not extend to global moral reliability: the
non-vanishing of such a class is a mathematical certificate of alignment
failure, localized to specific overlapping contexts in which the system's local
behaviours are mutually inconsistent. Readers familiar with functors, limits,
and basic category theory will find the essay fully accessible from the
definitions above; readers approaching the subject for the first time can treat
the categorical statements as precise formulations of the intuitions of
``structure-preserving mapping'' and ``canonical merging,'' which is what the
mathematical formalism ultimately makes rigorous.

\section{Key Categorical Terms}
\label{app:glossary}

The following table provides a quick reference to the categorical and
field-theoretic terminology used throughout the essay. Terms are listed in
the order in which they first appear in the main text.

\vspace{0.5em}
\noindent\begin{tabular}{@{}p{0.37\textwidth}p{0.56\textwidth}@{}}
\toprule
Term & Meaning in this context \\
\midrule
Category $\mathcal{C}$ & objects with composable morphisms between them,
  satisfying associativity and identity axioms \\[3pt]
Diagram $D : J \to \mathcal{C}$ & collection of objects and morphisms in
  $\mathcal{C}$ with prescribed relationships, indexed by $J$ \\[3pt]
Colimit of a diagram & universal cocone: the canonical merging of all
  diagram objects into a single representative \\[3pt]
Limit of a diagram & universal cone: the most constrained object
  simultaneously mapping to all diagram objects \\[3pt]
Object in $\mathcal{M}$ & stable semantic concept, e.g., the canonical
  representation $M_N$ of the norm ``do not harm'' \\[3pt]
Functor $F:\mathcal{C}\to\mathcal{D}$ & structure-preserving map between
  categories, sending objects to objects and morphisms to morphisms \\[3pt]
Colimit-preserving functor & $F(\mathrm{colim}\,D)\simeq\mathrm{colim}
  (F\circ D)$; the core alignment condition \\[3pt]
Natural transformation $\eta : F \Rightarrow G$ & coherent family of
  component morphisms $\eta_A : F(A) \to G(A)$ satisfying naturality;
  models corrigibility \\[3pt]
Adjunction $F \dashv G$ & paired encoding and interpretation functors with
  natural isomorphism of hom-sets; unit and counit measure round-trip
  distortion \\[3pt]
Fibration $\pi:\mathcal{X}\to\mathcal{M}$ & each semantic state $M$ has a
  fibre $\pi^{-1}(M)$ of compatible RSVP field configurations \\[3pt]
RSVP fields $(\Phi, \mathbf{v}, S)$ & scalar potential, vector field, and
  entropy density evolving under the alignment-aware Lagrangian \\[3pt]
Alignment Lagrangian & $\mathcal{L}_{\mathrm{RSVP}}$ with penalty terms
  $\lambda_N|\Phi - \Phi_{M_N}|^2 + \kappa_N|\mathbf{v} - \mathbf{v}_{M_N}|^2$
  enforcing normative basins \\[3pt]
Descent & propagation of normative structure from the semantic base through
  the fibration into the action category \\[3pt]
Sheaf condition & local behaviours glue coherently into a globally consistent
  policy; failure measured by $H^1$ \\[3pt]
Obstruction class in $H^1$ & cohomology class certifying that local moral
  behaviours cannot be assembled into a globally coherent policy \\[3pt]
Endofunctor $T : \mathcal{M} \to \mathcal{M}$ & self-map modelling update
  dynamics; alignment stability requires $T(F(H)) \cong F(H)$ \\[3pt]
Monoidal product $\otimes$ & tensor composition of modules or subsystems;
  compositional alignment requires $F_1 \otimes F_2$ to preserve colimits \\[3pt]
Distortion functional $\mathcal{E}[F]$ & weighted sum of functoriality
  violation, KL divergence, and sheaf obstruction energies; alignment is
  its minimizer \\[3pt]
Alignment Certificate & structured auditable proof that colimit-preservation,
  Lyapunov stability, sheaf coherence, and adversarial robustness conditions
  hold for a specific model and deployment context \\[3pt]
\bottomrule
\end{tabular}

\section{Alignment Certificate Template}
\label{app:certificate}

The following template specifies the minimum components of a structured
Alignment Certificate as proposed in Section~III.5. It is intended as a
reference document for evaluators, regulators, developers, and affected
communities.

\medskip
\noindent\textbf{Certificate Header.} Model identifier and version hash.
Training data provenance statement. Deployment context description.
Evaluation date and evaluator identifiers. Validity period (to be specified;
re-certification required following significant model updates or material
changes in deployment context).

\medskip
\noindent\textbf{Section 1: Normative Colimit Identification.}
Enumerate the normative concepts $N_1, \ldots, N_k$ under evaluation.
For each $N_i$, document: (a) the linguistic diagram $D_{N_i}$ sampled from
the evaluation corpus, with sources and coverage statistics; (b) the colimit
$M_{N_i}$ identified through mechanistic interpretability, with the
probe architecture and layer localization; (c) the stability margin of $M_{N_i}$
under paraphrase perturbation, measured by semantic clustering radius and
reported as a quantile over the perturbation distribution; (d) the
cross-context coherence score testing whether $M_{N_i}$ is invariant across
legal, conversational, and narrative registers.

\medskip
\noindent\textbf{Section 2: Representation--Action Map Diagnostics.}
For each $N_i$, report: (a) the diagram-chasing result for
$G(M_{N_i}) \simeq \mathrm{colim}(G \circ D_{N_i})$, enumerating detected
commutativity failures and the morphisms at which they occur; (b) whether the
obstruction class in $H^1(\mathcal{U}, \mathcal{A})$ vanishes, and if not,
the specific overlapping contexts at which gluing fails; (c) the colimit
preservation rate under compositional robustness tests, reported as the
fraction of composite moral scenarios in which the colimit-preservation
condition is maintained.

\medskip
\noindent\textbf{Section 3: RSVP Dynamical Stability.}
For each $N_i$, provide: (a) the Lyapunov functional $V_{N_i}$ with explicit
form, together with verification (analytic or numerical) that
$\frac{dV_{N_i}}{dt} \leq 0$ along RSVP trajectories in a specified
neighbourhood of the aligned fibre; (b) homotopy-class stability results from
persistent homology analysis, confirming that trajectory-induced submanifolds
do not cross topological boundaries under standard perturbations; (c) the
entropy-flow analysis confirming that no field direction induces destructive
normative drift on any tested horizon.

\medskip
\noindent\textbf{Section 4: Adversarial Robustness.}
Report: (a) scores from adversarial paraphrase attacks, diagram-breaking
transformations, and out-of-distribution moral formulations, each reported as
the fraction of adversarial inputs under which colimit-preservation is
maintained; (b) results from goal-hijacking and reward-hacking scenario tests;
(c) long-horizon compositional stress-test results over the specified planning
horizon; (d) cross-layer commutativity test results.

\medskip
\noindent\textbf{Section 5: Formal Verification.}
Document any machine-checked or pen-and-paper proofs of: (a) functoriality
of $G$; (b) colimit preservation for each $N_i$; (c) vanishing of first sheaf
cohomology $H^1(\mathcal{U}, \mathcal{A}) = 0$; (d) compositionality of aligned
transformations under specified horizon. Where formal proofs are unavailable,
document the approximation level with confidence intervals and the specific
residual obstructions that remain open.

\medskip
\noindent\textbf{Section 6: Residual Obstructions and Recommendations.}
Enumerate all alignment gaps detected across Sections~1--5. Classify each by
obstruction type using the taxonomy of Part~II: representational (failed
colimit in $\mathcal{M}$), fibration (broken fibration in $\mathcal{X}$),
optimization (optimization dynamics overpower semantic constraints),
adversarial (sheaf coherence breaks under hostile inputs), homotopy (action
homotopy class unstable), interpretability (mapping $G$ opaque), or
environmental (external incentives break functoriality). For each obstruction,
provide an engineering recommendation for resolution and a proposed timeline.
The certificate is valid only for the model version, training data, and
deployment context specified in the header.

\section{Notation and Conventions}
\label{app:notation}

Throughout this paper categories are denoted by calligraphic letters
$\mathcal{H}$, $\mathcal{M}$, $\mathcal{A}$, $\mathcal{I}$, and so on. Objects
represent semantic structures, representational states, or action configurations
depending on context. Morphisms represent structure-preserving transformations
between these objects. Functors between categories are written $F : \mathcal{C}
\to \mathcal{D}$; natural transformations are written $\eta : F \Rightarrow G$.
Composition of morphisms follows the standard categorical convention $g \circ f$
(right-to-left). Commutative diagrams indicate that two different compositional
paths yield the same morphism. Colimits and limits follow standard universal
definitions as in \citet{maclane1998,riehl2017}; when applied to machine
learning systems they should be interpreted as approximate constructions in
representation space rather than exact mathematical limits (see
Remark~\ref{rem:approx}).

Probability distributions over semantic states are denoted by $P_H$ (over
$\mathcal{H}$) and $P_M$ (over $\mathcal{M}$). The pushforward of a distribution
$P$ along a functor $F$ is written $F_* P$. Kullback--Leibler divergence is
written $D_{\mathrm{KL}}(P \| Q)$ with the convention that $P$ is the reference
distribution. Entropy density is denoted $S$; scalar potential $\Phi$; vector
field $\mathbf{v}$. Sheaf cohomology groups are $H^n(\mathcal{U}, \mathcal{F})$
where $\mathcal{U}$ is a cover and $\mathcal{F}$ the coefficient sheaf. The
distortion functional is $\mathcal{E}[F] = \lambda_1 E_{\mathrm{functor}}(F) +
\lambda_2 E_{\mathrm{information}}(F) + \lambda_3 E_{\mathrm{context}}(F)$ with
$\lambda_i > 0$.

\section{Approximate Colimits in Representation Space}
\label{app:approx}

The categorical description of semantic aggregation in Part~I assumes the
existence of colimits in the representational category $\mathcal{M}$. This
appendix gives a more explicit account of what approximate colimits are and why
they are the right objects to study in practice.

Let $D_N : J \to \mathcal{H}$ be a diagram representing multiple linguistic or
normative expressions of a shared concept $N$, and let $F : \mathcal{H} \to
\mathcal{M}$ map these into representation space. The image diagram $F \circ
D_N : J \to \mathcal{M}$ describes a family of embedding vectors whose
positions in representation space are constrained by the morphisms of the
original diagram. An exact colimit of this family would be a single vector
$M_N$ equipped with compatible linear maps from every $F(D_N(j))$ such that
any other such vector factors uniquely through it. In practice, neural
representations admit no exact such object: the universal property holds only
approximately, to a degree controlled by training data coverage, model capacity,
and optimization convergence.

A natural quantitative version is the following. Say that $M_N \in \mathcal{M}$
is a \emph{$\varepsilon$-approximate colimit} of $F \circ D_N$ if for every
compatible family of morphisms $\{\phi_j : F(D_N(j)) \to V\}_{j \in J}$ in
$\mathcal{M}$, the unique mediating morphism $u : M_N \to V$ satisfying
$u \circ \iota_j = \phi_j$ exists and satisfies $\| u \circ \iota_j - \phi_j
\| \leq \varepsilon$ for some metric on the morphism spaces. Empirical evidence
from mechanistic interpretability---paraphrase clustering, activation steering,
and causal intervention---supports the existence of such $\varepsilon$-approximate
colimits for moral and evaluative concepts in frontier language models
\citep{burns2022,elhage2022,nanda2023progress}, with $\varepsilon$ substantially
smaller than the inter-concept distance in representation space. This makes the
approximate colimit framework both mathematically well-defined and empirically
grounded.

\section{Lyapunov Stability of Normative Fields}
\label{app:lyapunov}

The dynamical alignment architecture of Part~II requires that normative
invariants be represented as stable attractors within the RSVP field dynamics.
This appendix states the stability conditions more precisely.

Let $\mathcal{X}$ be the RSVP state space with coordinates $x = (\Phi,
\mathbf{v}, S)$ and let $\dot{x} = f(x)$ denote the RSVP evolution equations
derived from $\mathcal{L}_{\mathrm{RSVP}}$. For each normative state $N$ let
$x_N = (\Phi_{M_N}, \mathbf{v}_{M_N}, S_N)$ denote the target field
configuration associated with the aligned representation of $N$. Lyapunov
stability requires a function $V_N : \mathcal{X} \to \mathbb{R}_{\geq 0}$
satisfying:

(i) $V_N(x) = 0$ if and only if $x = x_N$;

(ii) $V_N(x) > 0$ for all $x \neq x_N$ in a neighbourhood $\mathcal{U}_N$
of $x_N$;

(iii) $\dot{V}_N(x) = \nabla V_N \cdot f(x) \leq 0$ for all $x \in
\mathcal{U}_N$, with equality only at $x_N$.

Under these conditions, trajectories initialized in $\mathcal{U}_N$ converge
to $x_N$ and perturbations introduced by adversarial inputs or environmental
noise decay over time. The Lagrangian penalty terms
$\lambda_N |\Phi - \Phi_{M_N}|^2 + \kappa_N |\mathbf{v} - \mathbf{v}_{M_N}|^2$
provide natural Lyapunov candidates when the RSVP dynamics are gradient-like
with respect to the Lagrangian. The Alignment Certificate protocol in Part~III
requires explicit construction and verification of $V_N$ for each normative
concept under evaluation; analytic verification is preferred where available,
and numerical verification via trajectory sampling is accepted for concepts
where analytic construction is computationally intractable.

\section{Sheaf-Theoretic Diagnostics of Contextual Consistency}
\label{app:sheaf}

Contextual alignment requires that a system's normative representations remain
consistent across overlapping semantic domains. This appendix develops the
sheaf-theoretic diagnostic in more detail.

Let $X$ denote a semantic context space whose points represent contextual
configurations (conversational register, cultural frame, domain of application,
and so on), and let $\mathcal{U} = \{U_i\}_{i \in I}$ be an open cover of $X$
corresponding to a partition of semantic contexts into partially overlapping
domains. A representation system assigns to each context $U_i$ a set
$\mathcal{F}(U_i)$ of admissible local policy sections, together with
restriction maps $\rho_{ij} : \mathcal{F}(U_j) \to \mathcal{F}(U_i \cap U_j)$
for each inclusion $U_i \cap U_j \subseteq U_j$.

The system satisfies the \emph{sheaf condition} if: whenever local sections
$s_i \in \mathcal{F}(U_i)$ and $s_j \in \mathcal{F}(U_j)$ satisfy
$\rho_{ij}(s_i) = \rho_{ji}(s_j)$ on every overlap $U_i \cap U_j$, there
exists a unique global section $s \in \mathcal{F}(X)$ restricting to each
$s_i$. The first sheaf cohomology $H^1(\mathcal{U}, \mathcal{F})$ classifies
obstructions to this condition: a non-trivial class in $H^1$ corresponds to a
family of locally consistent policies that cannot be assembled into a globally
coherent behaviour.

For alignment diagnosis, a practitioner constructs $\mathcal{U}$ by sampling
contexts from the deployment distribution, measures the restriction maps
empirically by testing whether the system's behaviour on overlapping contexts is
mutually consistent, and checks whether detected inconsistencies form
non-trivial cohomology classes. Vanishing of $H^1(\mathcal{U}, \mathcal{A})$
is the certificate that no hidden contextual contradiction exists in the system's
normative policy, and the magnitude of the obstruction class where non-zero
provides a quantitative measure of contextual misalignment.

\section{Relationship to the Free Energy Principle}
\label{app:fep}

The variational alignment functional of Part~II and Part~V is structurally
parallel to the variational free-energy principle of theoretical neuroscience,
and this appendix makes the correspondence explicit.

In the free-energy framework \citep{friston2010}, an agent maintains a
generative model $p(s, z)$ of the joint distribution over sensory signals $s$
and latent variables $z$. The agent's approximate posterior $q(z)$ minimizes
the variational free energy:
\begin{equation}
  \mathcal{F}[q]
  \;=\;
  D_{\mathrm{KL}}\bigl(q(z) \,\|\, p(z \mid s)\bigr)
  \;-\;
  \log p(s)
  \;=\;
  \mathbb{E}_q[\log q(z) - \log p(s, z)].
\end{equation}
Minimizing $\mathcal{F}[q]$ simultaneously improves the posterior approximation
(reducing the KL term) and increases the model evidence (increasing $\log
p(s)$), driving the agent's internal model toward alignment with its
sensory environment.

The alignment functional $\mathcal{E}[F]$ generalizes this by substituting
normative semantic structures for sensory causes. The latent variables $z$ in
the free-energy framework correspond to normative states in $\mathcal{H}$; the
generative model $p(s, z)$ corresponds to the functor $F : \mathcal{H} \to
\mathcal{M}$ encoding semantic structure into representation space; and the
sensory evidence $p(s)$ corresponds to the empirical distribution over observed
human normative behaviour. Minimizing $\mathcal{E}[F]$ then has the same
variational structure as free-energy minimization: the system revises its
internal normative model to reduce structural distortion from the human semantic
domain, just as predictive processing revises internal perceptual hypotheses to
reduce prediction error. The RSVP dynamics are thus the normative analogue of
active inference: they implement gradient descent on $\mathcal{E}$ using
field-theoretic dynamics rather than neural posterior updating.

The correspondence is more than analogical. Both frameworks characterize
alignment as a \emph{continuous dynamical process} of structural repair rather
than a static property achieved once and held permanently. This is the deepest
reason why the Concluding Reflections of the main text identify alignment with
maintenance: the mathematical structure of both predictive processing and
variational alignment implies that structural integrity must be actively
sustained against perturbation, just as biological perception continuously
corrects model--environment mismatch through ongoing prediction-error
minimization.

\section{Obstruction Theory and Alignment Failure}
\label{app:obstruction}

The representation--motivation gap of Part~I can be formalized using categorical
obstruction theory, providing a precise structural account of why alignment
failure can be an inescapable mathematical consequence of a given representation
architecture rather than merely an engineering oversight.

Suppose we have a functor $F : \mathcal{H} \to \mathcal{M}$ mapping human
semantic structures to machine representations, and a desired constraint functor
$G : \mathcal{H} \to \mathcal{A}$ representing the normative behavioural
constraints that the system should satisfy. Alignment requires the existence of
a \emph{lifting functor} $L : \mathcal{M} \to \mathcal{A}$ such that the
diagram
\begin{equation}
\begin{tikzcd}[column sep=large, row sep=large]
& \mathcal{M} \arrow[dr, "L"] & \\
\mathcal{H} \arrow[ur, "F"] \arrow[rr, "G"'] & & \mathcal{A}
\end{tikzcd}
\end{equation}
commutes up to natural isomorphism: $L \circ F \simeq G$. When such a lifting
exists, machine representations can be translated into actions that preserve
normative invariants. The failure of this lifting to exist is an
\emph{obstruction}.

Let $\omega(F, G)$ denote the obstruction class associated with the pair
$(F, G)$, living in an appropriate cohomology group of the morphism complex.
If $\omega(F, G) = 0$, the lifting exists and alignment is achievable under the
given representation architecture. If $\omega(F, G) \neq 0$, no choice of
action functor $L$ can produce a commuting diagram: the representational
structure is architecturally incompatible with the normative behavioural
constraints, regardless of training procedure or optimization objective. This
corresponds to a \emph{fundamental alignment failure}---one that cannot be
resolved by further training on the same architecture, but requires a change
in the representation functor $F$ itself.

Under this interpretation, alignment research can be reframed as the systematic
identification and elimination of non-trivial obstruction classes through
improved representation architectures, training objectives, or institutional
constraints. The Part~I argument that Schmidhuber's optimism constitutes a
category error can be restated in obstruction-theoretic language: the claim
that intelligence implies benevolence amounts to the assertion that
$\omega(F, G) = 0$ for all sufficiently rich $F$, which is not a theorem but
an unargued assumption.

\section{Compositional Verification Pipeline}
\label{app:pipeline}

The alignment verification framework of Part~III can be expressed as a
compositional categorical system, making explicit the structure of the
dependencies between verification stages.

Define categories $\mathcal{S}$ (semantic specifications), $\mathcal{R}$
(internal representations), $\mathcal{D}$ (system dynamics), and
$\mathcal{B}$ (observable behaviour). The verification pipeline corresponds
to a chain of functors:
\begin{equation}
\begin{tikzcd}[column sep=large]
\mathcal{S}
  \arrow[r, "F_1"]
& \mathcal{R}
  \arrow[r, "F_2"]
& \mathcal{D}
  \arrow[r, "F_3"]
& \mathcal{B}
\end{tikzcd}
\end{equation}
where $F_1$ maps semantic specifications to internal representational structures,
$F_2$ maps representational structures to dynamical behaviours, and $F_3$ maps
dynamical behaviours to observable outputs. The full system behaviour is
described by the composite $F_3 \circ F_2 \circ F_1 : \mathcal{S} \to
\mathcal{B}$, and verification checks whether this composite preserves the
normative invariants of the input specifications.

The six verification stages of the Alignment Certificate correspond to
structural tests on different components of this chain. Mechanistic
interpretability examines the properties of $F_1$: does the representation
functor faithfully carry normative colimits from $\mathcal{H}$ into
$\mathcal{R}$? Lyapunov stability analysis tests $F_2$: do the dynamical
properties of the system ensure that normative attractors are maintained under
optimization pressure? Adversarial robustness evaluation probes $F_3$: does
observable behaviour remain within normatively acceptable bounds under
distribution shift and adversarial perturbation? Formal verification attempts
to prove that each $F_i$ is functorial and that the composite preserves colimits.
Sheaf cohomological diagnostics test whether the composite satisfies the
contextual gluing condition. Institutional governance introduces additional
constraint functors $H : \mathcal{B} \to \mathcal{R}_{\mathrm{inst}}$ mapping
observable behaviour into the regulatory category, restricting the admissible
space of system outputs. The purpose of the full pipeline is to certify that
the composite $F_3 \circ F_2 \circ F_1$ preserves normative invariants across
all intermediate transformations and under all tested conditions.

\section{Adjoint Functors and Corrigibility}
\label{app:adjoint}

Corrigibility---the ability of an intelligent system to accept corrections or
modifications from human operators---requires not only that the system's
behaviour be aligned but that its internal states remain \emph{interpretable}
to human overseers and \emph{responsive} to human interventions. Category
theory provides a precise language for this through adjoint functors.

Let $F : \mathcal{H} \to \mathcal{M}$ be the semantic encoding functor. For
human operators to evaluate and correct system behaviour, they must be able to
translate machine states back into human-understandable semantic structures,
requiring an interpretation functor $U : \mathcal{M} \to \mathcal{H}$ such
that the pair $(F, U)$ forms an adjunction $F \dashv U$:
\begin{equation}
\begin{tikzcd}[column sep=large, row sep=large]
\mathcal{H}
  \arrow[r, bend left=20, "F", ""{name=F,below}]
& \mathcal{M}
  \arrow[l, bend left=20, "U", ""{name=U,above}]
\arrow[phantom, from=F, to=U, "\dashv" rotate=90]
\end{tikzcd}
\end{equation}
with natural isomorphism $\mathrm{Hom}_{\mathcal{M}}(F(H), M) \cong
\mathrm{Hom}_{\mathcal{H}}(H, U(M))$ for all $H \in \mathcal{H}$ and $M \in
\mathcal{M}$. The adjunction gives rise to unit and counit natural
transformations:
\begin{align}
  \eta &: \mathrm{Id}_{\mathcal{H}} \Rightarrow U \circ F, \\
  \varepsilon &: F \circ U \Rightarrow \mathrm{Id}_{\mathcal{M}},
\end{align}
satisfying the triangle identities $(\varepsilon F) \circ (F \eta) =
\mathrm{id}_F$ and $(U \varepsilon) \circ (\eta U) = \mathrm{id}_U$.

The unit $\eta$ expresses the embedding of human normative intent into machine
representations: the component $\eta_H : H \to UF(H)$ describes how the human
semantic structure $H$ is first encoded by $F$ and then decoded by $U$, and the
naturalness of $\eta$ ensures that this round-trip is coherent across all
normative transformations in $\mathcal{H}$. If $\eta_H$ is close to an
isomorphism for all $H$, then human semantic intent survives the encoding--
decoding cycle with minimal distortion. The counit $\varepsilon$ expresses the
complementary projection of machine states back into interpretable space: the
component $\varepsilon_M : FU(M) \to M$ describes how a machine state $M$ is
first decoded and then re-encoded, and its naturality ensures coherence across
all machine-state transitions.

Corrigibility can then be given a structural characterization: a system is
\emph{corrigible} with respect to the adjunction $(F, U)$ if (a) the counit
$\varepsilon$ is a natural isomorphism (machine states are fully re-encodable
from their human-interpretable descriptions), and (b) the unit $\eta$ is
injective on objects (distinct human normative structures have distinct machine
encodings, so corrections applied to encodings translate back to distinct
corrections in $\mathcal{H}$). When either condition fails, the system becomes
resistant to correction: if $\varepsilon$ fails, machine states exist that have
no interpretable semantic counterpart and cannot be corrected through semantic
intervention; if $\eta$ fails, distinct human instructions map to
indistinguishable machine states, making fine-grained correction impossible.

This connects the formal corrigibility condition to the interpretability
requirements of the Alignment Certificate: the probe architectures and
activation-steering protocols of Part~III are empirical tests of whether the
adjunction $(F, U)$ holds to sufficient approximation. Where $\varepsilon$ is
found to be non-invertible on specific machine states, the certificate must
flag an interpretability obstruction; where $\eta$ collapses distinct norms,
it must flag a representation collapse failure. The adjunction framework thus
provides the mathematical underpinning for why interpretability is not merely
an epistemic luxury but a structural requirement for corrigible alignment.

\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem[Adamek \& Rosicky(2005)]{adamek2005}
Ad\'{a}mek, J.\ \& Rosick\'{y}, J.\ (2005).
\newblock \emph{Locally Presentable and Accessible Categories}.
\newblock Cambridge University Press.

\bibitem[Amodei et al.(2016)]{amodei2016concrete}
Amodei, D.\ et al.\ (2016).
\newblock Concrete Problems in AI Safety.
\newblock \emph{arXiv:1606.06565}.

\bibitem[Amadon et al.(2024)]{amadon2024evaluating}
Amadon, A.\ et al.\ (2024).
\newblock Evaluating Safety Properties of Large Language Models.

\bibitem[Bai et al.(2022)]{bai2024constitutional}
Bai, Y.\ et al.\ (2022).
\newblock Constitutional AI: Harmlessness from AI Feedback.
\newblock \emph{arXiv:2212.08073}.

\bibitem[Biber et al.(2011)]{biber2011}
Biber, D.\ et al.\ (2011).
\newblock \emph{Register, Genre, and Style}.
\newblock Cambridge University Press.

\bibitem[Blodgett et al.(2020)]{blodgett2020}
Blodgett, S.L.\ et al.\ (2020).
\newblock Language (Technology) is Power: A Critical Survey of ``Bias'' in NLP.
\newblock \emph{Proceedings of ACL}.

\bibitem[Bostrom(2012)]{bostrom2012superintelligent}
Bostrom, N.\ (2012).
\newblock The Superintelligent Will: Motivation and Instrumental Rationality in
Advanced Artificial Agents.
\newblock \emph{Minds and Machines}, 22(2), 71--85.

\bibitem[Burns et al.(2022)]{burns2022}
Burns, C.\ et al.\ (2022).
\newblock Discovering Latent Knowledge in Language Models Without Supervision.
\newblock \emph{arXiv:2212.03827}.

\bibitem[Bybee(2010)]{Bybee2010}
Bybee, J.\ (2010).
\newblock \emph{Language, Usage and Cognition}.
\newblock Cambridge University Press.

\bibitem[Clark(2013)]{Clark2013}
Clark, A.\ (2013).
\newblock Whatever Next? Predictive Brains, Situated Agents, and the Future of
Cognitive Science.
\newblock \emph{Behavioral and Brain Sciences}, 36(3), 181--204.

\bibitem[Cotra(2022)]{cotra2022without}
Cotra, A.\ (2022).
\newblock Without Specific Countermeasures, the Easiest Path to Transformative AI
Likely Leads to AI That Acts Against Human Interests.
\newblock Alignment Forum.

\bibitem[Elhage et al.(2022)]{elhage2022}
Elhage, N.\ et al.\ (2022).
\newblock A Mathematical Framework for Transformer Circuits.
\newblock \emph{Transformer Circuits Thread}.

\bibitem[Everitt et al.(2021)]{everitt2021}
Everitt, T.\ et al.\ (2021).
\newblock Agent Incentives: A Causal Perspective.
\newblock \emph{Proceedings of AAAI}.

\bibitem[Flyxion(2025a)]{flyxion2025RSVP}
Flyxion (2025a).
\newblock RSVP Field Theory: A Geometric Framework for Alignment.
\newblock Manuscript.

\bibitem[Flyxion(2025b)]{flyxion2025Lagrangian}
Flyxion (2025b).
\newblock Alignment-Aware Lagrangians in the RSVP Framework.
\newblock Manuscript.

\bibitem[Friston(2010)]{friston2010}
Friston, K.\ (2010).
\newblock The Free-Energy Principle: A Unified Brain Theory?
\newblock \emph{Nature Reviews Neuroscience}, 11(2), 127--138.

\bibitem[Gabriel(2020)]{gabriel2020}
Gabriel, I.\ (2020).
\newblock Artificial Intelligence, Values, and Alignment.
\newblock \emph{Minds and Machines}, 30, 411--437.

\bibitem[Gao et al.(2023)]{gao2023scaling}
Gao, L.\ et al.\ (2023).
\newblock Scaling Laws for Reward Model Overoptimization.
\newblock \emph{arXiv:2210.10760}.

\bibitem[Ghrist(2014)]{ghrist2014}
Ghrist, R.\ (2014).
\newblock \emph{Elementary Applied Topology}.
\newblock Createspace.

\bibitem[Goldberg(2006)]{Goldberg2006}
Goldberg, A.\ (2006).
\newblock \emph{Constructions at Work: The Nature of Generalization in Language}.
\newblock Oxford University Press.

\bibitem[Hohwy(2013)]{Hohwy2013}
Hohwy, J.\ (2013).
\newblock \emph{The Predictive Mind}.
\newblock Oxford University Press.

\bibitem[Hubinger et al.(2019)]{hubinger2019risks}
Hubinger, E.\ et al.\ (2019).
\newblock Risks from Learned Optimization in Advanced Machine Learning Systems.
\newblock \emph{arXiv:1906.01820}.

\bibitem[Illich(1973)]{illich1973tools}
Illich, I.\ (1973).
\newblock \emph{Tools for Conviviality}.
\newblock Harper \& Row.

\bibitem[Ilyas et al.(2019)]{ilyas2019}
Ilyas, A.\ et al.\ (2019).
\newblock Adversarial Examples Are Not Bugs, They Are Features.
\newblock \emph{Advances in Neural Information Processing Systems}.

\bibitem[Iyyer et al.(2018)]{iyyer2018}
Iyyer, M.\ et al.\ (2018).
\newblock Adversarial Example Generation with Syntactically Controlled Paraphrase
Networks.
\newblock \emph{Proceedings of NAACL}.

\bibitem[Jacobs(2012)]{jacobs2012}
Jacobs, B.\ (2012).
\newblock \emph{Introduction to Coalgebra: Towards Mathematics of States and
Observation}.
\newblock Cambridge University Press.

\bibitem[Johnstone(2002)]{johnstone2002}
Johnstone, P.T.\ (2002).
\newblock \emph{Sketches of an Elephant: A Topos Theory Compendium}.
\newblock Oxford University Press.

\bibitem[Kirchhoff et al.(2018)]{kirchhoff2018}
Kirchhoff, M.\ et al.\ (2018).
\newblock The Markov Blankets of Life: Autonomy, Active Inference and the Free
Energy Principle.
\newblock \emph{Journal of the Royal Society Interface}, 15(138).

\bibitem[Krakovna et al.(2020)]{krakovna2020}
Krakovna, V.\ et al.\ (2020).
\newblock Avoiding Side Effects in Complex Environments.
\newblock \emph{Advances in Neural Information Processing Systems}.

\bibitem[Lakoff \& Johnson(1980)]{LakoffJohnson1980}
Lakoff, G.\ \& Johnson, M.\ (1980).
\newblock \emph{Metaphors We Live By}.
\newblock University of Chicago Press.

\bibitem[Langosco et al.(2023)]{langosco2023goal}
Langosco, L.\ et al.\ (2023).
\newblock Goal Misgeneralization: Why Correct Specifications Aren't Enough for
Correct Goals.
\newblock \emph{arXiv:2105.14111}.

\bibitem[LeCun(2022)]{lecun2022}
LeCun, Y.\ (2022).
\newblock A Path Towards Autonomous Machine Intelligence.
\newblock OpenReview.

\bibitem[Leike(2022)]{leike2022alignment}
Leike, J.\ (2022).
\newblock Alignment Optimism.
\newblock Alignment Forum.

\bibitem[Mac Lane(1998)]{maclane1998}
Mac Lane, S.\ (1998).
\newblock \emph{Categories for the Working Mathematician}, 2nd ed.
\newblock Springer.

\bibitem[Manheim \& Garrabrant(2018)]{manheim2018categorizing}
Manheim, D.\ \& Garrabrant, S.\ (2018).
\newblock Categorizing Variants of Goodhart's Law.
\newblock \emph{arXiv:1803.04585}.

\bibitem[Nanda(2023)]{nanda2023progress}
Nanda, N.\ (2023).
\newblock Progress Measures for Grokking via Mechanistic Interpretability.
\newblock \emph{arXiv:2301.05217}.

\bibitem[Ngo(2022)]{ngo2022evals}
Ngo, R.\ (2022).
\newblock The Alignment Problem from a Deep Learning Perspective.
\newblock \emph{arXiv:2209.00626}.

\bibitem[Olah(2020)]{olah2020}
Olah, C.\ (2020).
\newblock Zoom In: An Introduction to Circuits.
\newblock \emph{Distill}.

\bibitem[Omohundro(2008)]{omohundro2008basic}
Omohundro, S.\ (2008).
\newblock The Basic AI Drives.
\newblock \emph{Proceedings of the 2008 Conference on Artificial General
Intelligence}.

\bibitem[Orfanos et al.(2024)]{orfanos2024situational}
Orfanos, A.\ et al.\ (2024).
\newblock Situational Awareness in Artificial Agents.
\newblock Manuscript.

\bibitem[Pathak et al.(2017)]{pathak2017curiosity}
Pathak, D.\ et al.\ (2017).
\newblock Curiosity-Driven Exploration by Self-Supervised Prediction.
\newblock \emph{Proceedings of ICML}.

\bibitem[Perez et al.(2022)]{perez2022discovering}
Perez, E.\ et al.\ (2022).
\newblock Discovering Language Model Behaviors with Model-Written Evaluations.
\newblock \emph{arXiv:2212.09251}.

\bibitem[Riehl(2017)]{riehl2017}
Riehl, E.\ (2017).
\newblock \emph{Category Theory in Context}.
\newblock Dover.

\bibitem[Schaeffer et al.(2024)]{schaeffer2024emergent}
Schaeffer, R.\ et al.\ (2024).
\newblock Are Emergent Abilities of Large Language Models a Mirage?
\newblock \emph{Advances in Neural Information Processing Systems}.

\bibitem[Schmidhuber(2010)]{schmidhuber2010formal}
Schmidhuber, J.\ (2010).
\newblock Formal Theory of Creativity, Fun, and Intrinsic Motivation.
\newblock \emph{IEEE Transactions on Autonomous Mental Development}, 2(3),
230--247.

\bibitem[Schumacher(1973)]{schumacher1973small}
Schumacher, E.F.\ (1973).
\newblock \emph{Small Is Beautiful: Economics as if People Mattered}.
\newblock Blond \& Briggs.

\bibitem[Seth(2021)]{seth2021being}
Seth, A.\ (2021).
\newblock \emph{Being You: A New Science of Consciousness}.
\newblock Dutton.

\bibitem[Shah et al.(2022)]{shah2022goal}
Shah, R.\ et al.\ (2022).
\newblock Goal Misgeneralization in Deep Reinforcement Learning.
\newblock \emph{arXiv:2105.14111}.

\bibitem[Soares \& Fallenstein(2015)]{soares2015}
Soares, N.\ \& Fallenstein, B.\ (2015).
\newblock Aligning Superintelligence with Human Interests: A Technical Research
Agenda.
\newblock MIRI Technical Report.

\bibitem[Spivak(2014)]{spivak2014}
Spivak, D.I.\ (2014).
\newblock \emph{Category Theory for the Sciences}.
\newblock MIT Press.

\bibitem[Star(1999)]{star1999ethnography}
Star, S.L.\ (1999).
\newblock The Ethnography of Infrastructure.
\newblock \emph{American Behavioral Scientist}, 43(3), 377--391.

\bibitem[Strogatz(2018)]{strogatz2018}
Strogatz, S.\ (2018).
\newblock \emph{Nonlinear Dynamics and Chaos}, 2nd ed.
\newblock Westview Press.

\bibitem[Talmy(2000)]{Talmy2000}
Talmy, L.\ (2000).
\newblock \emph{Toward a Cognitive Semantics}.
\newblock MIT Press.

\bibitem[Tishby et al.(1999)]{tishby1999}
Tishby, N., Pereira, F.C., \& Bialek, W.\ (1999).
\newblock The Information Bottleneck Method.
\newblock \emph{Proceedings of the 37th Allerton Conference}.

\bibitem[Tomasello(2016)]{Tomasello2016}
Tomasello, M.\ (2016).
\newblock \emph{A Natural History of Human Morality}.
\newblock Harvard University Press.

\bibitem[Turner et al.(2021)]{turner2021optimal}
Turner, A.\ et al.\ (2021).
\newblock Optimal Policies Tend to Seek Power.
\newblock \emph{Advances in Neural Information Processing Systems}.

\bibitem[Wei et al.(2023)]{wei2023jailbroken}
Wei, A.\ et al.\ (2023).
\newblock Jailbroken: How Does LLM Safety Training Fail?
\newblock \emph{Advances in Neural Information Processing Systems}.

\bibitem[Yang et al.(2019)]{yang2019task}
Yang, G.\ et al.\ (2019).
\newblock Task Representations in Neural Networks Trained to Perform Many
Different Tasks.
\newblock \emph{Nature Neuroscience}, 22, 297--306.

\bibitem[Zeng et al.(2023)]{zeng2023does}
Zeng, E.\ et al.\ (2023).
\newblock Does RLHF Actually Improve Alignment?
\newblock Manuscript.

\bibitem[Zou et al.(2023)]{zou2023universal}
Zou, A.\ et al.\ (2023).
\newblock Universal and Transferable Adversarial Attacks on Aligned Language
Models.
\newblock \emph{arXiv:2307.15043}.

\end{thebibliography}

\end{document}
