\documentclass[11pt]{article}

\usepackage{amsmath,amssymb,amsthm,amsfonts,mathtools}
\usepackage{bm}
\usepackage{stmaryrd}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{mathrsfs}
\usepackage{listings}
\usepackage{xcolor}
\geometry{margin=1in}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue!70!black},
  commentstyle=\color{gray},
  stringstyle=\color{teal},
  breaklines=true,
  breakatwhitespace=true,
  columns=flexible,
  keepspaces=true,
  showstringspaces=false,
  frame=single,
  framesep=4pt,
  xleftmargin=8pt,
  xrightmargin=8pt,
  aboveskip=8pt,
  belowskip=8pt
}

\setlength{\parskip}{1em}
\setlength{\parindent}{0pt}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{principle}{Principle}

% ------------------------------------------------------------
% THEORETICAL MONOGRAPH
% ------------------------------------------------------------

\title{\Huge\bfseries RSVP--Polyxan:\\
A Unified Field Theory of Semantic Hyperstructures}

\author{Flyxion\\[0.4em]
\normalsize Independent Researcher}
\date{2026}

\begin{document}

\maketitle

\begin{abstract}
This paper develops a unified theoretical framework coupling the
Relativistic Scalar--Vector Plenum (RSVP) field system, defined on a
semantic manifold of embeddings, with the Polyxan hypermedia substrate,
a typed, bidirectionally linked graph of content atoms and media quines
in the Xanadu tradition.
We construct an action functional for RSVP fields over semantic space,
derive Euler--Lagrange equations, and introduce a geometric interpretation
of polysemantic clusters.
The framework is then extended from a static semantic field theory into a
dynamical world model with persistent latent memory, projection-based
observation, and constrained field evolution.
Constraint projection is internalized into the variational structure via
Lagrange multipliers, yielding a principled dynamics in which neural
components act as generalized forcing terms rather than unconstrained
updates.
Noether's theorem applied to the augmented action produces semantic
conservation laws for coherence, directional agency, and causal residue.
The system is then reformulated in Hamiltonian phase-space terms,
equipped with gauge symmetries corresponding to semantically equivalent
representations, and finally extended to a stochastic path-integral
formulation that subsumes diffusion generative models as a special case.
A sheaf-theoretic treatment of galaxy-local worldviews is refined into a
derived sheaf valued in $\infty$-groupoids, and the Polyxan category is
lifted to an $\infty$-category in which semantic inconsistency is
characterized by non-vanishing cohomology.
The framework is then brought into precise contact with MultiGen, a
concrete diffusion-based generative game engine, which is shown to be a
degenerate projection of RSVP--Polyxan obtained by collapsing the full
latent field to a static scalar substrate and eliminating the write
operator.
The factored update $\Pi_{\mathcal{C}} \circ \mathcal{N}_\psi$ is
specified with six explicit field-level constraint equations governing
mass consistency, energy budgets, topological admissibility, causal write
locality, entropy accounting, and identity persistence.
Fast and slow update cadences, event-typed operator decomposition,
read-write cycle closure, and a four-phase training curriculum complete
the operational architecture.
Multiplayer interaction is characterized as causal field synchronization
over a shared substrate, with concurrent edits resolved by constraint
projection rather than consensus protocol.
Throughout, we present proofs, lemmas, and structural invariants that
guarantee coherence between continuous semantic fields and discrete
hypergraphs.
\end{abstract}

\newpage
\tableofcontents
\newpage

% =============================================================
\section{Introduction}

The RSVP--Polyxan framework fuses three mathematical layers.
The first is a \textbf{semantic manifold} $X$ where conceptual embeddings
live.
The second is a \textbf{field theory} consisting of a scalar potential
$\Phi$, vector flow $\mathbf{v}$, and entropy field $S$.
The third is a \textbf{typed hypergraph} of content atoms and links, whose
structure both sources and responds to RSVP fields.

The goal is to provide a unified variational, geometric, and categorical
description of semantic information flow, and to promote this description
into a full dynamical world engine with physics, memory, and rendering
separation.
The paper proceeds through three ascending layers of theoretical
elaboration.
At the classical layer we establish Lagrangian and Hamiltonian dynamics,
constraint manifolds, and Noether invariants.
At the stochastic layer we construct a path integral over field
configurations and identify diffusion generative models as sampling
trajectories from this integral.
At the derived layer we lift the Polyxan category to an
$\infty$-categorical structure, characterize semantic consistency as
homotopy-coherent gluing, and interpret hallucination as a cohomological
obstruction.

% =============================================================
\section{Semantic Manifold and RSVP Fields}

Let $(X,g)$ be a smooth Riemannian manifold representing semantic space.
Define three RSVP fields:
\[
\Phi: X\to\mathbb{R},\quad
\mathbf{v}: X\to T X,\quad
S: X\to \mathbb{R}.
\]
The scalar field $\Phi$ encodes semantic potential or coherence density.
The vector field $\mathbf{v}$ encodes directed agency or semantic drift.
The scalar field $S$ encodes entropy, uncertainty, or morphic degeneracy.
Field derivatives are defined using the Levi-Civita connection.

% =============================================================
\section{RSVP Lagrangian Formulation}

We propose an action functional:
\[
\mathcal{A}[\Phi,\mathbf{v},S]
= \int_X \mathcal{L}\, d\mu_g,
\]
where the Lagrangian density incorporates kinetic, elastic, and potential
terms:
\begin{align*}
\mathcal{L}
&= \frac{1}{2}(\partial_t\Phi)^2
 - \frac{c_\Phi^2}{2}\|\nabla\Phi\|^2 \\
&\quad + \frac{1}{2}\|\partial_t \mathbf{v}\|^2
 - \frac{c_v^2}{2}\|\nabla\mathbf{v}\|^2 \\
&\quad + \frac{1}{2}(\partial_t S)^2
 - \frac{c_S^2}{2}\|\nabla S\|^2 \\
&\quad - V(\Phi,\mathbf{v},S; \rho, \kappa).
\end{align*}
Here $\rho(x)$ denotes the node density induced by the Polyxan graph, and
$\kappa(x)$ denotes the local curvature induced by typed links.
Variation yields Euler--Lagrange equations:
\[
\frac{\delta \mathcal{A}}{\delta \Phi}
= \partial_t^2\Phi - c_\Phi^2\Delta \Phi
  - \frac{\partial V}{\partial\Phi} = 0,
\]
and similarly for $\mathbf{v}$ and $S$.

% =============================================================
\section{Coupling RSVP to the Polyxan Hypergraph}

Let $G=(V,E)$ be the Polyxan graph.
Each node $i\in V$ is assigned an embedding $\mathbf{z}_i\in X$.
Define discrete samples:
\[
\Phi_i = \Phi(\mathbf{z}_i),\quad
\mathbf{v}_i = \mathbf{v}(\mathbf{z}_i),\quad
S_i = S(\mathbf{z}_i).
\]
Define a discrete evolution law:
\[
\frac{d\mathbf{z}_i}{dt}
= -\alpha \nabla \Phi(\mathbf{z}_i)
+ \beta\, \mathbf{v}(\mathbf{z}_i)
- \gamma \nabla S(\mathbf{z}_i).
\]
Thus embeddings evolve in RSVP gradient flows.

\subsection{Topological Curvature from Links}

Define a link-curvature invariant:
\[
\kappa_i = \sum_{j,k \sim i} f_{\text{tri}}(i,j,k),
\]
counting typed 3-cycles weighted by link types.
This curvature term shapes the potential $V$.

% =============================================================
\section{Latent Field Memory and World-State Representation}

The classical RSVP formulation treats $(\Phi,\mathbf{v},S)$ as fields over
a semantic manifold $X$.
However, this representation remains insufficient for persistent world
modeling, as it does not explicitly encode revisitable state, appearance,
or causal residue.

We therefore introduce a time-indexed latent field memory:
\[
\mathcal{M}_t : X \to \mathbb{R}^d,
\]
which encodes the full world-state as a structured field over semantic space.
We decompose $\mathcal{M}_t$ into interpretable components:
\[
\mathcal{M}_t(x)
=
\big(
\Phi_t(x),\,
\mathbf{v}_t(x),\,
S_t(x),\,
A_t(x),\,
R_t(x),\,
z_t(x)
\big).
\]
The first three components are the scalar semantic potential $\Phi_t$,
the directed flow $\mathbf{v}_t$, and the entropy-uncertainty field $S_t$
already established by the RSVP formulation.
The remaining three are new: $A_t$ encodes affordance structure,
$R_t$ encodes causal residue accumulated through the system's history,
and $z_t$ encodes latent appearance, the field-theoretic analogue of a
rendered view.
This decomposition generalizes both continuous RSVP fields and discrete
Polyxan structures into a unified persistent substrate.

\subsection{State Decomposition}

We define the full system state:
\[
X_t = (\mathcal{M}_t, p_t, h_t),
\]
where $p_t \in TX$ is the agent trajectory coordinate and
$h_t = (o_{t-L+1},\dots,o_t)$ is a finite observation trace of length $L$.
Crucially, $h_t$ is no longer the world state but merely a local perceptual
history.
The world-state resides entirely in $\mathcal{M}_t$, and observations
are derived from it by projection.

% =============================================================
\section{Observation as Projection Operator}

In classical generative models the world is implicitly encoded in the
observation sequence.
In contrast, we treat observation as a projection from latent field memory.
Define the observation operator:
\[
\mathcal{O}_\theta :
(\mathcal{M}_t, p_t, h_t, a_t)
\;\longrightarrow\;
o_{t+1},
\]
where $a_t$ is an action input.
Observations are then sampled via:
\[
o_{t+1} \sim
p_\theta\big(o \mid \operatorname{Read}(\mathcal{M}_t,p_t), h_t, a_t\big).
\]
The operator $\operatorname{Read}(\mathcal{M}_t,p_t)$ extracts a
view-dependent slice of the field, analogous to a camera projection from
a three-dimensional scene.

\subsection{Observation Does Not Store State}

Unlike autoregressive systems, $\mathcal{O}_\theta$ does not maintain
world-state; it merely resolves latent uncertainty into visible form.

\begin{principle}
The world is stored in $\mathcal{M}_t$, not in the observation history.
\end{principle}

This separation is structurally fundamental.
Long-range consistency, persistent object identity, and causal coherence
are maintained not through the growth of an observation buffer but through
the evolution of the latent field.

% =============================================================
\section{Dynamics as Constrained Field Evolution}

We define a dynamics operator:
\[
\Delta_\psi :
(\mathcal{M}_t, p_t, a_t, o_{t+1})
\;\longrightarrow\;
\delta \mathcal{M}_t,
\]
which proposes updates to the latent field.
The world-state evolves according to:
\[
\mathcal{M}_{t+1}
=
\Pi_{\mathcal{C}}
\big(
\mathcal{M}_t + \Delta_\psi(\mathcal{M}_t,p_t,a_t,o_{t+1})
\big),
\]
where $\Pi_{\mathcal{C}}$ is a projection onto a constraint manifold
$\mathcal{C}$.

\subsection{Constraint Manifold}

The admissible set $\mathcal{C}$ enforces a system of global invariants.
Conservation constraints require that mass and energy be preserved across
field updates.
Topological consistency requires that the Polyxan hypergraph remain
well-formed under induced modifications.
Causal locality requires that updates propagate only within the future
causal cone of the acting agent.
Entropy monotonicity outside observation regions requires that $S_t$
not decrease in unobserved subsets of $X$, in accordance with the second
law interpreted at the field level.
The projection $\Pi_{\mathcal{C}}$ is therefore not a free deformation
but a physically and semantically constrained retraction.

\subsection{Invention-to-Memory Principle}

Generation may invent details only once; after invention, details must be
promoted into memory.
Formally, if a region $x \in X$ is observed at time $t+1$, then:
\[
z_{t+1}(x) \gets z_{t+1}(x) + \lambda\, \hat{z}(x),
\]
where $\hat{z}(x)$ is the inferred appearance extracted by
$\mathcal{O}_\theta$.
Thus appearance becomes persistent after first observation, and subsequent
observations of the same region are constrained by the accumulated field
value $z_t(x)$ rather than re-sampled from the prior.

% =============================================================
\section{Variational Constrained Field Evolution}

The update rule of the previous section may be interpreted as a projection
onto an admissible manifold of states.
We now internalize this projection into a variational formulation, so that
constraint enforcement arises continuously from the equations of motion
rather than as a post-hoc correction.

\subsection{Augmented Action Functional}

Define an augmented action:
\[
\mathcal{A}_{\text{aug}}[\mathcal{M},\lambda]
=
\int dt \int_X
\Big(
\mathcal{L}(\mathcal{M}, \partial_t \mathcal{M})
+
\langle \lambda, \mathcal{C}(\mathcal{M}) \rangle
\Big)\, d\mu_g,
\]
where $\mathcal{L}$ extends the RSVP Lagrangian to all components of
$\mathcal{M}_t$, the equation $\mathcal{C}(\mathcal{M})=0$ encodes
constraint equations, and $\lambda$ is a field of Lagrange multipliers
enforcing those constraints continuously throughout the evolution.

\subsection{Euler--Lagrange Equations with Constraints}

Variation of $\mathcal{A}_{\text{aug}}$ with respect to $\mathcal{M}$
yields:
\[
\frac{\delta \mathcal{A}_{\text{aug}}}{\delta \mathcal{M}}
=
\frac{\partial \mathcal{L}}{\partial \mathcal{M}}
-
\partial_t \left(\frac{\partial \mathcal{L}}{\partial (\partial_t \mathcal{M})}\right)
+
\lambda \cdot \frac{\partial \mathcal{C}}{\partial \mathcal{M}}
= 0,
\]
together with the constraint equation $\mathcal{C}(\mathcal{M}) = 0$.
The projection $\Pi_{\mathcal{C}}$ is thereby realized as a continuous
constraint force in the equations of motion rather than a discrete
retraction applied at each time step.

\subsection{Neural Proposal as Generalized Force}

We reinterpret the learned update $\Delta_\psi$ as an external forcing term
in the equations of motion:
\[
\partial_t^2 \mathcal{M}
=
-\nabla_{\mathcal{M}} E(\mathcal{M})
+
F_\psi(\mathcal{M},p_t,a_t,o_{t+1}),
\]
where $F_\psi$ is a learned generalized force field derived from $\Delta_\psi$.

\begin{principle}
The neural component proposes dynamics, but admissibility is enforced
by variational constraints.
\end{principle}

This separation guarantees that learned forcing can enrich the dynamics
without violating the invariants encoded in $\mathcal{C}$.

% =============================================================
\section{Noether Theorem for Semantic Field Invariants}

The augmented RSVP system admits conservation laws arising from symmetries
of the action functional.
These laws are not imposed externally but follow from the variational
structure established in the previous section.

\subsection{Symmetry Transformations}

Let $\mathcal{M}_t$ transform under a one-parameter group:
\[
\mathcal{M}_t \mapsto \mathcal{M}_t + \epsilon \, \delta \mathcal{M}_t,
\]
such that the augmented action satisfies $\delta \mathcal{A}_{\text{aug}} = 0$.

\subsection{Conserved Current}

Then there exists a conserved current $J^\mu$ satisfying:
\[
\partial_t J^0 + \nabla \cdot \mathbf{J} = 0.
\]

\subsection{Examples of Semantic Invariants}

\paragraph{Coherence Conservation.}
If $\mathcal{L}$ is invariant under $\Phi \mapsto \Phi + \epsilon$, then
\[
Q_\Phi = \int_X \Phi\, d\mu_g
\]
is conserved, representing total semantic coherence across the manifold.

\paragraph{Flow Conservation.}
If $\mathcal{L}$ is invariant under translations of $\mathbf{v}$, then
\[
Q_v = \int_X \mathbf{v}\, d\mu_g
\]
is conserved, representing total directional agency.

\paragraph{Entropy Balance.}
If $\mathcal{L}$ admits a scaling symmetry in $S$, then
\[
Q_S = \int_X S\, d\mu_g
\]
is conserved modulo boundary flux, representing entropy redistribution
rather than entropy creation.

\paragraph{Residue Persistence.}
For invariance under shifts of $R$ in unobserved regions, we obtain:
\[
\partial_t R(x) \ge 0 \quad \text{for } x \notin \operatorname{view}(p_t).
\]
Thus causal residue cannot spontaneously disappear from regions outside the
agent's perceptual cone.

\begin{theorem}[Semantic Conservation Law]
If the augmented action is invariant under a continuous transformation,
then the corresponding semantic quantity is conserved along admissible
field trajectories.
\end{theorem}

\begin{proof}
Follows from the standard Noether theorem applied to
$\mathcal{A}_{\text{aug}}$.
\end{proof}

The RSVP--Polyxan system thus satisfies variational evolution through
the field equations derived from $\mathcal{A}_{\text{aug}}$, constrained
admissibility enforced by the projection manifold $\mathcal{C}$, learned
perturbation through the neural forcing $F_\psi$, and conserved quantities
arising from Noether's theorem.
Together these properties establish the system as a genuine field theory
of semantic dynamics rather than a purely descriptive model.

% =============================================================
\section{Hamiltonian Formulation of RSVP Memory Dynamics}

We now reformulate RSVP--Polyxan dynamics in Hamiltonian form, providing
a phase-space description of latent field evolution.

\subsection{Canonical Variables}

Let $\mathcal{M}_t : X \to \mathbb{R}^d$ denote the latent field memory.
Define conjugate momenta:
\[
\Pi_t(x) = \frac{\partial \mathcal{L}}{\partial (\partial_t \mathcal{M}_t(x))}.
\]
The phase space is therefore $\mathcal{P} = \{(\mathcal{M}, \Pi)\}$,
a space of pairs of field configurations and their conjugate momenta.

\subsection{Hamiltonian Functional}

Define the Hamiltonian via Legendre transform:
\[
\mathcal{H}[\mathcal{M}, \Pi]
=
\int_X
\left(
\langle \Pi, \partial_t \mathcal{M} \rangle - \mathcal{L}
\right)
\, d\mu_g.
\]
Expanding the kinetic and gradient terms yields:
\[
\mathcal{H}
=
\int_X
\left(
\frac{1}{2}\|\Pi\|^2
+
\frac{c_\Phi^2}{2}\|\nabla \Phi\|^2
+
\frac{c_v^2}{2}\|\nabla \mathbf{v}\|^2
+
\frac{c_S^2}{2}\|\nabla S\|^2
+
V(\mathcal{M})
\right)
d\mu_g.
\]
The Hamiltonian thus decomposes into a kinetic term $\frac{1}{2}\|\Pi\|^2$
representing the rate of change of field configurations, structural gradient
terms $\|\nabla\Phi\|^2$, $\|\nabla\mathbf{v}\|^2$, $\|\nabla S\|^2$
representing the elastic energy stored in spatial variation, and the
interaction potential $V$ encoding coupling to the Polyxan graph.

\subsection{Hamilton's Equations}

The evolution equations are:
\[
\partial_t \mathcal{M}
=
\frac{\delta \mathcal{H}}{\delta \Pi},
\quad
\partial_t \Pi
=
- \frac{\delta \mathcal{H}}{\delta \mathcal{M}} + F_\psi.
\]
Here $F_\psi$ is the learned forcing term from Section~\ref{sec:variational}.

\subsection{Symplectic Structure}

Define the canonical symplectic form:
\[
\omega = \int_X d\Pi \wedge d\mathcal{M}.
\]

\begin{theorem}
In the absence of forcing and dissipation, the RSVP memory dynamics
preserve the symplectic form $\omega$.
\end{theorem}

\begin{proof}[Sketch]
Follows from the standard result that Hamiltonian flow preserves $\omega$,
since the unforced equations of motion are precisely the condition
$\iota_{X_H}\omega = d\mathcal{H}$.
\end{proof}

Symplectic preservation means that the phase-space volume of any ensemble
of field configurations is conserved by the unforced dynamics.
Learned forcing $F_\psi$ breaks symplecticity and introduces dissipation,
but the constraint projection $\Pi_{\mathcal{C}}$ acts as a restoring
force that keeps trajectories on the admissible manifold.
RSVP dynamics thus correspond to energy-conserving flow on the space of
semantic configurations, perturbed by learned forces that drive the system
toward observed evidence.

% =============================================================
\section{Polyxan as a Discrete Interface to Field Memory}

The Polyxan hypergraph $G=(V,E)$ provides a discrete interface into the
continuous field $\mathcal{M}_t$.
Each node $i \in V$ corresponds to a sample at $\mathbf{z}_i \in X$, with
the full field value:
\[
\mathcal{M}_t(\mathbf{z}_i)
=
(\Phi_i, \mathbf{v}_i, S_i, A_i, R_i, z_i).
\]
Edges in $E$ define constraints and transformations that influence the
local field geometry by modifying $\rho$ and $\kappa$ in the potential $V$.

\subsection{Bidirectional Coupling}

The coupling between field and graph is bidirectional.
In the field-to-graph direction, RSVP fields determine node dynamics through
the gradient flow equation established in Section~4.
In the graph-to-field direction, hypergraph structure modifies the density
$\rho$ and curvature $\kappa$ terms, which in turn influence the field
evolution through the potential $V(\Phi,\mathbf{v},S;\rho,\kappa)$.
Thus Polyxan acts as a read/write layer over $\mathcal{M}_t$: reading the
field to generate observable structure, and writing curvature and density
back into the field to record the effects of discrete interactions.

% =============================================================
\section{Gauge Symmetries and Semantic Equivalence}

We now introduce gauge symmetries corresponding to transformations that
preserve semantic equivalence classes.
These symmetries formalize the principle that multiple internal
representations may correspond to the same observable semantic reality,
and that a well-formed theory should be independent of which representative
is chosen.

\subsection{Gauge Transformations}

Let $\mathcal{G}$ be a group of transformations acting on $\mathcal{M}_t$:
\[
\mathcal{M}_t \mapsto \mathcal{M}_t^g
=
g \cdot \mathcal{M}_t,
\quad g \in \mathcal{G},
\]
subject to the requirement that observations be gauge-invariant:
\[
\mathcal{O}_\theta(\mathcal{M}_t^g, p_t) =
\mathcal{O}_\theta(\mathcal{M}_t, p_t).
\]

\subsection{Examples of Gauge Groups}

Three natural gauge groups arise from the structure of $\mathcal{M}_t$.
The latent appearance gauge acts by $z_t(x) \mapsto z_t(x) + \nabla \chi(x)$
for a scalar field $\chi$ preserving perceptual equivalence under
$\mathcal{O}_\theta$.
Embedding reparameterization acts by $\mathbf{z}_i \mapsto f(\mathbf{z}_i)$
where $f$ is a diffeomorphism of $X$ preserving semantic relations encoded
in the hypergraph.
Affordance relabeling acts by $A_t(x) \mapsto \sigma(A_t(x))$ where
$\sigma$ is a symmetry of the interaction algebra governing
action-affordance pairings.

\subsection{Gauge Invariance of the Action}

\begin{definition}
The RSVP action is gauge-invariant if
$\mathcal{A}[\mathcal{M}] = \mathcal{A}[\mathcal{M}^g]$
for all $g \in \mathcal{G}$.
\end{definition}

\begin{theorem}
Gauge invariance implies conservation laws corresponding to redundant
semantic degrees of freedom, via Noether's second theorem.
\end{theorem}

\subsection{Connection and Covariant Derivative}

Introduce a gauge connection $\mathcal{A}_\mu$ and define the covariant
derivative:
\[
D_\mu \mathcal{M}
=
\partial_\mu \mathcal{M}
+
\mathcal{A}_\mu \cdot \mathcal{M}.
\]
Gradients in the Lagrangian are replaced by covariant derivatives
$\nabla \to D$, ensuring that the action transforms covariantly under
local gauge transformations.
The field strength curvature is then:
\[
\mathcal{F}_{\mu\nu}
=
[D_\mu, D_\nu],
\]
which measures the obstruction to consistent semantic identification
across distinct regions of $X$.
Non-vanishing curvature $\mathcal{F}_{\mu\nu} \neq 0$ indicates that
representations in overlapping neighborhoods cannot be identified by a
flat connection, a field-theoretic signature of representational
inhomogeneity.

% =============================================================
\section{Sheaf-Theoretic Interpretation}

Define for each user $u$ an open neighborhood $U_u\subset X$.
Define a presheaf $\mathscr{G}$ of galaxy renderings:
\[
\mathscr{G}(U)=\{\text{layout functions over }U\}.
\]
Restriction maps obey $\rho_{UV}(G_U)=G_V$ for $V\subset U$.

\begin{theorem}
$\mathscr{G}$ is a sheaf if the layout map is determined by
$(\Phi,\mathbf{v},S)$, embeddings are globally indexed, and projections
are deterministic.
\end{theorem}

\begin{proof}[Sketch]
Local layouts glued along intersections produce a unique global layout due
to determinism and continuity of the underlying field.
\end{proof}

% =============================================================
\section{Derived and $\infty$-Categorical Structure of Polyxan}
\label{sec:derived}

We now lift the Polyxan--RSVP framework into derived and higher-categorical
form, providing a language adequate to the treatment of higher-order
equivalences and obstruction phenomena.

\subsection{Derived Field Space}

Instead of treating $\mathcal{M}_t$ as a point in a manifold, we consider
it as an object in a derived moduli space:
\[
\mathcal{M}_t \in \mathcal{X},
\]
where $\mathcal{X}$ is a derived stack encoding field configurations,
constraint solutions, and the higher-order equivalences between them.
The passage to derived geometry allows the moduli space to carry
cohomological information that is invisible at the classical level.

\subsection{Obstruction Theory}

Let $\mathrm{Obs}^k(\mathcal{M})$ denote the $k$-th obstruction class of a
configuration $\mathcal{M}$.

\begin{definition}
A configuration $\mathcal{M}$ is admissible if
$\mathrm{Obs}^k(\mathcal{M}) = 0$ for all $k \ge 1$.
\end{definition}

Non-vanishing obstruction classes correspond to three qualitatively distinct
failure modes.
Inconsistencies in world-state arise when local field patches fail to agree
on overlaps.
Multiplayer desynchronization arises when distinct agent perspectives induce
incompatible updates to $\mathcal{M}_t$.
Failure of semantic gluing arises when local Polyxan neighborhoods cannot
be assembled into a coherent global hypergraph.

\subsection{Polyxan as an $\infty$-Category}

We upgrade the category $\mathbf{C}$ of Section~\ref{sec:cat} to an
$\infty$-category $\mathbf{C}_\infty$, in which objects are content atoms,
1-morphisms are typed links, and higher morphisms encode equivalences of
transformations and coherence data for composites.
The passage from $\mathbf{C}$ to $\mathbf{C}_\infty$ is necessary because
the composition of semantic transformations is in general only associative
up to coherent homotopy, not strictly.

\subsection{Functor to Derived Fields}

Define:
\[
\mathsf{RSVP} :
\mathbf{C}_\infty \to \mathbf{Field}_\infty,
\]
mapping hypergraph structure into derived field configurations.
This functor preserves the homotopy-theoretic structure of both sides:
equivalences of content atoms map to quasi-isomorphisms of field
configurations.

\subsection{Homotopy-Coherent Gluing}

The sheaf $\mathscr{G}$ of Section~13 is refined to a derived sheaf:
\[
\mathscr{G}^\infty : \mathrm{Op}(X)^{op} \to \infty\text{-Groupoids}.
\]

\begin{theorem}
Global semantic coherence corresponds to the existence of a homotopy
limit:
\[
\mathcal{M} \simeq \operatorname{holim}\, \mathscr{G}^\infty(U_i).
\]
\end{theorem}

\subsection{Interpretation of Hallucination}

We interpret hallucination as non-vanishing first cohomology of the derived
sheaf:
\[
H^1(\mathscr{G}^\infty) \neq 0.
\]
This represents the failure of local semantic patches to glue into a
consistent global structure: the system generates locally coherent content
that cannot be assembled into a globally consistent world-state.
The cohomological characterization provides a mathematically precise
criterion for detecting and diagnosing hallucination without appeal to
external ground truth.

\subsection{Semantic Fixed Points}

A consistent world-state is a derived fixed point:
\[
\mathcal{M}
\simeq
\mathsf{RSVP}(\mathsf{Poly}(\mathcal{M})).
\]
Semantic stability therefore corresponds to homotopy-invariant closure:
the system's field configuration is a fixed point of the composed functor
up to coherent equivalence.

% =============================================================
\section{Category-Theoretic Architecture}
\label{sec:cat}

Define category $\mathbf{C}$ with objects given by Content Atoms and
Spans, and morphisms given by Typed Links.
Define the Polycompiler as an endofunctor:
\[
\mathsf{Poly}: \mathbf{C}\to\mathbf{C}.
\]
Define RSVP as a functor $\mathsf{RSVP}: \mathbf{C}\to\mathbf{F}$
where $\mathbf{F}$ is the category of field configurations.

\begin{theorem}
The composition $\mathsf{RSVP}\circ \mathsf{Poly}$ yields a natural
transformation describing generative semantic deformation.
\end{theorem}

% =============================================================
\section{Quantization of RSVP Field Dynamics}

We now construct a quantum, or equivalently stochastic, extension of RSVP
memory dynamics, interpreting the latent field evolution as a path integral
over field configurations.

\subsection{Configuration Space}

Let $\mathcal{M}(t)$ denote a trajectory in field space:
\[
\mathcal{M} : [0,T] \to \mathcal{F},
\]
where $\mathcal{F}$ is the space of admissible field configurations, that is,
the classical solutions to the constrained Euler--Lagrange equations.

\subsection{Path Integral}

Define the partition functional:
\[
Z = \int \mathcal{D}\mathcal{M}\;
\exp\left(
\frac{i}{\hbar} \mathcal{A}_{\text{aug}}[\mathcal{M}]
\right),
\]
where the functional integral is taken over all paths in $\mathcal{F}$.
Observables $\mathcal{O}$ are computed as:
\[
\langle \mathcal{O} \rangle
=
\frac{1}{Z}
\int \mathcal{D}\mathcal{M}\;
\mathcal{O}[\mathcal{M}]
\exp\left(
\frac{i}{\hbar} \mathcal{A}_{\text{aug}}
\right).
\]

\subsection{Stochastic Diffusion Limit}

Under Wick rotation $t \mapsto -i\tau$, the Lorentzian path integral
becomes a Euclidean one:
\[
Z_E
=
\int \mathcal{D}\mathcal{M}\;
\exp\left(
-\frac{1}{\hbar} \mathcal{A}_E[\mathcal{M}]
\right).
\]
This corresponds to the stochastic differential equation:
\[
\partial_\tau \mathcal{M}
=
-\nabla_{\mathcal{M}} E(\mathcal{M})
+
\eta(\tau),
\]
where $\eta$ is Gaussian white noise with variance $\hbar$.
The invariant measure of this process is the Boltzmann distribution
$p(\mathcal{M}) \propto e^{-E(\mathcal{M})/\hbar}$ over field configurations.

\subsection{Connection to Diffusion Generative Models}

Diffusion generative models correspond precisely to sampling trajectories
from this stochastic evolution.
The RSVP energy functional $E$ defines the score function
$\nabla_{\mathcal{M}} \log p(\mathcal{M})$ that guides the reverse diffusion
process.
Generation is then sampling from the Boltzmann distribution over fields,
and the denoising process is the time-reversal of the stochastic gradient
flow on $E$.

\begin{principle}
Diffusion models are discrete approximations to stochastic quantization
of RSVP field dynamics.
\end{principle}

A single generated world is one sampled trajectory through $\mathcal{F}$.
Consistency of the generated world emerges from the action functional,
which penalizes configurations that violate the constraints encoded in
$\mathcal{C}$.
Uncertainty at any point $x \in X$ corresponds to the entropy field
$S_t(x)$, which is high where the path integral measure is diffuse and
low where it concentrates on a narrow set of field configurations.

% =============================================================
\section{Stability and Energy Minimization}

Define energy:
\[
E = \int_X V(\Phi,\mathbf{v},S;\rho,\kappa)\, d\mu_g.
\]

\begin{lemma}
Fixed points of RSVP flow minimize $E$ subject to embedding constraints.
\end{lemma}

\begin{proof}[Sketch]
Gradient descent of embeddings follows $-\nabla E$ by design.
\end{proof}

% =============================================================
\section{Global Reset as Field Reconfiguration}

Define reset operator:
\[
\mathcal{R}: (\Phi,\mathbf{v},S,\mathbf{z})\mapsto
(\Phi',\mathbf{v}',S',\mathbf{z}'),
\]
where $\mathcal{R}$ recomputes fields and embeddings by re-estimating
$\rho$ and $\kappa$, relaxing the field equations to a new equilibrium,
and reprojecting embeddings into $X$.
Reset corresponds to a global reconfiguration of semantic geometry,
and is distinguished from the incremental update $\Delta_\psi$ in that
it does not preserve the prior field configuration but reconstructs it
from the current hypergraph topology.

% =============================================================
\section{MultiGen as a Degenerate Projection of RSVP}
\label{sec:multigen}

MultiGen \cite{po2026multigen} represents the most structurally
sophisticated system in the emerging class of diffusion-based generative
game engines.
We do not treat it merely as a compatible architecture.
We show that it is a degenerate projection of the RSVP trajectory-operator
framework, arising under severe dimensional restriction of the full field
state.

\subsection{The Projection Map}

MultiGen defines an explicit world-state:
\[
S_t = (M,\, p_t,\, o_{t-L+1:t}),
\]
consisting of a global geometric memory $M$, a player pose $p_t$, and a
finite visual trace of length $L$.
In RSVP terms, this bundle arises from the full state
$X_t = (\mathcal{M}_t, p_t, h_t)$ via a projection:
\[
\pi : X_t \;\longrightarrow\; S_t,
\]
which acts by (i) collapsing the rich latent field
$\mathcal{M}_t = (\Phi_t, \mathbf{v}_t, S_t, A_t, R_t, z_t)$ to its
scalar substrate $\Phi_t$ alone, (ii) retaining only the pose coordinate
$p_t$, and (iii) replacing the full observation history with a window of
length $L$.
MultiGen is therefore not a different kind of system but the RSVP system
evaluated at $\pi$.

\subsection{Field Correspondences}

The correspondence between MultiGen components and RSVP fields is exact.
The authored minimap $M = (V,E)$ corresponds to a coarse scalar constraint
field $M \sim \Phi$: it encodes where structure is geometrically possible
and where it is blocked, but carries no semantic content beyond occupancy.
The player pose $p_t$ is the agent trajectory coordinate, unchanged across
both frameworks.
The observation operator is a projection from field and pose:
\[
\mathcal{O}_\phi : (\Phi,\, p_t,\, h_t,\, a_t) \;\mapsto\; o_{t+1},
\]
and the dynamics operator updates pose only:
\[
\mathcal{D}_\psi : (p_t,\, a_t,\, r_t,\, f_t) \;\mapsto\; p_{t+1}.
\]
The full MultiGen rollout is therefore:
\[
M_{t+1} = M_t, \qquad
p_{t+1} = \mathcal{D}_\psi(p_t, a_t, \cdots), \qquad
o_{t+1} = \mathcal{O}_\phi(M_t, p_t, \cdots).
\]
The substrate is frozen: $M_{t+1} = M_t$ for all $t$.

\subsection{The Missing Write Operator}

The structural deficiency of MultiGen is made precise by the absence of a
write operator.
In the RSVP framework, the dynamics operator $\Delta_\psi$ writes
observation-derived information back into the field:
\[
\mathcal{W} : o_{t+1} \;\longrightarrow\; \mathcal{M}_{t+1}.
\]
MultiGen has no such operator.
The field $M$ is an immutable anchor produced by a level designer prior to
runtime; it is read by both observation and dynamics modules but never
updated by them.
This is not an engineering limitation but a structural one.
The absence of $\mathcal{W}$ means the system has no feedback channel from
observation back into substrate.
It is a one-way system: constraint flows toward projection, but projection
never flows back toward constraint.

\subsection{MultiGen as a Theorem}

The preceding analysis entitles the following proposition.

\begin{proposition}
MultiGen is the degenerate projection of RSVP--Polyxan obtained by
setting $\mathbf{v}_t = 0$, $A_t = 0$, $R_t = 0$, $z_t = 0$,
restricting $\Phi_t$ to a static geometric graph, and removing the
write operator $\mathcal{W}$.
Under this projection, the RSVP update equations reduce exactly to the
MultiGen rollout.
\end{proposition}

The RSVP generalization is therefore not merely an extension but a
structural completion.
MultiGen establishes that diffusion must be demoted from world-state to
observation operator.
RSVP establishes the missing half: observation must be coupled to a
write-back mechanism that promotes realized detail into the persistent
field, so that the system is bidirectional---constraint to projection,
and projection back to constraint---and the field can converge toward
a fixed point under its own dynamics.

\subsection{Multiplayer as Evidence}

MultiGen's multiplayer capability is the most direct empirical confirmation
of the substrate-first principle.
In a purely diffusion-based system, two agents cannot share a world because
there is no shared source of truth beyond the model weights.
MultiGen achieves cross-agent consistency by making all agents read from
the same $M$, so spatial agreement is guaranteed by geometric constraint
rather than model intelligence.
In the RSVP extension, multiplayer no longer reduces to reading from a
shared map.
It becomes co-evolution of a shared field: every action by any agent
induces a write into $\mathcal{M}_t$ via $\mathcal{W}$, and all subsequent
observations across all agents are constrained by the accumulated field
history.
Two agents agree not merely on where walls are but on what has happened to
those walls.
Consistency is achieved not through spatial anchoring alone but through
shared causal residue.

% =============================================================
\section{Factored Update: Proposal, Projection, and Commit}
\label{sec:factored}

We now refine the dynamics operator $\Delta_\psi$ introduced in
Section~7 into a three-stage factored update, and specify the constraint
set $\mathcal{C}$ with field-level equations rather than prose invariants.

\subsection{The Three-Stage Update}

A purely neural update $\Delta_\psi$ applied without constraint will
exhibit drift on long horizons: high-capacity updaters treat the field as
a writable canvas, producing visually plausible local edits that accumulate
global contradictions.
The correct formulation is a constrained neural operator: a learnable
update whose proposals live inside a space defined by explicit field-level
laws.
We factor the update into three stages:
\[
\mathcal{M}_{t+1}
=
\Pi_{\mathcal{C}}\!\Big(
\mathcal{M}_t
+
\mathcal{N}_\psi(\mathcal{M}_t, p_t, a_t, o_{t+1})
\Big),
\]
where $\mathcal{N}_\psi$ is the neural proposal network, which generates
local edits to the field; $\Pi_{\mathcal{C}}$ is the projection onto the
admissible constraint manifold; and the commit stage writes back only the
closest valid state.
This structure mirrors incompressible flow solvers: an unconstrained
velocity update is followed by a pressure solve that enforces
incompressibility.
The projection is realized as:
\[
\Pi_{\mathcal{C}}(\mathcal{M}_t + \delta)
\;\approx\;
\arg\min_{Y \in \mathcal{C}}\;
\|Y - (\mathcal{M}_t + \delta)\|^2.
\]

\subsection{Explicit Constraint Equations}

The constraint set $\mathcal{C}$ is defined by six field-level invariants,
each with a precise mathematical form.

\paragraph{Mass and Occupancy Consistency.}
The occupancy density $\rho$ must satisfy a continuity equation across
updates:
\[
\int_\Omega \rho_{t+1}(x)\, dx
\;\approx\;
\int_\Omega \rho_t(x)\, dx
+ \text{sources}(a_t)
- \text{sinks}(a_t).
\]
This prevents mass from appearing or disappearing without a corresponding
action.

\paragraph{Energy Budget.}
The total field energy is bounded by the work performed by the agent's
action and dissipated by entropy:
\[
E_{t+1} \;\le\; E_t + W(a_t) - \mathcal{D}(S_t),
\]
where $W(a_t)$ is the work budget of action $a_t$ and $\mathcal{D}(S_t)$
is the entropy-driven dissipation.

\paragraph{Topological Admissibility.}
Edits to the substrate $\Phi_t$ must preserve or change topology only via
admissible operations---cut, join, fracture---with topological validity
monitored through Betti number changes or a learned admissibility
classifier that vetoes proposals introducing illegal connectivity changes.

\paragraph{Causal Write Locality.}
The support of $\mathcal{N}_\psi$ must be spatially localized around the
causal influence cone of the action.
This is enforced by a kernel $K_\psi$ that decays with geodesic distance
from the action site in $X$:
\[
\mathcal{N}_\psi(\mathcal{M}_t, p_t, a_t, o_{t+1})(x)
\;\propto\;
K_\psi\!\left(d_X(x, \operatorname{support}(a_t))\right).
\]

\paragraph{Entropy Accounting.}
Entropy $S_t$ increases under occlusion and uncertainty, and decreases
under observation.
Write-backs must not reduce entropy outside the agent's observation cone:
\[
\Delta S_t(x) \ge 0 \quad \text{for all } x \notin \operatorname{view}(p_t).
\]
This enforces the second law at the field level and prevents information
from appearing in unobserved regions.

\paragraph{Identity Persistence.}
Object embeddings in $\mathcal{M}_t$ carry persistent identity labels.
Updates must be consistent with identity trajectories: spontaneous
duplication or destruction of objects is forbidden unless permitted by
the mass and energy budgets of the action.

\subsection{Composable Projections}

The projection $\Pi_{\mathcal{C}}$ need not be implemented as a monolithic
solver.
Differentiable projections handle occupancy and continuity via Poisson-like
solves, SDF clamping, and angle wrapping.
Lagrange multiplier terms in a local optimization adjust proposals to
satisfy energy budgets.
Learned validator networks score admissibility and reject or correct
proposals that fail.
Typed discrete operations with preconditions and postconditions govern
topology changes such as breaking a wall or opening a door.
Each projection component is composable: they are applied in sequence, and
the final committed state satisfies all constraints simultaneously.

% =============================================================
\section{Fast and Slow Dynamics, Event Operators, and Read-Write Closure}
\label{sec:fastslow}

\subsection{Fast and Slow Field Update Cadences}

Not all components of $\mathcal{M}_t$ should be updated at the same
temporal cadence.
The latent appearance field $z_t$ and short-range deformations of
$\mathbf{v}_t$ are updated at every step by the neural proposal
$\mathcal{N}_\psi$, as these fields must respond to high-frequency
perceptual change.
The scalar substrate $\Phi_t$, topological structure, affordances $A_t$,
and object identities are updated only via typed operations that pass
through the full constraint projection.
This two-cadence design prevents noisy neural proposals from catastrophically
editing structural fields while keeping the rendering layer responsive.

\subsection{Event-Typed Update Operators}

The dynamics operator $\Delta_\psi$ is decomposed as a mixture of
event-typed operators:
\[
\Delta_\psi
=
\sum_k \alpha_k\, \mathcal{E}_k(\mathcal{M}_t, p_t, a_t),
\]
where each $\mathcal{E}_k$ is a typed operator with known preconditions
and field effects, and $\alpha_k$ are learned mixture weights.
Examples of typed operators include displacement (translating an object
along a trajectory in $X$), fracture (splitting a connected component of
$\Phi_t$ into two, subject to energy budget), occlusion write (updating
$S_t$ and $z_t$ in regions newly blocked from the observation cone), and
residue deposit (incrementing $R_t$ at the action site).
The typed decomposition provides a scaffold that constrains the neural
proposal to a structured vocabulary of world-edits, sharply reducing the
surface area over which drift can accumulate.

\subsection{Read-Write Closure}

To prevent divergence between what is observed and what is stored, the
system must satisfy a closure condition relating the read operator
$\operatorname{Read}(\mathcal{M}_t, p_t)$ to the write-back induced by
$\mathcal{W}$.

The read operator must be Lipschitz in $\mathcal{M}_t$ locally: small
changes to the field must not produce discontinuous changes in the
rendered observation.
This ensures that the observation module $\mathcal{O}_\theta$ is
stable under incremental field updates.

The write-back must be consistent with read in the following cycle-closure
sense.
Given an observation $o_{t+1}$ produced by $\mathcal{O}_\theta$, the
encoding of that observation back into field space must be compatible with
the stored field:
\[
\mathcal{M}_{t+1}
\;\xrightarrow{\;\mathcal{O}_\theta\;}
\hat{o}_{t+1}
\;\xrightarrow{\;\mathrm{Enc}\;}
\widetilde{\mathcal{M}}_{t+1},
\]
where $\widetilde{\mathcal{M}}_{t+1}$ and $\mathcal{M}_{t+1}$ should
agree in the observed region.
This cycle loss is incorporated as a training objective to prevent the
field from storing representations that the observation module cannot
decode, and to prevent the observation module from generating content
that the field cannot absorb.

% =============================================================
\section{Multiplayer as Causal Field Synchronization and Training}
\label{sec:multiplayer}

\subsection{Multiplayer as Causal DAG Merging}

In the RSVP framework, multiplayer interaction is not a state-replication
problem but a causal field synchronization problem.
All agents operate over a shared field $\mathcal{M}_t$; each agent $i$
maintains a local pose $p_t^i$ and produces local observation
$o_{t+1}^i = \mathcal{O}_\theta(\mathcal{M}_t, p_t^i, h_t^i, a_t^i)$.
When agent $i$ takes action $a_t^i$, the corresponding neural proposal
$\mathcal{N}_\psi$ generates a local field delta, which after projection
is committed to $\mathcal{M}_{t+1}$.
All agents' subsequent reads are constrained by this updated field.

To handle concurrent actions from multiple agents, the system maintains a
causal directed acyclic graph of field-editing events, each stamped with
a logical timestamp and the identity of the agent that produced it.
Commutative field edits---residue deposits, appearance updates, entropy
writes in non-overlapping regions---are merged via CRDT-style operations
that are order-independent.
Non-commutative edits, such as two agents simultaneously attempting to
destroy the same structural component, are resolved through the energy
budget constraint: the action with sufficient energy at the time of
application succeeds; the other is vetoed by $\Pi_{\mathcal{C}}$ and
generates a null delta.
Because all agents project into the same admissible manifold $\mathcal{C}$,
divergent local proposals collapse to the same valid committed state,
and consistency is guaranteed by constraint structure rather than
consensus protocol.

\subsection{Training Curriculum}

Training the full RSVP update system proceeds in four phases.
In the first phase, the observation operator $\mathcal{O}_\theta$ is
pretrained as a renderer conditioned on $\mathcal{M}_t$, with the field
held fixed at ground-truth values.
In the second phase, the neural proposal network $\mathcal{N}_\psi$ is
pretrained on synthetic event logs with known ground-truth field deltas,
supervised by the typed operator structure.
In the third phase, the constraint projection $\Pi_{\mathcal{C}}$ is
introduced with soft penalties that gradually anneal toward hard
projections, so the system learns to generate proposals that are already
close to admissible before projection.
In the fourth phase, the full system is trained end-to-end using
self-consistency losses: when ground-truth field deltas are unavailable,
the cycle-closure condition between $\mathcal{O}_\theta$ and
$\operatorname{Enc}$ provides the supervision signal, and conservation
penalties enforce the field-level invariants of $\mathcal{C}$.
This curriculum ensures that the system acquires local expressivity from
the neural proposal, global coherence from the constraint projection, and
long-horizon persistence from the enforced invariants, in that order.

% =============================================================
\section{Computational Instantiation}
\label{sec:prototype}

The preceding formalism admits a direct computational realization.
To demonstrate that the proposed architecture is not merely interpretive
but executable, we construct a minimal prototype that instantiates the
separation between substrate, dynamics, and observation.

The implementation is deliberately restricted to the smallest system that
preserves the essential asymmetry: observation is stochastic and local,
while memory is persistent and global.
The system consists of a client-side rendering layer and a server-side
field evolution layer.
The former implements the observation operator $\mathcal{O}_\theta$, while
the latter implements a constrained update $\mathcal{T}$ together with an
explicit write channel $\mathcal{W}$.

The global state is represented as a discretized field
\[
X_t = (\Phi_t,\, S_t,\, \mathcal{R}_t),
\]
where $\Phi_t$ encodes scalar structure, $S_t$ encodes uncertainty, and
$\mathcal{R}_t$ accumulates residue.
This field is stored persistently on the server and evolves independently
of any particular observation.

The observation operator $\mathcal{O}_\theta$ is realized in the client
as a projection from field to pixel space.
Crucially, this projection is stochastic: each render introduces noise
proportional to the local entropy $S_t$.
In the absence of a write-back mechanism, this would produce temporally
incoherent outputs, analogous to diffusion-based frame prediction.

The write operator $\mathcal{W}$ is triggered by user interaction.
Localized perturbations are injected into $\mathcal{R}_t$ and
simultaneously reduce $S_t$, thereby promoting observed structure into
persistent memory.
This enforces the invention-to-memory principle of Section~7: generated
detail becomes constraint upon realization.

The dynamics operator $\mathcal{T}$ is implemented as a diffusion-like
update on $\Phi_t$ coupled to $\mathcal{R}_t$, together with a monotone
decay of entropy.
This realizes a minimal form of constraint propagation and stabilization,
corresponding to the Laplacian term in the RSVP Lagrangian of Section~3.

\subsection*{Client: Observation and Interaction}

\begin{lstlisting}[language=HTML]
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<title>RSVP Field Prototype</title>
<style>
body { margin:0; background:#000; overflow:hidden; }
canvas { display:block; }
</style>
</head>
<body>
<canvas id="c"></canvas>
<script>
const canvas = document.getElementById("c");
const ctx = canvas.getContext("2d");
canvas.width = window.innerWidth;
canvas.height = window.innerHeight;

const W = 128, H = 128;

// persistent field: phi (structure), entropy, residue
let field = {
  phi:     new Float32Array(W*H),
  entropy: new Float32Array(W*H),
  residue: new Float32Array(W*H)
};
for (let i=0; i<W*H; i++){
  field.phi[i]     = Math.random()*0.1;
  field.entropy[i] = 1.0;
}

function idx(x, y){ return y*W + x; }

// observation operator: stochastic projection from field to pixels
function render(){
  let img = ctx.createImageData(canvas.width, canvas.height);
  for (let y=0; y<canvas.height; y++){
    for (let x=0; x<canvas.width; x++){
      let gx = Math.floor(x/canvas.width*W);
      let gy = Math.floor(y/canvas.height*H);
      let i  = idx(gx, gy);

      let base  = field.phi[i]*255 + field.residue[i]*200;
      let noise = Math.random() * field.entropy[i] * 50;  // variance ~ S_t
      let c     = base + noise;

      let p = (y*canvas.width + x)*4;
      img.data[p]   = c;
      img.data[p+1] = c*0.7;
      img.data[p+2] = 255-c;
      img.data[p+3] = 255;
    }
  }
  ctx.putImageData(img, 0, 0);
}

// write operator: interaction injects residue, collapses local entropy
canvas.addEventListener("mousemove", (e)=>{
  let x = Math.floor(e.clientX/canvas.width*W);
  let y = Math.floor(e.clientY/canvas.height*H);
  for (let dy=-2; dy<=2; dy++){
    for (let dx=-2; dx<=2; dx++){
      let nx=x+dx, ny=y+dy;
      if (nx>=0 && nx<W && ny>=0 && ny<H){
        let i = idx(nx, ny);
        field.residue[i] += 0.1;   // promote residue
        field.entropy[i] *= 0.8;   // reduce uncertainty
      }
    }
  }
});

// dynamics operator: diffusion on phi coupled to residue
function step(){
  let newPhi = new Float32Array(W*H);
  for (let y=1; y<H-1; y++){
    for (let x=1; x<W-1; x++){
      let i   = idx(x,y);
      let lap =
        field.phi[idx(x+1,y)] + field.phi[idx(x-1,y)] +
        field.phi[idx(x,y+1)] + field.phi[idx(x,y-1)] -
        4*field.phi[i];

      newPhi[i]         = field.phi[i] + 0.1*lap + 0.05*field.residue[i];
      field.entropy[i] *= 0.999;   // monotone decay outside view
    }
  }
  field.phi = newPhi;
}

function loop(){ step(); render(); requestAnimationFrame(loop); }
loop();
</script>
</body>
</html>
\end{lstlisting}

\subsection*{Server: Persistent Field and Evolution}

\begin{lstlisting}[language=Python]
import numpy as np
from flask import Flask, jsonify, request

app = Flask(__name__)
W, H = 128, 128

# persistent field exists independently of any rendering pass
field = {
    "phi":     np.random.rand(H, W) * 0.1,
    "entropy": np.ones((H, W)),
    "residue": np.zeros((H, W))
}

# dynamics operator: constrained Laplacian update + residue coupling
def step():
    phi     = field["phi"]
    residue = field["residue"]
    lap = (
        np.roll(phi, 1, 0) + np.roll(phi,-1, 0) +
        np.roll(phi, 1, 1) + np.roll(phi,-1, 1) - 4*phi
    )
    field["phi"]     = phi + 0.1*lap + 0.05*residue
    field["entropy"] *= 0.999   # entropy monotone outside observed cone

@app.route("/step")
def api_step():
    step()
    return jsonify({"ok": True})

@app.route("/field")
def get_field():
    return jsonify({
        "phi":     field["phi"].tolist(),
        "residue": field["residue"].tolist(),
        "entropy": field["entropy"].tolist()
    })

# write operator: localized residue injection with entropy collapse
@app.route("/write", methods=["POST"])
def write():
    x, y = request.json["x"], request.json["y"]
    for dy in range(-2, 3):
        for dx in range(-2, 3):
            nx, ny = x+dx, y+dy
            if 0 <= nx < W and 0 <= ny < H:
                field["residue"][ny, nx] += 0.2
                field["entropy"][ny, nx] *= 0.7
    return jsonify({"ok": True})

if __name__ == "__main__":
    app.run(port=5000)
\end{lstlisting}

\subsection*{Interpretation}

This prototype realizes the minimal closure:
\[
\mathcal{O}_\theta
\;\longrightarrow\;
\mathcal{W}
\;\longrightarrow\;
\mathcal{T}
\;\longrightarrow\;
\mathcal{O}_\theta,
\]
in which observation produces realizations, realizations are written back
as residue, and the field evolves under those accumulated constraints.

The essential property is that stochastic variation in the observation
layer does not produce temporal incoherence, because the field retains the
history of prior realizations through $\mathcal{R}_t$.
The system does not remember via a buffer; it remembers via deformation of
the substrate itself.
This is precisely the structural feature absent in purely autoregressive
or diffusion-based generative systems, and it constitutes the minimal
condition for persistent, causally coherent world generation.

\subsection*{Discrete Constrained Field Flow}

The prototype admits a direct interpretation as a discrete approximation
to the constrained field evolution developed in Section~8.
Writing the state as $X_t = (\Phi_t, S_t, \mathcal{R}_t)$, the server
update implements an explicit Euler step:
\[
X_{t+1}^{\mathrm{prop}} = X_t + \delta_t,
\]
where the proposal $\delta_t$ is given by a local Laplacian term on
$\Phi_t$ coupled to $\mathcal{R}_t$, together with a monotone decay of
$S_t$.

The absence of an explicit $\Pi_{\mathcal{C}}$ in the code reflects the
choice of a minimal constraint set enforced implicitly.
The locality of the Laplacian update enforces causal support; the monotone
decay of $S_t$ via the factor $0.999$ at each step prevents spontaneous
reduction of uncertainty outside regions of interaction, realizing the
entropy accounting constraint of Section~\ref{sec:factored}.
The write operator $\mathcal{W}$ introduces localized, entropy-reducing
perturbations that act as admissible sources for structure.
Thus the full update may be written schematically as:
\[
X_{t+1} = \Pi_{\mathcal{C}}\!\big(X_t + \mathcal{N}_\psi(X_t, a_t)\big),
\]
with $\mathcal{N}_\psi$ realized by the discrete diffusion--residue
coupling, and $\Pi_{\mathcal{C}}$ realized implicitly by the restricted
form of admissible updates.

The observation operator $\mathcal{O}_\theta$ is a stochastic projection
from field space to pixel space with variance proportional to $S_t$.
In the absence of write-back, this yields temporally uncorrelated samples.
The inclusion of $\mathcal{R}_t$ closes the loop by promoting realized
structure into persistent constraint, converting stochastic observation
into deterministic evolution at the level of the field.

The prototype therefore constitutes a discrete, low-dimensional instance
of a projected gradient flow on the space of admissible field
configurations, in which observation acts as a probe of underdetermined
degrees of freedom and writing acts as a mechanism of constraint
accumulation.
It is not a toy in the usual sense but a collapsed limit of the general
RSVP--Polyxan theory: the system already inhabits the same category of
constrained dynamical systems as the full formalism, with the richness of
$\mathcal{M}_t$ replaced by the three-component field $({\Phi}_t, S_t,
\mathcal{R}_t)$ sufficient to exhibit the core asymmetry.

% =============================================================
\section{Conclusion}

We have extended RSVP--Polyxan from a static semantic field theory into a
fully dynamical world model defined by persistent latent memory,
projection-based observation, and constrained field evolution, and then
elevated that dynamical model through three ascending layers of mathematical
structure.

At the classical layer, the constraint projection $\Pi_{\mathcal{C}}$ is
internalized into an augmented action functional with Lagrange multipliers,
yielding constrained Euler--Lagrange equations in which admissibility is
enforced continuously rather than imposed as a post-hoc correction.
Noether's theorem applied to the augmented action produces four classes of
semantic conservation law: coherence conservation, flow conservation,
entropy balance, and residue persistence.
The system is then reformulated in Hamiltonian phase-space terms, and
symplectic preservation is established for the unforced dynamics.

At the stochastic layer, Wick rotation of the path integral yields a
stochastic diffusion process whose invariant measure is the Boltzmann
distribution over field configurations.
Diffusion generative models emerge as discrete approximations to stochastic
quantization of RSVP dynamics, with the RSVP energy functional playing the
role of the score function.

At the derived layer, the Polyxan category is lifted to an $\infty$-category
$\mathbf{C}_\infty$, the sheaf of galaxy renderings is refined to a derived
sheaf valued in $\infty$-groupoids, and semantic consistency is characterized
as homotopy-coherent gluing.
Hallucination is identified with non-vanishing first cohomology
$H^1(\mathscr{G}^\infty) \neq 0$, providing a cohomological criterion for
semantic failure that is intrinsic to the field-theoretic structure.

The relationship to MultiGen is established at the level of a formal
projection theorem.
MultiGen arises from RSVP--Polyxan by collapsing the full latent field to
a static scalar substrate, removing the affordance, residue, flow, and
appearance components, and eliminating the write operator $\mathcal{W}$.
The resulting one-way system---constraint flowing toward projection but
projection never flowing back toward constraint---achieves spatial
consistency but not causal consistency.
The RSVP completion adds the missing feedback channel: the factored update
$\Pi_{\mathcal{C}} \circ \mathcal{N}_\psi$ promotes realized observations
into the persistent field via six explicitly formulated field-level
invariants.
Multiplayer interaction becomes causal field synchronization over a shared
substrate, with concurrent edits resolved by the constraint projection
rather than by consensus protocol.

The resulting system constitutes a unified framework in which continuous
fields, discrete hypergraphs, and generative operators interact within a
single coherent structure governed by variational principles, conserved
under symmetry, and interpretable at every level from Hamiltonian mechanics
to derived algebraic geometry.
A generative system is not a model that produces worlds, but a
gauge-invariant, stochastic, Hamiltonian field theory over a persistent
semantic substrate, whose observable realizations arise as samples from a
path integral, whose consistency is governed by homotopy-coherent gluing
conditions, and whose dynamics are stabilized by explicit field-level
conservation laws enforced through constrained neural projection.
That this architecture is constructively realizable, and not merely
formally consistent, is demonstrated by the prototype of
Section~\ref{sec:prototype}, which instantiates the core closure
$\mathcal{O}_\theta \to \mathcal{W} \to \mathcal{T} \to \mathcal{O}_\theta$
in executable form and exhibits the field's capacity to accumulate
constraint from observation rather than discarding it.

\newpage
\begin{thebibliography}{99}

\bibitem{po2026multigen}
R.~Po, D.~J.~Zhang, A.~Hertz, G.~Wetzstein, N.~Wadhwa, and N.~Ruiz,
\newblock \emph{MultiGen: Level-Design for Editable Multiplayer Worlds in
  Diffusion Game Engines},
\newblock arXiv:2603.06679 (2026).

\bibitem{ho2020ddpm}
J.~Ho, A.~Jain, and P.~Abbeel,
\newblock \emph{Denoising Diffusion Probabilistic Models},
\newblock Advances in Neural Information Processing Systems (2020).

\bibitem{song2021score}
Y.~Song, J.~Sohl-Dickstein, D.~P.~Kingma, A.~Kumar, S.~Ermon, and B.~Poole,
\newblock \emph{Score-Based Generative Modeling through Stochastic Differential
  Equations},
\newblock International Conference on Learning Representations (2021).

\bibitem{anderson1984}
P.~W.~Anderson,
\newblock \emph{Basic Notions of Condensed Matter Physics},
\newblock Addison-Wesley (1984).

\bibitem{landauer1961}
R.~Landauer,
\newblock \emph{Irreversibility and Heat Generation in the Computing Process},
\newblock IBM Journal of Research and Development, 5(3), 183--191 (1961).

\bibitem{bennett1982}
C.~H.~Bennett,
\newblock \emph{The Thermodynamics of Computation---A Review},
\newblock International Journal of Theoretical Physics, 21(12), 905--940
  (1982).

\bibitem{lloyd2000}
S.~Lloyd,
\newblock \emph{Ultimate Physical Limits to Computation},
\newblock Nature, 406, 1047--1054 (2000).

\bibitem{maclane1998}
S.~Mac~Lane,
\newblock \emph{Categories for the Working Mathematician},
\newblock Springer (1998).

\bibitem{lurie2009}
J.~Lurie,
\newblock \emph{Higher Topos Theory},
\newblock Princeton University Press (2009).

\bibitem{hartshorne1977}
R.~Hartshorne,
\newblock \emph{Algebraic Geometry},
\newblock Springer (1977).

\bibitem{kashiwara1990}
M.~Kashiwara and P.~Schapira,
\newblock \emph{Sheaves on Manifolds},
\newblock Springer (1990).

\bibitem{baez2011}
J.~C.~Baez and J.~P.~Master,
\newblock \emph{Open Petri Nets},
\newblock Mathematical Structures in Computer Science (2011).

\bibitem{jacobson1995}
T.~Jacobson,
\newblock \emph{Thermodynamics of Spacetime: The Einstein Equation of State},
\newblock Physical Review Letters, 75(7), 1260--1263 (1995).

\bibitem{verlinde2011}
E.~Verlinde,
\newblock \emph{On the Origin of Gravity and the Laws of Newton},
\newblock Journal of High Energy Physics (2011).

\bibitem{ha2018worldmodels}
D.~Ha and J.~Schmidhuber,
\newblock \emph{World Models},
\newblock arXiv:1803.10122 (2018).

\bibitem{hafner2020dream}
D.~Hafner, T.~Lillicrap, J.~Ba, and M.~Norouzi,
\newblock \emph{Dream to Control: Learning Behaviors by Latent Imagination},
\newblock International Conference on Learning Representations (2020).

\bibitem{hafner2024worldmodels}
D.~Hafner et al.,
\newblock \emph{Mastering Diverse Domains through World Models},
\newblock arXiv (2024).

\bibitem{brooks2024video}
T.~Brooks et al.,
\newblock \emph{Video Generation Models as World Simulators},
\newblock OpenAI Research (2024).

\bibitem{haohen2024lumiere}
O.~Bar-Tal et al.,
\newblock \emph{Lumiere: A Space-Time Diffusion Model for Video Generation},
\newblock SIGGRAPH Asia (2024).

\bibitem{chen2024diffusionforcing}
B.~Chen et al.,
\newblock \emph{Diffusion Forcing: Next-Token Prediction Meets Full-Sequence
  Diffusion},
\newblock arXiv (2024).

\bibitem{huang2025selfforcing}
X.~Huang et al.,
\newblock \emph{Self-Forcing: Bridging the Train-Test Gap in Autoregressive
  Video Diffusion},
\newblock arXiv (2025).

\bibitem{guzdial2021pcg}
M.~Guzdial and M.~Riedl,
\newblock \emph{Game Level Generation from Gameplay Videos},
\newblock AIIDE (2021).

\bibitem{kim2020gamegan}
S.~W.~Kim et al.,
\newblock \emph{Learning to Simulate Dynamic Environments with GameGAN},
\newblock CVPR (2020).

\bibitem{nasir2023llmgames}
M.~U.~Nasir and J.~Togelius,
\newblock \emph{Practical PCG through Large Language Models},
\newblock IEEE Conference on Games (2023).

\bibitem{anand2024coherent}
A.~Anand et al.,
\newblock \emph{Coherent Multi-Agent Simulation with Shared Latent States},
\newblock arXiv (2024).

\bibitem{wadhwa2025genie}
N.~Wadhwa et al.,
\newblock \emph{Genie: Generative Interactive Environments},
\newblock arXiv (2025).

\bibitem{hong2025relic}
Y.~Hong et al.,
\newblock \emph{Relic: Interactive Video World Model with Long-Horizon Memory},
\newblock arXiv (2025).

\bibitem{deng2024streetscapes}
B.~Deng et al.,
\newblock \emph{Streetscapes: Large-Scale Consistent Street View Generation},
\newblock SIGGRAPH (2024).

\bibitem{kauffman1993origins}
S.~Kauffman,
\newblock \emph{The Origins of Order},
\newblock Oxford University Press (1993).

\bibitem{odum1988self}
H.~T.~Odum,
\newblock \emph{Self-Organization, Transformity, and Information},
\newblock Science (1988).

\bibitem{georgescu1971entropy}
N.~Georgescu-Roegen,
\newblock \emph{The Entropy Law and the Economic Process},
\newblock Harvard University Press (1971).

\bibitem{ayres1998industrial}
R.~U.~Ayres,
\newblock \emph{Industrial Ecology and the Flow of Materials},
\newblock Edward Elgar (1998).

\bibitem{daly1996beyondgrowth}
H.~E.~Daly,
\newblock \emph{Beyond Growth},
\newblock Beacon Press (1996).

\end{thebibliography}

\end{document}
