% ============================================================
% MASTER LATEX FILE FOR:
% Delta--Sigma Modulators and Data Converters:
% A Field-Theoretic Approach to Quantization, Noise Shaping, and Information Flow
% ============================================================

\documentclass[12pt,oneside]{book}

% ------------------------------------------------------------
%   ENCODING AND FONTS
% ------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% ------------------------------------------------------------
%   GEOMETRY AND LAYOUT
% ------------------------------------------------------------
\usepackage{geometry}
\geometry{
    paperwidth=8.5in,
    paperheight=11in,
    margin=1in,
    headheight=14pt
}

\usepackage{setspace}
\setstretch{1.15}

\usepackage{titlesec}
\titleformat{\chapter}{\Huge\bfseries}{\thechapter}{1em}{}

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[LO]{\nouppercase{\rightmark}}
\fancyhead[RE]{\nouppercase{\leftmark}}

% ------------------------------------------------------------
%   MATH PACKAGES
% ------------------------------------------------------------
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\usepackage{mathtools}
\usepackage{physics}
\usepackage{bm}

% Theorem environments
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\newtheorem{corollary}{Corollary}[chapter]

% ------------------------------------------------------------
%   HYPERREF
% ------------------------------------------------------------
\usepackage{xurl}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

% ------------------------------------------------------------
%   GRAPHICS
% ------------------------------------------------------------
\usepackage{graphicx}
\usepackage{float}

% ------------------------------------------------------------
%   BIBLIOGRAPHY
% ------------------------------------------------------------
\usepackage[
    backend=biber,
    style=authoryear,
    maxcitenames=2,
    maxbibnames=10,
    url=true,
    doi=true
]{biblatex}

\addbibresource{references.bib}

% ------------------------------------------------------------
%   TITLE AND FRONTMATTER
% ------------------------------------------------------------

\title{
\Huge Delta--Sigma Modulators and Data Converters:\\[6pt]
\Large A Field-Theoretic Approach to Quantization, Noise Shaping, and Information Flow
}

\author{Flyxion}
\date{\today}

\begin{document}

\frontmatter
\maketitle
\tableofcontents

% ------------------------------------------------------------
%   MAIN CONTENT
% ------------------------------------------------------------
\mainmatter

% ============================================================
% PART I
% ============================================================

\part{Foundations}

\chapter*{Introduction to Part I}
\addcontentsline{toc}{chapter}{Introduction to Part I}

Part I establishes the conceptual foundations upon which the remainder of this book is constructed. It begins by placing delta--sigma modulation within its historical and technological context, tracing the evolution of converter architectures from early sampling schemes to modern oversampling data converters. The central aim is to reveal that the classical viewpoint of quantization as a static nonlinearity is insufficient to describe the deeper structure underlying high-resolution conversion. By examining signals as fields, quantization as a dynamic projection, and feedback as an entropy-routing mechanism, Part I introduces the geometric and field-theoretic language that will guide the rest of the work.

\input{chapters/ch1}
\input{chapters/ch2}
\input{chapters/ch3}

% ============================================================
% PART II
% ============================================================

\part{RSVP-Inspired Reformulation}

\chapter*{Introduction to Part II}
\addcontentsline{toc}{chapter}{Introduction to Part II}

Part II presents the reformulation of delta--sigma modulation in the language of the RSVP field theory. Entropy fields, vector flows, and scalar potentials are used to reinterpret modulators’ internal dynamics. The feedback loop becomes an entropy-shaping engine, the integrator becomes a vector field guiding the descent of a potential, and the quantizer emerges as a derived geometric projection that collapses continuous trajectories onto discrete symbolic strata. This part reorients the reader toward a unified geometric interpretation that connects engineering intuition with modern mathematical structure.

\input{chapters/ch4}
\input{chapters/ch5}
\input{chapters/ch6}

% ============================================================
% PART III
% ============================================================

\part{Engineering the Modulator}

\chapter*{Introduction to Part III}
\addcontentsline{toc}{chapter}{Introduction to Part III}

Part III applies the geometric and field-theoretic concepts of earlier chapters to explicit engineering design. First- and higher-order modulators are examined in depth, and the quantizer is treated as a nonlinear projection with a derived boundary structure. Practical architectures arise from the interaction of entropy descent, symbolic projection, and structured vector flows. This part grounds the theory in realizable designs while demonstrating how manifold geometry dictates practical converter behavior.

\input{chapters/ch7}
\input{chapters/ch8}
\input{chapters/ch9}
\input{chapters/ch10}

% ============================================================
% PART IV
% ============================================================

\part{Advanced Themes}

\chapter*{Introduction to Part IV}
\addcontentsline{toc}{chapter}{Introduction to Part IV}

Part IV extends delta--sigma modulation into nonlinear, geometric, and inferential domains. Stability is expressed through entropy-balance criteria, noise shaping becomes a PDE on an evolving manifold, and information geometry reveals the modulators as inference engines. Nonlinear and chaotic regimes emerge naturally from curved manifolds and symbolic collapse. This part expands delta--sigma from an engineering method into a general mathematical paradigm.

\input{chapters/ch11}
\input{chapters/ch12}
\input{chapters/ch13}
\input{chapters/ch14}
\input{chapters/ch15}

% ============================================================
% PART V
% ============================================================

\part{Applications and Design Blueprints}

\chapter*{Introduction to Part V}
\addcontentsline{toc}{chapter}{Introduction to Part V}

Part V applies the theoretical machinery to real-world systems such as audio converters, precision metrology, switching power electronics, and embedded implementations. It shows how entropy shaping interacts with psychoacoustic masking, low-frequency drift, switching behaviors, and hardware constraints. Through these examples the field-theoretic framework becomes a practical design guide across multiple domains.

\input{chapters/ch16}
\input{chapters/ch17}
\input{chapters/ch18}
\input{chapters/ch19}

% ============================================================
% PART VI
% ============================================================

\part{Unification and Future Directions}

\chapter*{Introduction to Part VI}
\addcontentsline{toc}{chapter}{Introduction to Part VI}

Part VI unifies quantization, sampling, geometry, and field theory by interpreting delta--sigma modulators as thermodynamic controllers that convert continuous fields into discrete symbols. Sampling becomes a derived mapping problem, and quantization becomes a geometric projection. This part provides the conceptual core of the book’s synthesis.

\input{chapters/ch20}
\input{chapters/ch21}
\input{chapters/ch22}

% ============================================================
% PART VII
% ============================================================

\part{Delta--Sigma as Generalized Inference}

\chapter*{Introduction to Part VII}
\addcontentsline{toc}{chapter}{Introduction to Part VII}

Part VII develops the insight that delta--sigma modulators are specific instances of active inference systems. Prediction, correction, and symbolic collapse are governed by variational principles, and the modulator becomes a belief-updating mechanism operating on an error manifold. Quantization becomes symbolic perception, and loop filtering becomes cognitive propagation.

\input{chapters/ch23}
\input{chapters/ch24}
\input{chapters/ch25}
\input{chapters/ch26}
\input{chapters/ch27}

% ============================================================
% PART VIII
% ============================================================

\part{General Physics of Measurement and Feedback}

\chapter*{Introduction to Part VIII}
\addcontentsline{toc}{chapter}{Introduction to Part VIII}

Part VIII situates delta--sigma modulation within the physics of measurement. Scalar, vector, and entropy fields govern the loop dynamics exactly as they do in natural sensing processes. Quantizers act as entropy-sculpting operators, and modulators behave as thermodynamic engines that regulate information flow. Measurement emerges as an active geometric interaction with a field.

\input{chapters/ch28}
\input{chapters/ch29}
\input{chapters/ch30}
\input{chapters/ch31}
\input{chapters/ch32}

% ============================================================
% PART IX
% ============================================================

\part{Beyond Engineering: ΔΣ in Natural and Artificial Systems}

\chapter*{Introduction to Part IX}
\addcontentsline{toc}{chapter}{Introduction to Part IX}

Part IX explores how the invariants of delta--sigma modulation appear across biological, cognitive, and artificial systems. Neural coding, perception, symbolic learning, and adaptive control all exhibit the structural motifs of prediction, correction, and entropy routing. These similarities indicate that delta--sigma modulators embody a universal architectural principle governing coherent behavior.

\input{chapters/ch33}
\input{chapters/ch34}
\input{chapters/ch35}
\input{chapters/ch36}
\input{chapters/ch37}
\input{chapters/ch38}
\input{chapters/ch39}
\input{chapters/ch40}

% ------------------------------------------------------------
% APPENDICES
% ------------------------------------------------------------

\appendix

\input{appendices/appA}
\input{appendices/appB}
\input{appendices/appC}
\input{appendices/appD}
\input{appendices/appE}

% ------------------------------------------------------------
% BIBLIOGRAPHY
% ------------------------------------------------------------

\backmatter
\printbibliography

\end{document}

