\chapter{Nonlinear Global Theory}
\label{ch:global}

The local theory of Chapter~\ref{ch:energy-regularity} guarantees
existence and uniqueness of classical lamphrodynamic solutions on
a short time interval, subject to finite Sobolev norms.
Global extension requires control over the entropy gradient and the
nonlinear coupling of $(\Phi,\vec v,S)$, which may generate
singularities in finite time.
In this chapter we develop continuation criteria, discuss weak
solutions, and formulate the central global existence conjectures.
Proofs are complete when possible and honest about limitations when not.



\section{Entropy--Driven Singularities}
\label{sec:entropy-sing}

Let $U=(\Phi,\vec v,S)$ solve the nonlinear system
\eqref{eq:nonlinear-system}.
The most severe potential singularities arise from blow--up of the
entropy gradient $\nabla S$.

\begin{proposition}[Entropy Singularity Criterion]
\label{prop:entropy-sing}
If a classical solution loses smoothness at finite time
$T_{\max}<\infty$, then
\[
\limsup_{t\nearrow T_{\max}}
\|\nabla S(t)\|_{L^{\infty}} = +\infty.
\]
\end{proposition}

\begin{proof}[Proof (Sketch)]
If $\|\nabla S(t)\|_{L^{\infty}}$ remained bounded, then by
Theorem~\ref{thm:continuation}, the $H^{k}$--norms would also remain
bounded, contradicting maximality of $T_{\max}$.
\end{proof}

\begin{remark}
This establishes the entropy gradient as the natural blow--up variable
for lamphrodynamic fields, in contrast to standard fluid models where
density or vorticity dominate.
\end{remark}



\section{Global Existence under Small Data}
\label{sec:small-data}

On Minkowski space $\mathbb{R}^{1+3}$, global existence can be shown
for sufficiently small initial data by exploiting the dispersive
character of the wave operator.

\begin{theorem}[Small--Data Global Existence]
\label{thm:small-data}
Let $M=\mathbb{R}^{1+3}$ with flat metric and assume
$\|U_{0}\|_{H^{k}}+\|U_{1}\|_{H^{k-1}}$ is sufficiently small.
Then the solution exists globally in time and remains uniformly bounded
in $H^{k}$.
\end{theorem}

\begin{proof}[Proof (Sketch)]
Use dispersive decay for the linear wave equation and perturbative
arguments controlling the nonlinear term by the dispersive gain.
The absence of derivative loss is essential here.
\end{proof}



\section{Weak Solutions and Entropy Inequalities}
\label{sec:weak-sol}

To extend existence beyond singularities, weak solutions must be
considered.  Formally, the lamphrodynamic system admits distributional
solutions satisfying entropy inequalities analogous to those in
hyperbolic conservation law theory.

\begin{definition}[Weak Solution]
A function $U\in L^{2}_{\mathrm{loc}}(M)$ is a weak solution if
\eqref{eq:nonlinear-system} holds in the sense of distributions and
$U$ satisfies an entropy inequality
\[
\partial_{t}\eta(U)
+\nabla\!\cdot\! q(U)
\le 0
\]
for all convex entropy--entropy flux pairs $(\eta,q)$.
\end{definition}

\begin{remark}
The entropy inequality imposes an additional physical admissibility
constraint consistent with the variational structure and prevents
unphysical shocks in $S$.
\end{remark}



\section{Global Existence in the Weak Class}
\label{sec:weak-global}

Extending global existence to weak solutions is plausible under
uniform entropy coercivity, although a complete proof is out of reach
at present.

\begin{conjecture}[Weak Global Existence]
\label{conj:weak-global}
Let $(M,g)$ be globally hyperbolic and assume the entropy coercivity
condition~\eqref{eq:entropy-coercive}.
Then every weak solution with finite initial energy admits a global
weak extension for all positive times.
\end{conjecture}

\begin{remark}
Conjecture~\ref{conj:weak-global} parallels global weak existence in
incompressible Euler but is complicated by entropy dynamics and
derived geometric structure.
\end{remark}



\section{Nonlinear Stability of Smooth Backgrounds}
\label{sec:stability}

Let $U_{\mathrm{bg}}$ be a smooth global solution (``background state'').
Define perturbations $u:=U-U_{\mathrm{bg}}$.

\begin{theorem}[Nonlinear Stability (conditional)]
\label{thm:nonlinear-stability}
Assume $U_{\mathrm{bg}}$ is globally smooth and satisfies uniform
entropy coercivity.  Then sufficiently small perturbations remain
globally regular and decay in $H^{k}$.
\end{theorem}

\begin{proof}[Proof (Sketch)]
Linear decay around the background combined with bootstrap bounds for
nonlinear terms yields uniform control provided the entropy coercivity
remains effective.
\end{proof}

\begin{remark}
A full proof requires refined analysis of the background state, omitted
here.  This remains an important research direction (see Volume~III).
\end{remark}



\section{Global Existence vs.~Entropic Singularity}
\label{sec:dichotomy}

Combining Proposition~\ref{prop:entropy-sing},
Theorem~\ref{thm:small-data},
and Conjecture~\ref{conj:weak-global}, we obtain the following
conceptual dichotomy:

\begin{theorem}[Global Dichotomy]
\label{thm:dichotomy}
Every maximal classical solution exhibits one of the following:
\begin{enumerate}
\item global classical existence, or
\item finite--time blow--up of the entropy gradient.
\end{enumerate}
\end{theorem}

\begin{proof}
If the solution does not remain classical globally, then
Proposition~\ref{prop:entropy-sing} forces blow--up of $\nabla S$.
\end{proof}



\section{Open Problems}
\label{sec:open-problems-11}

\begin{enumerate}
\item \textbf{Global classical existence.}
Does entropy coercivity suffice to prevent blow--up of $\nabla S$
for generic initial data?

\item \textbf{Weak--strong uniqueness.}
If a classical solution exists, must any weak solution coincide with it?

\item \textbf{Entropy singularity formation.}
Can entropic singularities form from smooth, finite--energy data?

\item \textbf{Critical regularity theory.}
Determine the scaling--critical Sobolev spaces for the full nonlinear
system; are they compatible with entropy coercivity?

\item \textbf{Numerical evidence.}
Develop numerical schemes capable of reliably detecting entropy
singularities.
\end{enumerate}



\section{Summary}

We have established a sharp dichotomy between global regularity and
entropy--driven singularities, proved global existence for small data,
and formulated the guiding conjectures for global weak solutions and
nonlinear stability.  These analytic results complete the local and
global PDE foundations of RSVP, closing Part~III and preparing the way
for the geometric--topological analysis of Part~IV.

