% File: Volume-I-Foundations/main-vol1.tex

\documentclass[11pt]{book}

% Relative paths assume this file is in Volume-I-Foundations/
\input{../shared/macros.tex}

% ------------------------------------------------------------
% Bibliography setup (volume-specific)
% ------------------------------------------------------------
\usepackage[backend=biber,style=alphabetic,maxbibnames=10]{biblatex}
\addbibresource{bibliography-vol1.bib}

% ------------------------------------------------------------
% Title metadata
% ------------------------------------------------------------
\title{\rsvptitle:\\
Mathematical Foundations of Lamphrodynamic Field Theory}
\author{Flyxion}
\date{\today}

\begin{document}

\frontmatter
\maketitle

\chapter*{Preface}

This volume is the first of a three--volume monograph developing the
Relativistic Scalar--Vector Plenum (\rsvpabbr{}) as a rigorous mathematical
framework. Its primary audience is mathematicians and theoretical physicists
with a background in differential geometry, category theory, and quantum field
theory. A secondary aim is pedagogical: the material is written so that
physicists willing to learn derived geometry can follow, provided they are
comfortable with homological algebra and gauge theory.

The central goal of Volume~I is to construct the lamphrodynamic field theory
underlying \rsvpabbr{}, to formulate it in the language of derived algebraic
geometry and shifted symplectic structures, and to establish fundamental
analytic results (well--posedness, energy estimates, regularity). Later volumes
will build on this foundation to explore cosmology, emergent gravity, geometric
computation, and philosophical implications.

We have tried to adhere to the following design principles:
\begin{itemize}
  \item \emph{Rigor first in foundations.} The core constructions (derived
  stacks, shifted symplectic forms, AKSZ action) are presented with as much
  mathematical rigor as possible.

  \item \emph{Motivation before formalism.} Each chapter begins with an
  informal explanation of why the subsequent technical machinery is needed.

  \item \emph{Explicit status of claims.} Theorems are proved or referenced;
  conjectures and open problems are clearly labeled and accompanied by
  evidence or heuristic arguments.
\end{itemize}

Volume~I is structured in four parts. Part~I reviews the necessary background
in derived geometry and shifted symplectic structures, oriented toward
physicists. Part~II introduces the lamphrodynamic plenum and constructs the
relevant derived moduli stacks and AKSZ action. Part~III develops the analytic
foundations of the lamphrodynamic PDEs. Part~IV investigates geometric and
topological properties, including entropic singularities and symmetries.

Subsequent volumes will not repeat this material, but will refer back to it
frequently.

\vspace{1em}
\noindent\textbf{Acknowledgements.} 
The author thanks numerous interlocutors --- human and artificial --- for
many conversations that shaped this framework. Any errors are, of course,
the author’s responsibility.

\tableofcontents

\chapter*{Notation and Conventions}

Unless otherwise stated, we adopt the following conventions:
\begin{itemize}
  \item $(M,g)$ denotes a smooth oriented Lorentzian or Riemannian manifold,
  with metric $g$ of signature $(-,+,\dots,+)$ in the Lorentzian case.

  \item Greek indices $\mu,\nu,\dots$ range over spacetime coordinates;
  Latin indices $i,j,\dots$ are reserved for spatial coordinates when a
  splitting is chosen. We freely use the abstract index notation of Penrose.

  \item The Levi--Civita connection of $g$ is denoted by $\nabla$ and the
  associated Laplace--Beltrami operator by $\Delta = \nabla_a\nabla^a$.

  \item The symbol $\Map(X,Y)$ denotes a (derived) mapping stack when $X$ and
  $Y$ are derived stacks, and an ordinary mapping space when they are topological
  or smooth manifolds; the intended sense is clear from context.

  \item All tensor contractions use the Einstein summation convention.

  \item The core lamphrodynamic fields of \rsvpabbr{} are denoted
  $(\PhiField,\vField,\SField)$: a scalar potential, a vector flow, and an
  entropy density. We often group them into a single configuration field
  $\Psi := (\PhiField,\vField,\SField)$.
\end{itemize}

We recommend that readers unfamiliar with derived stacks or shifted symplectic
geometry skim \cref{chap:intro,chap:inf-cat} first, and then consult
\cref{chap:shifted-symplectic,chap:aksz} as needed when these notions appear
later.

\mainmatter

% ------------------------------------------------------------
% Part I: Derived Algebraic Geometry for Physicists
% ------------------------------------------------------------
\part{Derived Algebraic Geometry for Physicists}

\input{chapters/ch01-intro.tex}
\input{chapters/ch02-infinity-categories.tex}
\input{chapters/ch03-shifted-symplectic.tex}
\input{chapters/ch04-aksz-bv.tex}

% ------------------------------------------------------------
% Part II: Lamphrodynamic Field Theory
% ------------------------------------------------------------
\part{Lamphrodynamic Field Theory}

\input{chapters/ch05-lamphrodynamic-axioms.tex}
\input{chapters/ch06-plenum-stack.tex}
\input{chapters/ch07-aksz-rsvp.tex}
\input{chapters/ch08-variational-hamiltonian.tex}

% ------------------------------------------------------------
% Part III: Well-Posedness and Analysis
% ------------------------------------------------------------
\part{Well-Posedness and Analysis}

\input{chapters/ch09-linear-theory.tex}
\input{chapters/ch10-energy-regularity.tex}
\input{chapters/ch11-nonlinear-analysis.tex}

% ------------------------------------------------------------
% Part IV: Geometric and Topological Properties
% ------------------------------------------------------------
\part{Geometric and Topological Properties}

\input{chapters/ch12-entropic-singularities.tex}
\input{chapters/ch13-symmetries-reduction.tex}
\input{chapters/ch14-comparison-classical.tex}

% ------------------------------------------------------------
% Appendices
% ------------------------------------------------------------
\appendix

\input{appendices/appA-diff-geometry.tex}
\input{appendices/appB-homological-algebra.tex}
\input{appendices/appC-functional-analysis.tex}
\input{appendices/appD-notation.tex}
\input{appendices/appE-technical-lemmas.tex}

\backmatter
\printbibliography

\end{document}

