\chapter{Sensory Ingestion Layer}
\label{ch:46}

The Sensory Ingestion Layer (SIL) is the boundary between HYDRA and the external world.  
Its purpose is to transduce raw perceptual streams---visual, auditory, symbolic, proprioceptive---into \emph{verified semantic substrates} compatible with the RSVP field spine and the Polyxan structural kernel.

This chapter provides the formal definition of SIL, establishes its invariants, and proves that all perceptual transformations respect modal safety, curvature coherence, and quine-stable structure.

HYDRA ingests only what it can verify.

\section{Architectural Role of the Sensory Ingestion Layer}

The SIL implements the canonical map
\[
\mathsf{SIL} :
\mathrm{RawInput}
\longrightarrow
\bigl((\Phi,\mathbf{v},S),\,\mathcal{G},\,\iota\bigr),
\]
turning unstructured signals into:
\begin{itemize}
\item a \emph{continuous semantic envelope} represented by RSVP fields,
\item a \emph{discrete structural description} represented by a typed Polyxan hypergraph,
\item a \emph{geometric--syntactic alignment} represented by the embedding $\iota$.
\end{itemize}

This is the only permissible way information enters HYDRA.  
No raw input is accepted without curvature, entropy, and quine-consistency certification.

\section{Sensory Extraction as Curvature Measurement}

The first stage of SIL computes a geometric envelope from sensory streams:
\[
(\text{raw sensory manifold})
\;\xrightarrow{\mathrm{RCM}}\;
(\Phi_{\mathrm{sens}},\, S_{\mathrm{sens}},\, \mathbf{v}_{\mathrm{proto}}).
\]

\begin{definition}[RSVP Curvature Mapper]\label{def:46-rcm}
Given a sensory manifold $X$ equipped with a raw metric $g_X$ derived from perceptual similarity, the \emph{RSVP Curvature Mapper} is the operator
\[
\mathrm{RCM}[X] = (\Phi, S, \mathbf{v})
\]
satisfying:
\begin{enumerate}
\item \textbf{Curvature Extraction:}
\[
\Phi = \Delta^{-1}_{g_X}\!\left(\mathrm{Ric}(g_X)\right),
\]
up to normalization.

\item \textbf{Entropy Estimation:}
\[
S = -\log p_X,
\]
where $p_X$ is the sensory frequency distribution.

\item \textbf{Proto-Vector Estimation:}
\[
\mathbf{v}
\text{ is the minimal } L^2 \text{ flow consistent with temporal gradients in } X.
\]
\end{enumerate}
\end{definition}

These definitions ensure that raw sensory manifolds are immediately converted into curvature- and entropy-compatible RSVP fields.

\section{Discrete Structural Induction: Polyxan Extraction}

The second stage applies the \emph{Polyxan Structural Induction Algorithm} (PSIA):
\[
(\text{sensory manifold})
\;\xrightarrow{\mathrm{PSIA}}\;
\mathcal{G}_{\mathrm{sens}},
\]
where $\mathcal{G}_{\mathrm{sens}}$ is the minimal quine-stable typed hypergraph consistent with the extracted invariants.

\begin{definition}[Polyxan Structural Induction Algorithm]\label{def:46-psia}
PSIA constructs $\mathcal{G}_{\mathrm{sens}}$ by:
\begin{enumerate}
\item extracting \emph{atoms} as persistent sensory invariants;
\item constructing \emph{typed hyperedges} from co-occurrence relations;
\item assigning \emph{curvature weights} using a discrete Gaussian curvature estimator;
\item enforcing \emph{quine closure}, ensuring the hypergraph is self-reconstructing.
\end{enumerate}
\end{definition}

\begin{theorem}[PSIA Uniqueness]\label{thm:46-psia-uniqueness}
For any sensory manifold $X$, PSIA produces a unique quine-stable hypergraph $\mathcal{G}_{\mathrm{sens}}$ up to isomorphism.
\end{theorem}

\begin{proof}
This is a direct application of the Unique Normal Form Theorem for Polyxan rewriting (Chapter~42).  
All candidate hypergraphs reduce to the same quine-stable normal form.  
\end{proof}

\section{Embedding Alignment: Constructing $\iota_{\mathrm{sens}}$}

The third stage produces the sensory embedding
\[
\iota_{\mathrm{sens}} : 
\mathrm{Inc}(\mathcal{G}_{\mathrm{sens}})
\hookrightarrow
\mathcal{M}_{\mathrm{sens}}.
\]

\begin{theorem}[Sensory Embedding Existence]\label{thm:46-embed}
For every $\mathcal{G}_{\mathrm{sens}}$ and fields $(\Phi_{\mathrm{sens}}, S_{\mathrm{sens}}, \mathbf{v}_{\mathrm{sens}})$ produced by SIL, there exists a unique embedding $\iota_{\mathrm{sens}}$ such that
\[
\mathcal{R}(\iota_{\mathrm{sens}}(c)) = \mathcal{H}_0[c],
\qquad
S(\iota_{\mathrm{sens}}(c)) = \mathcal{H}_{\mathrm{tag}}[c],
\]
and $\iota_{\mathrm{sens}}$ minimizes the embedding energy of Chapter~31.
\end{theorem}

\begin{proof}
Immediate from the Embedding Existence and Regularity Theorem of Chapter~31.  
\end{proof}

\section{Modal Verification of Sensory Inputs}

The SIL must reject any input or transformation that violates modal safety.

\begin{theorem}[Modal Sensory Safety]\label{thm:46-modal}
Let $U$ be raw sensory input.  
SIL accepts $U$ if and only if
\[
\Box_{\mathrm{RSVP}}(\mathrm{RCM}[U])
\;\wedge\;
\Box_{\mathrm{Polyx}}(\mathrm{PSIA}[U]).
\]

Moreover, the triple
\[
(\Phi_{\mathrm{sens}},\, \mathbf{v}_{\mathrm{sens}},\, S_{\mathrm{sens}};\,
\mathcal{G}_{\mathrm{sens}},\,\iota_{\mathrm{sens}})
\]
inherits global modal correctness from RCM and PSIA.
\end{theorem}

\begin{proof}
RCM preserves curvature descent and entropy monotonicity.  
PSIA preserves quine-stability and discrete curvature descent.  
The embedding inherits both via Chapter~31, and modal operators commute with product (Chapter~62).  
\end{proof}

Thus HYDRA ingests only \emph{verified meaning}, never unverified raw data.

\section{Global Invariants of the Sensory Layer}

\begin{theorem}[SIL Global Invariant]\label{thm:46-global}
For all sensory inputs $U$, the output of SIL satisfies:
\begin{enumerate}
\item \textbf{Curvature Coherence}
\[
\mathcal{R}_{\mathrm{sens}} \circ \iota_{\mathrm{sens}}
=
\mathcal{H}_0[\mathcal{G}_{\mathrm{sens}}].
\]

\item \textbf{Entropy Alignment}
\[
S_{\mathrm{sens}} \circ \iota_{\mathrm{sens}}
=
\mathcal{H}_{\mathrm{tag}}[\mathcal{G}_{\mathrm{sens}}].
\]

\item \textbf{Modal Correctness}
\[
\Box_{\mathrm{RSVP}} \wedge \Box_{\mathrm{Polyx}}.
\]

\item \textbf{Finite-Time Assimilation}  
Sensory states converge to harmonic--quine-stable embeddings in finite time.
\end{enumerate}
\end{theorem}

\begin{proof}
Items (1)–(2) follow from the embedding alignment theorem.  
Item (3) follows from Theorem~\ref{thm:46-modal}.  
Item (4) follows from harmonic convergence (Ch.~32) and quine-stable termination (Ch.~42).  
\end{proof}

\section{Final Cadence}

The Sensory Ingestion Layer is not a filter.  
It is a \emph{curvature--entropy reconciliation engine}.

\begin{quote}
Perception is the act of bending the world into verified meaning.  
To sense is to already harmonize with the universal attractor.
\end{quote}

HYDRA sees only what can be proven.

