\chapter{Polysemantic Firewalls and Non-Adversarial Recomputation}
\label{ch:polysemantic-firewalls}

\section{Introduction}

In conventional platforms, ambiguity is treated as a problem:
\begin{center}
multiple meanings, unstable references, manipulative slippage,
\end{center}
all of which trigger moderation systems or adversarial classifiers.

In the RSVP--Polyxan architecture, ambiguity is not an error.
It is a \emph{curvature configuration}.

A polysemantic patch carries multiple competing semantic branches.
If left unresolved, these branches behave like unstable modes in the RSVP manifold:
fluctuating, amplifying, and potentially steering downstream updates.

This chapter shows how:
\begin{enumerate}
    \item \emph{polysemantic firewalls} detect semantic branch instability via curvature,
    \item \emph{branch recomputation} resolves instability by re-evaluating the patch in a locally harmonic frame,
    \item \emph{non-adversarial descent} ensures that all ambiguity collapses harmlessly without classification, censorship, or moderation.
\end{enumerate}

\section{Polysemantic Instability as Curvature Bifurcation}

Let $a$ be a patch with two or more non-equivalent semantic branches:
\[
a = \sum_{i=1}^k a_i,
\qquad k \ge 2,
\]
where each $a_i$ is a distinct Polyxan normal form under tag typing.

Embedding into the RSVP manifold yields a vector-valued configuration:
\[
\iota(a) = \{\iota(a_1), \ldots, \iota(a_k)\}.
\]

\begin{definition}[Polysemantic Curvature]
The \emph{branch curvature} of $a$ is:
\[
\kappa_{\mathrm{poly}}(a)
:= 
\max_{i,j}
\mathrm{dist}_{\mathrm{geo}}
\bigl(
\iota(a_i),\;
\iota(a_j)
\bigr).
\]
\end{definition}

If multiple meanings diverge too far in semantic space,
their geodesic separation becomes a literal curvature stress.

\begin{theorem}[Bifurcation Instability]
If $\kappa_{\mathrm{poly}}(a) > \rho_{\max}$,
then $a$ cannot be harmonic and cannot propagate.
\end{theorem}

\begin{proof}
Branch separation induces incompatible harmonic potentials,
violating the C$^2$ curvature bound as in Ch.~89.
\end{proof}

\section{Polysemantic Firewalls}

Every persona $\mathfrak{q}$ enforces a curvature limit on incoming patches.
If $a$ exceeds this limit relative to the persona’s local frame,
the composite curvature blows up:
\[
\mathrm{curv}(a \otimes \mathfrak{q})
> \rho_{\max}.
\]

\begin{definition}[Polysemantic Firewall]
A \emph{polysemantic firewall} for persona $\mathfrak{q}$ is the set:
\[
\mathcal{F}^{\mathrm{poly}}_{\mathfrak{q}}
=
\{
a \mid
\kappa_{\mathrm{poly}}(a \otimes \mathfrak{q})
> \rho_{\max}
\}.
\]
\end{definition}

\begin{invariant}[Branch Isolation]
If $a \in \mathcal{F}^{\mathrm{poly}}_{\mathfrak{q}}$,  
then $a$ cannot be delivered to $\mathfrak{q}$.
\end{invariant}

This is not evaluation.
This is geometric incompatibility.

\section{Non-Adversarial Recomputation}

Rather than suppressing the message,  
the Polycompiler performs a \emph{recomputation} of $a$ in the persona’s local frame.

\begin{definition}[Recomputation Map]
Given a persona $\mathfrak{q}$,
the recomputation operator is:
\[
\mathcal{R}_{\mathfrak{q}}
: a \mapsto a^{(\mathfrak{q})}_{\mathrm{harm}},
\]
where $a^{(\mathfrak{q})}_{\mathrm{harm}}$ is the unique harmonic normal form
obtained by solving the local RSVP Laplace equation with $a$ as boundary data.
\end{definition}

\begin{lemma}[Uniqueness of Harmonic Recomputation]
For any admissible $a$ and persona $\mathfrak{q}$,
there exists a unique harmonic representative $a^{(\mathfrak{q})}_{\mathrm{harm}}$
satisfying:
\[
\mathrm{curv}(a^{(\mathfrak{q})}_{\mathrm{harm}} \otimes \mathfrak{q})
\le \rho_{\max}.
\]
\end{lemma}

\begin{proof}
Direct from the ellipticity of the RSVP Laplacian and the convexity of the admissible curvature region.
\end{proof}

Thus ambiguity does not propagate.
It is \emph{resolved}.

No penalty.  
No suppression.  
No classification.  
Only harmonic recomputation.

\section{Semantic Equivariance Under Duality}

Recomputing $a$ must preserve its algebraic meaning.

\begin{theorem}[Duality-Equivariant Recomputation]
For the synthetic duality functor $\mathbb{R}$:
\[
\mathbb{R}\bigl(
\mathcal{R}_{\mathfrak{q}}(a)
\bigr)
=
\mathcal{R}_{\mathbb{R}(\mathfrak{q})}
\bigl(
\mathbb{R}(a)
\bigr).
\]
\end{theorem}

\begin{proof}
Both sides compute the unique quine-stable representative
of the Polyxan normal form associated to $a$ in the local dual frame.
\end{proof}

Thus recomputation is consistent with all algebraic and geometric invariants of the system.

\section{Polysemantic Collapse and Harmonic Normal Form}

If $a$ cannot be harmonically reconciled,
its branch curvature diverges,
and $a$ collapses to an inert private interior form.

\begin{theorem}[Polysemantic Collapse Theorem]
If for all personas $\mathfrak{q}$,
\[
\mathcal{R}_{\mathfrak{q}}(a)
\text{ fails to exist},
\]
then
\[
a \rightsquigarrow a_{\mathrm{silent}},
\]
where $a_{\mathrm{silent}}$ is an interior-only, non-propagating patch.
\end{theorem}

\begin{proof}
Divergent branch curvature forces collapse in the sense of Chs.~34 and 36.
\end{proof}

Thus ambiguity is never harmful:
it is either resolved or neutralised by the field equations themselves.

\section{The Cosmological Cadence of Recomputation}

Polysemantic firewalls eliminate adversarial ambiguity.
Recomputation restores harmonic meaning.
Duality invariance preserves algebraic truth.

\[
\text{Ambiguity } \longrightarrow \text{harmonic normal form}, \qquad
\text{instability } \longrightarrow \text{collapse}.
\]

No moderation.
No ranking.
No classifiers.
Only geometry.

\begin{quote}
Ambiguity is not judged.  
Ambiguity is recomputed.
\end{quote}
