\chapter{Real-Time Synthetic Media Pipeline}
\label{ch:53}

This chapter constructs the full real-time synthetic media pipeline built on the
RSVP--Polyxan substrate.  
Where previous chapters established the verification kernels, patch algebra,
and subsystem architecture, here we provide the continuous, executable pipeline
that converts RSVP fields and Polyxan hypergraphs into coherent audiovisual
streams.  
The pipeline operates under strict curvature descent, quine stability,
and modal correctness, producing media that is both generative and
interpretable.

The real-time pipeline is not merely a renderer.  
It is a \emph{semantic flow transformer} that carries meaning through the RSVP
manifold, the Polyxan hypergraph, the embedding layer, and the synthetic media
outputs, while preserving curvature invariants and inter-frame quine
consistency.


\section{Overview of the Pipeline Architecture}
\label{sec:53:overview}

The real-time pipeline is organised into four layers:

\begin{enumerate}
  \item \textbf{Field Evolution Layer:} advances the RSVP fields
        $(\Phi,\mathbf{v},S)$ under harmonic descent.
  \item \textbf{Hypergraph Frame Layer:} updates the Polyxan hypergraph
        $\mathcal{G}$ through EHR rewriting driven by curvature gradients.
  \item \textbf{Embedding Synchronisation Layer:} maintains a stratified,
        curvature-aligned embedding $\iota$ across frames.
  \item \textbf{Media Synthesis Layer:} generates real-time images, audio, and
        symbolic outputs consistent with the semantic state.
\end{enumerate}

Each layer is separately verified and collectively constrained by the modal
operators $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.


\section{Field Evolution Layer}
\label{sec:53:fields}

The RSVP fields advance according to the verified RSVP engine (Chapter~40),
ensuring curvature descent at each timestep.

\begin{definition}[Real-Time RSVP Step]
A real-time RSVP step is a function
\[
\mathsf{evolve}_{\mathrm{RSVP}} :
(\Phi_t,\mathbf{v}_t,S_t) \longrightarrow (\Phi_{t+\Delta t},
\mathbf{v}_{t+\Delta t},S_{t+\Delta t})
\]
that satisfies:
\begin{enumerate}
  \item Lyapunov monotonicity:
  \[
  \mathcal{E}_{\mathrm{RSVP}}(t+\Delta t) \le
  \mathcal{E}_{\mathrm{RSVP}}(t),
  \]
  \item harmonicity preservation,
  \item modal correctness: $\Box_{\mathrm{RSVP}}$.
\end{enumerate}
\end{definition}

The RSVP flow supplies the continuous substrate for visual and auditory
curvature fields in the media outputs.


\section{Hypergraph Frame Layer}
\label{sec:53:hypergraph}

The Polyxan hypergraph drives the symbolic structure of the media stream.

\begin{definition}[Real-Time Hypergraph Update]
An update step
\[
\mathsf{update}_{\mathcal{G}} :
\mathcal{G}_t \longrightarrow \mathcal{G}_{t+\Delta t}
\]
consists of a finite EHR rewrite sequence that strictly decreases discrete
curvature and respects quine stability.
\end{definition}

Each such update is the minimal symbolic change compatible with
curvature descent and patch algebra normalisation.


\section{Embedding Synchronisation Layer}
\label{sec:53:embedding}

The embedding $\iota_t : \mathrm{Inc}(\mathcal{G}_t) \hookrightarrow \mathcal{M}$
must evolve in synchrony with both the RSVP fields and the hypergraph frame.

\begin{definition}[Synchronous Embedding Step]
A synchronous embedding update
\[
\mathsf{sync}_{\iota} :
(\mathcal{G}_t,\iota_t) \longrightarrow (\mathcal{G}_{t+\Delta t},\iota_{t+\Delta t})
\]
is the harmonic map between consecutive frames minimising local distortion:
\[
\iota_{t+\Delta t} = \arg\min_{\iota}
\|\iota - \iota_t\|_{H^1}
\quad
\text{subject to curvature matching.}
\]
\end{definition}

\begin{theorem}[Embedding Stability]
Synchronous embedding preserves stratification and curvature alignment for all
frames.
\end{theorem}


\section{Media Synthesis Layer}
\label{sec:53:media-synthesis}

The media synthesis layer converts the semantic state
$(\Phi,\mathbf{v},S;\mathcal{G},\iota)$ into real-time audiovisual outputs.

\subsection{Visual Output}

Visual generation is curvature-driven.

\begin{definition}[Curvature-Based Render Field]
The render field is
\[
R_t(x) :=
f\bigl(\Phi_t(x),\,\|\mathbf{v}_t(x)\|,\,S_t(x),\,\mathcal{R}(x)\bigr),
\]
where $f$ is a monotone curvature-preserving transfer function.
\end{definition}

\begin{theorem}[Visual Consistency]
Between any two consecutive frames,
\[
\|R_{t+\Delta t} - R_t\|_{L^2} \le C \Delta t,
\]
ensuring stable real-time synthesis.
\end{theorem}


\subsection{Audio Output}

Audio synthesis is based on the divergence and vorticity of the RSVP vector
field.

\begin{definition}[Audio Signal]
\[
A_t(\omega) :=
\int_{\mathcal{M}} 
\mathrm{div}\,\mathbf{v}_t(x)\,
e^{-i\omega t} \, dV_x.
\]
\end{definition}

The divergence encodes harmonic convergence; the vorticity encodes quine
rotations.


\subsection{Symbolic Output}

Symbolic data are generated by projecting $\mathcal{G}_t$ to its quine-stable
core.

\begin{definition}[Symbolic Frame]
\[
\Sigma_t := \mathrm{Core}_{\mathrm{quine}}(\mathcal{G}_t).
\]
\end{definition}

\begin{theorem}[Symbolic Stability]
\[
\Sigma_t = \Sigma_{t+\Delta t}
\quad\text{whenever $\mathcal{G}_t$ is quine-stable.}
\]
\end{theorem}


\section{Pipeline Guarantees}
\label{sec:53:guarantees}

The real-time pipeline satisfies three core guarantees:

\begin{enumerate}
  \item \textbf{Modal Safety:}
        Every frame satisfies $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.
  \item \textbf{Curvature Monotonicity:}
        Total curvature is non-increasing across time.
  \item \textbf{Frame-Level Convergence:}
        Each frame is closer to the universal embedding of the quine $\mathfrak{U}$.
\end{enumerate}

\begin{theorem}[Real-Time Convergence]
The synthetic media stream converges to a steady-state representation of the
universal quine as $t\to\infty$.
\end{theorem}


\section{Conclusion}
\label{sec:53:conclusion}

We have constructed the full real-time synthetic media pipeline of the
RSVP--Polyxan system: field evolution, hypergraph updates, synchronised
embedding maintenance, and audiovisual synthesis under curvature and modal
constraints.  
This pipeline is the operative interface between verified meaning and human
perception: every pixel, waveform, and symbol is backed by harmonic curvature,
quine stability, and modal necessity.

With the pipeline complete, we now proceed to the media systems that refine and
extend it, beginning with the Polyxan Narrative Generator in Chapter~54.

