\chapter{RSVP Jet Bundles and Gauge Conditions}
\label{ch:jet-bundles}

This chapter develops the \emph{jet bundle formalism} for RSVP fields and introduces the associated \emph{gauge conditions} necessary for analytical well-posedness and numerical stability. Building on the semantic manifold (Chapter~\ref{ch:semantic-manifold}) and the RSVP Lagrangian (Chapter~\ref{ch:rsvp-lagrangian}), jet bundles provide a coordinate-free structure for handling derivatives of arbitrary order. Gauge conditions fix redundancies inherent in the RSVP field representation, enabling stable integration of PDEs and coherent embedding flows.

% ============================================================
\section{RSVP Jet Bundles}
% ============================================================

Let $(\mathcal{M},g)$ be the semantic manifold, and let
\[
E \;\to\; \mathcal{M}
\]
be the total RSVP field bundle whose sections are triples
\[
\sigma = (\Phi,\mathbf{v},S).
\]
The fields are assumed smooth.

\begin{definition}[k-th Jet Bundle]
The \emph{$k$-th jet bundle} $J^k E$ is the bundle over $\mathcal{M}$ whose fiber at $x \in \mathcal{M}$ consists of equivalence classes of sections whose partial derivatives agree up to order $k$:
\[
j_x^k\sigma \sim j_x^k\tau
\quad\Longleftrightarrow\quad
D^\alpha \sigma(x) = D^\alpha \tau(x)
\quad\text{for all }|\alpha|\le k.
\]
\end{definition}

Local coordinates on $J^k E$ consist of
\[
(x^i,\;\Phi,\;v^i,\;S,\;\Phi_{i_1},\;v^i_{i_1},\;S_{i_1},\dots,\Phi_{i_1\cdots i_k},\dots).
\]

Jet bundles serve as the natural arena for:
\begin{itemize}
\item higher-order Lagrangians,
\item symmetry analysis and prolongations,
\item systematic derivation of Euler–Lagrange operators,
\item gauge-fixing in overparameterized field theories.
\end{itemize}

% ============================================================
\section{Prolongation of Vector Fields}
% ============================================================

Let $X$ be a vector field on $E$, generating infinitesimal transformations
\[
\delta\Phi,\qquad \delta v^i,\qquad \delta S.
\]
The action of $X$ on derivatives of the fields must be compatible with total derivatives on $\mathcal{M}$.

\begin{definition}[k-th Prolongation]
The \emph{$k$-th prolongation} of $X$ is the vector field
\[
\mathrm{pr}^{(k)}X
=
X
+
\sum_{|\alpha|\le k}
\Bigl(
D_\alpha(\delta\Phi)\,\frac{\partial}{\partial\Phi_\alpha}
+
D_\alpha(\delta v^i)\,\frac{\partial}{\partial v^i_\alpha}
+
D_\alpha(\delta S)\,\frac{\partial}{\partial S_\alpha}
\Bigr),
\]
where $D_\alpha$ denotes the iterated total derivative operator.
\end{definition}

Prolongations are required for:
\begin{itemize}
\item constructing symmetries of $\mathcal{L}_{\mathrm{RSVP}}$,
\item formulating Noether currents,
\item implementing gauge symmetry in higher-order PDEs.
\end{itemize}

% ============================================================
\section{Contact Structure and Differential Invariants}
% ============================================================

The jet bundle $J^k E$ carries a canonical \emph{contact structure} governed by the 1-forms:
\[
\theta_\Phi = d\Phi - \Phi_i\,dx^i,\qquad
\theta_{v^i} = dv^i - v^i_j\,dx^j,\qquad
\theta_S = dS - S_i\,dx^i.
\]

These 1-forms vanish on the image of the $k$-jet prolongation of a genuine section.  
They encode:
\begin{itemize}
\item the constraint that derivative coordinates originate from actual fields,
\item differential invariants under gauge transformations,
\item the geometric separation between physical and redundant degrees of freedom.
\end{itemize}

% ============================================================
\section{Gauge Redundancy in RSVP Fields}
% ============================================================

Several transformations leave RSVP observables invariant, producing gauge redundancy:

\begin{itemize}
\item \textbf{Velocity potential gauge}:
\[
\mathbf{v} \mapsto \mathbf{v} + \nabla\chi.
\]
\item \textbf{Scalar shift gauge}:
\[
\Phi \mapsto \Phi + f(t).
\]
\item \textbf{Entropy flux gauge}:
\[
S \mapsto S + \mathrm{div}\,\mathbf{w}.
\]
\end{itemize}

These represent vertical directions in $J^k E$ along which the physical state does not change. Without gauge fixing, variational derivatives and PDE solvers exhibit degeneracy and numerical drift.

% ============================================================
\section{Gauge Conditions and Fixing}
% ============================================================

We impose the following canonical gauge conditions:

\begin{enumerate}
\item \textbf{Divergence-free flow} (in incompressible regimes):
\[
\nabla\cdot\mathbf{v}=0.
\]

\item \textbf{Zero-mean scalar potential}:
\[
\int_{\mathcal{M}} \Phi\, d\mathrm{vol} = 0.
\]

\item \textbf{Entropy-flux orthogonality} (boundary layers):
\[
\mathbf{w}\cdot\nabla S = 0.
\]
\end{enumerate}

These conditions:
\begin{itemize}
\item remove gauge freedom,
\item enforce uniqueness of variational solutions,
\item stabilize numerical integration,
\item and reduce the effective dimension of $J^k E$.
\end{itemize}

% ============================================================
\section{Jet-Bundle Euler--Lagrange Operators}
% ============================================================

Let $\mathcal{L}_{\mathrm{RSVP}}$ depend on fields and their derivatives up to order $k$. The Euler–Lagrange operator is:

\begin{definition}[Jet-Bundle Euler--Lagrange Operator]
\[
\mathcal{E}_\Phi(\mathcal{L})
=
\sum_{|\alpha|\le k}
(-D)_\alpha
\Bigl(
\frac{\partial\mathcal{L}}{\partial \Phi_\alpha}
\Bigr),
\]
with analogous operators $\mathcal{E}_{v^i}$ and $\mathcal{E}_S$.
\end{definition}

The RSVP field equations (Chapter~\ref{ch:rsvp-EL}) are:
\[
\mathcal{E}_\Phi(\mathcal{L}_{\mathrm{RSVP}})=0,\qquad
\mathcal{E}_{v^i}(\mathcal{L}_{\mathrm{RSVP}})=0,\qquad
\mathcal{E}_S(\mathcal{L}_{\mathrm{RSVP}})=0.
\]

This expression is:
\begin{itemize}
\item coordinate-free,
\item gauge-covariant,
\item valid for arbitrary derivative order,
\item and ideal for formal verification and symbolic computation.
\end{itemize}

% ============================================================
\section{Applications to Embedding Flows and Simulation}
% ============================================================

The jet-bundle framework is essential for:

\begin{itemize}
\item computing $\delta\mathcal{F}/\delta\iota$ in embedding flows (Chapter~\ref{ch:embedding-flow}),
\item evaluating sheaf-consistency obstructions (Chapter~\ref{ch:global-galaxy-maps}),
\item defining symplectic or variational integrators for RSVP PDEs,
\item implementing automatic differentiation in simulations.
\end{itemize}

Proper gauge-fixing prevents the appearance of spurious modes and ensures stable convergence.

% ============================================================
\section{Summary}
% ============================================================

In this chapter we introduced:
\begin{itemize}
\item the jet bundle $J^k E$ of RSVP fields,
\item prolonged vector fields and contact structures,
\item RSVP gauge redundancies and canonical gauge conditions,
\item the jet-bundle Euler–Lagrange formalism,
\item applications to embedding dynamics and simulation.
\end{itemize}

Jet bundles provide the higher-order geometric machinery necessary for precise formulation of RSVP dynamics and pave the way for Polyxan coupling and global sheaf-theoretic structures in subsequent chapters.
