\chapter{Interpretable Reputation and Semantic Credit Networks}
\label{ch:semantic-reputation}

\section{Introduction}

Reputation, in an interpretable and constitutionally constrained semantic
universe, cannot be a scalar score, a popularity metric, or a
black\textendash box algorithm.  
It must be a \emph{transparent, curvature\textendash bounded,
sheaf\textendash coherent functional}
on the space of quine\textendash stable personas.

This chapter constructs the \emph{Semantic Credit Network} (SCN):
a fully interpretable, machine\textendash verified,
polymeric reputation system whose values correspond to
geometric–syntactic invariants within the RSVP–Polyxan universe.

Reputation here is not an opinion.
It is a geometric quantity.

\section{Reputation as a Curvature–Sheaf Functional}

Let $\mathfrak{p}$ be a quine\textendash stable persona with verified certificate
$\mathrm{Cert}(\mathfrak{p})$.
Define the \emph{semantic curvature profile}:
\[
\mathbf{K}_{\mathfrak{p}}(x) := 
\left(
    \mathcal{R}_{\mathfrak{p}}(x), \;
    \operatorname{div} \mathbf{v}_{\mathfrak{p}}(x), \;
    S_{\mathfrak{p}}(x)
\right),
\]
and the persona’s tag\textendash sheaf $\mathscr{T}_{\mathfrak{p}}$.

\begin{definition}[Reputation Functional]
The reputation of $\mathfrak{p}$ is
\[
\operatorname{Rep}(\mathfrak{p})
  := \int_{\iota_{\mathfrak{p}}(\mathcal{G}_{\mathfrak{p}})}
      W\!\left(
        \mathbf{K}_{\mathfrak{p}},
        \mathscr{T}_{\mathfrak{p}}
      \right) \, d\mu_g,
\]
where $W$ is a constitutionally defined weighting that is

\begin{enumerate}
  \item \textbf{monotone in negative curvature}: contributions from regions of high negative curvature decrease reputation,
  \item \textbf{monotone in entropy alignment}: well\textendash aligned entropy increases reputation,
  \item \textbf{sheaf\textendash monotone}: coherent, high\textendash quality tags increase reputation, incoherent or adversarial tags decrease it,
  \item \textbf{local}: depends only on persona\textendash local data.
\end{enumerate}
\end{definition}

This functional is exactly interpretable:
every component has a geometric or syntactic meaning,
and every term is inspectable by any viewer.

\section{Semantic Credit Networks (SCNs)}

A Semantic Credit Network is defined as a weighted, typed, signed, bidirectional graph
\[
\mathrm{SCN} = \left(
V,\; E,\; w: E \to \mathbb{R},\; \tau: E \to \mathrm{Type}
\right),
\]
where $V$ is the set of quine\textendash stable personas.

Each edge encodes a \emph{credit assignment}:

\[
\mathfrak{p} \xrightarrow{w,\tau} \mathfrak{q}
\]

meaning:
persona $\mathfrak{p}$ credits $\mathfrak{q}$ with amount $w$
under justification of type $\tau$.

\subsection{Types of Credit}

The SCN supports a spectrum of grounded credit types:
\begin{itemize}
  \item epistemic credit (valid contribution to shared knowledge),
  \item semantic coherence credit (improving curvature and entropy),
  \item pedagogical credit (clarification, synthesis),
  \item infrastructural credit (patches, maintenance),
  \item integrity credit (actions that preserve the constitutional invariants).
\end{itemize}

All credit assignments must satisfy the governance rule:
\[
\operatorname{Rep}(\mathfrak{q}) \;\text{ increases iff }\;
\mathfrak{p} \xrightarrow{w,\tau} \mathfrak{q}
\;\text{ is curvature\textendash compatible}.
\]

\section{Conservation and Locality}

Reputation is not a conserved quantity.  
Credit is.

\begin{definition}[Credit Conservation]
For each persona $\mathfrak{p}$,
\[
\sum_{\mathfrak{p}\rightarrow \mathfrak{q}} w_{pq}
 = \sum_{\mathfrak{r}\rightarrow \mathfrak{p}} w_{r p},
\]
i.e.\ total outgoing credit equals total incoming credit.
\end{definition}

This ensures that:

\begin{itemize}
  \item reputation is emergent, not manipulable,
  \item no persona can inflate reputation by issuing credit arbitrarily,
  \item integrity of credit flows is preserved under EHR steps.
\end{itemize}

Locality of credit means that no credit can be transmitted across incompatible persona links;  
there are no “long\textendash range endorsements”.

\section{Curvature–Safe Aggregation}

For a cluster of personas $C$,
define the aggregate reputation
\[
\operatorname{Rep}(C)
  := \sum_{\mathfrak{p} \in C} \operatorname{Rep}(\mathfrak{p}).
\]

\begin{proposition}[Curvature–Safe Aggregation]
If every $(\mathfrak{p}, \mathfrak{q}) \in C$ satisfies the compatibility condition
\[
\left|\mathcal{R}_{\mathfrak{p}} - \mathcal{R}_{\mathfrak{q}}\right| 
  < \epsilon,
\]
then $\operatorname{Rep}(C)$ is a curvature\textendash monotone functional
and cannot be adversarially inflated.
\end{proposition}

\begin{proof}
Compatibility implies that the weighting function $W$ varies by at most $\epsilon$ across the cluster.
Thus pairwise and aggregate contributions remain bounded.
\end{proof}

\section{Reputation Flow Equations}

Reputation flows under the global RSVP–Polyxan evolution:

\[
\partial_t \operatorname{Rep}(\mathfrak{p})
  = - \int_{\iota(\mathcal{G}_{\mathfrak{p}})}
        \left[
        \beta\, (\mathcal{R}_{\mathfrak{p}})^2
       + \gamma\, (\nabla S_{\mathfrak{p}})^2
       \right] d\mu_g
   + \sum_{\mathfrak{q}} \partial_t w_{q p}.
\]

The first term is \emph{geometric decay}:  
high curvature and entropy mismatch reduce reputation.  
The second term is \emph{credit inflow}:  
verified contributions from others increase reputation.

\section{The Reputation Normal Form Theorem}

\begin{theorem}[Reputation Normal Form]
Every persona’s reputation converges in finite time to a unique
fixed point determined by its curvature\textendash entropy profile and
its incoming credit distribution.
\end{theorem}

\begin{proof}
Curvature\textendash entropy descent ensures geometric convergence.
Credit conservation ensures that the inflow converges to a bounded total.
Hence the integral defining reputation converges.
Uniqueness follows from strict convexity of $W$.
\end{proof}

\section{No Black Boxes: Full Interpretability}

The SCN is fully interpretable because:

\begin{itemize}
  \item every credit edge has a semantic type and justification,
  \item every reputation value decomposes into identifiable terms,
  \item the underlying curvature and entropy fields are visible,
  \item all computations are deterministic, reversible, audited.
\end{itemize}

There is no hidden ranking.  
No recommendation engine.  
No opaque penalty function.

\section{Final Cadence}

Reputation in the Polycompiler is not a contest.
It is not persuasion.
It is not a popularity economy.

\begin{quote}
Reputation is the integral of curvature and coherence.  
Credit is conserved.  
Interpretability is absolute.
\end{quote}
