\chapter{Hamiltonian Structure, Canonical Variables, and Constraint Algebra}
\label{ch:hamiltonian}

This chapter develops the canonical phase-space formulation of RSVP field
theory, placing the Lagrangian framework of
Chapters~\ref{ch:rsvp-fields}–\ref{ch:noether} into the symplectic
language required for renormalization, quantization, and coupling to
Polyxan hypergraphs.

We construct the canonical momenta for the RSVP triple
$(\Phi, v^a, S)$, identify the primary and secondary constraints arising
from gauge redundancies, develop the full constraint algebra, and obtain
the Hamiltonian evolution equations in their covariant form.

This Hamiltonian structure will serve as the canonical backbone for the
semantic RG flow (Ch.~\ref{ch:jet-renormalization}) and the coupled RSVP–Polyxan
dynamics developed in Part~III.

% --------------------------------------------------------------------
\section{Canonical Variables on the Semantic Manifold}
% --------------------------------------------------------------------

Let $(\mathcal{M},g)$ be a semantic manifold of dimension $d$,
decomposed as
\[
\mathcal{M} \simeq \Sigma \times \mathbb{R},
\]
with $\Sigma$ a spatial slice and $t$ a semantic time coordinate.

Write the RSVP Lagrangian density introduced earlier as
\[
\mathcal{L} = \tfrac{1}{2} |\nabla \Phi|^2
             + \tfrac{1}{2} |v|^2
             + \tfrac{1}{2} |\nabla S|^2
             + \beta \Phi\, \nabla_a v^a
             - V(\Phi,S).
\]

Define canonical momenta:
\[
\pi_\Phi := \frac{\partial \mathcal{L}}{\partial (\partial_t \Phi)}, \qquad
\pi^i_v := \frac{\partial \mathcal{L}}{\partial (\partial_t v_i)}, \qquad
\pi_S := \frac{\partial \mathcal{L}}{\partial (\partial_t S)}.
\]

Since $\mathcal{L}$ contains no $\partial_t v_0$ or $\partial_t g_{ab}$,
we immediately identify *primary constraints*.

\[
\pi^0_v \approx 0, \qquad
\pi^{ab}_g \approx 0.
\]

These are the canonical shadows of the gauge and diffeomorphism symmetries
studied in Chapter~\ref{ch:noether}.

% --------------------------------------------------------------------
\section{Hamiltonian Density and Legendre Transform}
% --------------------------------------------------------------------

The Hamiltonian density is
\[
\mathcal{H}
  = \pi_\Phi \,\partial_t \Phi
  + \pi^i_v \,\partial_t v_i
  + \pi_S \,\partial_t S
  - \mathcal{L}.
\]

After explicit evaluation:
\[
\mathcal{H}
 = \tfrac{1}{2} \pi_\Phi^2
 + \tfrac{1}{2} \pi_S^2
 + \tfrac{1}{2} \pi^i_v \pi_{v,i}
 + \tfrac{1}{2} |\nabla \Phi|^2
 + \tfrac{1}{2} |\nabla S|^2
 + \tfrac{1}{2} |v|^2
 + V(\Phi,S)
 - \beta \Phi\, \nabla_i v^i.
\]

The last term is the canonical expression of *semantic gravitation*, which
will control relaxation and galaxy formation in Part~III.

Define the canonical Hamiltonian:
\[
H = \int_\Sigma \mathcal{H}\; d\mu_\Sigma.
\]

% --------------------------------------------------------------------
\section{Primary and Secondary Constraints}
% --------------------------------------------------------------------

The vanishing momenta $\pi_v^0$ and $\pi_g^{ab}$ generate primary constraints:
\[
\phi_1 := \pi_v^0 \approx 0,\qquad
\phi_2^{ab} := \pi_g^{ab} \approx 0.
\]

Time preservation (Dirac consistency) yields secondary constraints:
\[
\chi_1 := \nabla_i \pi^i_v + \beta \nabla_i \Phi^i \approx 0,
\]
\[
\chi_2^a := \nabla_b T^{ab} \approx 0.
\]

Thus RSVP fields obey:

- a **Gauss-like constraint** coupling $\Phi$ and $v^a$,  
- a **momentum constraint** identical to the diffeomorphism constraint
  of gravitational systems.

\begin{theorem}[RSVP Constraint Algebra]
The set of constraints $\{\phi, \chi\}$ closes under the Poisson bracket:
\[
\{\chi_2(\xi), \chi_2(\eta)\}
    = \chi_2([\xi,\eta]),
\]
\[
\{\chi_2(\xi), \chi_1(f)\}
    = \chi_1(\mathcal{L}_\xi f),
\]
\[
\{\chi_1(f), \chi_1(g)\}
    = 0.
\]
\end{theorem}

Thus the RSVP constraint algebra is a semidirect product:
\[
\mathfrak{g}_\mathrm{RSVP}
     \simeq \mathfrak{diff}(\Sigma) \ltimes C^\infty(\Sigma).
\]

% --------------------------------------------------------------------
\section{Poisson Brackets and Canonical Commutators}
% --------------------------------------------------------------------

Canonical Poisson brackets:
\[
\{\Phi(x), \pi_\Phi(y)\}
   = \delta(x,y),
\]
\[
\{S(x), \pi_S(y)\}
   = \delta(x,y),
\]
\[
\{v_i(x), \pi^j_v(y)\}
   = \delta_i^j\,\delta(x,y).
\]

All other brackets vanish up to constraints.

These determine the symplectic structure needed for quantization
(in Part IV, Ch.~72–79).

% --------------------------------------------------------------------
\section{Hamiltonian Evolution Equations}
% --------------------------------------------------------------------

The canonical equations,
\[
\partial_t q = \{q, H\}, \qquad \partial_t p = \{p, H\},
\]
yield:

\[
\partial_t \Phi = \pi_\Phi,
\]
\[
\partial_t S = \pi_S,
\]
\[
\partial_t v_i = \pi^i_v + \beta \nabla_i \Phi,
\]

and conjugate momentum equations:
\[
\partial_t \pi_\Phi
 = \Delta \Phi - V_\Phi + \beta \nabla_i v^i,
\]
\[
\partial_t \pi_S
 = \Delta S - V_S,
\]
\[
\partial_t \pi^i_v
 = - v^i + \beta \nabla^i \Phi.
\]

These reproduce the Euler–Lagrange equations of Chapter~\ref{ch:rsvp-fields}.

% --------------------------------------------------------------------
\section{Semantic Time and Hamiltonian Flow}
% --------------------------------------------------------------------

The Hamiltonian generates semantic time evolution $t$ on $\mathcal{M}$:
\[
\mathcal{T} := \frac{d}{dt} = \{\,\cdot\,,H\}.
\]

\begin{proposition}[Semantic Time as Entropic Gradient Flow]
On-shell,
\[
\mathcal{T} = \mathrm{grad}_{\mathrm{symp}}\,\mathcal{H}
\]
coincides with the entropy-increasing flow:
\[
\nabla_a J^a_S \ge 0,
\]
established in Chapter~\ref{ch:noether}.
\end{proposition}

Thus Hamiltonian evolution is fundamentally entropic.

% --------------------------------------------------------------------
\section{Phase-Space Geometry and Semiclassical Limits}
% --------------------------------------------------------------------

Define the phase space:
\[
\Gamma_\mathrm{RSVP}
 := \{(\Phi,v_i,S,\pi_\Phi,\pi^i_v,\pi_S)\}/\text{constraints}.
\]

Its symplectic 2-form is:
\[
\Omega
 = \int_\Sigma
   \big(
     \delta \pi_\Phi \wedge \delta \Phi
    +\delta \pi^i_v \wedge \delta v_i
    +\delta \pi_S \wedge \delta S
   \big)\, d\mu_\Sigma.
\]

In the semiclassical limit (Part IV), the RSVP phase space develops
a Kähler-like structure induced by the RG fixed point $\mathcal{L}^*$.

This will serve as the foundation for:

- BV quantization,  
- cohomological gauge elimination,  
- derived geometric quantization of semantic galaxies.

% --------------------------------------------------------------------
\section{Summary}
% --------------------------------------------------------------------

This chapter has established:

\begin{itemize}
\item the canonical phase-space variables of RSVP fields;
\item the primary and secondary constraints generated by gauge and diffeomorphism invariance;
\item the full RSVP constraint algebra;
\item Poisson brackets and Hamiltonian evolution equations;
\item the identification of semantic time with entropic Hamiltonian flow;
\item the geometric structure of the RSVP phase space.
\end{itemize}

With this, the analytic and geometric foundation of Part~I is complete:
RSVP fields now possess both a covariant variational formulation and a
canonical phase-space structure.

The next chapter, Chapter~\ref{ch:renormalization-flow}, completes Part~I by
developing the renormalization group flow, the semantic c-theorem, and the
structure of fixed points.
