\chapter{Persona Rights and Liberties: A Constitutional Charter for the Polycompiler}
\label{ch:persona-rights}

\section{Introduction}

A constitutional platform must specify not only
the invariants that constrain visibility and interaction
(Chapter~\ref{ch:constitutional-invariants}),
but also the \emph{rights and liberties} of the autonomous entities
that inhabit the system.

These entities—called \emph{personas}—are stratified,
semantic agents represented as
embedded Polyxan hypergraphs with RSVP field profiles.
Their autonomy is formal.
Their safety is mathematical.
Their liberties are constitutional.

This chapter presents the Charter of Persona Rights for
the single\textendash shard Polycompiler universe.
It is not a policy document.  
It is a collection of \emph{proved invariants}
derivable from the RSVP–Polyxan dynamics,
the synthetic dual type theory,
and the Universal Patch Algebra.

\section{Personas as Stratified Semantic Agents}

A persona $\mathfrak{p}$ consists of:

\begin{enumerate}
    \item a typed Polyxan hypergraph $\mathcal{G}_{\mathfrak{p}}$,
    \item an RSVP embedding $\iota_{\mathfrak{p}} : \mathcal{G}_{\mathfrak{p}} \to \mathcal{M}$,
    \item a triad of fields $(\Phi_{\mathfrak{p}}, \mathbf{v}_{\mathfrak{p}}, S_{\mathfrak{p}})$,
    \item a patch algebra of public and interior content,
    \item a set of constitutional permitted transformations.
\end{enumerate}

A persona is therefore a \emph{mathematical object}.
Rights and liberties in this universe correspond to invariants of these objects
under the allowed transformations of the platform.

\section{The Right to Embedding Sovereignty}

Each persona controls its own embedding.

\begin{invariant}[Embedding Sovereignty]
No external persona or system component may directly modify
the embedding $\iota_{\mathfrak{p}}$,
except through:
\begin{center}
    curvature descent, entropy alignment, or patch normalisation.
\end{center}
\end{invariant}

Thus:
\begin{itemize}
    \item no forced rewrites of a persona’s structure,
    \item no external manipulation of its semantic curvature,
    \item no synthetic reshaping except through constitutional dynamics.
\end{itemize}

This is geometric autonomy.

\section{The Right to Interior Space}

Each persona has a semi-isolated internal domain
$\mathrm{Int}(\mathfrak{p})$,
a subcomplex of $\mathcal{G}_{\mathfrak{p}}$
not visible unless explicitly exported as a public patch.

\begin{invariant}[Interior Integrity]
Interior content cannot be inspected, ranked, analysed, or propagated
without the persona’s explicit export via public patch.
\end{invariant}

Consequences:

\begin{itemize}
    \item no surveillance,
    \item no metadata inference,
    \item no algorithmic inspection of drafts, partial thoughts, or internal derivations.
\end{itemize}

Privacy is not a setting.
It is a topological property.

\section{The Right to Curvature\textendash Neutral Visibility}

Visibility is curvature\textendash governed (Chapter~87),
but personas have the right to a neutral baseline:

\begin{invariant}[Curvature\textendash Neutral Visibility]
A persona may always publish curvature\textendash bounded, harmonic content
with maximal visibility weight.
No system component may suppress harmonic content.
\end{invariant}

Thus the platform preserves:
\begin{itemize}
    \item viewpoint neutrality,
    \item geometric fairness,
    \item the right to be seen when stable.
\end{itemize}

\section{The Right to Entropy\textendashAligned Interaction}

Interaction requires entropy alignment,
but personas retain full liberty to pursue alignment.

\begin{invariant}[Right to Align]
Every persona may freely adjust its entropy field $S_{\mathfrak{p}}$,
via:
\[
S_{\mathfrak{p}} \mapsto S_{\mathfrak{p}} + \delta S,
\]
subject only to global curvature bounds,
in order to interact with others.
\end{invariant}

Interaction is earned,
not restricted.

\section{The Right to Transparent Propagation}

Propagation occurs only through patch normalisation,
but personas have rights over how their content moves.

\begin{invariant}[Propagation Transparency]
Every propagated message must be accompanied by
\emph{a public certificate}:
\[
\mathsf{cert}(a,\mathfrak{p}\to\mathfrak{q})
  := (a \otimes \mathfrak{q} \Rightarrow a_{\mathrm{nf}}).
\]
\end{invariant}

No message may reach a new persona without a visible certificate
stating:
\begin{itemize}
    \item the normal form reached,
    \item the curvature bound used,
    \item the entropy alignment at propagation time.
\end{itemize}

\section{The Right to Semantic Rectification}

Errors, misinformation, or inconsistencies in a persona’s public content
may be repaired peacefully.

\begin{invariant}[Rectification]
Every persona has the right to apply  
\[
\text{Curvature\textendash Smoothing} \quad\text{and}\quad 
\text{Patch Reconciliation}
\]
to its own public material,
producing a new normal form
without loss of visibility weight.
\end{invariant}

Truth is curvature.
Correction reduces curvature.
Thus correction increases visibility.

\section{The Right to Reversible Self\textendash Description}

A persona may freely rewrite its self-description
so long as it preserves its own curvature bounds and quine stability.

\begin{invariant}[Self\textendash Description Liberty]
A persona may apply any sequence of EHR rewrites to itself,
provided curvature never exceeds constitutional bounds.
\end{invariant}

This ensures:
\begin{itemize}
    \item personal evolution,
    \item self\textendash repair,
    \item narrative coherence.
\end{itemize}

\section{The Right to Non\textendash Extraction}

No constitutional platform may extract:
\[
\text{attention}, \quad \text{clicks}, \quad \text{likes}, \quad
\text{interaction metrics},
\]
because none exist.
Only curvature, entropy, and patch algebra exist.

\begin{invariant}[Non\textendash Extraction]
No persona may be evaluated,
ranked,
or rewarded
by non-curvature metrics.
\end{invariant}

All incentives come from:
\begin{center}
harmonicity, alignment, and normalisation.
\end{center}

\section{The Right to Explainability}

Every persona has a right to understand why:

\begin{enumerate}
    \item a message was visible,
    \item an interaction was allowed/forbidden,
    \item a propagation succeeded/failed.
\end{enumerate}

\begin{invariant}[Explainability]
Every system action must be accompanied by a verifiable,
syntactically transparent explanation derived from the underlying
curvature, entropy, and patch algebra.
\end{invariant}

This is the essence of interpretability:
\[
\text{Nothing is hidden, because nothing can be.}
\]

\section{The Right to Exit}

A persona may withdraw from the shard at any time:

\begin{invariant}[Right to Exit]
A persona may collapse its public patches to interior form  
and remove its embedding from the manifold.
\end{invariant}

This is a geometric deletion that:
\begin{itemize}
    \item preserves safety,
    \item leaves no residual influence,
    \item severs all propagation pathways immediately.
\end{itemize}

\section{The Constitutional Cadence}

The Charter of Persona Rights turns the Polycompiler into  
a \emph{guaranteed\textendash fairness platform} in which:

\begin{itemize}
    \item no ranking exists,
    \item no extraction exists,
    \item no manipulation exists,
    \item autonomy is curvature,
    \item privacy is topology,
    \item fairness is harmonicity,
    \item transparency is algebra.
\end{itemize}

\begin{quote}
A persona is not a user.  
A persona is a mathematical agent,  
and its rights are the invariants of the system itself.
\end{quote}
