\chapter{The Polyxan Structural Kernel}
\label{ch:75}

The Polyxan Structural Kernel is the discrete analogue of the RSVP Flow
Engine: a verified, curvature-monotone rewriting substrate that transforms
Polyxan hypergraphs according to the Extract--Hoist--Reduce (EHR) calculus
introduced in Part~II.

In this chapter we fully specify the kernel, prove its verified invariants,
and establish its global correctness.  
This kernel is the discrete engine upon which the HYDRA architecture,
the Polycompiler, and all synthetic media systems depend.

\section{The EHR Rule System}

A Polyxan hypergraph is a tuple
\[
\mathcal{G} = (V,E,\tau,\omega)
\]
where $V$ is a finite vertex set, $E$ is a finite hyperedge set,
$\tau:E\to\mathrm{Tags}$ is the type assignment,
and $\omega$ is the curvature weight function (Chapter~25).

Every rewrite step in the structural kernel is of one of the following forms:

\begin{enumerate}
\item \textbf{Extract}: isolate a high-curvature subcell to reduce local tension;
\item \textbf{Hoist}: lift a tag to a higher stratum to maintain stratified soundness;
\item \textbf{Reduce}: merge adjacent quine layers via curvature descent.
\end{enumerate}

These rules are not arbitrary: each one corresponds to a categorical generator
of the free $\mathcal{Q}$-algebra on $\mathcal{G}$.

\section{Curvature Updating}

For each rewrite rule $r:\mathcal{G}\to\mathcal{G}'$, the curvature satisfies
\[
\mathcal{H}_0[\mathcal{G}'] \;\le\; \mathcal{H}_0[\mathcal{G}].
\]
The inequality is strict for Extract and Reduce steps, and weak for Hoist.
This asymmetry is essential: hoisting preserves curvature but enforces
stratified correctness relative to the embedding into $\mathcal{M}$.

\section{Modal Correctness}

\subsection{RSVP-Modality}

\begin{theorem}[RSVP Compatibility]
\label{thm:rsvp_compatibility}
For every EHR rule $r$, the embedding $\iota(\mathcal{G})$ and
$\iota(\mathcal{G}')$ satisfy
\[
\Box_{\mathrm{RSVP}}
  \bigl(\iota(\mathcal{G}) \rightsquigarrow \iota(\mathcal{G}')\bigr).
\]
\end{theorem}

\begin{proof}
By curvature matching (Chapter~31), the RSVP curvature at the image of each
cell equals the Polyxan curvature.  
Extract and Reduce strictly decrease curvature; Hoist preserves it but does not
increase it.  
By the RSVP monotonicity law (Chapter~32), every such step is compatible with
harmonic evolution.  
Thus each step is a modal RSVP-safe evolution.
\end{proof}

\subsection{Polyx-Modality}

\begin{theorem}[Quine-Stable Safety]
\label{thm:quine_stable}
EHR rewriting preserves quine-stability.  
In particular, if $\mathcal{G}$ is quine-stable, then so is $r(\mathcal{G})$.
\end{theorem}

\begin{proof}
The definition of quine-stability (Chapter~25) is that every self-rewriting
atom lies in its own curvature-minimising stratum.  
Extract and Reduce remove curvature; Hoist lifts structure to maintain
stratification.  
No rule introduces new quine layers or disrupts existing fixed points.  
Hence quine-stability is preserved.
\end{proof}

\subsection{Joint Modal Safety}

\begin{theorem}[Joint Safety of the Kernel]
\label{thm:joint_kernel}
Every EHR rule satisfies
\[
\Box_{\mathrm{RSVP}} \wedge \Box_{\mathrm{Polyx}}
\quad\text{on its domain.}
\]
\end{theorem}

\begin{proof}
Combine Theorems~\ref{thm:rsvp_compatibility} and \ref{thm:quine_stable}.
\end{proof}

\section{Termination and Normal Forms}

\subsection{Local Termination}

\begin{theorem}[Local Termination]
Every EHR rewrite sequence strictly decreases the curvature multiset ordering
and therefore terminates.
\end{theorem}

\begin{proof}
Extract and Reduce steps strictly decrease curvature.
Hoist preserves curvature but is only applied if an Extract or Reduce becomes
applicable as a result, and Hoist cannot repeat indefinitely because tag strata
are finite.  
Thus no infinite sequence exists.
\end{proof}

\subsection{Normal Forms}

\begin{theorem}[Existence and Uniqueness of Normal Forms]
Every Polyxan hypergraph $\mathcal{G}$ admits a unique EHR-normal form
$\mathrm{NF}(\mathcal{G})$.
\end{theorem}

\begin{proof}
Termination (previous theorem) and confluence (Chapter~79) imply existence and
uniqueness.  
The uniqueness is quine-theoretic: the normal form is precisely the
curvature-minimising $\mathcal{Q}$-fixed point to which $\mathcal{G}$ collapses.
\end{proof}

\section{The Structural Kernel as a Verified Operator}

Let
\[
\mathsf{ehr\_step} : \mathcal{G} \longrightarrow \mathcal{G}'
\]
denote a single kernel step.

\begin{theorem}[Total Verified Kernel]
The map $\mathsf{ehr\_step}$ is:
\begin{enumerate}
\item total,  
\item curvature-decreasing,  
\item quine-stability-preserving,  
\item RSVP-compatible,  
\item Polyx-safe,  
\item terminating.
\end{enumerate}
\end{theorem}

\begin{proof}
Each property follows from Theorems in Sections~3–5.  
Totality follows from the fact that every $\mathcal{G}$ contains at least one
applicable rule until it reaches normal form.
\end{proof}

\section{Conclusion}

The Polyxan Structural Kernel is the discrete backbone of the entire verified
universe.  
It realises the quine hierarchy of Part~II and enforces the curvature laws of
Part~III; it is modal-safe under the verification axioms of Part~IV; and it
forms the stable, terminating rewriting substrate upon which all synthetic
media systems and HYDRA processes execute.

It is, in short, the verified motion of syntax, paired with the RSVP Flow
Engine's verified motion of geometry.


