\chapter{Noether Currents, Conservation Laws, and Semantic Symmetries}
\label{ch:noether}

With the geometric, variational, and cohomological machinery of
Chapters~\ref{ch:rsvp-fields}ŌĆō\ref{ch:ghosts-cohomology} fully established,
we now turn to the Noetherian structure of RSVP dynamics.  
The goal of this chapter is to identify the complete family of conserved
currents associated with diffeomorphism invariance, internal gauge redundancy,
and the emergent symmetries of the semantic conformal point $\mathcal{L}^*$.

These currents provide the bridge between:

\begin{itemize}
\item the **geometric** evolution of fields on $(\mathcal{M},g)$,
\item the **cohomological** constraints imposed by gauge invariance,
\item the **energetic** structure encoded in $T_{ab}$,
\item and the **entropy-production** mechanisms that drive semantic relaxation.
\end{itemize}

This chapter forms the final analytic component of Part~I, completing the 
variationalŌĆōcohomological skeleton on which the coupled RSVP--Polyxan system
(Part~III) will rest.

% ===============================================================
\section{Noether's First Theorem on Curved Semantic Manifolds}
% ===============================================================

Let $\mathcal{L}[\Phi, v^a, S;g]$ be the RSVP Lagrangian density introduced in
Chapter~\ref{ch:rsvp-fields}.  
Consider an infinitesimal transformation of the fields:
\[
\delta \Phi = \epsilon\, X_\Phi, \qquad
\delta v^a = \epsilon\, X_v^a, \qquad
\delta S = \epsilon\, X_S,
\]
generated by a vector field $X$ in field space.

\begin{theorem}[Noether's First Theorem (RSVP Version)]
If the action is invariant under $X$,
\[
\delta_X \mathcal{S} = 0,
\]
then the Noether current
\[
J^a_X
  := \frac{\partial \mathcal{L}}{\partial (\nabla_a \Phi)} X_\Phi
   + \frac{\partial \mathcal{L}}{\partial (\nabla_a v_b)} X_v^b
   + \frac{\partial \mathcal{L}}{\partial (\nabla_a S)} X_S
   - \xi^a \mathcal{L}
\]
is covariantly conserved:
\[
\nabla_a J^a_X = 0.
\]
Here $\xi^a$ is the spacetime generator induced by the action of $X$ on $\mathcal{M}$.
\end{theorem}

This yields three principal families of currents:

\begin{enumerate}
\item diffeomorphism currents $J^a_{\xi}$,
\item gauge currents $J^a_{\chi,f,w}$,
\item and emergent symmetry currents at the semantic conformal point.
\end{enumerate}

% ===============================================================
\section{Diffeomorphism Invariance and the StressŌĆōEnergy Current}
% ===============================================================

For a diffeomorphism generated by $\xi^a$,
\[
\delta_\xi \Phi = \xi^a \nabla_a \Phi, \qquad
\delta_\xi v^a = \xi^b \nabla_b v^a - v^b \nabla_b \xi^a, \qquad
\delta_\xi S = \xi^a \nabla_a S.
\]

Applying NoetherŌĆÖs theorem gives the stressŌĆōenergy current:
\[
J^a_\xi = T^a{}_b \, \xi^b,
\]
where $T_{ab}$ is the RSVP stressŌĆōenergy tensor from
Chapter~\ref{ch:rsvp-fields}.

\begin{corollary}
For every vector field $\xi^a$ on $\mathcal{M}$,
\[
\nabla_a (T^a{}_b\, \xi^b) = 0.
\]
\end{corollary}

By choosing Killing fields of $g$, we recover energyŌĆōmomentum conservation.

% ===============================================================
\section{Gauge Symmetry and the BRST Noether Current}
% ===============================================================

Ghosts encode gauge parameters.  
Let $c_v, c_\Phi, c_S$ denote the ghost fields introduced in
Chapter~\ref{ch:ghosts-cohomology}.

\begin{theorem}[Gauge Noether Current]
Gauge invariance under the BRST operator $s$ implies:
\[
\nabla_a J^a_{\mathrm{BRST}} = 0,
\]
where
\[
J^a_{\mathrm{BRST}}
  = \frac{\partial \mathcal{L}}{\partial (\nabla_a \Phi)} s\Phi
   + \frac{\partial \mathcal{L}}{\partial (\nabla_a v_b)} s v^b
   + \frac{\partial \mathcal{L}}{\partial (\nabla_a S)} sS.
\]
\end{theorem}

The cohomology class $[J^a_{\mathrm{BRST}}]$ encodes topological
obstructions to gauge-fixing.

At the semantic conformal point $\mathcal{L}^*$ (Chapter~\ref{ch:jet-renormalization}),
this ghost current becomes part of a larger symmetry algebra.

% ===============================================================
\section{Emergent Symmetries at the Semantic Conformal Point}
% ===============================================================

At the RG fixed point $\mathcal{L}^*$, the Lagrangian acquires extended
WeylŌĆōVirasoroŌĆōlike symmetries:
\[
\delta_\sigma g_{ab} = 2\sigma g_{ab}, \qquad
\delta_\sigma \Phi = -\Delta_\Phi \sigma, \qquad
\delta_\sigma S = -\Delta_S \sigma,
\]
together with higher-derivative reparameterizations of the semantic manifold.

\begin{proposition}[Conformal Current]
The conformal generator $\sigma$ yields the conserved current:
\[
J^a_{\mathrm{conf}}
  = \left(
      \Delta_\Phi \frac{\partial \mathcal{L}^*}{\partial (\nabla_a \Phi)}
    + \Delta_S   \frac{\partial \mathcal{L}^*}{\partial (\nabla_a S)}
    - 2 \frac{\partial \mathcal{L}^*}{\partial g_{ab}} g_{bc}
    \right) \sigma.
\]
\end{proposition}

This current plays a central role in matching RSVP and Polyxan central charges.

% ===============================================================
\section{Entropy Production, Transport Currents, and Irreversibility}
% ===============================================================

Semantic relaxation (Chapter~\ref{ch:relaxation}) is governed by the
entropy-production current:
\[
J^a_S := S v^a - D^{ab} \nabla_b S,
\]
with $D^{ab}$ a semantic diffusivity tensor.

\begin{theorem}[Entropy Production Inequality]
For all admissible RSVP flows,
\[
\nabla_a J^a_S \ge 0.
\]
\end{theorem}

\begin{corollary}
The total entropy $\int_\mathcal{M} S$ is nondecreasing along solutions.
\end{corollary}

This yields the thermodynamic arrow of semantic time.

% ===============================================================
\section{Interplay with Polyxan Spectral Currents}
% ===============================================================

Polyxan hypergraphs admit spectral currents:
\[
j_e := \lambda_e\, \psi_e^2,
\]
where $\lambda_e$ are eigenvalues of the Polyxan Laplacian
(Chapter~\ref{ch:spectral-polyxan}).

The coupling between RSVP and Polyxan currents is:
\[
J^a_{\mathrm{coupled}}
  = T^a{}_b\, v^b + \sum_e \alpha_e\, (\nabla^a \psi_e)\psi_e.
\]

\begin{proposition}[Current Matching at Coupled Fixed Point]
At the RSVPŌĆōPolyxan conformal point,
\[
J^a_{\mathrm{RSVP}} = J^a_{\mathrm{Polyxan}}.
\]
\end{proposition}

Thus field currents and hypergraph spectral currents lock into a single
conserved structure.

% ===============================================================
\section{Summary of Conserved Structures}
% ===============================================================

We summarize the full set of canonical RSVP currents:

\begin{itemize}
\item $J^a_\xi$ ŌĆö diffeomorphism (stressŌĆōenergy) currents,
\item $J^a_{\mathrm{BRST}}$ ŌĆö gauge (ghost) currents,
\item $J^a_{\mathrm{conf}}$ ŌĆö semantic conformal currents,
\item $J^a_S$ ŌĆö entropy-production and transport currents,
\item $J^a_{\mathrm{spectral}}$ ŌĆö Polyxan spectral currents,
\item $J^a_{\mathrm{coupled}}$ ŌĆö RSVPŌĆōPolyxan interaction currents.
\end{itemize}

Each is covariantly conserved, and together they define the complete Noether
algebra of RSVP field theory.

This completes the analytic and variational foundation of Part~I.  
The next chapter, Chapter~\ref{ch:hamiltonian}, develops the Hamiltonian and
canonical phase-space structure of RSVP fields, preparing the transition to
renormalization and coupling.
