\chapter{The L-System of Cognitive Rewriting}
\label{ch:lsystem}

\section{Overview}

Chapter~\ref{ch:semantic-manifold} established the semantic manifold
$\mathcal{M}$, its stratified and sheaf-theoretic structure, and the RSVP field
content $(\Phi,\mathbf{v},S)$. Chapter~\ref{ch:rsvp-fields} then introduced the
RSVP variational principles and the geometric differential operators governing
semantic smoothing, directional bias, and entropy diffusion.

In the present chapter we introduce a generative, rule-based dynamical system
that acts on semantic configurations. This is the \emph{cognitive L-system},
a sheaf-valued generalization of the classical Lindenmayer rewriting formalism.
In contrast with string-based systems, the rewrite units here are local sheaf
sections of Polyxan atoms, and rewrite rules are natural transformations
compatible with the RSVP geometry, entropy gradients, and BV antibracket
structure.

The cognitive L-system defines:
\begin{itemize}
  \item a \emph{semantic alphabet} of rewrite atoms,
  \item a category of semantic configurations,
  \item a family of sheaf-morphic rewrite rules,
  \item an unfolding (growth) functor,
  \item and an induced dynamical flow on $\mathcal{M}$.
\end{itemize}

This layer mediates between RSVP thermodynamics and Polyxan hypergraph
evolution (cf.\ Chapter~\ref{ch:coupling}). It plays a foundational role in
semantic morphogenesis, cognitive development, and the formal ethics of
rewrite constraints (Chapter~\ref{ch:ethics}).

\section{The Semantic Alphabet}

\subsection{Polyxan atoms}

Let $\mathscr{P}$ denote the Polyxan sheaf from Chapter~\ref{ch:semantic-manifold}.
For each open set $U\subseteq\mathcal{M}$,
a section $p\in\Gamma(U,\mathscr{P})$ represents a local, typed semantic atom.
Via the embedding
\[
\iota: \mathscr{P} \hookrightarrow T^*\mathcal{M}\otimes\Lambda^k T\mathcal{M},
\]
each atom inherits geometric and differential structure from $\mathcal{M}$.

\subsection{Primary definition: geometric alphabet}
\label{subsec:alphabet-primary}

\begin{definition}[Semantic Alphabet]
The \emph{semantic alphabet} $\mathcal{A}$ is defined as
\[
\mathcal{A} := \mathrm{Aut}(\mathcal{M}) \backslash \mathscr{P},
\]
the set of orbits of Polyxan atoms under local diffeomorphisms of the semantic
manifold.
\end{definition}

Thus, each symbol $a\in\mathcal{A}$ corresponds to an equivalence class of
geometrically identical semantic atoms, invariant under semantic coordinate
changes. An element $a$ comes equipped with:
\begin{itemize}
  \item \emph{entropy charge} $q_S(a)$ inherited from $S$,
  \item \emph{valence vector} $q_v(a)$ encoding coupling to $\mathbf{v}$,
  \item \emph{semantic mass} $m(a)$, a curvature-induced inertia functional.
\end{itemize}

These values will participate in rewrite dynamics and energy estimates.

\begin{remark}[Alternative categorical interpretation]
A common alternative defines the alphabet as
\[
\mathcal{A}_{\mathrm{cat}} := \mathrm{Ob}(\mathcal{C}_{\mathscr{P}}),
\]
the object set of the small semantic category generated by Polyxan atoms.
This viewpoint is useful in homotopy-coherent L-systems (Chapter~\ref{ch:verification}),
but the geometric definition above is canonical for RSVP coupling.
\end{remark}

\subsection{Semantic strings}

Instead of 1D words, a \emph{semantic string} is a section of the configuration
sheaf:
\[
w \in \Gamma(U, \mathcal{A}^{\otimes n}),
\]
where the tensor product is fiberwise and respects geometric stratification.
Ordering is replaced by fiberwise monoidal structure and adjacency relations.

\section{Rewrite Rules as Sheaf Morphisms}

\subsection{Local rewrite rules}

A rewrite rule replaces one semantic atom by a structured configuration.

\begin{definition}[Rewrite Rule]
A \emph{rewrite rule} is a natural transformation of sheaves
\[
\rho : \mathscr{P} \Longrightarrow \mathscr{P}^{\otimes k}
\]
for some $k\ge 1$, satisfying:
\begin{enumerate}
  \item \textbf{Metric compatibility:}
  \[
  g(p_i,p_j) \le g(p,p)\,(1+\varepsilon).
  \]
  \item \textbf{Entropy monotonicity:}
  \[
  S(p_1)+\cdots+S(p_k)\ge S(p).
  \]
  \item \textbf{Functoriality under automorphisms:}
  \[
  \rho\circ f = f^{\otimes k}\circ\rho
  \quad \forall f\in\mathrm{Aut}(\mathcal{M}).
  \]
\end{enumerate}
\end{definition}

These constraints ensure that rewrite steps respect underlying RSVP geometry.

\subsection{Rewrite rule algebra}

Let $\mathcal{R}$ denote the free strict monoidal category generated by the
rewrite rules $\{\rho_i\}_{i\in I}$,
\[
\mathcal{R} := \mathrm{FreeMonCat}(\{\rho_i\}).
\]

\begin{itemize}
  \item Sequential composition is categorical composition.
  \item Parallel application corresponds to the monoidal product.
\end{itemize}

The algebra $\mathcal{R}$ admits:
\begin{itemize}
  \item an \emph{entropy grading},
  \item a BV extension $(\mathcal{R},\Delta,\{\cdot,\cdot\})$,
  \item homotopy completions ensuring coherence of higher rewrites.
\end{itemize}

The BV Laplacian encodes second-order rewrite interactions, and the antibracket
imposes nilpotence of ghost-induced symmetries (cf.\ Chapter~\ref{ch:rsvp-fields}).

\section{Growth Operators and Cognitive Morphogenesis}

\subsection{Configuration space}

Let $\mathsf{Conf}(\mathcal{M})$ denote the category of semantic configurations:
objects are sheaf sections
\[
X \in \Gamma(U, \mathscr{P}^{\otimes n}),
\]
and morphisms are restriction maps and natural transformations.

\subsection{Unfolding functor}

The core operator of the cognitive L-system is the \emph{unfolding functor}
\[
\mathsf{U} : \mathsf{Conf}(\mathcal{M}) \longrightarrow \mathsf{Conf}(\mathcal{M}),
\]
defined by
\[
\mathsf{U}(X)
= 
(\nabla \circ \rho \circ \pi_{\mathscr{P}})(X),
\]
where:
\begin{itemize}
  \item $\pi_{\mathscr{P}}$ extracts Polyxan atoms,
  \item $\rho$ applies a rewrite rule,
  \item $\nabla$ transports the expanded configuration using the RSVP connection.
\end{itemize}

\subsection{Discrete and continuous dynamics}

Discrete-time evolution:
\[
X_{t+1} = \mathsf{U}(X_t).
\]

In the continuous limit, this induces a vector field
\[
\mathcal{V}_{\mathsf{U}} := \lim_{\Delta t\to 0}
\frac{\mathsf{U}(X)-X}{\Delta t},
\]
and a corresponding flow
\[
\Phi_t = \exp(t\,\mathcal{V}_{\mathsf{U}}).
\]

This is the \emph{cognitive morphogenesis flow}.

\subsection{Stability analysis}

Fixed points satisfy
\[
\mathsf{U}(X)=X.
\]

In practice, semantic systems exhibit:
\begin{itemize}
  \item stable fixed points (persistent schemas),
  \item periodic orbits (cognitive loops),
  \item quasi-periodic regions,
  \item chaotic generative zones (high entropy, strong rewrite expansion).
\end{itemize}

Intentionality manifests as components of $\mathcal{V}_{\mathsf{U}}$
aligned with $\mathbf{v}$.

\section{Entropy and Affect in Rewrite Dynamics}

\subsection{Affect-weighted rules}

Define the affect-weighted rewrite tensor:
\[
\mathcal{T}_{\rho}(p) := \exp(S(p))\;\rho(p),
\]
which amplifies rewrite rules applied to high-salience atoms.

\subsection{Entropy coupling}

Rewrite steps change entropy via
\[
S(\rho(p)) - S(p)
= \nabla\cdot\mathbf{v} + \sigma,
\]
where $\sigma$ represents semantic dissipation.

This expresses RSVP thermodynamics in cognitive form.

\section{Ethical Rewrite Constraints}

\subsection{Ethical gradient}

Following the L-system sigma-model formalism, ethical constraints arise as a
subfunctor
\[
\mathsf{E} \subseteq \mathsf{R},
\]
where permitted rewrites satisfy a positivity condition:
\[
\langle J,\,\mathcal{V}_{\mathsf{U}}\rangle \ge 0,
\]
for an \emph{ethical current} $J$ (a vector field or 1-form on $\mathcal{M}$).

\subsection{Colimit-preserving rewrites}

An equivalent categorical condition:
\[
\mathsf{E}(\mathrm{colim}\,D) = \mathrm{colim}\,\mathsf{E}(D),
\]
ensuring that ethical rewrites preserve system-wide semantic coherence.

In cognitive terms, these are operations that maintain:
\begin{itemize}
  \item self-consistency,
  \item intersubjective agreement,
  \item long-term stability.
\end{itemize}

\section{Summary}

This chapter developed:
\begin{itemize}
  \item the semantic alphabet $\mathcal{A}$ as equivalence classes of Polyxan atoms,
  \item sheaf-morphic rewrite rules compatible with RSVP geometry and entropy,
  \item the monoidal rewrite algebra with BV extension,
  \item the unfolding functor and continuous cognitive morphogenesis flow,
  \item entropic and affective modulation of rewrite dynamics,
  \item and ethical rewrite constraints as a colimit-preserving subfunctor.
\end{itemize}

These structures complete the generative layer connecting RSVP fields to Polyxan
hypergraph dynamics. Chapter~\ref{ch:rsvp-lagrangian} develops the Lagrangian
governing rewrite evolution and its coupling to RSVP curvature.
