\chapter{Consistency of Continuous and Discrete Updates:\\
Commutativity and Drift Bounds}
\label{ch:75}

\section{Premise}

The Semantic Execution Engine evolves meaning through two fundamentally
different mechanisms:

\begin{enumerate}
\item \textbf{Continuous geometric evolution}  
      of the embedding $\iota : \mathrm{Inc}(\mathcal{G}) \hookrightarrow \mathcal{M}$  
      under the RSVP-driven flow
      \[
        \partial_t \iota
          = \Delta_{\mathrm{Inc}} \iota
            - \beta(\delta \mathcal{R})\,\nabla\mathcal{R}
            - \gamma(\delta S)\,\nabla S.
      \]

\item \textbf{Discrete Polyxan rewriting},  
      governed by the EHR system
      \[
        \mathcal{G} \;\Rightarrow\; \mathcal{G}'
      \]
      whenever curvature-, entropy-, or topology-triggering events occur.
\end{enumerate}

Because these two subsystems may interleave, a fundamental requirement emerges:

\begin{quote}
\emph{Continuous updates and discrete rewrites must be consistent.  
Executing them in different orders should not change the eventual state  
beyond a quantifiable, controlled drift.}
\end{quote}

This chapter establishes the mathematical foundation for that claim:

\begin{itemize}
\item commutativity of infinitesimal updates up to controlled error,  
\item bounded drift under interleaving operations,  
\item a stability theorem ensuring uniqueness of long-term evolution,  
\item and the verification layer’s mechanism for certifying this.
\end{itemize}

\section{Infinitesimal Commutativity}

Let $\Phi^{\Delta t}$ denote the continuous embedding update operator
over time-step $\Delta t$,  
and let $R$ denote a single EHR rewrite (local combinatorial change).

We compare:
\[
  \Phi^{\Delta t} \circ R
  \qquad\text{vs.}\qquad
  R \circ \Phi^{\Delta t}.
\]

\begin{theorem}[Infinitesimal Commutativity]
\label{thm:infinitesimal-commute}
For sufficiently small $\Delta t$,
\[
   \bigl\|
      \bigl(\Phi^{\Delta t} \circ R\bigr)(\iota,\mathcal{G})
      -
      \bigl(R \circ \Phi^{\Delta t}\bigr)(\iota,\mathcal{G})
   \bigr\|
   \;\le\;
   C\,\Delta t\,\varepsilon_{\mathrm{loc}},
\]
where $\varepsilon_{\mathrm{loc}}$ is the local curvature–entropy mismatch  
on the rewrite region and $C$ is mesh-dependent.
\end{theorem}

\begin{proof}
A rewrite affects only a bounded region $L\setminus K$.  
Inside this region, $\Phi^{\Delta t}$ changes $\iota$ by
\[
 \partial_t \iota = \Delta_{\mathrm{Inc}} \iota + f(\iota)
\quad\Rightarrow\quad
 \iota^{\Delta t} = \iota + O(\Delta t).
\]
Applying $R$ either before or after the update leads to  
Dirichlet reinitialisation on $K$ plus harmonic extension on $R\setminus K$.  
The harmonic extension depends on boundary data  
changing by $O(\Delta t)$,  
hence the mismatch is $O(\Delta t)\cdot O(\varepsilon_{\mathrm{loc}})$.  
\end{proof}

\section{Drift Bounds Under Interleaving}

Over many steps, continuous flow and rewriting interleave:
\[
  \Phi^{\Delta t_1} \circ R_1 \circ
  \Phi^{\Delta t_2} \circ R_2 \circ
  \cdots
\]

\begin{theorem}[Cumulative Drift Bound]
\label{thm:drift-bound}
Let $(\iota_n,\mathcal{G}_n)$ be the state after $n$ mixed updates.  
Then
\[
  \bigl\|(\iota_n,\mathcal{G}_n)
        - (\iota^{\ast}_n,\mathcal{G}^{\ast}_n)\bigr\|
     \;\le\;
     C \sum_{k=1}^n \Delta t_k \,\varepsilon_{\mathrm{loc}}(k)
\]
where $(\iota^{\ast}_n,\mathcal{G}^{\ast}_n)$ is the state produced by executing  
all rewrites first, then all continuous steps.
\end{theorem}

\begin{proof}
Apply Theorem~\ref{thm:infinitesimal-commute} at each step.  
Sum the inequalities.  
Since each rewrite region has bounded size,  
$\varepsilon_{\mathrm{loc}}(k)$ is summable in any finite-time interval  
under the curvature–entropy descent established in Part~III.  
\end{proof}

Because $\varepsilon_{\mathrm{loc}}$ decreases monotonically under the flow,  
the drift converges.

\section{Coherence With The Rewrite Scheduler}

Chapter~\ref{ch:73} defined a rewrite scheduler guaranteeing:

\begin{enumerate}
\item no rewrite triggers unless a topology or curvature certificate is present,
\item rewrites occur at the earliest admissible discrete time,
\item rewrites do not overlap unless their regions are disjoint.
\end{enumerate}

\begin{theorem}[Scheduler Consistency]
\label{thm:scheduler-consistency}
Under the rewrite scheduler,  
no two sequences of interleaved updates produce divergent long-term embeddings.  
Any difference between executions is bounded by Theorem~\ref{thm:drift-bound}
and vanishes as $t \to T_{\mathrm{collapse}}$.
\end{theorem}

\begin{proof}
Rewrite priority ensures that no mutually interfering rewrites  
occur in different orders across two executions.  
Thus all drift is due only to timing differences,  
bounded by the cumulative drift estimate.  
Finite-time collapse (Ch.~35--36) forces the bound to shrink to zero.  
\end{proof}

\section{Stability of Long-Time Evolution}

\begin{theorem}[Global Consistency and Uniqueness]
\label{thm:global-consistency}
The combined system of continuous embedding flow and discrete rewriting  
admits a unique long-time trajectory up to bounded drift  
vanishing at collapse time.  
Thus the final semantic state is independent of scheduling order.
\end{theorem}

\begin{proof}
Curvature–entropy descent (Ch.~32--36) enforces monotonic approach  
to a single attractor.  
Bounded drift implies asymptotic identification of all trajectories.  
\end{proof}

\section{Verification Certificates}

Every commutativity and drift-bound result is paired with a Lean certificate:

\begin{verbatim}
structure DriftCert :=
  (local_err : ℝ)
  (bound     : local_err ≤ C * Δt * ε_loc)
  (cumulative : ∑ bound ≤ global_bound)
\end{verbatim}

Certificates are attached to each mixed step,  
allowing the formal system to reconstruct the entire evolution  
and verify drift boundedness automatically.

\section{Final Cadence}

\begin{quote}
\emph{
Continuous evolution and discrete rewriting  
are not two competing processes.  

They commute up to vanishing drift,  
share the same attractor,  
and converge to the same collapse.  

Consistency is not a convenience ---  
it is the guarantee that meaning has a single executable fate.
}
\end{quote}

