\chapter{RSVP Audio Field Synthesis}
\label{ch:56}

This chapter develops the RSVP Audio Field Synthesis engine: the continuous,
curvature-driven method by which RSVP fields generate coherent, interpretable,
time-varying auditory structures.  
In the same way that Chapter~\ref{ch:55} constructs visual scenes as harmonic
cross-sections of the RSVP manifold, the present chapter constructs audio as
\emph{vibrational modes} of the scalar--vector--entropy fields.

The key principle is that all sound is curvature and flow:  
\[
\text{audio} = \text{oscillation of curvature under entropy modulation}.
\]

\section{Acoustic Manifolds and Vibrational Modes}
\label{sec:56:manifold}

\begin{definition}[Acoustic Manifold]
Given an RSVP field configuration $(\Phi,\mathbf{v},S)$, the \emph{acoustic
manifold} is the stratified space
\[
\mathcal{A}_t := \{ x \in \mathcal{M} \mid \partial_t S_t(x) \neq 0 \},
\]
i.e.\ the region of active entropy modulation.
\end{definition}

Entropy change is the universal signal of dynamical activity; thus audio lives
on $\mathcal{A}_t$.

\begin{definition}[Vibrational Mode]
A vibrational mode is an eigenfunction of the curvature-weighted Laplacian:
\[
\Delta_{\mathcal{R}} u = -\omega^2 u,
\qquad
\Delta_{\mathcal{R}} := \nabla\cdot ( e^{-\mathcal{R}} \nabla ).
\]
\end{definition}

These modes generalize physical acoustics: curvature replaces density.


\section{Audio Field Equation}
\label{sec:56:wave-equation}

The RSVP audio field $A_t : \mathcal{M} \to \mathbb{R}$ is defined by the
coupled wave equation
\[
\partial_{tt} A_t
= \Delta_{\mathcal{R}} A_t
  + \alpha\,\nabla\Phi_t\cdot\nabla A_t
  - \gamma\,\partial_t S_t,
\]
where:
\begin{itemize}
  \item $\Delta_{\mathcal{R}}$ governs harmonic oscillation,
  \item the $\nabla\Phi$ term introduces semantic directionality,
  \item the $\partial_t S$ term injects episodic emphasis.
\end{itemize}

\begin{theorem}[Well-Posedness]
For any smooth, finite-energy initial data, the audio field equation has a
unique global solution on any finite time interval.
\end{theorem}

This ensures that sound generation is stable under RSVP flow.


\section{Harmonic Timbre Extraction}
\label{sec:56:timbre}

Given the vibrational modes $\{ u_k \}$, audio timbre is defined by curvature
weights.

\begin{definition}[Harmonic Timbre]
The harmonic timbre profile at time $t$ is
\[
\mathsf{timbre}_t(k)
=
\int_{\mathcal{A}_t} e^{-\mathcal{R}(x)} u_k(x)^2 \, dV_g.
\]
\end{definition}

\begin{theorem}[Curvature--Timbre Monotonicity]
If curvature decreases under RSVP flow, then
\[
t < t' \;\Longrightarrow\;
\mathsf{timbre}_{t'}(k) \le \mathsf{timbre}_t(k)
\quad
\text{for all modes $k$.}
\]
\end{theorem}

Thus harmonic timbre becomes smoother as the universe moves toward the
universal attractor.


\section{Narrative-Driven Modulation}
\label{sec:56:narrative}

As in Chapter~\ref{ch:55}, audio synthesis is synchronised with the Polyxan
narrative layer (Chapter~\ref{ch:54}).

Let $a_t$ denote the active narrative atom.

\begin{definition}[Narrative Modulation Envelope]
The modulation envelope of $a_t$ is
\[
M_t(x) :=
\exp\!\bigl( -\|\nabla\Phi_t(x) - \nabla\Phi_t(L(a_t))\|^2 \bigr),
\]
where $L(a_t)$ is the harmonic lift of $a_t$.
\end{definition}

\begin{theorem}[Narrative–Audio Synchronisation]
The dominant frequencies in $A_t$ are exactly the modes whose spatial support
overlaps $L(a_t)$ after curvature weighting.
\end{theorem}

Thus sound follows meaning, not vice versa.


\section{Amplitude, Spatialisation, and Motion}
\label{sec:56:spatial}

Audio amplitude corresponds to entropy flux, and spatialisation to vector flow.

\begin{definition}[Amplitude Field]
The amplitude field is
\[
\mathsf{amp}_t(x) := |\partial_t S_t(x)|.
\]
\end{definition}

\begin{definition}[Spatialisation Map]
The spatialisation of a listener-centred coordinate frame is given by
\[
\mathsf{spat}_t(x) =
\frac{\mathbf{v}_t(x)}{\|\mathbf{v}_t(x)\| + \varepsilon},
\]
mapping vector flow to perceived direction of sound.
\end{definition}

\begin{theorem}[Curvature-Preserving Spatialisation]
Spatialisation is invariant under harmonic smoothing and quine-frame
transformations.
\end{theorem}

Thus audio perspective matches narrative perspective.


\section{Canonical Audio Forms}
\label{sec:56:canonical}

\begin{definition}[Canonical Audio Form]
A canonical audio form is an audio field $A_\infty$ satisfying
\[
\partial_{tt} A_\infty = \Delta_{\mathcal{R}} A_\infty
\quad
\text{and supported on harmonic layers $H_0,H_1$.}
\]
\end{definition}

\begin{theorem}[Canonical Form Uniqueness]
Each quine frame determines a unique canonical audio form.
\end{theorem}

This parallels the canonical visual scenes of Chapter~\ref{ch:55}.


\section{Implementation Architecture}
\label{sec:56:architecture}

The RSVP Audio Field Synthesis engine integrates with the synthetic media
pipeline via the following stages:

\begin{enumerate}
  \item Extract $\mathcal{A}_t$ from RSVP fields.
  \item Compute curvature-weighted Laplacian eigenfunctions.
  \item Solve the audio field equation.
  \item Apply narrative modulation envelope.
  \item Render spatialised audio using $\mathbf{v}_t$.
  \item Harmonically normalise amplitude using entropy profile $S_t$.
\end{enumerate}

Each step is curvature-monotone and quine-stable.


\section{Long-Time Behaviour}
\label{sec:56:long-time}

\begin{theorem}[Audio Convergence]
Under RSVP flow, the audio field converges in the $C^\infty$ topology to the
canonical audio form associated with the universal quine $\mathfrak{U}$.
\end{theorem}

Thus \emph{all sound} collapses, in finite time, to the minimal harmonic audio
signature of the universal attractor.


\section{Conclusion}
\label{sec:56:conclusion}

The RSVP Audio Field Synthesis engine is the auditory counterpart to the
Harmonic Scene Composer.  
Where the latter reveals curvature as image, the former reveals curvature as
sound.  
Both are driven by entropy, both converge to canonical harmonic configurations,
and both express meaning through curvature descent.

Chapter~\ref{ch:57} now fuses these into the Semantic Galaxy Visualiser:  
a multimodal harmonic depiction of the RSVP--Polyxan universe.

