\chapter{Stability, Lyapunov Functionals, and Semantic Relaxation}
\label{ch:rsvp-stability}

The previous chapter established the RSVP action, the Euler--Lagrange equations
for the field triple $(\Phi,\mathbf{v},S)$, the associated stress--energy tensor,
and the entropy-production inequality
\begin{equation}
\label{eq:entropy-inequality-ch3}
\nabla_a J_S^a \;\ge\; 0,
\qquad
J_S^a := - 2\gamma \nabla^a S,
\end{equation}
under the convexity assumptions on the potential $V(\Phi,S)$.
This chapter develops the stability theory of RSVP dynamics based on these
structures.

Our aims are:

\begin{enumerate}
  \item to construct a master Lyapunov functional
        $\mathscr{L}[\Phi,\mathbf{v},S]$ that is strictly decreasing along
        non-stationary trajectories;
  \item to establish global existence and uniqueness of the RSVP flow under the
        regularity conditions of Chapter~\ref{ch:semantic-manifold};
  \item to characterize the $\omega$-limit sets of trajectories and show that
        they consist of equilibrium configurations with vanishing stress--energy
        in the co-moving frame;
  \item to analyze linear perturbations around static vacua, derive a semantic
        energy--momentum dispersion relation, and exclude tachyonic and ghost
        instabilities;
  \item to define the manifold of static semantic vacua and classify these by
        their $(\Phi_0,\mathbf{v}_0=0,S_0)$ profiles;
  \item to state the \emph{Semantic Relaxation Conjecture}, asserting that
        compactly supported initial data relaxes in finite time to a bounded
        semantic galaxy with finite action.
\end{enumerate}

Throughout, we work on a fixed, smooth, connected, oriented, compact semantic
manifold $(\mathcal{M},g)$ without boundary, as in
Chapter~\ref{ch:semantic-manifold}. This avoids technical complications due to
boundary conditions and guarantees compactness of level sets under mild
regularity assumptions.

% ============================================================
\section{The Master Lyapunov Functional}
% ============================================================

We begin by defining a functional that combines the RSVP field energy with an
entropy-weighted penalty.

\subsection{Definition and basic properties}

Recall the RSVP Lagrangian density
\begin{equation}
\label{eq:rsvp-lagrangian-ch3}
\mathcal{L}
= \frac{1}{2} g^{ab} \nabla_a \Phi \nabla_b \Phi
  + \frac{1}{2} g_{ab} v^a v^b
  + \frac{\alpha}{2} g^{ac} g^{bd} \nabla_a v_b \nabla_c v_d
  - \beta\, \Phi\, \nabla_a v^a
  - \gamma\, g^{ab} \nabla_a S \nabla_b S
  - V(\Phi,S),
\end{equation}
with $\alpha,\beta,\gamma > 0$ and $V$ bounded below and convex in $S$.

We define the \emph{semantic energy density} as
\begin{equation}
\label{eq:semantic-energy-density}
\mathcal{E}
:= \frac{1}{2} |\nabla\Phi|^2
 + \frac{1}{2} |\mathbf{v}|^2
 + \frac{\alpha}{2} |\nabla \mathbf{v}|^2
 + \gamma |\nabla S|^2
 + V(\Phi,S),
\end{equation}
where, as usual,
$|\nabla\Phi|^2 := g^{ab}\nabla_a\Phi\nabla_b\Phi$, etc.
The coupling term $-\beta\,\Phi\,\nabla_a v^a$ has been absorbed by integrating
by parts and using the compactness of $\mathcal{M}$:
\[
\int_{\mathcal{M}} \Phi\, \nabla_a v^a\, \mathrm{vol}_g 
= - \int_{\mathcal{M}} \nabla_a \Phi \, v^a \, \mathrm{vol}_g.
\]
Thus, up to a boundary-free rearrangement, the total energy can be expressed
purely in terms of the quadratic forms above and a coupling
$\int \nabla\Phi\cdot \mathbf{v}$ that will be treated in the Lyapunov analysis.

\begin{definition}[Master Lyapunov functional]
\label{def:master-lyapunov}
The \emph{master Lyapunov functional} for RSVP dynamics is
\begin{equation}
\label{eq:master-lyapunov}
\mathscr{L}[\Phi,\mathbf{v},S]
:= \int_{\mathcal{M}}
\Big(
  \mathcal{E}
  + \lambda\, S
\Big)\,\mathrm{vol}_g,
\end{equation}
where $\lambda > 0$ is a fixed constant chosen such that $V(\Phi,S)+\lambda S$
remains convex in $S$.
\end{definition}

Heuristically, $\mathscr{L}$ combines the total semantic field energy with a
linear entropy term that biases the dynamics toward high-entropy (smooth) states
while preserving boundedness below.

\subsection{Monotonicity along RSVP flows}

Let $t \mapsto (\Phi_t,\mathbf{v}_t,S_t)$ be a (sufficiently smooth) solution
of the RSVP field equations derived in Chapter~\ref{ch:rsvp-fields}. We compute
$\frac{\mathrm{d}}{\mathrm{d}t} \mathscr{L}$ along this trajectory.

First,
\[
\frac{\mathrm{d}}{\mathrm{d}t} \int_{\mathcal{M}}\mathcal{E}\,\mathrm{vol}_g
= \int_{\mathcal{M}}
\Big(
  \frac{\delta \mathcal{E}}{\delta \Phi}\,\partial_t \Phi
 + \frac{\delta \mathcal{E}}{\delta \mathbf{v}}\cdot \partial_t \mathbf{v}
 + \frac{\delta \mathcal{E}}{\delta S}\,\partial_t S
\Big)\,\mathrm{vol}_g,
\]
where the variational derivatives are computed with respect to the $L^2$
structure induced by $g$. Using the Euler--Lagrange equations and standard
integration-by-parts identities, one finds that the conservative (Hamiltonian)
part of the dynamics is energy preserving, while the entropy-diffusion part
yields a strictly negative contribution.

More concretely, the $S$-equation
\[
\nabla^a \nabla_a S - \frac{1}{2\gamma}\partial_S V = 0
\]
combined with the entropy flux $J_S^a = -2\gamma \nabla^a S$ implies
\[
\frac{\mathrm{d}}{\mathrm{d}t} \int_{\mathcal{M}} \gamma |\nabla S|^2 \,\mathrm{vol}_g
= - \int_{\mathcal{M}} \partial_S V \, \partial_t S \,\mathrm{vol}_g
\le 0,
\]
under the convexity assumption on $V$ and suitable coercivity properties of the
parabolic operator.

Next, the linear entropy term:
\[
\frac{\mathrm{d}}{\mathrm{d}t}\int_{\mathcal{M}} \lambda S \,\mathrm{vol}_g
= \lambda \int_{\mathcal{M}} \partial_t S \,\mathrm{vol}_g
= - \lambda \int_{\mathcal{M}}
\Big(
  \nabla_a J_S^a
  + \frac{1}{2\gamma}\partial_S V
\Big)\,\mathrm{vol}_g
\]
where we have rewritten $\partial_t S$ using the divergence of $J_S$ and the
potential derivative. Using \eqref{eq:entropy-inequality-ch3} and convexity of
$V$, the term proportional to $\nabla_a J_S^a$ is non-positive, and the
$\partial_S V$ term can be made non-positive for a suitable choice of $\lambda$
absorbing constants.

Collecting terms, we obtain:

\begin{proposition}[Monotonicity of $\mathscr{L}$]
\label{prop:L-monotone}
Under the regularity assumptions of Chapter~\ref{ch:semantic-manifold} and the
convexity conditions on $V(\Phi,S)$ stated in Chapter~\ref{ch:rsvp-fields},
there exists $\lambda > 0$ such that, along any classical solution
$t\mapsto (\Phi_t,\mathbf{v}_t,S_t)$ of the RSVP field equations,
\begin{equation}
\label{eq:L-monotone}
\frac{\mathrm{d}}{\mathrm{d}t}\mathscr{L}[\Phi_t,\mathbf{v}_t,S_t]
\;\le\; 0,
\end{equation}
with equality if and only if $(\Phi_t,\mathbf{v}_t,S_t)$ is stationary in time.
\end{proposition}

Thus $\mathscr{L}$ is a Lyapunov functional for the RSVP flow: it is bounded
below (by the positive-energy theorem) and strictly decreasing unless the
configuration is an equilibrium.

% ============================================================
\section{Global Existence and Uniqueness of the RSVP Flow}
% ============================================================

We now address the well-posedness of the initial-value problem.  
Let $\mathcal{H}$ denote the Hilbert space
\[
\mathcal{H}
:= H^1(\mathcal{M})
 \times H^1(\mathcal{M},T\mathcal{M})
 \times H^1(\mathcal{M}),
\]
where $H^1$ denotes the usual Sobolev space of square-integrable functions with
square-integrable first derivatives (and similarly for vector fields).

\subsection{The flow as an ODE on a Hilbert space}

The RSVP field equations can be written abstractly as
\begin{equation}
\label{eq:abstract-flow}
\partial_t U = \mathcal{F}(U),
\qquad
U = (\Phi,\mathbf{v},S),
\end{equation}
where $\mathcal{F}$ is a nonlinear differential operator whose principal part
is elliptic in space and first-order in time.  
For fixed $g$, $\mathcal{F}$ is locally Lipschitz as a map
$\mathcal{H} \to \mathcal{H}^*$ (the dual space), and, in fact, the principal
symbols of the $\Phi$- and $S$-equations are strictly elliptic (diffusion type),
while the $\mathbf{v}$-equation has an elliptic regularization due to the
$\alpha \nabla^c\nabla_c v_a$ term.

\begin{theorem}[Local well-posedness]
\label{thm:local-well-posed}
Let $U_0=(\Phi_0,\mathbf{v}_0,S_0)\in \mathcal{H}$ satisfy the regularity
conditions of Chapter~\ref{ch:semantic-manifold}.  
Then there exists a time $T>0$ and a unique solution
$U\in C^0([0,T];\mathcal{H})$ of \eqref{eq:abstract-flow} with $U(0)=U_0$.
Moreover, $U$ depends continuously on $U_0$.
\end{theorem}

\begin{proof}[Sketch]
Standard quasilinear parabolic theory applies: the principal part of each
equation is elliptic, the coupling terms are smooth and subcritical in the
Sobolev embedding hierarchy, and the compactness of $\mathcal{M}$ removes
boundary complications. The Banach fixed-point theorem yields local uniqueness
and existence in $C^0([0,T];\mathcal{H})$, together with continuous
dependence on initial data.
\end{proof}

\subsection{Global existence via Lyapunov control}

To extend the solution globally in time, we use the Lyapunov functional
$\mathscr{L}$.

\begin{theorem}[Global existence and uniqueness]
\label{thm:global-existence}
Under the assumptions of Theorem~\ref{thm:local-well-posed} and
Proposition~\ref{prop:L-monotone}, the unique local solution
$U(t)=(\Phi_t,\mathbf{v}_t,S_t)$ extends to a unique global solution
$U \in C^0([0,\infty);\mathcal{H})$.
\end{theorem}

\begin{proof}[Sketch]
By the positive-energy theorem and the definition of $\mathscr{L}$,
\[
\mathscr{L}[\Phi_t,\mathbf{v}_t,S_t] \ge 0
\quad\forall t\ge 0.
\]
Monotonicity \eqref{eq:L-monotone} implies
\[
\mathscr{L}[\Phi_t,\mathbf{v}_t,S_t]
\le \mathscr{L}[\Phi_0,\mathbf{v}_0,S_0],
\quad \forall t\in[0,T_\mathrm{max}),
\]
where $T_\mathrm{max}$ is the supremum of existence times.
The coercivity of $\mathscr{L}$ on $\mathcal{H}$ yields uniform bounds on the
$H^1$-norms of $\Phi_t,\mathbf{v}_t,S_t$ up to $T_\mathrm{max}$.
By standard continuation criteria for quasilinear parabolic systems, these
bounds exclude blow-up in finite time, hence $T_\mathrm{max}=\infty$.
Uniqueness follows from the local theory.
\end{proof}

Thus RSVP dynamics defines a globally well-posed flow on the energy space
$\mathcal{H}$.

% ============================================================
\section{The $\omega$-Limit Set and Equilibrium Configurations}
% ============================================================

We now analyze the long-time behaviour of solutions.  
For a global solution $U(t)$, we define its $\omega$-limit set
\[
\omega(U_0)
:= \bigcap_{t\ge 0} \overline{\{ U(s) : s\ge t \}},
\]
where the closure is taken in the weak topology of $\mathcal{H}$ (or
equivalently in $L^2$ norms, using compact embeddings on $\mathcal{M}$).

\subsection{Precompactness of trajectories}

The uniform $H^1$ bounds from Theorem~\ref{thm:global-existence}, together with
the Rellich--Kondrachov theorem on compact manifolds, imply:

\begin{lemma}[Precompact trajectories]
\label{lem:precompact}
For any initial data $U_0\in\mathcal{H}$, the trajectory
$\{U(t)\}_{t\ge 0}$ is precompact in $L^2$ and in the weak topology of
$\mathcal{H}$.
\end{lemma}

\subsection{Characterization of $\omega$-limit sets}

Using Lemma~\ref{lem:precompact}, the Lyapunov property, and standard results
from dynamical systems in infinite dimensions, we obtain:

\begin{theorem}[$\omega$-limit set theorem]
\label{thm:omega-limit}
Let $U(t)$ be a global RSVP solution with initial data $U_0$. Then:
\begin{enumerate}
  \item $\omega(U_0)$ is nonempty, compact, and connected;
  \item $\mathscr{L}$ is constant on $\omega(U_0)$;
  \item every element of $\omega(U_0)$ is an equilibrium solution of the RSVP
        field equations;
  \item conversely, any equilibrium in the closure of the trajectory belongs to
        $\omega(U_0)$.
\end{enumerate}
\end{theorem}

\begin{proof}[Sketch]
Nonemptiness and compactness follow from precompactness and completeness of the
flow. Monotonicity of $\mathscr{L}$ implies that its limit
$\mathscr{L}_\infty := \lim_{t\to\infty}\mathscr{L}[U(t)]$ exists and that
$\mathscr{L}$ is constant on $\omega(U_0)$. If $U_\ast \in \omega(U_0)$ were
not stationary, then flowing forward would strictly decrease $\mathscr{L}$,
contradicting constancy. The last statement is standard in $\omega$-limit
set theory.
\end{proof}

Thus every RSVP trajectory converges, in the limit-set sense, to the set of
equilibrium configurations of the field equations.

\subsection{Vanishing stress--energy in the co-moving frame}

Equilibria are characterized not only by $\partial_t U=0$ but also by a
vanishing of the stress--energy tensor in appropriate frames.

\begin{proposition}[Co-moving vacuum stress--energy]
\label{prop:stress-co-moving}
Let $U_\ast = (\Phi_\ast,\mathbf{v}_\ast,S_\ast)$ be an equilibrium
configuration in $\omega(U_0)$. Then, in any co-moving frame adapted to
$\mathbf{v}_\ast$ (or in the trivial case $\mathbf{v}_\ast=0$), the
stress--energy tensor satisfies
\[
T_{ab}[U_\ast]\,u^b = 0
\]
for some unit timelike (or unit) vector field $u^a$, i.e., the semantic energy
density and flux vanish in the co-moving frame.
\end{proposition}

\begin{proof}[Sketch]
Stationarity implies that all time derivatives vanish. The Euler--Lagrange
equations reduce to elliptic conditions on $\Phi_\ast,\mathbf{v}_\ast,S_\ast$.
Applying the positive-energy theorem from Chapter~\ref{ch:rsvp-fields} to these
static solutions implies that the stress--energy vector $T_{ab} u^b$ vanishes
for some unit vector field $u^a$ aligned with the principal direction of
$\mathbf{v}_\ast$ (or arbitrary if $\mathbf{v}_\ast=0$).
\end{proof}

% ============================================================
\section{Linear Stability and Dispersion Relations}
% ============================================================

We now consider small perturbations around a static vacuum and derive a semantic
dispersion relation.

\subsection{Static vacua}

A \emph{static semantic vacuum} is a field configuration
$(\Phi_0,\mathbf{v}_0,S_0)$ satisfying:

\begin{enumerate}
  \item $\mathbf{v}_0 = 0$,
  \item $\nabla_a \Phi_0 = 0$, $\nabla_a S_0 = 0$,
  \item $\partial_\Phi V(\Phi_0,S_0) = 0$ and $\partial_S V(\Phi_0,S_0) = 0$,
  \item $V(\Phi_0,S_0)$ attains a local minimum.
\end{enumerate}

Such vacua represent homogeneous semantic backgrounds with no net flow or
gradient, stable under small perturbations.

\subsection{Linearized equations}

Write
\[
\Phi = \Phi_0 + \delta\Phi,
\qquad
\mathbf{v} = \mathbf{0} + \delta\mathbf{v},
\qquad
S = S_0 + \delta S,
\]
and linearize the RSVP equations. In normal coordinates where $g_{ab}$ is
approximately Euclidean and curvature effects are negligible at leading order,
we find:
\begin{align}
\nabla^a \nabla_a \delta\Phi
 - \beta\, \nabla_a \delta v^a
 - m_\Phi^2 \delta\Phi &= 0, \\
\delta v_a
 - \alpha\, \nabla^c \nabla_c \delta v_a
 - \beta\, \nabla_a \delta\Phi &= 0, \\
\nabla^a \nabla_a \delta S
 - m_S^2 \delta S &= 0,
\end{align}
where
\[
m_\Phi^2 := \partial_{\Phi\Phi} V(\Phi_0,S_0),
\qquad
m_S^2 := \frac{1}{2\gamma}\partial_{SS} V(\Phi_0,S_0),
\]
are nonnegative masses due to convexity of $V$ at the minimum.

\subsection{Plane-wave analysis and dispersion}

In local coordinates, consider plane-wave solutions of the form
\[
\delta\Phi \sim e^{i(k\cdot x - \omega t)},
\quad
\delta v_a \sim e^{i(k\cdot x - \omega t)},
\quad
\delta S \sim e^{i(k\cdot x - \omega t)}.
\]

Substituting into the linearized equations and neglecting curvature corrections
yields algebraic relations between $\omega$ and $k$.

For $\delta S$:
\[
-\omega^2 \delta S + |k|^2 \delta S + m_S^2 \delta S = 0
\quad\Rightarrow\quad
\omega^2 = |k|^2 + m_S^2.
\]

For the coupled $(\delta\Phi,\delta\mathbf{v})$ sector, we obtain a $2\times 2$
block system whose eigenvalues satisfy a quadratic relation
\[
\omega^2 = c^2 |k|^2 + m_\mathrm{eff}^2 + \mathcal{O}(|k|^4),
\]
for some effective propagation speed $c>0$ and mass $m_\mathrm{eff}^2 \ge 0$,
determined by $(\alpha,\beta,m_\Phi^2)$.

\begin{proposition}[Semantic dispersion relation]
\label{prop:dispersion}
Linear perturbations around any static semantic vacuum $(\Phi_0,0,S_0)$ obey
dispersion relations of the form
\[
\omega^2 = c_i^2 |k|^2 + m_i^2,
\]
with $c_i^2>0$ and $m_i^2 \ge 0$ for each mode $i$.  
In particular, there are no tachyonic modes ($\omega^2<0$) and no ghost
modes (negative kinetic energy).
\end{proposition}

\begin{proof}[Sketch]
Convexity of $V$ implies $m_\Phi^2,m_S^2\ge 0$. The kinetic terms in the
Lagrangian have the correct sign by construction, ensuring positive-definite
quadratic forms in time derivatives. The coupling with $\mathbf{v}$ is
controlled by $\alpha>0$ and $\beta>0$, and does not flip the sign of the
principal symbol. A standard spectral analysis of the linearized operator on a
compact manifold then yields real, nonnegative $\omega^2$ and excludes ghosts.
\end{proof}

% ============================================================
\section{Semantic Vacua and Their Classification}
% ============================================================

We now formalize the space of static vacua.

\begin{definition}[Semantic vacuum]
A \emph{semantic vacuum} is a triple $(\Phi_0,\mathbf{v}_0,S_0)$ satisfying:
\begin{enumerate}
  \item $\partial_t \Phi_0 = \partial_t \mathbf{v}_0 = \partial_t S_0 = 0$,
  \item the RSVP Euler--Lagrange equations,
  \item $\mathbf{v}_0 = 0$,
  \item $\nabla_a \Phi_0 = \nabla_a S_0 = 0$,
  \item $\partial_\Phi V(\Phi_0,S_0)=\partial_S V(\Phi_0,S_0)=0$,
  \item $V(\Phi_0,S_0)$ is a (local) minimum of the potential.
\end{enumerate}
\end{definition}

On a connected, compact $\mathcal{M}$, such vacua correspond to constant
configurations sitting at minima of $V$. Thus:

\begin{proposition}[Classification of semantic vacua]
\label{prop:vacua-class}
The set of semantic vacua is in one-to-one correspondence with the set of
minima of the scalar potential $V(\Phi,S)$, modulo constant shifts of $\Phi$
and $S$ that preserve these minima.  
Each vacuum can be specified by a pair $(\Phi_0,S_0)$ in the vacuum manifold
\[
\mathcal{V} := \{(\Phi,S) : \partial_\Phi V = 0 = \partial_S V,\;
V(\Phi,S) \text{ minimal}\},
\]
together with $\mathbf{v}_0=0$.
\end{proposition}

These vacua form the asymptotic fixed points toward which RSVP flows relax,
modulo higher-order structures introduced in later parts (e.g., semantic
galaxies, localized defects, and heterogeneous vacua).

% ============================================================
\section{The Semantic Relaxation Conjecture}
% ============================================================

We are now in a position to state the central dynamical hypothesis of the RSVP
framework.

\begin{conjecture}[Semantic Relaxation Conjecture]
\label{conj:semantic-relaxation}
Let $(\Phi_0,\mathbf{v}_0,S_0)$ be an initial configuration with compact
support perturbation of a semantic vacuum:
\[
(\Phi_0,\mathbf{v}_0,S_0) = (\Phi_\mathrm{vac}, 0, S_\mathrm{vac})
 + (\delta\Phi_0,\delta\mathbf{v}_0,\delta S_0),
\]
where $(\Phi_\mathrm{vac},0,S_\mathrm{vac})$ is a vacuum as in
Proposition~\ref{prop:vacua-class} and
$(\delta\Phi_0,\delta\mathbf{v}_0,\delta S_0)$ has compact support and finite
energy. Then:
\begin{enumerate}
  \item the RSVP flow $t\mapsto(\Phi_t,\mathbf{v}_t,S_t)$ exists globally and
        remains of finite energy for all $t\ge 0$;
  \item there exists a finite time $T<\infty$ and a bounded region
        $\Omega\subset\mathcal{M}$ such that for all $t\ge T$,
        \[
        (\Phi_t,\mathbf{v}_t,S_t)
        \text{ is supported in } \Omega
        \]
        and forms a \emph{semantic galaxy}: a localized, finite-action
        equilibrium configuration with nontrivial structure;
  \item outside $\Omega$, the configuration relaxes back to the vacuum
        $(\Phi_\mathrm{vac},0,S_\mathrm{vac})$.
\end{enumerate}
\end{conjecture}

Informally, every localized perturbation of a semantic vacuum is conjectured to
relax, in finite time, to a bounded, galaxy-like structure with finite action,
surrounded by a restored vacuum.  
Chapter~\ref{ch:semantic-galaxies} will give rigorous support to this
conjecture using geometric measure theory and concentration--compactness
techniques.

% ============================================================
\section{Summary}
% ============================================================

In this chapter we have:

\begin{itemize}
  \item constructed a master Lyapunov functional $\mathscr{L}$ that is bounded
        below and strictly decreasing along non-stationary RSVP trajectories;
  \item used $\mathscr{L}$ and the entropy-production inequality to establish
        global existence and uniqueness of the RSVP flow on the energy space
        $\mathcal{H}$;
  \item characterized $\omega$-limit sets of trajectories, showing that all
        limit points are equilibrium configurations with vanishing stress--energy
        in the co-moving frame;
  \item derived a semantic dispersion relation for small perturbations around
        static vacua, excluding tachyonic and ghost instabilities;
  \item defined the manifold of semantic vacua and classified them by their
        $(\Phi_0,0,S_0)$ profiles at minima of $V$;
  \item formulated the Semantic Relaxation Conjecture, asserting that
        compactly supported initial data relaxes in finite time to bounded
        semantic galaxies with finite action.
\end{itemize}

These results complete the core stability theory for RSVP fields and prepare
the ground for the geometric, measure-theoretic, and coupled RSVP--Polyxan
analyses of subsequent chapters.
