\chapter{Harmonic Scene Composer}
\label{ch:55}

This chapter constructs the Harmonic Scene Composer: a continuous, RSVP-driven
visual synthesis engine that converts stratified curvature patterns in
$\mathcal{M}$ into dynamic, interpretable scenes.  
Where Chapter~\ref{ch:54} established narrative at the discrete, Polyxan
level, the present chapter provides its continuous geometric counterpart:
scene composition as harmonic descent in the RSVP manifold.

The Harmonic Scene Composer is not a generative renderer layered on top of the
system.  
It \emph{is} the RSVP flow visualised: the geometric signature of semantic
curvature evolving in real time.

\section{Scene Manifolds and Harmonic Layers}
\label{sec:55:scene-manifolds}

A scene is the visible cross-section of the RSVP manifold.

\begin{definition}[Scene Manifold]
Given an RSVP field configuration $(\Phi, \mathbf{v}, S)$, the scene manifold
at time $t$ is
\[
\mathcal{S}_t := \{ x \in \mathcal{M} \mid \Phi_t(x) \ge \lambda_t \},
\]
where $\lambda_t$ is the dynamic visibility threshold.
\end{definition}

Thus scenes are not arbitrary projections: they are curvature-selected domains.

\begin{definition}[Harmonic Layer]
A harmonic layer is the level set
\[
H_k := \{ x \in \mathcal{M} \mid \mathcal{R}(x) = k \},
\]
for curvature stratum index $k$ (Chapter~\ref{ch:13}).
\end{definition}

Layers form the structural backbone of scene composition.


\section{Curvature-Driven Scene Geometry}
\label{sec:55:geometry}

Scene geometry is determined directly by RSVP curvature.

\begin{definition}[Harmonic Geometry Map]
The harmonic geometry map is the function
\[
\mathsf{geom}_t : \mathcal{S}_t \to \mathbb{R}^3
\]
defined by
\[
\mathsf{geom}_t(x) =
\bigl(
  \nabla\Phi_t(x),
  \nabla\cdot\mathbf{v}_t(x),
  S_t(x)
\bigr),
\]
interpreted as spatial, motion, and intensity coordinates.
\end{definition}

\begin{theorem}[Curvature Monotonicity]
Under RSVP flow,
\[
t < t' \;\Longrightarrow\;
\mathsf{geom}_{t'}(\mathcal{S}_{t'}) \subseteq
\mathsf{geom}_t(\mathcal{S}_t)
\]
up to diffeomorphism.
\end{theorem}

Scenes become simpler, smoother, and more interpretable over time, mirroring
semantic convergence.


\section{Scene Elements and Their Harmonic Lifts}
\label{sec:55:elements}

Visual elements arise as harmonic lifts of discrete narrative atoms.

Let $a$ be a narrative atom (Chapter~\ref{ch:54}).  
Its lifted region $L(a)$ inherits a harmonic feature signature.

\begin{definition}[Scene Element]
A scene element corresponding to $a$ is the tuple
\[
E(a) :=
\bigl(
  L(a),
  \mathsf{geom}_t(L(a)),
  w(a)
\bigr),
\]
where $w(a)$ is the narrative weight of $a$.
\end{definition}

Scene elements automatically reflect the narrative structure because both are
driven by the same curvature descent.

\begin{theorem}[Harmonic Coherence]
If $a_1$ and $a_2$ share a quine frame, then their scene elements satisfy
\[
\mathrm{dist}\bigl(E(a_1),E(a_2)\bigr) = O\!\left( w(a_1)+w(a_2) \right).
\]
\end{theorem}

Thus visual coherence emerges from quine stability.


\section{Scene Composition as Harmonic Minimisation}
\label{sec:55:composition}

Scene composition is phrased as a variational problem.

\begin{definition}[Scene Energy]
The scene energy at time $t$ is
\[
\mathcal{E}_{\mathrm{scene}}(t)
=
\int_{\mathcal{S}_t}
\Bigl(
  \|\nabla \Phi_t\|^2
  + \|\nabla\mathbf{v}_t\|^2
  + (S_t - \bar{S}_t)^2
\Bigr) \, dV.
\]
\end{definition}

\begin{theorem}[Scene Harmonisation Equation]
Critical points of $\mathcal{E}_{\mathrm{scene}}$ satisfy
\[
\Delta \mathsf{geom}_t = 0,
\]
the harmonic scene equation.
\end{theorem}

Thus the Harmonic Scene Composer generates scenes by solving the harmonic
equation on the visible submanifold.


\section{Canonical Scene Structures}
\label{sec:55:canonical}

As curvature descends, scenes converge to a finite catalogue of canonical
forms.

\begin{definition}[Canonical Scene]
A canonical scene is a harmonic scene
\[
\mathcal{S}_\infty := \lim_{t\to\infty} \mathcal{S}_t
\]
whose geometry is supported entirely on harmonic layers $H_0$ and $H_1$.
\end{definition}

\begin{theorem}[Canonical Scene Uniqueness]
Each quine frame determines exactly one canonical scene up to isometry.
\end{theorem}

Thus setting, character, and narrative frame determine visual appearance.


\section{Scene–Narrative Synchronisation}
\label{sec:55:synchronisation}

The Harmonic Scene Composer is synchronised with the Polyxan Narrative
Generator (Chapter~\ref{ch:54}) through the shared curvature and quine layers.

\begin{theorem}[Narrative–Scene Synchronisation]
Let $a_t$ be the active narrative atom at time $t$.
Then the dominant scene element is $E(a_t)$, and the visible scene satisfies
\[
\mathcal{S}_t =
\mathrm{nbhd}\bigl( L(a_t) \bigr)
\]
up to harmonic smoothing.
\end{theorem}

This ensures that visual and symbolic layers evolve coherently.


\section{Implementation Architecture}
\label{sec:55:architecture}

The Harmonic Scene Composer integrates with the real-time synthetic media
pipeline (Chapter~\ref{ch:53}) as follows:

\begin{enumerate}
  \item Input RSVP fields $(\Phi,\mathbf{v},S)$.
  \item Compute curvature strata and harmonic layers.
  \item Determine scene manifold $\mathcal{S}_t$.
  \item Lift narrative atoms to scene elements.
  \item Solve the harmonic scene equation on $\mathcal{S}_t$.
  \item Render canonical scenes with minimal curvature distortion.
\end{enumerate}

All steps are guaranteed to preserve curvature descent and quine stability.


\section{Long-Time Behaviour}
\label{sec:55:long-time}

As $t\to\infty$, scenes converge to harmonic limit forms.

\begin{theorem}[Scene Convergence]
Under RSVP flow, the scene manifold converges in the $C^\infty$ topology to
the canonical scene $\mathcal{S}_\infty$ associated with the universal quine
$\mathfrak{U}$.
\end{theorem}

Thus the visual universe collapses to the unique, minimal scene determined by
the universal media quine.


\section{Conclusion}
\label{sec:55:conclusion}

The Harmonic Scene Composer provides the continuous, geometric backbone of
synthetic visual media.  
Scenes are not painted; they emerge.  
Narrative is not scripted; it descends in curvature.  
The RSVP manifold becomes visible, harmonic, and interpretable in real time.

Chapter~\ref{ch:56} extends this curvature-driven synthesis to the auditory
domain through RSVP audio field generation.

