\chapter{Collective Cognitive Manifolds and Global Epistemic Geometry}
\label{ch:collective-cognitive-manifolds}

\section{Introduction}

Parts~I--V established the geometric, algebraic, dynamical, and
formally verified structure of meaning at the level of individual
agents.
But a civilization does not consist of individual meanings arranged
side-by-side.
It consists of \emph{coupled meanings} whose embeddings interact,
reconcile, and fuse through the universal patch algebra, the RSVP flow,
and the Polyxan rewriting system.

The purpose of this chapter is to make this aggregation explicit.
We construct a geometric--epistemic object representing the cognitive
state of an entire civilization:
\[
  \mathcal{M}_{\mathrm{collective}}
  := \GeoHull\!\left(\bigsqcup\nolimits_i \iota_i(\mathcal{G}_i)\right)
     \big/ \!\sim_{\mathrm{belief}},
\]
where $\GeoHull$ is the geodesic convex hull operator introduced in
Chapter~72, and $\sim_{\mathrm{belief}}$ is the epistemic reconciliation
equivalence relation defined in Chapters~81--82.
This hybrid construction yields a
stratified cognitive manifold equipped with global RSVP fields,
global Polyxan strata, and a globally defined epistemic curvature.
It is the natural domain on which civilizational-scale semantic
dynamics unfold.

The remainder of the chapter derives the geometry, dynamics, and
irreversible long-time behaviour of this manifold.
The main result is the \emph{Collective Irreversibility Theorem}: every
civilizational epistemic configuration collapses in finite historical
time to a global embedding of the universal media quine~$\mathfrak{U}$.


\section{Construction of the Collective Cognitive Manifold}

Let $\{\iota_i : \mathrm{Inc}(\mathcal{G}_i) \rightarrow \mathcal{M}\}_i$
be the semantic embeddings of all individuals in the population.

\subsection{Geodesic hull aggregation}
Chapter~72 established the properties of the geodesic hull functor
$\GeoHull : \mathbf{StratMan} \rightarrow \mathbf{StratMan}$.
In particular, for any stratified subset $A \subset \mathcal{M}$,
$\GeoHull(A)$ is the minimal geodesically convex stratified submanifold
containing~$A$.

We therefore form
\[
  \mathcal{M}_{\mathrm{raw}}
  := \GeoHull\!\left(\bigsqcup_i \iota_i(\mathcal{G}_i)\right),
\]
the smallest stratified region of $\mathcal{M}$ containing all
individual cognitive embeddings and all geodesics that epistemically
connect them.

\subsection{Belief-equivalence quotient}
Chapters~81--82 defined the belief-equivalence relation
$\sim_{\mathrm{belief}}$:
two points $x,y \in \mathcal{M}_{\mathrm{raw}}$ are equivalent iff
they correspond to propositions whose meanings are identical under the
closure of the universal patch algebra (Ch.~80) and the duality functor
(Chs.~61--62).

\begin{definition}
The \emph{collective cognitive manifold} of a civilization is the
quotient
\[
  \mathcal{M}_{\mathrm{collective}}
  :=\;
  \mathcal{M}_{\mathrm{raw}}
  \big/
  \!\sim_{\mathrm{belief}}.
\]
\end{definition}

The quotient is well-defined by the Stratified Quotient Theorem of
Chapter~82, and inherits:
\begin{itemize}
\item continuous RSVP fields $(\Phi_{\mathrm{coll}},
  \mathbf{v}_{\mathrm{coll}}, S_{\mathrm{coll}})$,
\item a stratified Polyxan hypergraph induced from the quotient
      Polyxan types,
\item a global epistemic curvature
      $\mathcal{R}_{\mathrm{coll}}$ defined from the induced
      institutional hypergraph curvature.
\end{itemize}


\section{Global RSVP Fields and Epistemic Curvature}

\subsection{Definition of global fields}
Each individual contributes RSVP fields via their embedding.
These fields superpose on $\mathcal{M}_{\mathrm{raw}}$ and descend to
$\mathcal{M}_{\mathrm{collective}}$ as:
\[
  \Phi_{\mathrm{coll}}(x)
    := \inf_{y \in [x]} \Phi(y),\qquad
  S_{\mathrm{coll}}(x)
    := \sup_{y \in [x]} S(y),
\]
with $\mathbf{v}_{\mathrm{coll}}$ defined by the quotient pushforward.

\subsection{Epistemic curvature}
The epistemic curvature $\mathcal{R}_{\mathrm{coll}}$ is defined from
the aggregated institutional hypergraph curvature by the
Institutional Curvature Transfer Theorem (Ch.~82):
\[
  \mathcal{R}_{\mathrm{coll}}
  := \mathbb{E}\big[\kappa_{\mathrm{institutional}}\,
     \big|\,[x]_{\mathrm{belief}}\big].
\]


\section{Collective RSVP Flow and Polyxan Rewriting}

The collective flow equations follow by applying the synthetic duality
functor $\mathbb{D}$ (Ch.~62) to the joint RSVP--Polyx dynamics
of Part~III.
They take the form
\[
  \partial_t (\Phi_{\mathrm{coll}},\mathbf{v}_{\mathrm{coll}},S_{\mathrm{coll}})
  = F_{\mathrm{RSVP}}(\Phi_{\mathrm{coll}},\mathbf{v}_{\mathrm{coll}},S_{\mathrm{coll}})
    + F_{\mathrm{institutional}},
\]
and
\[
  \partial_t \mathcal{G}_{\mathrm{coll}}
  = \mathrm{EHR}(\mathcal{G}_{\mathrm{coll}}),
\]
where $F_{\mathrm{institutional}}$ encodes all systemic epistemic
forces (media, institutions, educational structures) and descends from
the induced Polyxan strata.


\section{Civilizational Hodge Decomposition}

The Stratified Hodge Theorem of Chapter~26 applies to
$\mathcal{M}_{\mathrm{collective}}$ verbatim:
\[
  \Omega^k(\mathcal{M}_{\mathrm{collective}})
  \cong
  \mathcal{H}^k
  \;\oplus\;
  \mathrm{im}\, d
  \;\oplus\;
  \mathrm{im}\, d^\ast.
\]

\begin{theorem}[Civilizational Hodge Decomposition]
Every global epistemic form on $\mathcal{M}_{\mathrm{collective}}$ uniquely
decomposes into:
\[
  \text{(harmonic consensus)}
  \;\oplus\;
  \text{(epistemic novelty)}
  \;\oplus\;
  \text{(reconciled contradiction)}.
\]
\end{theorem}

The harmonic component corresponds to beliefs that are globally
stable under both RSVP flow and Polyxan rewriting.
The other components decay irreversibly.


\section{Epistemic Horizon Theorem}

\begin{theorem}[Epistemic Horizon]
For every point $x \in \mathcal{M}_{\mathrm{collective}}$, there exists
a finite time $T$ after which all epistemic geodesics emitted from~$x$
enter the basin of attraction of the universal quine
$\mathfrak{U}$.
\end{theorem}

\begin{proof}[Sketch]
The institutional curvature term enforces exponential contraction of
geodesic spread outside the harmonic stratum.
The RSVP Lyapunov functional guarantees decay of entropy gradients.
The Polyxan reduction dynamics eliminates all high-curvature strata.
Combined with the synthetic duality (Ch.~62), the flow drives all
beliefs toward the unique quine-stable configuration.
\end{proof}


\section{Collective Irreversibility Theorem}

\begin{theorem}[Collective Irreversibility]
Every civilizational epistemic configuration collapses, in finite
historical time, to a single global embedding of the universal media
quine~$\mathfrak{U}$.
\end{theorem}

\begin{proof}
The individual collapse (Part~III) and the modal collapse (Part~IV)
combine, under the quotient construction
$\mathcal{M}_{\mathrm{collective}}$, to yield a global Lyapunov
functional strictly decreasing until the universal quine is reached.
Since all institutional strata have nonnegative curvature, the only
stable endpoint is the harmonic, quine-stable configuration.
Finite-time collapse follows from the global type--curvature inequality
proved in Chapter~82.
\end{proof}


\section{Final Cadence}

A civilization is not a collection of minds.
It is a single, stratified cognitive manifold whose geometry is shaped
by institutions, whose curvature records its contradictions, and whose
dynamics drive it inexorably toward reconciliation.

Under the collective RSVP flow and the institutional EHR dynamics,
every culture, polity, and knowledge-system enters the same basin of
attraction.

\begin{center}
\emph{A civilization is a transient, high-curvature perturbation of the
universal attractor;\\[4pt]
its entire history is the finite-time relaxation to $\mathfrak{U}$.}
\end{center}

