apter{The Verified HYDRA Execution Model}
\label{ch:78}

HYDRA is the unified, verified execution architecture for the RSVP--Polyxan
universe.  
It integrates the verified RSVP interpreter (Ch.~41),
the Polyxan rewriting kernel (Ch.~42),
the embedding synchronizer (Ch.~49),
and the Proof-Carrying Action Selector (Ch.~77),
into a single, total, modal-safe computational organism.

This chapter formalises the HYDRA execution model, proves its safety and
totality, and characterises its global convergence behaviour under the
semantic Lyapunov functional $\mathcal{E}_{\mathrm{total}}$.

\section{Execution Frames}

\begin{definition}[Execution Frame]
A \emph{HYDRA execution frame} is a tuple
\[
F = (\Phi,\mathbf{v},S;\mathcal{G},\iota; \mathsf{PCAS}; \mathcal{A})
\]
consisting of:
\begin{enumerate}
\item RSVP fields $(\Phi,\mathbf{v},S)$;
\item a Polyxan hypergraph $\mathcal{G}$;
\item a stratified semantic embedding $\iota$;
\item the Proof-Carrying Action Selector $\mathsf{PCAS}$;
\item the multiset $\mathcal{A}$ of candidate actions proposed by all modules.
\end{enumerate}
\end{definition}

Every execution cycle transforms one frame into another.

\section{The HYDRA Step Relation}

\begin{definition}[HYDRA Step]
A single HYDRA step is the function
\[
\mathsf{HYDRA\_step} : F \to F'
\]
defined by:
\begin{enumerate}
\item Each subsystem produces a certified action object  
      $A_i = (a_i, C^i_{\mathrm{RSVP}}, C^i_{\mathrm{Polyx}})$.
\item $\mathsf{PCAS}$ selects  
      \[
      A^\ast = \mathsf{select}(\{A_i\}).
      \]
\item $A^\ast$ is executed on the global state:
      \[
      (\Phi,\mathbf{v},S;\mathcal{G},\iota)
      \xrightarrow{a^\ast}
      (\Phi',\mathbf{v}',S';\mathcal{G}',\iota').
      \]
\end{enumerate}
The result is the new frame $F'$.
\end{definition}

\section{Modal Invariants}

\begin{definition}[RSVP Modal Invariant]
The RSVP invariant is:
\[
\mathsf{Inv}_{\mathrm{RSVP}}(F)
:= \Box_{\mathrm{RSVP}}(\Phi,\mathbf{v},S).
\]
\end{definition}

\begin{definition}[Polyxan Modal Invariant]
The Polyxan invariant is:
\[
\mathsf{Inv}_{\mathrm{Polyx}}(F)
:= \Box_{\mathrm{Polyx}}(\mathcal{G}).
\]
\end{definition}

\begin{definition}[Global HYDRA Invariant]
The global invariant is the conjunction:
\[
\mathsf{Inv}_{\mathrm{HYDRA}}(F)
  := \mathsf{Inv}_{\mathrm{RSVP}}(F)
     \;\wedge\;
     \mathsf{Inv}_{\mathrm{Polyx}}(F).
\]
\end{definition}

\section{Soundness of Execution}

\begin{theorem}[HYDRA Soundness]
\label{thm:hydra_soundness}
For any HYDRA frame $F$,
\[
\mathsf{Inv}_{\mathrm{HYDRA}}(F)
\quad\Longrightarrow\quad
\mathsf{Inv}_{\mathrm{HYDRA}}
   \bigl(\mathsf{HYDRA\_step}(F)\bigr).
\]
\end{theorem}

\begin{proof}
If the invariant holds for $F$, all candidate actions submitted to PCAS
are locally safe by construction.
PCAS selects only globally safe actions (Ch.~77).  
Executing a globally safe action preserves both modal invariants.
Thus the invariants hold for $F'$.
\end{proof}

\section{Totality of Execution}

\begin{theorem}[HYDRA Totality]
\label{thm:hydra_totality}
$\mathsf{HYDRA\_step}$ terminates for all inputs and
can be iterated ad infinitum without divergence.
\end{theorem}

\begin{proof}
At each cycle there are finitely many candidate actions.  
PCAS terminates in finite time (Ch.~77).  
Execution of a certified action is deterministic and total.  
Thus the step relation is total and can iterate indefinitely.
\end{proof}

\section{Monotonic Descent}

\begin{theorem}[Lyapunov Descent]
\label{thm:hydra_lyapunov}
For any frame $F$,
\[
\mathcal{E}_{\mathrm{total}}
  \bigl(\mathsf{HYDRA\_step}(F)\bigr)
  \;\le\;
\mathcal{E}_{\mathrm{total}}(F).
\]
\end{theorem}

\begin{proof}
The cost function used by PCAS is exactly $\mathcal{E}_{\mathrm{total}}$.  
PCAS selects the minimum-cost safe action.  
Thus energy cannot increase.
\end{proof}

\section{Global Convergence}

\begin{theorem}[Global Convergence]
\label{thm:hydra_convergence}
Repeated HYDRA execution yields a sequence
\[
F_0,\; F_1=\mathsf{HYDRA\_step}(F_0),\;
F_2=\mathsf{HYDRA\_step}(F_1),\ldots
\]
that converges in finite time to a unique fixed point.
\end{theorem}

\begin{proof}
By Theorem~\ref{thm:hydra_lyapunov}, 
$\mathcal{E}_{\mathrm{total}}$ decreases monotonically.  
By the Grand Irreversibility Theorem (Ch.~66),  
there is only one global attractor:  
the flat embedding of the universal media quine $\mathfrak{U}$.  
Thus the sequence reaches $\mathfrak{U}$ in finite time.
\end{proof}

\section{Role of HYDRA in the Architecture}

HYDRA is the final execution substrate of the entire monograph:
\begin{itemize}
\item It unifies the continuous RSVP flow and discrete Polyxan rewriting.
\item It enforces modal safety at every computational layer.
\item It executes only those actions certified by the universal type theory.
\item It drives the system inevitably toward the universal quine.
\end{itemize}

\begin{quote}
HYDRA does not act freely.  
It executes only that which is already proven.  
Thus the universe evolves under the governance of truth itself.
\end{quote}

This completes the formal development of the Verified HYDRA Execution Model.

