\chapter{Semantic Patch Algebra Implementation}
\label{ch:52}

This chapter realises the Semantic Patch Algebra introduced abstractly in
Chapter~79 as an executable, verified subsystem of the RSVP--Polyxan
architecture.  
The goal is to construct a fully operational implementation of the algebra
$(\mathcal{P},\otimes,1)$, together with normalisation routines, modal safety
guards, and compilers that translate patches into verified rewrite programs.

The Semantic Patch Algebra is the mechanism by which distributed or concurrent
semantic artefacts --- posts, narratives, embeddings, hypergraphs, media frames,
or RSVP field segments --- are reconciled into a single, curvature-consistent
meaning object.  
The implementation given here is the canonical realisation of that algebra.


\section{Patch Objects and Their Structural Fields}
\label{sec:52:patch-objects}

A patch is the primitive semantic delta from which reconciliation is built.

\begin{definition}[Patch Object]
A \emph{patch} is a tuple
\[
P = (\Delta\mathcal{G},\Delta\iota,\Delta\Phi,\Delta \mathbf{v},\Delta S)
\]
consisting of:
\begin{enumerate}
  \item a finite Polyxan hypergraph increment $\Delta\mathcal{G}$,
  \item a stratified embedding increment $\Delta\iota$,
  \item RSVP field increments $(\Delta\Phi,\Delta \mathbf{v},\Delta S)$,
\end{enumerate}
satisfying the curvature-matching and quine-consistency invariants of
Chapter~79.
\end{definition}

Patches therefore represent consistent semantic updates across both the discrete
and continuous layers of the system.


\section{Patch Composition}
\label{sec:52:patch-composition}

The monoidal product $\otimes$ defines the fundamental composition law for
patches.

\begin{definition}[Patch Product]
Given patches $P_1$ and $P_2$, their product is
\[
P_1 \otimes P_2 :=
\bigl(
\Delta\mathcal{G}_1 \uplus \Delta\mathcal{G}_2,
\;
\Delta\iota_1 \circledast \Delta\iota_2,
\;
\Delta\Phi_1+\Delta\Phi_2,
\;
\Delta\mathbf{v}_1+\Delta\mathbf{v}_2,
\;
\Delta S_1+\Delta S_2
\bigr),
\]
where each component operation is curvature-respecting and quine-preserving.
\end{definition}

\begin{theorem}[Associativity]
\[
(P_1\otimes P_2)\otimes P_3 = P_1 \otimes (P_2 \otimes P_3).
\]
\end{theorem}

\begin{proof}
Set-theoretic disjoint union is associative; stratified embedding merge
$\circledast$ is associative by Chapter~32; field increments add associatively.
\end{proof}

\begin{theorem}[Unit Patch]
There exists a unique patch $1$ such that
\[
1\otimes P = P = P\otimes 1.
\]
\end{theorem}

The unit patch represents the zero semantic delta.


\section{Patch Normalisation}
\label{sec:52:normalisation}

Normalisation ensures that arbitrary compositions of patches collapse to a
canonical representative, guaranteeing convergence and preventing semantic drift.

\begin{definition}[Normal Form]
A patch $P$ is in \emph{normal form} if:
\begin{enumerate}
  \item $\Delta\mathcal{G}$ is in EHR-normal form,
  \item $\Delta\iota$ is harmonic,
  \item $(\Delta\Phi,\Delta\mathbf{v},\Delta S)$ minimise local Lyapunov energy.
\end{enumerate}
\end{definition}

\begin{theorem}[Patch Normalisation]
Every patch admits a unique normal form $\mathrm{NF}(P)$, and
\[
\mathrm{NF}(P_1 \otimes P_2)
  = \mathrm{NF}\bigl(\mathrm{NF}(P_1)\otimes \mathrm{NF}(P_2)\bigr).
\]
\end{theorem}

\begin{proof}
Uniqueness follows from confluence of EHR rewriting (Chapter~42), uniqueness of
harmonic embeddings (Chapter~32), and strict convexity of the Lyapunov energy
functional (Chapter~12).
\end{proof}


\section{Patch Compiler}
\label{sec:52:compiler}

The patch compiler translates a patch into a verified rewrite program.

\begin{definition}[Patch Compiler]
The patch compiler
\[
\mathsf{compile}_{\mathcal{P}} : \mathcal{P} \longrightarrow \Lambda_{\mathrm{syn}}
\]
maps a patch to a finite sequence of EHR rewrites and RSVP field updates,
each annotated with modal correctness certificates.
\end{definition}

\begin{theorem}[Compiler Correctness]
If $\Pi = \mathsf{compile}_{\mathcal{P}}(P)$ then executing $\Pi$ yields a
semantic state differing from the original by exactly $P$, and satisfies
\[
\Box_{\mathrm{RSVP}} \land \Box_{\mathrm{Polyx}}.
\]
\end{theorem}


\section{Patch Algebra Engine}
\label{sec:52:engine}

The Patch Algebra Engine (PAE) is the executable system implementing
$(\mathcal{P},\otimes,1)$.

\begin{definition}[Patch Algebra Engine]
The PAE is a total, deterministic function
\[
\mathsf{apply\_patch} :
(\mathrm{State},\mathcal{P}) \longrightarrow \mathrm{State}
\]
that preserves harmonicity, quine stability, and curvature descent.
\end{definition}

\begin{theorem}[Totality]
$\mathsf{apply\_patch}$ is total and terminates for all inputs.
\end{theorem}

\begin{proof}
EHR rewriting terminates (Chapter~42), harmonic RSVP flow is finite-time
convergent (Chapter~31), and all modal guards are terminating predicates.
\end{proof}

\begin{theorem}[Soundness]
Applying a patch never increases curvature or breaks quine stability.
\end{theorem}

\begin{proof}
Patch normalisation ensures all patch objects are curvature-decreasing and
quine-preserving; modal certificates guarantee that all intermediate states
respect $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.
\end{proof}


\section{Global Reconciliation via Patch Algebra}
\label{sec:52:global-reconciliation}

The Semantic Patch Algebra is the reconciliation substrate for the entire
RSVP--Polyxan ecosystem: all distributed updates, local edits, user-level
semantic artefacts, and media flows reduce to patch composition followed by
normalisation.

\begin{theorem}[Global Convergence]
Any finite set of patches $\{P_i\}$ reduces to a unique global reconciliation:
\[
\mathrm{NF}(P_1\otimes\cdots\otimes P_n).
\]
\end{theorem}

\begin{proof}
Associativity and confluence yield a unique normal form.  
\end{proof}

\begin{quote}
\textit{Through patch algebra, a universe of fragments becomes one meaning.  
Every delta descends.  
Every narrative converges.}
\end{quote}


\section{Conclusion}
\label{sec:52:conclusion}

This chapter has constructed the executable form of the Semantic Patch Algebra:
patch objects, patch composition, normalisation, compilation, and the global
reconciliation engine.  
It is the algebraic backbone of the entire implementation layer.

With this machinery in place, we now turn to the real-time generative substrates
of synthetic media (Chapter~53).

