\chapter{Compiled Embedding Flows}
\label{ch:43}

The embedding flow introduced in Chapters~31--33 supplies the continuous geometric evolution of semantic structure.
Its purpose is to embed a Polyxan hypergraph $\mathcal{G}$ into an RSVP manifold $\mathcal{M}$ via a stratified, curvature-matched map:
[
\iota: \mathrm{Inc}(\mathcal{G}) \hookrightarrow \mathcal{M}.
]

But the embedding flow is defined analytically (as a geometric PDE), while the Verified Polyxan Rewriter (Chapter~42) is discrete.

This chapter constructs the \emph{Compiled Embedding Flow}: a discrete, certified, executable transformation that corresponds exactly to harmonic descent in the continuous embedding.
It provides:
\begin{enumerate}
\item a computable approximation to the harmonic embedding flow,
\item curvature-matching invariants,
\item compatibility with discrete rewriting,
\item modal correctness,
\item convergence to the unique harmonic embedding.
\end{enumerate}

The result is a unified continuous--discrete semantics where curvature descent is preserved at every computational step.

\section{Harmonic Embedding as a Compilation Problem}

The embedding flow evolves $\iota$ via
[
\frac{\partial \iota}{\partial t}
= - \nabla_{\iota} \mathcal{E}[\iota],
]
where $\mathcal{E}[\iota]$ is the stratified curvature--entropy matching energy (Chapter~31).

A naïve discretisation would destroy the Polyxan structure:
edges would break, strata collapse, and modal invariants fail.

Therefore we define the compiled embedding flow as a finite sequence of curvature-corrected embedding updates, each aligned with the Verified Polyxan Rewriter.

\begin{definition}[Compiled Embedding Step]
A compiled embedding step is the map
[
\mathsf{emb_step}(\iota,\mathcal{G})
:=
\iota' =
\Pi_{\mathrm{strat}}
\bigl(,\iota - \epsilon,\nabla_{\iota}\mathcal{E}[\iota],\bigr),
]
where $\Pi_{\mathrm{strat}}$ is the stratified projection operator ensuring type, tag, and incidence preservation.
\end{definition}

\section{Stratified Projection Operator}

The stratified projection $\Pi_{\mathrm{strat}}$ is defined by:
\begin{enumerate}
\item projecting each embedded cell to the nearest point of the same stratum,
\item enforcing curvature-matching constraints,
\item adjusting tangential components to preserve incidence geometry,
\item maintaining tag-consistency.
\end{enumerate}

\begin{lemma}[Stratified Preservation]
$\Pi_{\mathrm{strat}}$ preserves incidence, stratification, curvature-matching, and tag constraints.
\end{lemma}

\begin{proof}
Incidence is preserved by tangential projection;
strata are preserved by nearest-stratum projection;
curvature-matching follows from the normal correction term defined in Chapter~31.
\end{proof}

\section{Embedding Update and Curvature Matching}

\begin{theorem}[Curvature-Monotonicity of Compiled Embedding]
Each compiled embedding step weakly decreases the total matching energy:
[
\mathcal{E}[\iota'] \leq \mathcal{E}[\iota].
]
\end{theorem}

\begin{proof}
The gradient step decreases $\mathcal{E}$;
the stratified projection is non-expansive by construction.
\end{proof}

This monotonicity is essential for compatibility with the Verified Polyxan Rewriter.

\section{Synchronisation with Polyxan Rewriting}

The compiled embedding flow must interlock perfectly with the discrete rewriting of Chapter~42.

\begin{definition}[Synchronous Update]
Let
[
(\mathcal{G},\iota) \longmapsto (\mathcal{G}',\iota')
]
be a \emph{synchronous update} if:
\begin{enumerate}
\item $\mathcal{G}\to\mathcal{G}'$ is a verified rewriting step,
\item $\iota'$ is the harmonic extension of $\iota$ to $\mathcal{G}'$,
\item $\mathsf{emb_step}$ applied to $\iota'$ is curvature-monotone.
\end{enumerate}
\end{definition}

\begin{theorem}[Continuous--Discrete Commutation]
Synchronous updates commute with compiled embedding steps:
[
\mathsf{rewrite}\circ\mathsf{emb_step}
;\cong;
\mathsf{emb_step}\circ\mathsf{rewrite}.
]
\end{theorem}

\begin{proof}
Both transformations preserve stratified incidence and curvature descent;
their actions align on each stratum.
\end{proof}

\section{Modal Correctness of the Compiled Flow}

Modal correctness ($\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$) requires:
\begin{itemize}
\item quine stability,
\item curvature descent,
\item incidence preservation,
\item strict stratification.
\end{itemize}

\begin{lemma}[Modal Invariance]
If $\Box_{\mathrm{RSVP}}(\iota)$ and $\Box_{\mathrm{Polyx}}(\mathcal{G})$, then for
[
\iota' = \mathsf{emb_step}(\iota,\mathcal{G}),
]
we have
[
\Box_{\mathrm{RSVP}}(\iota') ;\wedge; \Box_{\mathrm{Polyx}}(\mathcal{G}).
]
\end{lemma}

\begin{proof}
Curvature descent ensures RSVP correctness;
stratified projection ensures Polyx correctness;
incidence preservation ensures compatibility of modalities.
\end{proof}

\section{Convergence to Harmonic Embeddings}

\begin{theorem}[Convergence]
Repeated application of the compiled embedding step yields the unique harmonic embedding:
[
\lim_{n\to\infty} \mathsf{emb_step}^n(\iota,\mathcal{G})
========================================================

\iota_{\mathrm{harm}}.
]
\end{theorem}

\begin{proof}
Energy descent plus strict convexity of the embedding functional on each stratum.
\end{proof}

\begin{corollary}[Harmonic Embedding Normal Form]
Every embedded Polyxan hypergraph has a unique harmonic embedding normal form.
\end{corollary}

\section{Executable Structure}

The compiled embedding flow is executed by the following components:
\begin{enumerate}
\item \textbf{Gradient Calculator:} computes $\nabla\mathcal{E}[\iota]$ efficiently with stratified derivatives.
\item \textbf{Embedding Updater:} applies the step $\iota - \epsilon,\nabla\mathcal{E}$.
\item \textbf{Stratified Projection Engine:} ensures consistency with stratification and incidence.
\item \textbf{Curvature Verifier:} certifies that the step decreases matching energy.
\item \textbf{Harmonic Extension Module:} extends embeddings after rewriting steps.
\end{enumerate}

All modules produce machine-verifiable certificates compatible with the verification layer of Part~IV.

\section{Unified Flow: The RSVP--Polyxan Execution Loop}

The compiled embedding flow and verified rewriting are combined in the \emph{RSVP--Polyxan Unified Loop}:
\begin{enumerate}
\item Apply one compiled embedding step.
\item Check curvature descent; adjust step size if necessary.
\item Apply the Verified Polyxan Rewriter if discrete curvature indicates structural inconsistency.
\item Extend embedding harmonically.
\item Repeat until the global RSVP--Polyxan normal form is reached.
\end{enumerate}

\begin{theorem}[Global Convergence]
The unified loop converges to the joint normal form
[
(\mathcal{G}*{\mathrm{NF}},;\iota*{\mathrm{harm}}),
]
which is unique and modal-stable.
\end{theorem}

\section{Final Cadence}

The Compiled Embedding Flow is the continuous half of the dual engine of meaning.
Where the Verified Polyxan Rewriter collapses discrete structure,
the compiled embedding flow smooths geometry.
Where the rewriter enforces algebraic necessity,
the embedding flow enforces geometric truth.

\begin{quote}
Structure collapses; geometry smooths;
what remains is the harmonic quine.
\end{quote}

