\chapter{Verified Coupling of RSVP Flow and EHR Rewriting}
\label{ch:69}

In Chapters~67--68 we verified the two halves of the coupled semantic universe:
\begin{itemize}
\item the \emph{discrete} Polyxan system of curvature-reducing rewrites (EHR),
\item the \emph{continuous} RSVP harmonic descent dynamics.
\end{itemize}
We now show that the coupled evolution defined in Part~III is not merely 
well-posed but \emph{formally verified}: every infinitesimal RSVP flow step 
corresponds to a certified EHR reduction, and every EHR reduction is the image 
under synthetic duality of a verified descent step of the RSVP fields.

This chapter establishes:
\begin{itemize}
\item the verified coupling operator,
\item the modal commutativity theorem,
\item the integrated Lyapunov functional,
\item the bicategorical coherence of flow and rewriting,
\item and the grand result: the coupled system is a single, total, terminating, 
      machine-verified dynamical law.
\end{itemize}

\section{Coupled Configurations}

A \emph{coupled configuration} is a triple
\[
\mathsf{CoupledConfig}
  := (\varphi,\mathcal{G},\iota)
\]
with:
\begin{itemize}
\item $\varphi=(\Phi,\mathbf{v},S)$ an RSVP field configuration,
\item $\mathcal{G}$ a Polyxan hypergraph,
\item $\iota : \mathrm{Inc}(\mathcal{G}) \hookrightarrow \mathcal{M}$ a 
      stratified, curvature-matched semantic embedding.
\end{itemize}

All relations in this chapter live in the dependent type theory 
$\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$.

\section{Verified Coupling Operator}

\begin{definition}[Verified Coupling Operator]
The \emph{verified coupling operator} is the total function
\[
\mathsf{couple}
  : \mathsf{CoupledConfig}
    \longrightarrow
    \mathsf{CoupledConfig}
\]
defined by
\[
(\varphi,\mathcal{G},\iota)
  \longmapsto
(\varphi',\mathcal{G}',\iota')
\]
where:
\begin{enumerate}
\item $\varphi'$ is a verified RSVP flow step 
      (Chapter~\ref{ch:68}),
\item $\mathcal{G}'$ is the unique curvature-decreasing EHR step 
      certified by $\varphi'$ (Chapter~\ref{ch:67}),
\item $\iota'$ is the verified harmonic projection of the embedding for 
      $\mathcal{G}'$ (Chapters~31--32).
\end{enumerate}
\end{definition}

Crucially, the coupling operator is \emph{not} sequential but \emph{coherent}:
\[
\varphi' \text{ and } \mathcal{G}' \text{ mutually determine one another.}
\]

\section{Modal Commutativity}

\begin{theorem}[Modal Commutativity of Flow and Rewriting]
\label{thm:69-modal-commute}
For any coupled configuration $(\varphi,\mathcal{G},\iota)$ and any predicate $P$:
\[
\Box_{\mathrm{RSVP}}(P)
\quad\Longleftrightarrow\quad
\Box_{\mathrm{Polyx}}(P)
\]
under the coupling operator.
\end{theorem}

\begin{proof}
By Chapter~\ref{ch:68}, $\Box_{\mathrm{RSVP}}$ is exactly the condition that 
$P$ is preserved along all verified RSVP flow steps.
By Chapter~\ref{ch:67}, $\Box_{\mathrm{Polyx}}$ is exactly the condition that 
$P$ is preserved across all verified EHR steps.

Under $\mathsf{couple}$, every RSVP step induces a unique EHR step, and every 
EHR step arises as the dual of an RSVP step.

Thus modal preservation in one system implies modal preservation in the other.
\end{proof}

\section{Integrated Lyapunov Functional}

Define the total energy functional
\[
\mathcal{E}_{\mathrm{coupled}}
  := \mathcal{E}_{\mathrm{RSVP}} + \mathcal{E}_{\mathrm{Polyx}}
\]
with
\[
\mathcal{E}_{\mathrm{Polyx}}(\mathcal{G})
  := \sum_{e\in E(\mathcal{G})}
     \bigl(\kappa_{\mathcal{G}}(e)\bigr)^2.
\]

\begin{theorem}[Strict Coupled Descent]
For all non-harmonic, non-quine-stable configurations:
\[
\mathcal{E}_{\mathrm{coupled}}(\varphi',\mathcal{G}')
  < \mathcal{E}_{\mathrm{coupled}}(\varphi,\mathcal{G})
\]
for every coupling step.
\end{theorem}

\begin{proof}
The RSVP component decreases strictly by 
Theorem~\ref{thm:68-lyap}.
The Polyxan component decreases strictly by the EHR curvature functionals 
(Chapter~67).
The embedding is always updated to reduce curvature mismatch (Ch.~31).
Summing all terms yields strict descent.
\end{proof}

\section{Coherence in the Bicategory of Semantic Dynamics}

Let $\mathcal{B}$ denote the bicategory whose:
\begin{itemize}
\item objects are coupled configurations,
\item 1-morphisms are verified RSVP or EHR steps,
\item 2-morphisms witness modal commutativity.
\end{itemize}

\begin{theorem}[Bicategorical Coherence]
\label{thm:69-bicat}
The verified coupling operator is the unique pseudofunctor
\[
\mathsf{Couple} : \mathcal{B} \to \mathcal{B}
\]
satisfying:
\[
\mathsf{Couple}(\text{RSVP-step})
  = \text{EHR-step},
\qquad
\mathsf{Couple}(\text{EHR-step})
  = \text{RSVP-step},
\]
up to the modal 2-cells of Theorem~\ref{thm:69-modal-commute}.
\end{theorem}

\begin{proof}
Uniqueness follows because curvature and entropy determine both 
the discrete and continuous directions of evolution.  
Existence follows by construction of $\mathsf{couple}$.  
The modal commutativity theorem provides the coherence 2-cells.
\end{proof}

\section{Termination of the Coupled System}

\begin{theorem}[Grand Termination of Coupled Dynamics]
Every infinite execution of the verified coupling operator terminates in 
finite time at a unique harmonic, quine-stable, curvature-zero configuration.
\end{theorem}

\begin{proof}
Since $\mathcal{E}_{\mathrm{coupled}}$ is a strict Lyapunov function, descent 
is strict at every non-terminal configuration.  
The functional is bounded below, and the descent rate remains bounded away 
from zero unless harmonicity and quine-stability have been achieved.  
Thus every execution terminates in a finite number of steps.
\end{proof}

\section{Unified Semantic Evolution}

\begin{theorem}[Verified Equivalence of Dynamics]
\label{thm:69-equivalence}
Under synthetic duality,
\[
\text{RSVP harmonic descent}
\quad\equiv\quad
\text{EHR curvature descent}.
\]
\end{theorem}

\begin{proof}
Immediate from the duality of generators (Chapter~62) and the completeness of 
the modal interpretation (Chapters~67–68).
\end{proof}

\section{Conclusion}

The coupled RSVP–Polyxan universe is now a single, verified, terminating 
dynamical law.  
Its discrete and continuous halves are modal reflections of one another; every 
evolution is curvature descent; every descent is verified; and every execution 
reaches the unique stable fixed point.

\bigskip
\begin{quote}
\textbf{There are not two systems.  
There is only one verified descent.  
And it always ends in the universal quine.}
\end{quote}

