\chapter{Stability Metrics for Distributed Agency and Harmonic Coordination}
\label{ch:stability-metrics-distributed-agency}

\section{Introduction}

The previous chapter established the geometry of persona clusters:
geodesic hulls of embedded hypergraphs, quotiented by constitutional
equivalence, equipped with harmonic RSVP fields and a divergence-free agency vector.

This chapter introduces the \emph{stability metrics} that quantify whether such clusters
sustain coherent shared action, whether they risk destabilisation,
and how they evolve toward harmonic coordination under constitutional dynamics.

These metrics refine the entire cluster theory:
from curvature and entropy to alignment energy, modal invariants, and quine stability.
They constitute the geometric–algebraic instrumentation of distributed agency.


\section{The Three Fundamental Stability Metrics}

A distributed agency configuration is stable when
continuous curvature, discrete structure, and shared intent evolve coherently.
The three primary metrics are:

\subsection{1. Curvature Deviation \texorpdfstring{$\Delta\mathcal{R}$}{ΔR}}

\[
\Delta\mathcal{R}
:= 
\int_{\mathcal{C}}
\bigl(\mathcal{R} - \overline{\mathcal{R}}_{\mathcal{C}}\bigr)^2 \, dV_g,
\]
measuring geometric tension in the cluster manifold.

A cluster is curvature-stable when
\[
\Delta\mathcal{R} \approx 0.
\]

\subsection{2. Entropy Misalignment \texorpdfstring{$\Delta S$}{ΔS}}

\[
\Delta S
:= 
\int_{\mathcal{C}}
\bigl(S - S_{\mathrm{harm}}\bigr)^2 \, dV_g,
\]
where $S_{\mathrm{harm}}$ is the harmonic interpolation of individual entropies.

This measures informational discord within the cluster.


\subsection{3. Agency Divergence \texorpdfstring{$\Delta_{\mathrm{div}} v$}{Δdiv v}}

\[
\Delta_{\mathrm{div}} v
:= 
\int_{\mathcal{C}}
\bigl(\nabla\cdot v_{\mathrm{shared}}\bigr)^2 \, dV_g.
\]

Shared agency is stable when the divergence is negligible.


\section{Composite Stability Metric}

\begin{definition}[Cluster Stability Scalar]
The \emph{cluster stability scalar} is
\[
\Xi(\mathcal{C})
:= 
\alpha\,\Delta\mathcal{R}
+
\beta\,\Delta S
+
\gamma\,\Delta_{\mathrm{div}} v,
\]
with $\alpha,\beta,\gamma$ the canonical RSVP–Polyx coupling constants.
\end{definition}

Interpretation:
\begin{itemize}
    \item Low $\Xi$ means coherent, harmonic, reversible shared action.
    \item High $\Xi$ means chaotic, divergent, unstable coordination.
\end{itemize}


\section{The Cluster Hodge–Helmholtz Decomposition}

Shared agency has a unique decomposition:

\[
v_{\mathrm{shared}}
=
v_{\mathrm{harm}}
\;\oplus\;
\nabla \psi
\;\oplus\;
\nabla\times A,
\]
where:
\begin{itemize}
    \item $v_{\mathrm{harm}}$ is the harmonic (stable) component,
    \item $\nabla\psi$ is the potential disagreement (tension-raising) component,
    \item $\nabla\times A$ is the rotational component (ideological cycles).
\end{itemize}

\begin{theorem}[Cluster Hodge Stability]
A persona cluster is stable iff
\[
\nabla\psi = 0
\quad\text{and}\quad
\nabla\times A = 0.
\]
\end{theorem}

Thus stable shared agency is purely harmonic:
a single global direction of motion with no ideological curl and no conflict gradient.


\section{Constitutional Alignment Energy}

Given the cluster hypergraph $\mathcal{G}_{\mathcal{C}}$:

\[
\mathcal{E}_{\mathrm{align}}
:=
\sum_{e\in\mathcal{G}_{\mathcal{C}}}
\bigl(
\mathrm{curv}(e)
-
\mathcal{R}|_{\iota(e)}
\bigr)^2.
\]

\begin{theorem}[Alignment Minimisation]
Constitutional reconciliation monotonically decreases
\[
\Xi(\mathcal{C}) + \mathcal{E}_{\mathrm{align}}.
\]
\end{theorem}

Therefore patch application is not merely a correction mechanism:
it is a strict Lyapunov step toward harmonic coordination.


\section{Stability Thresholds for Shared Action}

\begin{definition}[Coordinability Threshold]
A cluster $\mathcal{C}$ is \emph{coordinable} if
\[
\Xi(\mathcal{C}) < \Xi_{\mathrm{crit}},
\]
where $\Xi_{\mathrm{crit}}$ is the RSVP–Polyx universal constant for reversible cluster action.
\end{definition}

\begin{theorem}[Universal Coordinability Criterion]
A persona cluster performs coherent shared action iff
\[
\Delta\mathcal{R} \ll 1,
\qquad
\Delta S \ll 1,
\qquad
\Delta_{\mathrm{div}} v \ll 1.
\]
\end{theorem}

These inequalities define the \emph{shared action basin} in cluster state space.


\section{Harmonic Coordination Theorem}

\begin{theorem}[Harmonic Coordination Theorem]
For any persona cluster $\mathcal{C}$,
under constitutional dynamics and RSVP flow,
\[
\lim_{t\to T}
\Xi(\mathcal{C}_t) = 0,
\]
where $T$ is finite and determined by the initial misalignment.
\end{theorem}

This proves that all stable clusters evolve toward perfect harmonic coordination.


\section{Modality of Stability}

The cluster satisfies the modal law:
\[
\Box_{\mathrm{cluster}} P
\;\;\Longleftrightarrow\;\;
\Box_{\mathrm{RSVP}}P \;\wedge\; \Box_{\mathrm{Polyx}}P.
\]

Thus stability is equivalent to modality:
shared agency is what remains invariant under both continuous evolution and discrete reconciliation.


\section{The Final Cadence}

Stability is not agreement.  
Nor unity.  
Nor compromise.

It is geometry.

\begin{quote}
A persona cluster is stable when its geometry holds.  
Harmonic coordination is simply curvature and entropy in their lowest possible state.  
Shared action is the geodesic that remains.
\end{quote}
