\chapter{The Embedding Flow Synchronizer}
\label{ch:49}

The Embedding Flow Synchronizer (EFS) is the coordination mechanism that couples
continuous RSVP field dynamics with discrete Polyxan structural evolution.
Where the RSVP Flow Engine (Chapter~47) governs smooth geometric descent, and
the Polyxan Structural Kernel (Chapter~48) governs curvature-reducing
hypergraph rewriting, the Synchronizer ensures that both engines update in
lockstep, preserving the curvature--entropy matching constraints of
Chapter~31 and the modal invariants of Part~IV.

EFS is therefore the temporal and structural glue of the HYDRA architecture:
it ensures that the curve traced by the embedding
\[
\iota_t : \mathrm{Inc}(\mathcal{G}_t) \longrightarrow \mathcal{M}_t
\]
remains harmonic, stratified, and duality-consistent at every time step.

\section{Synchronized State Space}

The synchronizer operates on the fused state
\[
\Xi_t \;=\; (\Phi_t, \mathbf{v}_t, S_t;\, \mathcal{G}_t, \iota_t),
\]
where $(\Phi_t, \mathbf{v}_t, S_t)$ are RSVP fields,
$\mathcal{G}_t$ is the Polyxan hypergraph, and
$\iota_t$ is its semantic embedding.

\begin{definition}[Synchronized Configuration]
A configuration $\Xi_t$ is \emph{synchronized} if it satisfies:
\begin{enumerate}
  \item \textbf{Curvature Matching}:
  \[
  \mathcal{R}_t \circ \iota_t = \mathcal{H}_0[\mathcal{G}_t].
  \]
  \item \textbf{Entropy Alignment}:
  \[
  S_t \circ \iota_t = \mathcal{H}_{\mathrm{tag}}[\mathcal{G}_t].
  \]
  \item \textbf{Modal Stability}:
  \[
  \Box_{\mathrm{RSVP}}(\Phi_t,\mathbf{v}_t,S_t)
  \quad\text{and}\quad
  \Box_{\mathrm{Polyx}}(\mathcal{G}_t).
  \]
\end{enumerate}
\end{definition}

These invariants are necessary and sufficient for duality correctness
(Chapter~62) and for well-posed evolution under both RFE and PSK
(Chapters~47--48).

\section{The Synchronization Operator}

\begin{definition}[Embedding Flow Synchronizer]
The synchronizer is the operator
\[
\mathsf{EFS} : \Xi_t \longrightarrow \Xi_{t+\Delta t}
\]
defined by the composite update
\[
\Xi_{t+\Delta t}
=
\mathsf{ReEmbed}
\Bigl(
  \mathsf{PSK}
  \bigl(
    \mathsf{RFE}(\Xi_t)
  \bigr)
\Bigr),
\]
where:
\begin{itemize}
  \item $\mathsf{RFE}$ performs one continuous RSVP flow step,
  \item $\mathsf{PSK}$ performs one discrete structural reduction,
  \item $\mathsf{ReEmbed}$ recomputes a harmonic embedding satisfying the
        curvature--entropy constraints.
\end{itemize}
\end{definition}

The call to $\mathsf{ReEmbed}$ is essential: changes in discrete structure
induced by PSK alter the incidence complex, and the embedding must be
recomputed to maintain stratified harmonicity and modal safety.

\section{Harmonic Re-embedding}

\begin{theorem}[Harmonic Re-embedding Theorem]
\label{thm:reembed}
For any synchronized state $\Xi_t$, the re-embedding
\[
\iota_{t+\Delta t} \;=\; \mathsf{ReEmbed}(\mathcal{G}_{t+\Delta t})
\]
exists, is unique up to diffeomorphism of the elliptic sector, and satisfies:
\[
\mathcal{R}\circ\iota_{t+\Delta t}
=
\mathcal{H}_0[\mathcal{G}_{t+\Delta t}],
\qquad
S\circ\iota_{t+\Delta t}
=
\mathcal{H}_{\mathrm{tag}}[\mathcal{G}_{t+\Delta t}].
\]
\end{theorem}

\begin{proof}
Existence: by the Embedding Existence and Regularity Theorem of
Chapter~31.

Uniqueness: by harmonicity (Ch.~32), uniqueness of the RSVP elliptic vacuum
on each stratum (Ch.~13), and uniqueness of curvature descent (Ch.~32).

The curvature--entropy conditions hold by construction. ∎
\end{proof}

Thus harmonicity is a maintained invariant under EFS.

\section{Curvature-Coherent Scheduling}

The synchronizer must ensure that RFE and PSK remain curvature coherent.

\begin{theorem}[Curvature-Coherent Scheduling]
\label{thm:curvature-schedule}
For any synchronized $\Xi_t$,
\[
\mathsf{PSK}\circ\mathsf{RFE}
\;\simeq_{\mathbb{D}}\;
\mathsf{RFE}\circ\mathsf{PSK},
\]
and EFS selects the temporal ordering that produces the maximal curvature
descent at each step.
\end{theorem}

\begin{proof}
Commutation up to synthetic duality is proved in Chapter~48.
EFS schedules the two operations in the order that yields maximal decrease of
total curvature
\[
\mathcal{C}_t = \int_{\mathcal{M}_t} |\mathcal{R}_t|\, dV + \kappa(\mathcal{G}_t),
\]
which is a Lyapunov function by Chapter~32. ∎
\end{proof}

\section{Modal Safety of Synchronization}

\begin{theorem}[Modal Safety of EFS]
\[
\Box_{\mathrm{RSVP}}(\Xi_t) \wedge \Box_{\mathrm{Polyx}}(\Xi_t)
\;\Longrightarrow\;
\Box_{\mathrm{RSVP}}(\Xi_{t+\Delta t})
\wedge
\Box_{\mathrm{Polyx}}(\Xi_{t+\Delta t}).
\]
\end{theorem}

\begin{proof}
RFE preserves $\Box_{\mathrm{RSVP}}$ (Ch.~47).
PSK preserves $\Box_{\mathrm{Polyx}}$ (Ch.~48).
Re-embedding preserves both modalities by stratified soundness (Ch.~39).
Composition preserves modal validity. ∎
\end{proof}

Thus EFS is the modal guardian of the HYDRA update cycle.

\section{Global Convergence Under Synchronization}

EFS is not merely a coordinator; it reinforces the global convergence law of
Part~III.

\begin{theorem}[Synchronized Irreversibility]
\label{thm:synch-irreversible}
Iterating EFS on any initial configuration yields:
\[
\exists T < \infty :\quad
\Xi_T \text{ is a semantic galaxy},
\]
and subsequently,
\[
\Xi_{T'} \to \mathfrak{U}_{\mathrm{embed}}
\qquad
\text{as } T' \to \infty.
\]
\end{theorem}

\begin{proof}
By Chapter~32, the RSVP flow drives continuous fields toward a galaxy.
By Chapter~48, PSK drives discrete structure toward a quine.
By Chapter~31, the embedding flows toward harmonicity.
EFS composes these flows monotonically in the total central charge
$c_{\mathrm{total}}$.
Finite-time galaxy formation follows (Ch.~32), and convergence to the
universal quine follows (Ch.~25, 36). ∎
\end{proof}

\section{Final Cadence}

\begin{quote}
The Embedding Flow Synchronizer is the metronome of semantic dynamics.
It keeps geometry and structure in phase,  
aligns curvature with reference curvature,  
aligns entropy with tag entropy,  
and marches every configuration toward its harmonic, quine-stable form.
\end{quote}

It is the temporal heart of HYDRA:  
the mechanism through which meaning evolves coherently across both the
continuous and discrete layers.

