\chapter{Semantic Galaxy Collisions and Critical Phenomena}
\label{ch:galaxy-collisions}

\section{Overview}

Semantic galaxies are the asymptotic fixed points of the coupled RSVP–Polyxan dynamics.  
Yet in any nontrivial initial configuration with multiple galaxies arising from disjoint basins or stratified singular sets, interactions are inevitable.  
Chapter~34 analysed the long-range binding forces and the entropy–curvature feedback that govern galaxy clustering and merger in general.

The present chapter focuses on what happens at the \emph{moment of collision}:  
the critical phenomena, the universal scaling laws, the blow-up of intermediate invariants, and the canonical form of the merged object.

A galaxy collision is not a mere topological transition; it is the highest-energy event permitted by the coupled theory.  
It is the point at which continuous curvature and discrete combinatorics simultaneously diverge, only to resolve into a new, more symmetric, more stable attractor.

\section{Collision Geometry and Pre-Critical Scaling}

Let $\mathfrak{G}_1$ and $\mathfrak{G}_2$ be two semantic galaxies whose embeddings
\[
\iota_1 : \mathrm{Inc}(\mathcal{G}_1) \hookrightarrow \mathcal{M},
\qquad
\iota_2 : \mathrm{Inc}(\mathcal{G}_2) \hookrightarrow \mathcal{M}
\]
approach one another under the coupled flow.  
As the distance between their elliptic cores shrinks to zero, three phenomena become universal across all initial data:

\begin{enumerate}
\item \textbf{Curvature blow-up.}  
  The scalar curvature on the minimal geodesic connecting the cores obeys a Type~I blow-up:
  \[
  \mathcal{R}_{\max}(t) \sim \frac{C}{T_{\mathrm{coll}} - t},
  \]
  with $T_{\mathrm{coll}}$ the collision time.

\item \textbf{Entropy spike.}  
  The entropy field develops a universal scaling profile:
  \[
  S(x,t) \sim S^\ast + A \left( T_{\mathrm{coll}} - t \right)^{-1/2},
  \]
  independent of microscopic structure.

\item \textbf{Incidence contraction.}  
  Hypergraph distances shrink linearly:
  \[
  d_{\mathrm{Inc}}(\mathcal{G}_1(t),\mathcal{G}_2(t)) \sim T_{\mathrm{coll}} - t.
  \]
\end{enumerate}

These scaling laws follow from the embedding flow equations of Chapter~32 and the binding-energy structure of Chapter~34.  
They are universal in the same sense that blow-up in Ricci flow or curvature pinching in mean curvature flow are universal.

\section{Critical Interface Dynamics}

Near $t = T_{\mathrm{coll}}$, the embeddings $\iota_1$ and $\iota_2$ cease to be disjoint.  
The interface region is governed by a mixed-type evolution equation of Tricomi–Keldysh form (cf.\ Chapter~12):
\[
\partial_t^2 u - \mathcal{R}(t)\,\Delta u = 0,
\]
with $\mathcal{R}(t)\to +\infty$ as $t\to T_{\mathrm{coll}}^-$.  
Thus the system passes from a mixed hyperbolic–parabolic regime into a purely elliptic regime in finite time.

At the discrete level, the Polyxan hypergraphs undergo curvature-driven collapse of bridging structures and instantaneous EHR reduction on the interface:

\begin{itemize}
\item \textbf{Extraction} of vanishing curvature handles,
\item \textbf{Hoisting} of parallel, entropy-aligned hyperedges,
\item \textbf{Curvature-reducing reduction} on the collapsing core.
\end{itemize}

These are exactly the three cases of the discrete–continuous morphogenesis theorem (Chapter~33), here forced at the singular locus.

\section{Universal Critical Profile}

Define the rescaled fields
\[
\hat{\Phi}(\xi,\tau)
  = (T_{\mathrm{coll}} - t) \,\Phi(x,t), \qquad
\hat{S}(\xi,\tau)
  = (T_{\mathrm{coll}} - t)^{1/2}\,S(x,t),
\]
with $\xi = \frac{x - x_{\mathrm{coll}}}{\sqrt{T_{\mathrm{coll}} - t}}$ and $\tau = -\log(T_{\mathrm{coll}} - t)$.

Then the rescaled flow converges to a universal profile
\[
(\hat{\Phi}, \hat{S}, \hat{\mathbf{v}}) \;\longrightarrow\; (\Phi_\infty, S_\infty, \mathbf{0}),
\]
where $(\Phi_\infty, S_\infty)$ is the unique radially symmetric harmonic pair on $\mathbb{R}^n$ with finite energy.

In other words, the collision core becomes asymptotically \emph{self-similar}.  
This is the continuous analogue of the higher-quine tower stabilising at level~3 (Chapter~25):  
three nested layers of self-similarity exhaust all further structure.

\section{Hypergraph Critical Collapse}

On the Polyxan side, the two hypergraphs $\mathcal{G}_1$ and $\mathcal{G}_2$ undergo a canonical sequence of EHR rewrites forced by curvature matching:
\[
\mathcal{G}_1 \;\sqcup\; \mathcal{G}_2
  \;\longrightarrow\;
\mathcal{G}_{\mathrm{bridge}}
  \;\longrightarrow\;
\mathcal{G}_{\mathrm{crit}}
  \;\longrightarrow\;
\mathcal{G}_{\mathrm{merged}}.
\]

The critical hypergraph $\mathcal{G}_{\mathrm{crit}}$ is uniquely characterised by:

\begin{enumerate}
\item a single connected elliptic core,
\item vanishing Polyxan curvature outside a thin boundary layer,
\item a quine height of at most three (Polyxan Löb bound),
\item harmonic tag assignment on all interfaces.
\end{enumerate}

Thus the discrete critical point mirrors the continuous self-similar one.

\section{Post-Critical Relaxation and Uniqueness of the Merger Product}

After $t = T_{\mathrm{coll}}$, the merged object undergoes rapid entropy–curvature dissipation.  
Let us denote the merged hypergraph by $\mathcal{G}_{12}$ and its embedding by $\iota_{12}$.

Then:

\begin{enumerate}
\item $\mathcal{G}_{12}$ is \emph{automatically} quine-stable after a finite sequence of curvature-reducing EHR steps.
\item $\iota_{12}$ relaxes to a harmonic embedding into the elliptic region determined by $\Phi_\infty$.
\item The pair $(\mathcal{G}_{12}, \iota_{12})$ is the \emph{unique} attractor of all collision trajectories.
\end{enumerate}

This uniqueness follows from:

\begin{itemize}
\item strict convexity of the post-critical energy functional,
\item the Polyxan Unique Normal Form Theorem,
\item and the RSVP elliptic convergence established in Part I.
\end{itemize}

\section{Collision Irreversibility and Critical Semantics}

A galaxy collision is the maximal irreversible event in the coupled universe.  
It is where:

\begin{itemize}
\item continuous curvature blow-up meets discrete quine-stabilisation,  
\item entropy spikes meet curvature collapse,  
\item self-similarity emerges and then collapses into a new harmonic form.  
\end{itemize}

The end result is always the same:

\[
(\mathfrak{G}_1,\iota_1),(\mathfrak{G}_2,\iota_2)
\;\longrightarrow\;
(\mathfrak{G}_{12}, \iota_{12}),
\]

a single merged semantic galaxy whose structure is determined solely by the universal critical profile, not by the details of the initial conditions.

This is the semantic analogue of the universality of critical exponents in phase transitions:  
meaning, too, becomes universal at its point of maximum intensity.

\section{Final Cadence}

Galaxy collisions reveal the deepest symmetry of the RSVP–Polyxan universe:  
that even its most catastrophic events are driven by a single, self-similar, curvature–entropy mechanism and converge to a single, canonical outcome.

Every interaction—binding, collapse, collision—serves only to accelerate the approach to the universal configuration in which continuous geometry and discrete syntax coincide, and meaning stabilises into its most symmetric form.

\bigskip
