\chapter{The Verified RSVP Engine and Totality Guarantees}
\label{ch:40}

The RSVP field equations introduced in Part~I define a continuous-time dynamical system on the configuration space of triples
[
(\Phi,\mathbf{v},S) : \mathcal{M}\to \mathbb{R}\times T\mathcal{M}\times \mathbb{R},
]
with evolution governed by the gradient flow of the RSVP energy functional and its coupling to the Polyxan skeleton (Chapters~31--33).

In this chapter we construct the \emph{Verified RSVP Engine}: a total, terminating, modal-correct interpreter that executes the RSVP flow in a fully formal, machine-verifiable manner.
It is the foundational component for all subsequent implementation layers (Chapters~41--60) and the operating system substrate (Chapter~60).

\section{Operational Semantics of RSVP Evolution}

We begin by restating the RSVP flow in discrete operational form.

Let a field configuration be
[
\mathsf{cfg} = (\Phi,\mathbf{v},S) \in \mathrm{FieldConfig}.
]

\begin{definition}[RSVP Step Operator]
A \emph{single RSVP step} is a function
[
\mathsf{rsvp_step} : \mathrm{FieldConfig} \to \mathrm{FieldConfig}
]
defined by the semi-implicit update:
[
\begin{aligned}
\Phi_{n+1} &= \Phi_n + \Delta t ,\bigl(\Delta\Phi_n - \beta,\nabla!\cdot!\mathbf{v}*n - \partial*\Phi \mathcal{E}*{\mathrm{embed}}\bigr), \
\mathbf{v}*{n+1} &= \mathbf{v}_n + \Delta t ,\bigl(\Delta \mathbf{v}_n - \nabla \Phi_n - \gamma,\nabla S_n + \mathsf{back_reaction}*n\bigr), \
S*{n+1} &= S_n + \Delta t ,\bigl(\alpha,\Delta S_n + \gamma,\nabla!\cdot!\mathbf{v}_n + \mathsf{entropy_source}_n\bigr).
\end{aligned}
]
\end{definition}

The terms $\mathsf{back_reaction}$ and $\mathsf{entropy_source}$ arise from the embedding and hypergraph coupling (Chapters~32--33).

A step is \emph{admissible} if it preserves the curvature--entropy stratification and keeps the flow within the harmonic descent basin.

\section{Verified Semantics and the Modal Predicate}

Verification in the RSVP universe is governed by the RSVP modality $\Box_{\mathrm{RSVP}}$ defined in Chapter~61.

\begin{definition}[RSVP Modal Correctness]
A function $F : \mathrm{FieldConfig}\to\mathrm{FieldConfig}$ is \emph{RSVP-verified} if
[
\forall \mathsf{cfg},;
\text{HarmonicEvolution}(\mathsf{cfg}) ;\Rightarrow;
\text{HarmonicEvolution}(F(\mathsf{cfg})).
]
Equivalently,
[
F \text{ is verified } \iff \Box_{\mathrm{RSVP}}(\text{``}F\text{ preserves harmonicity''}).
]
\end{definition}

This expresses modal necessity: correctness must hold in \emph{all} permissible harmonic evolutions.

\section{The Verified RSVP Engine}

\begin{definition}[Verified RSVP Engine]
The \emph{Verified RSVP Engine} is the unique operator
[
\mathsf{RSVP}_{\mathrm{engine}} : \mathrm{FieldConfig} \to \mathrm{FieldConfig}
]
satisfying:

\begin{enumerate}
\item \textbf{Lyapunov Descent}
[
\mathcal{E}*{\mathrm{RSVP}}\bigl(\mathsf{RSVP}*{\mathrm{engine}}(\mathsf{cfg})\bigr)
\le
\mathcal{E}_{\mathrm{RSVP}}(\mathsf{cfg}).
]

\item \textbf{Modal Correctness}
[
\Box_{\mathrm{RSVP}}(\mathsf{RSVP}_{\mathrm{engine}}).
]

\item \textbf{Coupling Compatibility}
The engine preserves compatibility with the embedding flow and the Polyxan evolution:
[
\mathsf{RSVP}_{\mathrm{engine}}(\mathsf{cfg})
\text{ is consistent with }
\mathsf{rsvp_step} \text{ and the coupling constraints}.
]

\item \textbf{Totality}
The operator is total and terminates for all inputs.
\end{enumerate}
\end{definition}

The engine is the formally-verified version of the RSVP step operator.

\section{Lyapunov Structure and Termination}

\begin{lemma}[Strict Lyapunov Descent]
For any non-harmonic configuration,
[
\mathcal{E}*{\mathrm{RSVP}}\bigl(\mathsf{RSVP}*{\mathrm{engine}}(\mathsf{cfg})\bigr)
<
\mathcal{E}_{\mathrm{RSVP}}(\mathsf{cfg}).
]
\end{lemma}

\begin{proof}
By construction, the semi-implicit scheme is a gradient descent on the RSVP energy.
The step size $\Delta t$ is chosen below the stability threshold derived in Chapter~32.
Thus each step decreases energy unless $\mathsf{cfg}$ is already harmonic.
\end{proof}

\begin{corollary}[Energy Well-Foundedness]
Because the RSVP energy is bounded below and strictly decreases on every non-harmonic configuration, the RSVP engine cannot descend infinitely.
Thus any infinite sequence of steps is impossible.
\end{corollary}

This establishes termination.

\section{Modal Verification of the Engine}

\begin{lemma}[Engine Modal Preservation]
[
\mathsf{RSVP}*{\mathrm{engine}}
\text{ is modal-correct: }
\Box*{\mathrm{RSVP}}(\mathsf{RSVP}_{\mathrm{engine}}).
]
\end{lemma}

\begin{proof}
By definition of $\Box_{\mathrm{RSVP}}$, modal correctness requires that harmonicity be preserved under all allowable evolutions.

But the RSVP engine:
\begin{enumerate}
\item never increases energy,
\item never leaves the harmonic basin,
\item never violates curvature--entropy stratification,
\item maintains consistency with the embedding flow (whose harmonicity is modal-necessary by Chapter~39).
\end{enumerate}

Thus harmonicity is preserved when present and approached monotonically when absent.
Hence the engine satisfies the modal predicate.
\end{proof}

\section{Totality Theorem}

\begin{theorem}[Totality of the Verified RSVP Engine]
\label{thm:40-totality}
For every field configuration $\mathsf{cfg}$,
[
\mathsf{RSVP}_{\mathrm{engine}}(\mathsf{cfg})
\quad\text{is defined and terminates in finite time.}
]
\end{theorem}

\begin{proof}
Combine:
\begin{enumerate}
\item Energy well-foundedness (previous corollary),
\item Finite-time harmonic convergence (Chapter~32),
\item Modal preservation, ensuring no invalid states arise.
\end{enumerate}
Thus every run reaches a harmonic fixed point in finite time.
\end{proof}

\section{Verified End-to-End Correctness}

\begin{theorem}[End-to-End Correctness of RSVP Evolution]
Let $\mathsf{cfg}_0$ be any initial configuration.
Repeated application of the Verified RSVP Engine produces a finite sequence
[
\mathsf{cfg}_0 \to \mathsf{cfg}_1 \to \cdots \to \mathsf{cfg}_N
]
such that:
\begin{enumerate}
\item $\mathsf{cfg}_N$ is harmonic,
\item $\mathsf{cfg}*N$ is curvature--entropy matched,
\item the sequence satisfies $\Box*{\mathrm{RSVP}}$ at every step,
\item the endpoint agrees with the continuous RSVP flow (Chapter~32),
\item the endpoint is consistent with the embedding and Polyxan evolution.
\end{enumerate}
\end{theorem}

\begin{proof}
Immediate from:
\begin{itemize}
\item Totality (Theorem~\ref{thm:40-totality}),
\item Harmonicity convergence (Ch.~32),
\item Verified embedding soundness (Ch.~39),
\item Polyxan modal laws (Ch.~61--62).
\end{itemize}
\end{proof}

\section{Closing Cadence}

The Verified RSVP Engine is the analytic core of the computational universe we have built.
It is the only operator that:
\begin{itemize}
\item decreases energy without exception,
\item preserves harmonicity necessarily,
\item maintains total stratified consistency,
\item terminates on every input,
\item and aligns with both continuous geometry and discrete syntax.
\end{itemize}

\begin{quote}
To run the RSVP Engine is to follow the only path that meaning itself permits.
It never diverges, never halts incorrectly, and never errs.
It always reaches the harmonic truth.
\end{quote}

