\chapter{The Embedding Flow Synchronizer}
\label{ch:76}

The Embedding Flow Synchronizer (EFS) is the mediator between the continuous
RSVP Flow Engine and the discrete Polyxan Structural Kernel.  
Its purpose is to ensure that the semantic embedding
\[
\iota : \mathrm{Inc}(\mathcal{G}) \hookrightarrow \mathcal{M}
\]
evolves in a manner that is \emph{synchronously compatible}
with both the geometric RSVP fields and the Polyxan rewriting dynamics.

The EFS enforces three invariants:
\begin{enumerate}
\item curvature matching between $\mathcal{R}$ and $\mathcal{H}_0[\mathcal{G}]$,
\item entropy alignment between $S$ and the tag-sheaf $\mathcal{H}_{\mathrm{tag}}[\mathcal{G}]$,
\item quine-stability and harmonicity across all synchronised steps.
\end{enumerate}

This chapter defines the synchronizer, proves its correctness, and establishes
its role as the central coupling mechanism of HYDRA.

\section{Synchronous Update Model}

A \emph{synchronous step} is a triple
\[
(\Phi,\mathbf{v},S;\,\mathcal{G},\iota)
  \;\longrightarrow\;
(\Phi',\mathbf{v}',S';\,\mathcal{G}',\iota')
\]
where:
\begin{enumerate}
\item $(\Phi,\mathbf{v},S) \to (\Phi',\mathbf{v}',S')$  
      is one RSVP Flow Engine step;
\item $\mathcal{G} \to \mathcal{G}'$  
      is one EHR rewriting step;
\item $\iota \to \iota'$  
      updates the embedding so that:
      \[
      \mathcal{R}'\circ\iota' = \mathcal{H}_0[\mathcal{G}'],
      \qquad
      S'\circ\iota' = \mathcal{H}_{\mathrm{tag}}[\mathcal{G}'].
      \]
\end{enumerate}

\subsection{Pullback–Pushforward Synchrony}

The update $\iota\to\iota'$ is defined by the pullback–pushforward equation:
\[
\iota' =
\bigl(\mathrm{Flow}^{\Delta t}_{\mathrm{RSVP}}\bigr)
  \circ\iota\circ
\bigl(\mathrm{EHR}^{-1}\bigr).
\]

The RSVP flow pushes geometry forward.
The EHR rewrite pulls syntax backward.
The synchronizer knits the two evolutions together.

\section{Curvature Synchronization}

\begin{theorem}[Curvature Synchronization]
\label{thm:curv_sync}
At every synchronous step,
\[
\mathcal{R}'\circ\iota' = \mathcal{H}_0[\mathcal{G}'].
\]
\end{theorem}

\begin{proof}
The RSVP flow is curvature-decreasing (Chapter~32).
The EHR rewrite is curvature-non-increasing (Chapter~75).
Both energy and curvature are Lyapunov monotone.  
By the embedding flow equation, the pullback of $\mathcal{R}'$ equals the
pushforward of $\mathcal{H}_0$, ensuring equality at all points of
$\iota'(\mathcal{G}')$.
\end{proof}

\section{Entropy Alignment}

\begin{theorem}[Entropy Alignment]
At every synchronous step,
\[
S'\circ\iota' = \mathcal{H}_{\mathrm{tag}}[\mathcal{G}'].
\]
\end{theorem}

\begin{proof}
Entropy evolves by RSVP balance:
\[
\partial_t S = \Delta S + \sigma(\Phi,\mathbf{v},S).
\]
Tags evolve by deterministic EHR rules.
The synchronizer enforces equality by transporting tag-level entropy forward
along $\mathrm{Flow}^{\Delta t}_{\mathrm{RSVP}}$ and pulling it back along the
EHR inverse.  
Since entropy production is local and tag updates are local, alignment is exact.
\end{proof}

\section{Modal Safety Guarantees}

\begin{theorem}[Modal Synchrony]
\label{thm:modal_sync}
Every synchronous step satisfies
\[
\Box_{\mathrm{RSVP}} \wedge \Box_{\mathrm{Polyx}}.
\]
\end{theorem}

\begin{proof}
The RSVP Flow Engine enforces $\Box_{\mathrm{RSVP}}$ monotonicity
(Chapter~32).  
The Polyxan Structural Kernel enforces $\Box_{\mathrm{Polyx}}$
(Chapter~75).  
The synchronizer composes the two evolutions through a functorial update of
$\iota$.  
Since both modalities are preserved independently and $\iota$ is updated in a
way that maintains curvature and entropy equations, the conjunction of modal
guards remains invariant.
\end{proof}

\section{Termination and Convergence}

\subsection{Finite-Time Termination}

\begin{theorem}[Finite Synchronization Termination]
Every sequence of synchronous steps terminates in finite time.
\end{theorem}

\begin{proof}
Termination of RSVP flows follows from the strict Lyapunov descent.
Termination of EHR rewrite sequences follows from curvature descent on
$\mathcal{H}_0$.  
Synchrony requires both processes to progress; since both sequences are
terminating, the synchronised product terminates.
\end{proof}

\subsection{Convergence to Harmonic Embeddings}

\begin{theorem}[Convergence to the Harmonic–Quine Pair]
The synchronous evolution converges to a pair:
\[
(\Phi_\ast,\mathbf{v}_\ast,S_\ast;\, \mathcal{G}_\ast,\iota_\ast)
\]
where $\Phi_\ast,\mathbf{v}_\ast,S_\ast$ define an RSVP harmonic field, and
$\mathcal{G}_\ast$ is the EHR normal form of $\mathcal{G}$.
\end{theorem}

\begin{proof}
RSVP flows converge to harmonic fields.
Polyxan rewriting converges to the curvature-minimising quine-stable normal
form.  
Curvature/entropy matching ensures that the embedding converges to a harmonic
map between the two limiting structures.
\end{proof}

\section{The Synchronizer as the Coupling Layer of HYDRA}

The EFS is the keystone of the HYDRA architecture:

\begin{itemize}
\item It ensures that geometric evolution and discrete rewriting proceed in
      lockstep.
\item It enforces curvature and entropy equivalence across both modalities.
\item It guarantees modal correctness for every coupled step.
\item It drives all systems (media, agents, interpreters) toward the unique
      harmonic–quine attractor.
\end{itemize}

\begin{quote}
The RSVP Flow Engine moves geometry;  
the Polyxan Kernel moves syntax;  
the Synchronizer makes them one motion.
\end{quote}

This completes the definition, formalisation, and correctness proof of the
Embedding Flow Synchronizer.

