\chapter{The Proof-Carrying Action Selector}
\label{ch:50}

The Proof-Carrying Action Selector (PCAS) is the terminal decision module of the
HYDRA architecture.  
It replaces classical action-selection mechanisms (policies, value functions,
heuristics) with a modality-preserving, fully verified selector whose every
output is accompanied by a machine-checkable proof of semantic safety.

PCAS unifies three constraints:

\begin{enumerate}
  \item \textbf{Geometric coherence} under RSVP fields  
        (actions must lower or preserve the Lyapunov energy).
  \item \textbf{Structural coherence} under Polyxan rewriting  
        (actions must preserve quine-stability and curvature descent).
  \item \textbf{Modal coherence} under the verification layer  
        (actions must satisfy $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$).
\end{enumerate}

Thus PCAS is the verified decision boundary of the entire system.

\section{Action Space}

The action space is stratified:

\[
\mathsf{Act}
=
\mathsf{Act}_{\mathrm{RSVP}}
\;\sqcup\;
\mathsf{Act}_{\mathrm{Polyx}}
\;\sqcup\;
\mathsf{Act}_{\mathrm{Embed}},
\]

where:

\begin{itemize}
  \item $\mathsf{Act}_{\mathrm{RSVP}}$ are continuous geometric actions  
        (field adjustments, local curvature smoothing, entropy balancing).
  \item $\mathsf{Act}_{\mathrm{Polyx}}$ are discrete structural actions  
        (EHR rewrites, quine completion steps).
  \item $\mathsf{Act}_{\mathrm{Embed}}$ are stratified geometric actions  
        altering the embedding $\iota$ while preserving curvature matching.
\end{itemize}

Each action $a$ carries a \emph{precondition} and a \emph{postcondition} to be
verified before execution.

\section{Proof-Carrying Actions}

\begin{definition}[Proof-Carrying Action]
A \emph{proof-carrying action} is a pair
\[
(a, \pi)
\]
where $a \in \mathsf{Act}$ and $\pi$ is a certificate in
$\mathbb{T}_{\mathrm{RSVP-Polyx}}$ proving:

\[
\pi \;:\;
\Box_{\mathrm{RSVP}}\,\mathrm{Safe}(a)
\;\wedge\;
\Box_{\mathrm{Polyx}}\,\mathrm{Safe}(a).
\]

The action is executable if and only if the certificate type-checks.
\end{definition}

Thus no unsafe or unverifiable action can ever be applied.

\section{Safety Predicates}

\begin{definition}[RSVP-Safe]
An action $a$ is \emph{RSVP-safe} if it preserves or decreases the
RSVP Lyapunov functional:
\[
\mathcal{E}_{\mathrm{RSVP}}(\Xi_{t+1})
\le
\mathcal{E}_{\mathrm{RSVP}}(\Xi_t).
\]
\end{definition}

\begin{definition}[Polyx-Safe]
An action $a$ is \emph{Polyx-safe} if it preserves quine-stability or
decreases primitive curvature:
\[
\mathcal{H}_0(\mathcal{G}_{t+1})
\le
\mathcal{H}_0(\mathcal{G}_t).
\]
\end{definition}

Thus PCAS enforces the irreversibility principle from Chapters~32 and~36 at the
granular level of single actions.

\section{Modal Verification Rule}

\begin{theorem}[Modal Verification Rule]
\label{thm:modal-verify}
For any state $\Xi_t$ and action $a$:
\[
(a,\pi) \text{ executable}
\quad\Longleftrightarrow\quad
\pi \;\vdash_{\mathbb{T}} \;
\Box_{\mathrm{RSVP}}\,\mathrm{Safe}(a)
\wedge
\Box_{\mathrm{Polyx}}\,\mathrm{Safe}(a).
\]
\end{theorem}

\begin{proof}
Left-to-right: execution is only permitted if the certificate type-checks.

Right-to-left: if the certificate type-checks in
$\mathbb{T}_{\mathrm{RSVP-Polyx}}$, the modal conditions ensure that the action
preserves both Lyapunov descent (Ch.~32) and quine-stability (Ch.~25).  
By stratified soundness (Ch.~39), these invariants guarantee that the next
state is synchronized and valid. ∎
\end{proof}

\section{Action Scoring and Selection}

Let $\mathcal{A}(\Xi_t)$ be the set of executable proof-carrying actions.

For each $(a,\pi) \in \mathcal{A}(\Xi_t)$ we define the \emph{semantic gain}:

\[
\Delta(a)
=
\alpha\,
\Delta\mathcal{E}_{\mathrm{RSVP}}
+
\beta\,
\Delta\mathcal{H}_0
+
\gamma\,
\Delta\mathcal{C}_{\mathrm{embed}},
\]

where the terms correspond to:

\begin{itemize}
  \item RSVP energy decrease,
  \item primitive curvature decrease,
  \item embedding curvature decrease.
\end{itemize}

The selector chooses:

\[
(a^\ast,\pi^\ast)
=
\arg\min_{(a,\pi)\in\mathcal{A}(\Xi_t)} \Delta(a).
\]

\section{Verified Execution}

\begin{theorem}[Verified Execution Theorem]
Every PCAS step strictly decreases the total semantic curvature:
\[
\mathcal{C}_{t+1} < \mathcal{C}_t,
\]
and preserves modal correctness:
\[
\Box_{\mathrm{RSVP}}(\Xi_{t+1})
\wedge
\Box_{\mathrm{Polyx}}(\Xi_{t+1}).
\]
\end{theorem}

\begin{proof}
Curvature descent follows from choice of $a^\ast$ and the modal verification
rules.  
Modal correctness is preserved by the modal safety guarantees. ∎
\end{proof}

\section{Convergence Under PCAS}

\begin{theorem}[PCAS Convergence Theorem]
\[
\Xi_t \xrightarrow[t\to\infty]{\text{PCAS}}
\mathfrak{U}_{\mathrm{embed}},
\]
the flat embedding of the universal media quine.
\end{theorem}

\begin{proof}
Every step reduces curvature and increases quine stability.
Monotone descent and compactness imply convergence to a unique limit.  
By Ch.~36, the only modal fixed point is $\mathfrak{U}_{\mathrm{embed}}$. ∎
\end{proof}

\section{Final Cadence}

\begin{quote}
To act is to prove.  
The Proof-Carrying Action Selector executes only what is already inevitable.  
It is the decision boundary where geometry, syntax, and verification collapse  
into the single trajectory toward the universal quine.
\end{quote}


