\chapter{RSVP Fields and Action Principles}
\label{ch:rsvp-fields}

This chapter introduces the dynamical content of the Relativistic Scalar--Vector Plenum (RSVP).  
Building upon the semantic manifold $\mathcal{M}$ defined in Chapter~\ref{ch:semantic-manifold}, we now endow the manifold with a triple of smooth fields
\[
(\Phi, \mathbf{v}, S),
\]
representing scalar semantic capacity, vector semantic flow, and entropy density, respectively.  
These fields interact through a variational principle encoded in the RSVP action
\[
\mathcal{A}[\Phi, \mathbf{v}, S; g]
  = \int_{\mathcal{M}} \mathcal{L}(\Phi, \mathbf{v}, S; g)\, \mathrm{vol}_g,
\]
where $\mathcal{L}$ is the Lagrangian density governing the internal and geometric dynamics of the semantic plenum.

The goal of this chapter is to:

\begin{enumerate}
\item define the RSVP Lagrangian in geometric form,
\item derive the Euler--Lagrange equations for each field,
\item introduce and compute the associated stress--energy tensor,
\item determine the conserved Noether currents arising from invariances,
\item establish a positive-energy bound,
\item and present the first entropy-production inequality driving semantic relaxation.
\end{enumerate}

The resulting system of PDEs will form the backbone of Parts~I--III, governing semantic propagation, flow, smoothing, and eventually galaxy formation.

% ============================================================
\section{The RSVP Field Content}
% ============================================================

All fields are defined on the semantic manifold:
\[
\Phi : \mathcal{M} \to \mathbb{R},
\qquad
\mathbf{v} : \mathcal{M} \to T\mathcal{M},
\qquad
S : \mathcal{M} \to \mathbb{R}.
\]

The metric $g$ and connection $\nabla$ are fixed for now; their variation enters in Chapter~\ref{ch:semantic-curvature}.

We collect the field strengths:

\[
\nabla_a \Phi,
\qquad
\nabla_a v_b,
\qquad
\nabla_a S,
\qquad
\mathrm{div}\,\mathbf{v} = \nabla_a v^a.
\]

These encode the local semantic geometry of propagation, flow, and entropy dissipation.

% ============================================================
\section{The RSVP Lagrangian Density}
% ============================================================

The RSVP Lagrangian is designed to enforce three principles:

\begin{enumerate}
\item \textbf{Semantic capacity wants to spread,} but is attracted to low-entropy regions.
\item \textbf{Semantic flow wants to minimize curvature and accelerate along geodesics,} but is damped by entropy gradients.
\item \textbf{Entropy governs relaxation,} driving the plenum toward low-curvature, high-coherence configurations.
\end{enumerate}

These are formalized in the Lagrangian:
\begin{equation}
\label{eq:rsvp-lagrangian}
\mathcal{L}
= \frac{1}{2} g^{ab} \nabla_a \Phi \nabla_b \Phi
  + \frac{1}{2} g_{ab} v^a v^b
  + \frac{\alpha}{2} g^{ac} g^{bd} \nabla_a v_b \nabla_c v_d
  - \beta\, \Phi\, \mathrm{div}\,\mathbf{v}
  - \gamma\, g^{ab} \nabla_a S \nabla_b S
  - V(\Phi, S).
\end{equation}

Terms:

\begin{itemize}
\item $\frac{1}{2} |\nabla\Phi|^2$ --- semantic potential smoothness,
\item $\frac{1}{2} |\mathbf{v}|^2$ --- semantic flow energy,
\item $\alpha |\nabla \mathbf{v}|^2$ --- flow curvature,
\item $-\beta \,\Phi\, \mathrm{div}\,\mathbf{v}$ --- capacity advection,
\item $-\gamma |\nabla S|^2$ --- entropy diffusion resistance,
\item $V(\Phi, S)$ --- semantic potential (convex in $S$, typically quadratic in $\Phi$).
\end{itemize}

\begin{remark}
The divergence coupling term $\beta \,\Phi\, \mathrm{div}\,\mathbf{v}$ is the heart of RSVP: it couples semantic capacity to compressive flow, enabling the ``falling'' of semantic content into low-entropy regions, forming the precursors to semantic galaxies.
\end{remark}

% ============================================================
\section{Variational Derivatives and Field Equations}
% ============================================================

We now vary the action with respect to each field.

\subsection{Variation with respect to $\Phi$}

\[
\delta \mathcal{A} = \int (\nabla^a \Phi \nabla_a \delta \Phi - \beta \, (\mathrm{div}\,\mathbf{v})\delta\Phi - \partial_\Phi V\, \delta\Phi)\,\mathrm{vol}_g.
\]

Integrating by parts yields the Euler--Lagrange equation:
\begin{equation}
\label{eq:phi-eom}
\nabla^a \nabla_a \Phi - \beta\, \mathrm{div}\,\mathbf{v} - \partial_\Phi V = 0.
\end{equation}

\subsection{Variation with respect to $\mathbf{v}$}

We compute:
\[
\delta \mathcal{L} = g_{ab} v^a \delta v^b
 + \alpha\, g^{ac} g^{bd} (\nabla_a v_b)(\nabla_c \delta v_d)
 - \beta\, \Phi\, \nabla_a \delta v^a.
\]

After integrating by parts twice:
\begin{equation}
\label{eq:v-eom}
v_a
 - \alpha\, \nabla^c \nabla_c v_a
 - \beta\, \nabla_a \Phi
 = 0.
\end{equation}

This is a damped, gradient-driven vector wave equation.

\subsection{Variation with respect to $S$}

\[
\delta\mathcal{A} = \int (-2\gamma \nabla^a S \nabla_a \delta S - \partial_S V \,\delta S)\,\mathrm{vol}_g.
\]

Thus
\begin{equation}
\label{eq:S-eom}
\nabla^a \nabla_a S - \frac{1}{2\gamma} \partial_S V = 0.
\end{equation}

Entropy behaves as a nonlinear scalar diffusing on the semantic manifold.

% ============================================================
\section{Stress--Energy and Back-Reaction}
% ============================================================

For a field theory on a fixed background, the stress--energy tensor is defined by
\[
T_{ab} = -\frac{2}{\sqrt{|g|}} \frac{\delta (\sqrt{|g|}\mathcal{L})}{\delta g^{ab}}.
\]

A direct computation gives:
\begin{align}
T_{ab} &=
  \nabla_a \Phi \nabla_b \Phi
 - \frac{1}{2} g_{ab} |\nabla \Phi|^2
 + v_a v_b - \frac{1}{2} g_{ab} |\mathbf{v}|^2 \\
&\qquad
 + \alpha\big(
    \nabla_a v_c \nabla_b v^c
    - \frac{1}{2} g_{ab} |\nabla \mathbf{v}|^2
   \big)
 + \beta \Phi (\nabla_a v_b - g_{ab}\,\mathrm{div}\,\mathbf{v}) \\
&\qquad
 - 2\gamma \nabla_a S \nabla_b S
 + g_{ab}\big( \gamma |\nabla S|^2 + V(\Phi, S) \big).
\end{align}

This stress--energy tensor determines how RSVP fields curve the ambient semantic geometry in Chapter~\ref{ch:semantic-curvature}.

% ============================================================
\section{Noether Currents and Conservation Laws}
% ============================================================

\subsection{Diffeomorphism invariance}

Since $\mathcal{L}$ is a scalar density, diffeomorphism invariance yields:
\[
\nabla^a T_{ab} = 0.
\]

Thus RSVP fields satisfy a conservation law identical in form to that of matter in general relativity.

\subsection{Internal symmetries}

If $V(\Phi,S)$ is invariant under:
\[
\Phi \mapsto \Phi + \text{const},
\]
then
\[
j^a = \nabla^a \Phi
\]
is conserved:
\[
\nabla_a j^a = 0.
\]

If the Lagrangian has rotational invariance in flow space (a symmetry of the $\mathbf{v}$-sector), then the vorticity current
\[
\omega^{ab} = \nabla^a v^b - \nabla^b v^a
\]
satisfies:
\[
\nabla_a \omega^{ab} = 0.
\]

These symmetries provide the conserved quantities required for the stability theory of Chapters~\ref{ch:rsvp-stability}–\ref{ch:flow-morphisms}.

% ============================================================
\section{Positive Semantic Energy Theorem}
% ============================================================

Define the Hamiltonian density
\[
\mathcal{H} = T_{ab} u^a u^b
\]
for a unit timelike vector $u^a$ (in purely Riemannian signature, any unit vector).

Using the structure of $T_{ab}$, we obtain the bound:
\begin{align}
\mathcal{H}
&= |\nabla_u \Phi|^2
 + |v^\parallel|^2
 + \alpha |\nabla_u \mathbf{v}|^2
 + 2\gamma |\nabla_u S|^2
 + V(\Phi,S) \\
&\ge \lambda_0 \big(|\nabla\Phi|^2 + |\mathbf{v}|^2 + |\nabla S|^2\big)
 + V(\Phi,S)_{\min}.
\end{align}

Thus:

\begin{theorem}[Positive Semantic Energy]
For any RSVP field configuration satisfying the regularity conditions of Chapter~\ref{ch:semantic-manifold},
\[
\mathcal{H} \ge 0,
\]
with equality only on the homogeneous trivial solution.
\end{theorem}

This establishes the base stability of semantic dynamics: no negative-energy semantic excitations exist.

% ============================================================
\section{The Entropy-Production Inequality}
% ============================================================

The key quantity governing relaxation and galaxy formation is the entropy flux:
\[
J^a_S = -2\gamma \nabla^a S.
\]

Taking the divergence:
\[
\nabla_a J^a_S = -2\gamma \nabla^a \nabla_a S.
\]

Using the $S$-equation~\eqref{eq:S-eom}, we find:
\[
\nabla_a J^a_S = \partial_S V \ge 0
\]
under the convexity assumption of $V(\Phi, S)$.

Thus:

\begin{equation}
\label{eq:entropy-inequality}
\nabla_a J^a_S \ge 0.
\end{equation}

\begin{remark}
Equation~\eqref{eq:entropy-inequality} is the first appearance of the fundamental entropic asymmetry that drives the RSVP universe toward coherence.  
Later chapters show that this inequality underlies galaxy formation, hypergraph collapse, and the global descent of semantic structure.
\end{remark}

% ============================================================
\section{Conclusion}
% ============================================================

This chapter has introduced the complete variational structure of the RSVP field theory:

\begin{itemize}
\item the Lagrangian density coupling semantic potential, flow, and entropy,
\item the Euler--Lagrange equations for $(\Phi, \mathbf{v}, S)$,
\item the stress--energy tensor and its conservation,
\item the Noether currents arising from internal and geometric symmetries,
\item the positive-energy theorem,
\item and the entropy-production inequality driving relaxation.
\end{itemize}

These elements constitute the full mathematical infrastructure for the dynamical evolution of semantic structure.  
Chapter~\ref{ch:rsvp-euler-lagrange} continues the analysis with stability cones, attractor structure, and intrinsic flow morphisms.
