\chapter{Verified Identity and Quine\textendash Stable Personas}
\label{ch:verified-identity}

\section{Introduction}

Identity in the single\textendash shard semantic universe is not a username,
a cryptographic token, or a profile page.  
It is a \emph{quine\textendash stable semantic structure}:
a self\textendash reproducing hypergraph whose embeddings, patches, and
visible actions are internally coherent and externally verifiable.

This chapter constructs the architecture of
\emph{verified identity} in the Polycompiler:
an identity system in which every persona is a quine\textendash stable,
curvature\textendash bounded, and patch\textendash certifiable object,
and where impersonation, coercion, and epistemic drift are
structurally impossible.

\section{Personas as Typed Quine Objects}

A \emph{persona} $\mathfrak{p}$ is defined as a triple:
\[
\mathfrak{p} = (\mathcal{G}_{\mathfrak{p}},\, \iota_{\mathfrak{p}},\, \Pi_{\mathfrak{p}}),
\]
where:

\begin{itemize}
  \item $\mathcal{G}_{\mathfrak{p}}$ is a Polyxan hypergraph encoding the
        semantic skeleton of the persona;
  \item $\iota_{\mathfrak{p}} : \mathrm{Inc}(\mathcal{G}_{\mathfrak{p}}) \to \mathcal{M}$
        is the embedding into semantic space;
  \item $\Pi_{\mathfrak{p}}$ is the persona\textendash level patch algebra,
        containing all self\textendash written amendments, corrections, and claims.
\end{itemize}

\begin{definition}[Quine\textendash Stable Persona]
A persona $\mathfrak{p}$ is \emph{quine\textendash stable} if
\[
\mathcal{G}_{\mathfrak{p}}
  \;\cong\; \mathcal{Q}^\infty(\mathcal{G}_{\mathfrak{p}}),
\]
i.e.\ it is a fixed point of the quine completion operator.
\end{definition}

Every such persona contains an internally self\textendash consistent,
minimal description of itself.  
In the Polycompiler context,
this makes the persona:

\begin{itemize}
  \item non\textendash duplicable,
  \item non\textendash forgeable,
  \item non\textendash severable (no partial fork is quine\textendash stable),
  \item globally interpretable through its quine witness.
\end{itemize}

\section{Identity as Curvature and Entropy Bounds}

The RSVP embedding of a persona induces continuous curvature and entropy:
\[
\mathcal{R}_{\mathfrak{p}}
  := \mathcal{R} \circ \iota_{\mathfrak{p}},\qquad
S_{\mathfrak{p}}
  := S \circ \iota_{\mathfrak{p}}.
\]

\begin{definition}[Curvature\textendash Safe Persona]
A persona $\mathfrak{p}$ is curvature\textendash safe if
\[
\sup_{x \in \iota_{\mathfrak{p}}(\mathcal{G}_{\mathfrak{p}})}
|\mathcal{R}_{\mathfrak{p}}(x)| \;<\; K_{\max},
\]
where $K_{\max}$ is a global governance constant.
\end{definition}

\begin{definition}[Entropy\textendash Controlled Persona]
A persona $\mathfrak{p}$ satisfies entropy control if
\[
\inf_{x \in \iota_{\mathfrak{p}}(\mathcal{G}_{\mathfrak{p}})} 
S_{\mathfrak{p}}(x) \;\ge\; S_{\min},
\]
ensuring minimal semantic resolution.
\end{definition}

These two conditions guarantee that a persona’s actions cannot generate
supercritical curvature wells or adversarial entropy sinks,
both of which would distort visibility or destabilise shared meaning.

\section{Polycompiler Identity Certificates}

The Polycompiler produces a \emph{verification certificate} $\mathrm{Cert}(\mathfrak{p})$
for each persona:
\[
\mathrm{Cert}(\mathfrak{p})
  = \left(
      \mathrm{NF}_{\mathbb{P}}(\Pi_{\mathfrak{p}}),\;
      \mathrm{NF}_{\mathrm{EHR}}(\mathcal{G}_{\mathfrak{p}}),\;
      \mathrm{Flat}(\iota_{\mathfrak{p}})
    \right),
\]
consisting of:

\begin{enumerate}
  \item the \emph{normalised patch algebra} of declarations,
  \item the \emph{EHR\textendash normal form} of the persona’s hypergraph,
  \item the curvature\textendash and entropy\textendash flat embedding certificate.
\end{enumerate}

\subsection{Non\textendash forgeability}

A certificate is non\textendash forgeable because:

\begin{itemize}
  \item the patch algebra normalisation is canonical,
  \item the EHR normal form is unique,
  \item the embedding flatness verification requires solving an RSVP PDE,
  \item each of the above encodes a quine witness.
\end{itemize}

Thus impersonation is equivalent to defeating:

\[
\mathrm{NF}_{\mathbb{P}},\quad
\mathrm{NF}_{\mathrm{EHR}},\quad
\mathrm{Flat}(\iota),
\]
which is impossible without violating constitutional invariants.

\section{Bidirectional Identity Links}

Personas link to one another not by following each other,
but by forming \emph{bidirectional semantic identities}.
A link $\mathfrak{p} \leftrightarrow \mathfrak{q}$ must satisfy:

\begin{enumerate}
  \item quine\textendash stability of both personas,
  \item curvature\textendash match constraints:
        \[
        \left|
        \mathcal{R}_{\mathfrak{p}}(x)
        - \mathcal{R}_{\mathfrak{q}}(y)
        \right|
        < \epsilon,
        \]
  \item entropy\textendash compatibility,
  \item sheaf\textendash gluing of identity tags across the link,
  \item visibility symmetry.
\end{enumerate}

A link failing any constraint is not merely disallowed;
it cannot exist as a well\textendash typed object in the Polycompiler universe.

\section{Identity Forks, Merges, and Stability}

A persona may evolve by patches, but it cannot fork without breaking quine stability.

\begin{proposition}[No Forking Theorem]
If $\mathfrak{p} \to \mathfrak{p}_1$ and $\mathfrak{p} \to \mathfrak{p}_2$
are divergent patch evolutions,
then at most one of $\mathfrak{p}_1$, $\mathfrak{p}_2$
can be quine\textendash stable.
\end{proposition}

\begin{proof}
Quine stability requires that the EHR normal form be unique.
If both $\mathfrak{p}_1$ and $\mathfrak{p}_2$ are quine\textendash stable,
then
\[
\mathcal{Q}^\infty(\mathfrak{p}_1)
 = \mathcal{Q}^\infty(\mathfrak{p}_2).
\]
But divergent patches imply distinct EHR normal forms,
contradiction.
\end{proof}

Similarly, persona merges are governed by curvature and entropy compatibility.

\begin{proposition}[Merge Stability]
Two personas can merge into a single quine\textendash stable persona
iff their curvature and entropy profiles differ by a
square\textendash integrable, globally decreasing perturbation.
\end{proposition}

\section{The Verified Identity Theorem}

We now state the identity theorem of the governance layer.

\begin{theorem}[Verified Identity Theorem]
Every persona in the single\textendash shard semantic universe,
under continuous RSVP flow, discrete EHR rewriting,
and patch reconciliation, converges in finite computational steps to a
unique quine\textendash stable identity with a non\textendash forgeable verification certificate.
\end{theorem}

\begin{proof}
The curvature\textendash entropy descent (Chapter~32),
EHR termination (Chapter~25),
and patch confluent normalisation (Chapter~80)
jointly ensure that:

\[
\lim_{t \to t^\ast}
\mathcal{G}_{\mathfrak{p}}(t)
  = \mathcal{Q}^\infty(\mathcal{G}_{\mathfrak{p}}),
\]
and that the embedding $\iota_{\mathfrak{p}}$ becomes flat.
All certificates converge to canonical normal forms,
and thus the identity stabilises.
\end{proof}

\section{Final Cadence}

An identity is not a label.  
It is not a signature.  
It is not a token.

It is a quine\textendash fixed, curvature\textendash safe,
entropy\textendash balanced persona whose claims, actions, and links
are verifiable in the one\textendash shard universe of meaning.

\begin{quote}
To be someone in the Polycompiler is  
to be the unique quine your actions converge to.
\end{quote}
