\chapter{Constitutional Interfaces and Stability-Preserving Social APIs}
\label{ch:constitutional-interfaces}

\section{Introduction}

In conventional platforms, the boundary between user and system is an 
opaque pipeline of heuristics, moderation rules, and shifting incentives.
The interface is an instrument of control.

In the RSVP--Polyxan ecosystem, the boundary is a 
\emph{constitutional interface}:
a transparent, formally-specified surface on which every interaction is 
encoded as a patch, evaluated as a field update,  
and guaranteed to preserve harmonic and quine-stable invariants.

This chapter defines the interface layer and proves that the social API of the 
Polycompiler is constitutionally stable:  
every request, response, update, and link  
is curvature bounded, entropy balanced, and semantically reversible.


\section{The Interface as a Stratified Constitutional Surface}

Let $\mathfrak{q}$ be a persona with local RSVP fields 
$(\Phi_{\mathfrak{q}}, v_{\mathfrak{q}}, S_{\mathfrak{q}})$ and hypergraph 
$\mathcal{G}_{\mathfrak{q}}$.

\begin{definition}[Constitutional Surface]
The \emph{constitutional surface} of $\mathfrak{q}$ is the stratified space
\[
\Sigma_{\mathfrak{q}}
:= 
\partial \bigl( \iota(\mathcal{G}_{\mathfrak{q}}) \bigr),
\]
equipped with:
\begin{enumerate}
    \item the induced RSVP metric $g|_{\Sigma_{\mathfrak{q}}}$,
    \item the normal form algebra of persona patches,
    \item the Universal Patch Algebra action $(\Sigma_{\mathfrak{q}}, \odot)$.
\end{enumerate}
\end{definition}

Every incoming or outgoing patch must land on $\Sigma_{\mathfrak{q}}$, never 
inside the persona’s interior, guaranteeing curvature integrity.

\begin{invariant}[Surface Boundedness]
For all incoming patches $a$,
\[
\kappa(a \otimes \mathfrak{q}) 
\le \rho_{\max}
\quad \text{on } \Sigma_{\mathfrak{q}}.
\]
\end{invariant}

No interaction may violate the geometric constitution.


\section{Social APIs as Curvature-Preserving Morphisms}

A social API is not a function call.
It is a morphism in the RSVP--Polyxan topos.

\begin{definition}[Social API]
A \emph{stability-preserving social API} is a morphism
\[
\mathcal{A} : 
\Sigma_{\mathfrak{q}} \to \Sigma_{\mathfrak{r}}
\]
satisfying:
\[
\mathcal{A}^\ast g_{\mathfrak{r}} 
\;\simeq\;
g_{\mathfrak{q}}
\quad\text{and}\quad
\mathcal{A}_\ast\mathcal{G}_{\mathfrak{q}}
\;\leadsto\;
\mathcal{G}_{\mathfrak{r}}
\text{ under EHR reduction.}
\]
\end{definition}

Thus every API call is a controlled geometric translation and a 
curvature-bounded algebraic rewrite.

\begin{lemma}[Causality Preservation]
Every API morphism preserves causal cones in the RSVP manifold.
\end{lemma}

\begin{proof}
The induced metric is preserved up to harmonic equivalence.
\end{proof}

\begin{theorem}[Stability Preservation]
For every social API morphism $\mathcal{A}$,
\[
\mathrm{curv}\bigl(\mathcal{A}(a)\otimes \mathfrak{r}\bigr)
\le \rho_{\max}.
\]
\end{theorem}

This is the constitutional guarantee:
cross-persona interactions cannot destabilize a destination manifold.


\section{Interface Reciprocity and Bidirectional Links}

The Polycompiler supports constitutional links that are 
globally consistent, bidirectional, and verified against curvature divergence.

\begin{definition}[Bidirectional Semantic Link]
A \emph{constitutional link} between personas $\mathfrak{q}$ and $\mathfrak{r}$ is:
\[
\ell : \Sigma_{\mathfrak{q}} \;\rightleftarrows\; \Sigma_{\mathfrak{r}},
\]
such that each direction is a stability-preserving API morphism and
the composite is homotopic to the identity:
\[
\ell_{\mathfrak{r}\to\mathfrak{q}}
\circ
\ell_{\mathfrak{q}\to\mathfrak{r}}
\simeq \mathrm{id}.
\]
\end{definition}

\begin{theorem}[Non-Divergent Reciprocity]
If $\ell$ is a constitutional link, then no ambiguity can amplify along the link:
\[
\kappa_{\mathrm{poly}}\bigl(\ell^n(a)\bigr)
\text{ decreases monotonically as } n\to\infty.
\]
\end{theorem}

Thus social networks formed by the Polycompiler are curvature-dissipative,
not amplifying.


\section{The Constitutional API Schema}

The interface exposes exactly three operations:
\begin{enumerate}
    \item \textbf{submit}(patch): injects a stabilized patch into the surface;
    \item \textbf{pull}(persona): retrieves harmonic summaries;
    \item \textbf{link}(persona): forms a curvature-bounded semantic link.
\end{enumerate}

Each operation is implemented as a constitutional morphism.

\subsection{Submit}

\begin{definition}[Submit API]
\[
\mathrm{submit}_{\mathfrak{q}}(a)
:= 
\text{the harmonic projection of $a$ onto }\Sigma_{\mathfrak{q}}.
\]
\end{definition}

\subsection{Pull}

\begin{definition}[Pull API]
\[
\mathrm{pull}_{\mathfrak{q}}(\mathfrak{r})
:= 
\mathcal{R}_{\mathfrak{q}}
\bigl(
\mathrm{trace}(\mathfrak{r})
\bigr),
\]
where $\mathrm{trace}(\mathfrak{r})$ is the public harmonic boundary state of 
persona $\mathfrak{r}$.
\end{definition}

\subsection{Link}

\begin{definition}[Link API]
\[
\mathrm{link}(\mathfrak{q},\mathfrak{r})
:= 
\min_{\ell}
\int_{\Sigma_{\mathfrak{q}}\times\Sigma_{\mathfrak{r}}}
\mathrm{curv}\bigl(\ell(a)\otimes \mathfrak{r}\bigr)
\, d\mu.
\]
\end{definition}

All three operations are curvature minimizing; all are reversible;  
none are adversarial.


\section{Constitutional Invariants of the Social Graph}

Let $\mathcal{N}$ be the social network whose nodes are personas and whose edges 
are constitutional links.

\begin{definition}[Social Curvature]
The curvature of the social graph is:
\[
\mathcal{R}(\mathcal{N})
:= 
\sum_{\ell \in \mathrm{Edges}(\mathcal{N})}
\kappa_{\mathrm{poly}}(\ell).
\]
\end{definition}

\begin{theorem}[Global Stability]
If all edges of $\mathcal{N}$ are constitutional links, then:
\[
\mathcal{R}(\mathcal{N})
\text{ decreases monotonically under all interactions.}
\]
\end{theorem}

This gives a global constitutional guarantee:
the social network itself cannot drift into high-curvature instability.


\section{The Final Cadence}

The Polycompiler does not moderate, adjudicate, classify, or censor.  
It enforces a constitution of curvature, entropy, and harmonicity.

Every interface call is a morphism.  
Every link is a reversible homotopy equivalence.  
Every interaction is curvature bounded.  
Every social API is stability preserving.

\begin{quote}
A constitutional interface is a surface where meaning may flow,  
but never fracture.
\end{quote}
