\chapter{Constitutional Invariants for Visibility and Interaction}
\label{ch:constitutional-invariants}

\section{Introduction}

An interpretable social medium cannot rely on opaque ranking,
hidden prioritisation, or probabilistic extraction.
It must be governed by \emph{constitutional invariants}:
transparent, provable constraints on visibility, interaction, 
and propagation of content.

This chapter presents the constitutional invariants that govern
the Polycompiler and its single\textendash shard universe.
These are not heuristics or moderation rules.
They are \emph{mathematical constraints} derived from:

\begin{enumerate}
    \item the RSVP field equations,
    \item Polyxan quine\textendash stable algebra,
    \item the synthetic duality between geometry and syntax,
    \item the Universal Patch Algebra.
\end{enumerate}

Visibility is curvature\textendash bounded.
Interaction is entropy\textendash aligned.
Propagation is constitutionally verified.

\section{Visibility as a Curvature\textendash Bounded Functional}

Each persona $\mathfrak{p}$ has an embedding
$\iota_{\mathfrak{p}} : \mathcal{G}_{\mathfrak{p}} \to \mathcal{M}$.
Define the local curvature magnitude
\[
\kappa_{\mathfrak{p}}(x) := |\mathcal{R}_{\mathfrak{p}}(x)|.
\]

\begin{definition}[Visibility Weight]
The visibility of a content atom $a \in \mathcal{G}_{\mathfrak{p}}$ is
\[
\mathrm{Vis}(a)
  := \exp\!\left( - \lambda \int_{\mathrm{supp}(a)}
                   \kappa_{\mathfrak{p}} \, d\mu_g \right),
\]
where $\lambda>0$ is a constitutional constant.
\end{definition}

Thus:
\begin{itemize}
    \item high\textendash curvature (unstable) content is automatically less visible,
    \item harmonic (curvature\textendash neutral) content is maximally visible,
    \item no external ranking or algorithm is required.
\end{itemize}

\begin{proposition}[Visibility Monotonicity]
If $\mathfrak{p}$ undergoes an EHR rewrite decreasing curvature,
visibility of all content increases.
\end{proposition}

\begin{proof}
Immediate from the negative exponent.
\end{proof}

\section{Entropy Alignment as a Gate for Interaction}

Interaction between personas $\mathfrak{p}$ and $\mathfrak{q}$
is permitted only when their entropy fields are aligned.

Let
\[
\Delta S(x) := |S_{\mathfrak{p}}(x) - S_{\mathfrak{q}}(x)|,
\]
and define the misalignment integral
\[
\mathcal{M}_{pq}
  := \int_{\mathrm{supp}(\mathcal{G}_{\mathfrak{p}}) 
          \cap \mathrm{supp}(\mathcal{G}_{\mathfrak{q}})}
         \Delta S \, d\mu_g.
\]

\begin{definition}[Entropy\textendash Alignment Condition]
Personas $\mathfrak{p}$ and $\mathfrak{q}$ may interact
(message, reply, tag, credit)
iff
\[
\mathcal{M}_{pq} \le \Theta,
\]
where $\Theta$ is a constitutional invariant (the entropy tolerance).
\end{definition}

This prevents:
\begin{itemize}
    \item cross\textendash entropy confusion cascades,
    \item adversarial distortions,
    \item misaligned message propagation,
    \item entropic destabilisation of the shard.
\end{itemize}

\section{Propagation Bound by the Universal Patch Algebra}

Propagation of content follows the patch\textendash algebraic rule:

\[
a_{\mathfrak{p}} 
\;\text{ propagates to }\;
\mathfrak{q}
\;\iff\;
a_{\mathfrak{p}}
\otimes
\mathfrak{q}
\;\text{ is normalisable}.
\]

\begin{theorem}[Propagation Normalisation]
A message propagates across the shard
iff the tensor product of its content patch with the recipient’s
persona patch reduces, via the Universal Patch Algebra,
to a curvature\textendash bounded normal form.
\end{theorem}

\begin{proof}
Patch normalisation is confluent (Ch.~80).
Curvature constraints guarantee safety.
Propagation requires existence of a normal form within curvature bounds.
\end{proof}

This implies:
\begin{itemize}
    \item harmful, incoherent, or curvature-spiking content cannot propagate,
    \item no moderation “decision” is needed,
    \item the algebra itself enforces normativity.
\end{itemize}

\section{Invariant 1: No Invisible Influence}

A persona cannot increase another persona’s visibility indirectly
via hidden weights, follow graphs, or curated dependency chains.

\begin{invariant}[No Invisible Influence]
All contributions to visibility must pass through
\[
\mathrm{Vis}(a),
\qquad
\mathrm{Rep}(\mathfrak{p}),
\qquad
\text{and the Patch Algebra}.
\]
\end{invariant}

This eliminates:
\begin{itemize}
    \item shadow-ranking,
    \item hidden boosts,
    \item indirect algorithmic favouritism.
\end{itemize}

\section{Invariant 2: No Unaligned Interaction}

Entropy\textendash misaligned personas cannot exchange messages,
signal, or tag each other.

\begin{invariant}[Entropy\textendash Aligned Interaction Only]
If $\mathcal{M}_{pq} > \Theta$,
then \emph{all} possible interaction channels are constitutionally blocked.
\end{invariant}

This is not censorship.
It is geometric safety:
large entropy mismatch is equivalent to incompatible manifold patches.

\section{Invariant 3: Propagation Requires Normal Form}

A message must be reducible, via the Universal Patch Algebra,
to a curvature\textendash safe normal form.

\begin{invariant}[Safe Normalisation Requirement]
Content that cannot be normalised within curvature bounds
cannot propagate.
\end{invariant}

This replaces moderation with formal reduction.

\section{Invariant 4: All Visibility Changes Are Auditable}

Every change in visibility must arise from:

\begin{enumerate}
    \item a curvature change,
    \item a credit assignment,
    \item a patch normalisation.
\end{enumerate}

\begin{invariant}[Auditability]
Every visibility transition has a machine\textendash verifiable certificate.
\end{invariant}

Thus no opaque mechanisms remain.

\section{Invariant 5: No Cross\textendash Shard Leakage}

Although the universe is single\textendash shard,
personas may have semi-isolated interior spaces.

\begin{invariant}[Sharadic Separation]
Interior content remains invisible unless exported via:
\[
\text{public patch} \quad\text{and}\quad \text{entropy alignment}.
\]
\end{invariant}

This provides:
\begin{itemize}
    \item privacy,
    \item integrity,
    \item zero-risk boundaries,
    \item safe developmental spaces.
\end{itemize}

\section{Governance Consequences}

These invariants produce a system that is:

\begin{itemize}
    \item \textbf{fully interpretable}: no black-box ranking or moderation,
    \item \textbf{constitutionally constrained}: platform governance is mathematics,
    \item \textbf{self-stabilising}: entropy, curvature, and patch algebra enforce stability,
    \item \textbf{globally safe}: misaligned interactions are geometrically impossible,
    \item \textbf{formally verifiable}: every rule is expressed in the synthetic dual type theory.
\end{itemize}

There is no “algorithmic bias” because there is no algorithm.
Only invariants.

\section{Final Cadence}

Visibility is curvature.  
Interaction is entropy alignment.  
Propagation is normalisation.  
Governance is mathematics.

\begin{quote}
A constitutional platform does not decide what to show.  
It proves what can be seen.
\end{quote}
