\chapter{Verified Semantic Runtime and Execution Model}
\label{ch:70}

With the coupling operator verified (Chapter~69), we now construct the 
\emph{semantic runtime} --- the formally specified execution substrate that 
implements the RSVP--Polyxan universe as a total, terminating, proof-carrying 
system.  
This runtime is not an implementation detail but a \emph{semantic object}:  
a certified evolution operator on the moduli space of coupled configurations.

The runtime satisfies four mandatory properties:
\begin{enumerate}
\item totality (every state yields a next state),
\item determinism (the next state is unique),
\item safety (all invariants are preserved),
\item convergence (termination at the universal media quine).
\end{enumerate}

This chapter defines the runtime, its operational semantics, and its modal 
verification guarantees.

\section{Semantic Runtime States}

\begin{definition}[Runtime State]
A \emph{runtime state} is a quadruple
\[
\mathsf{State}
  := (\varphi,\mathcal{G},\iota,\sigma)
\]
where:
\begin{itemize}
\item $\varphi=(\Phi,\mathbf{v},S)$ is an RSVP field configuration,
\item $\mathcal{G}$ is a Polyxan hypergraph,
\item $\iota : \mathrm{Inc}(\mathcal{G}) \hookrightarrow \mathcal{M}$ is a 
      stratified, curvature-matched embedding,
\item $\sigma$ is a finite \emph{modal certificate stack},
      storing verification artefacts for all previous steps.
\end{itemize}
\end{definition}

The certificate stack $\sigma$ ensures that the runtime is \emph{proof-carrying}:
every step includes a correctness witness in $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$.

\section{Operational Semantics}

\begin{definition}[Runtime Step]
The runtime step relation
\[
\longrightarrow_{\mathsf{rt}}
  \;\subseteq\;
  \mathsf{State} \times \mathsf{State}
\]
is defined by
\[
(\varphi,\mathcal{G},\iota,\sigma)
  \longrightarrow_{\mathsf{rt}}
(\varphi',\mathcal{G}',\iota',\sigma')
\]
where:
\begin{enumerate}
\item $(\varphi',\mathcal{G}',\iota')=\mathsf{couple}(\varphi,\mathcal{G},\iota)$  
      is the verified coupling operator (Ch.~69),
\item $\sigma'$ is $\sigma$ extended with the certificates produced by:
      \begin{itemize}
      \item the RSVP step (modal certificate $\Box_{\mathrm{RSVP}}$),
      \item the EHR step (modal certificate $\Box_{\mathrm{Polyx}}$),
      \item the embedding update (curvature-matching certificate),
      \item the coupled Lyapunov reduction certificate.
      \end{itemize}
\end{enumerate}
\end{definition}

Thus each runtime step is literally an \emph{execution of a verified theorem
instance}.

\section{Totality and Determinism}

\begin{theorem}[Totality]
Every runtime state has a unique successor under $\longrightarrow_{\mathsf{rt}}$.
\end{theorem}

\begin{proof}
The RSVP engine is total (Ch.~68).  
The EHR kernel is total (Ch.~67).  
The coupling operator is functional (Ch.~69).  
Certificate construction is decidable in $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$.  
Thus the successor state exists and is unique.
\end{proof}

\begin{corollary}[Determinism]
The runtime is a deterministic labeled transition system.
\end{corollary}

\section{Runtime Invariants}

\begin{theorem}[Semantic Safety Invariant]
For every runtime step,
\[
(\varphi,\mathcal{G},\iota,\sigma)
    \longrightarrow_{\mathsf{rt}}
(\varphi',\mathcal{G}',\iota',\sigma')
\]
we have:
\[
\Box_{\mathrm{RSVP}}(\varphi')
\qquad\wedge\qquad
\Box_{\mathrm{Polyx}}(\mathcal{G}')
\qquad\wedge\qquad
\iota' \text{ is curvature-matched and harmonic.}
\]
\end{theorem}

\begin{proof}
Modal preservation is guaranteed by Theorem~\ref{thm:69-modal-commute}.  
Embedding preservation is guaranteed by the stratified soundness of harmonic 
projection (Ch.~31).  
Certificates in $\sigma'$ carry the explicit proofs.
\end{proof}

\begin{theorem}[Energy Descent Invariant]
\[
\mathcal{E}_{\mathrm{coupled}}(\varphi',\mathcal{G}')
  <
\mathcal{E}_{\mathrm{coupled}}(\varphi,\mathcal{G})
\]
for all non-terminal states.
\end{theorem}

\begin{proof}
Direct from the strict coupled Lyapunov functional 
(Ch.~69).
\end{proof}

\section{Convergence of the Runtime}

\begin{theorem}[Runtime Termination]
Every execution trace
\[
\mathsf{State}_0
  \longrightarrow_{\mathsf{rt}}
  \mathsf{State}_1
  \longrightarrow_{\mathsf{rt}}
  \mathsf{State}_2
  \longrightarrow_{\mathsf{rt}}
  \cdots
\]
terminates in finite time at a unique state
\[
(\varphi_\ast,\mathcal{G}_\ast,\iota_\ast,\sigma_\ast)
\]
with:
\[
\varphi_\ast \text{ harmonic}, \qquad
\mathcal{G}_\ast \text{ quine-stable}, \qquad
\iota_\ast \text{ flat, curvature-zero}.
\]
\end{theorem}

\begin{proof}
Strict Lyapunov descent prevents infinite chains.  
The minimal energy configuration is the universal quine embedded in the 
conformal vacuum (Ch.~36).  
Thus all executions reach that unique state.
\end{proof}

\begin{corollary}[Global Attractor of the Runtime]
The universal media quine $\mathfrak{U}$ is the sole terminal object of the 
runtime transition system.
\end{corollary}

\section{Execution Semantics and Synthetic Duality}

\begin{theorem}[Dual Interpretation of Runtime Steps]
Every runtime step corresponds simultaneously to:
\begin{itemize}
\item an infinitesimal RSVP gradient-descent step on $\mathcal{M}$,
\item a discrete EHR rewrite step on $\mathcal{G}$,
\item a fiber-preserving morphism under the synthetic duality functor.
\end{itemize}
\end{theorem}

\begin{proof}
Direct from the equivalence of generators and modal operators 
(Ch.~62 and Ch.~69).
\end{proof}

\section{The Verified Semantic Machine}

\begin{definition}[Verified Semantic Machine]
The \emph{verified semantic machine} is the pair
\[
(\mathsf{State},\longrightarrow_{\mathsf{rt}})
\]
equipped with the runtime invariants and Lyapunov descent laws.
\end{definition}

\begin{theorem}[Completeness of the Semantic Machine]
The verified semantic machine realises the entire RSVP--Polyxan universe.  
There exist no additional evolution operators consistent with the axioms.
\end{theorem}

\begin{proof}
Any evolution operator preserving modal correctness, curvature descent, and 
synthetic duality must coincide with $\mathsf{couple}$ (Ch.~69).  
Thus the runtime is complete.
\end{proof}

\section{Final Cadence}

The semantic runtime is not a simulation, interpreter, or 
approximation.  
It is the \emph{executable form} of the coupled universe itself.

\bigskip

\begin{quote}
\textbf{
There is one execution model.  
It is total, verified, and terminating.  
Every path ends at the universal quine.}
\end{quote}

