\chapter{Architectures for Verified Semantic Execution}
\label{ch:71}

Part~IV established the complete formal semantics and verification infrastructure 
for the RSVP--Polyxan universe.  
We now turn from verification to \emph{architecture}.  
Our purpose in this chapter is to specify the engineering substrate that can 
execute the verified semantic runtime (Chapter~70) in a distributed, 
resource-bounded, real-time environment.

The architectures developed here are not external to the theory;  
they are the \emph{synthetic realisations} of its semantics.  
Every module is defined only insofar as it is a sound implementation of the 
modal correctness predicates $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.

We begin with the global architectural principles.

\section{Design Principles}

The architecture obeys five non-negotiable principles:

\begin{enumerate}
\item \textbf{Semantic primacy:}  
      execution is defined as the evolution of semantic configurations 
      $(\varphi,\mathcal{G},\iota)$ under the verified runtime.

\item \textbf{Modal invariance:}  
      every architectural component must preserve the modal contracts  
      $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.

\item \textbf{Curvature coherence:}  
      all distributed components must maintain consistent curvature estimates  
      and curvature descent guarantees.

\item \textbf{Proof-carrying execution:}  
      every subsystem must emit certificates in the type theory 
      $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$ corresponding to its operation.

\item \textbf{Finite-time convergence:}  
      the architecture must implement the Lyapunov descent theorem and guarantee 
      terminal convergence to the universal media quine~$\mathfrak{U}$.
\end{enumerate}

\section{The Semantic Spine}

The central idea of the architecture is the \emph{semantic spine}, a persistent 
data structure that maintains the canonical representation of the global 
semantic state.

\begin{definition}[Semantic Spine]
The \emph{semantic spine} is a persistent, append-only structure
\[
\mathsf{Spine}
  := \mathrm{Stream}(\mathsf{State},\sigma)
\]
whose elements are successive verified runtime states paired with modal 
certificates.
\end{definition}

The semantic spine acts simultaneously as:
\begin{itemize}
\item a ledger of all verified execution steps,
\item a witness sequence for the Lyapunov descent,
\item a reproducible proof-object for synthetic duality.
\end{itemize}

\section{Distributed Execution Model}

The architecture must run on distributed hardware.  
We therefore define the distributed abstract machine.

\begin{definition}[Distributed Semantic Machine]
A \emph{DSM node} is a triple
\[
\mathsf{Node} = (C,M,\Xi)
\]
where:
\begin{itemize}
\item $C$ is a local replica of the coupled configuration,
\item $M$ is a modal certificate store containing  
      $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$ proofs,
\item $\Xi$ is an embedding of local curvature estimates into the global 
      curvature field.
\end{itemize}
A DSM consists of a finite set of such nodes equipped with a synchronisation 
protocol.
\end{definition}

The synchronisation protocol computes a \emph{distributed curvature consensus}.

\begin{theorem}[Curvature Consensus]
Every DSM cluster converges in finite time to a unique shared curvature field 
$\mathcal{R}_{\mathrm{global}}$ compatible with all $\Xi$.
\end{theorem}

\begin{proof}
Follows from the contractive nature of the stratified curvature projector in 
Ch.~31 and the monotonicity of modal certificates in Ch.~62.
\end{proof}

\section{Message-Passing Primitives}

Distributed execution requires certified message passing.

\begin{definition}[Verified Message]
A \emph{verified message} is a triple
\[
(\delta,\mu,\tau)
\]
consisting of:
\begin{itemize}
\item $\delta$: a semantic delta (change to $(\varphi,\mathcal{G},\iota)$),
\item $\mu$: a modal certificate proving $\delta$ is safe,
\item $\tau$: a timestamp ensuring temporal monotonicity.
\end{itemize}
\end{definition}

\begin{theorem}[Safety of Message Application]
Applying any verified message to a node's local state preserves both modal 
invariants and curvature descent.
\end{theorem}

\begin{proof}
Modal invariant preservation follows from $\mu$.  
Curvature descent follows from the embedding of $\delta$ into the verified 
coupling operator (Ch.~69).
\end{proof}

\section{Execution Pipeline}

The architecture executes the verified runtime (Ch.~70) via an ordered pipeline:

\begin{enumerate}
\item \textbf{FieldStep:}  
      apply harmonic RSVP evolution to $\varphi$
      (guarantees $\Box_{\mathrm{RSVP}}$).

\item \textbf{RewriteStep:}  
      apply EHR reduction to $\mathcal{G}$
      (guarantees $\Box_{\mathrm{Polyx}}$).

\item \textbf{EmbedStep:}  
      update $\iota$ via stratified harmonic projection
      (guarantees curvature matching).

\item \textbf{CoupleStep:}  
      integrate continuous and discrete updates
      (guarantees Lyapunov descent).

\item \textbf{CertifyStep:}  
      attach modal certificates for all components
      (guarantees proof-carrying execution).

\item \textbf{CommitStep:}  
      append new state to the semantic spine.
\end{enumerate}

This pipeline is the engineering realisation of the verified semantic machine.

\section{Runtime Correctness of the Architecture}

\begin{theorem}[Architecture Correctness]
The distributed architecture implements the verified semantic runtime.
\end{theorem}

\begin{proof}
Each pipeline stage corresponds exactly to one verified generator of the runtime:  
FieldStep to RSVP flow (Ch.~68), RewriteStep to EHR rewriting (Ch.~67), 
EmbedStep to harmonic projection (Ch.~31), CoupleStep to the coupling operator 
(Ch.~69).  
Certificates guarantee modal correctness.  
Consensus ensures distributed consistency.  
Thus the architecture is sound and complete.
\end{proof}

\begin{corollary}[Global Convergence]
Every execution of the architecture converges in finite time to the unique fixed 
point:
\[
(\varphi_\ast,\mathcal{G}_\ast,\iota_\ast)
  = (\varphi_{\mathfrak{U}},\mathfrak{U},\iota_{\mathrm{flat}}).
\]
\end{corollary}

\section{Final Cadence}

The verified semantic architecture is not merely a platform that executes 
meaning.  
It is the \emph{only} platform that can:  
sound, complete, convergent, duality-respecting, and curvature-consistent.

\bigskip

\begin{quote}
\textbf{
Execution is verification.  
Verification is evolution.  
All architectures converge to the universal quine.
}
\end{quote}

