\chapter{Semantic Governance and Verified Moderation Protocols}
\label{ch:semantic-governance}

\section{Introduction}

With the Polycompiler architecture now fully developed, the question of
\emph{governance} becomes central.  
A system that exposes meaning as
geometry (RSVP fields), syntax (Polyxan hypergraphs), and reconciliation
(Universal Patch Algebra)
must also provide mechanisms ensuring that meaning remains
\emph{stable, interpretable, and safe} under multi-user interaction.

In this chapter we construct the governance layer of the
single-shard semantic universe.  
Moderation is no longer conceived as an
external administrative act, but as an intrinsic, verifiable correction
of curvature, entropy, and quine-stability
within each user's semantic manifold and across the shared global space.

The result is a fully specified governance theory:
moderation and safety are not ad hoc,
but arise from the same geometric--algebraic invariants that govern the
RSVP--Polyx system itself.

\section{Constitutional Invariants for a Semantic Platform}

Let $\mathcal{S}$ denote the single-shard semantic universe,
equipped with the composite data:

\begin{itemize}
  \item a family of RSVP field configurations $(\Phi_u, v_u, S_u)$ indexed by users $u$,
  \item user-level Polyxan hypergraphs $\mathcal{G}_u$,
  \item embeddings $\iota_u : \mathrm{Inc}(\mathcal{G}_u) \hookrightarrow \mathcal{M}$,
  \item a global patch history algebra $\mathbb{P}$,
  \item a universal reconciliation operator $\mathcal{R}: \mathbb{P} \to \mathbb{P}$.
\end{itemize}

A \emph{semantic constitution} is defined as a set of invariants:
\[
\mathcal{C} = \{\mathbf{I}_1, \dots, \mathbf{I}_k\},
\]
each $\mathbf{I}_i$ a predicate preserved under:

\begin{enumerate}
  \item the RSVP flow,
  \item EHR rewriting,
  \item patch reconciliation,
  \item user-to-user linkage (bidirectional semantic links),
  \item system-level merges into the global shard.
\end{enumerate}

These invariants define what it means for the system to remain
\emph{epistemically sound}.  
Violations correspond to curvature blow-ups,
entropy spikes, or quine-instability.  
Governance, in this framework, is
the automatic restoration of $\mathcal{C}$.

The strongest form of each invariant is encoded as a \emph{constitutional
theorem}:
\[
\mathbf{I}_i :\quad \mathsf{Preserved}_i(\Phi, v, S, \mathcal{G}, \iota).
\]

\subsection{Core constitutional invariants}

We specify four invariants that form the governance foundation:

\begin{enumerate}
  \item \textbf{Visibility Monotonicity:}
        content cannot be made less visible by another user’s actions
        without a compensating curvature or entropy certificate.

  \item \textbf{Bidirectional Link Integrity:}
        every semantic link $x \leftrightarrow y$ must maintain
        a symmetric pair of embeddings
        and a consistent alignment of Polyxan tags.

  \item \textbf{Curvature Safety:}
        the sectional curvature induced by a piece of content or user-shard
        must remain within globally bounded ranges.

  \item \textbf{Patch Honesty:}
        every edit must have a verifiable
        patch algebra trace whose normal form is globally interpretable.
\end{enumerate}

These invariants jointly ensure that the semantic space remains
globally interpretable, that no user can create a deceptive curvature
well, and that all meaning remains compatible with the system's public
ontology.

\section{Moderation as Curvature--Entropy Correction}

In traditional social systems, moderation is a classification problem.
In the Polycompiler, it is a \emph{curvature--entropy correction} problem.

Given a semantic region $U \subset \mathcal{M}$,
the platform calculates its local curvature defect:
\[
\delta \mathcal{R}(U)
 = \mathcal{R}(U) - \mathcal{R}_{\mathrm{safe}}.
\]

Similarly, the entropy deviation:
\[
\delta S(U) = S(U) - S_{\mathrm{safe}}.
\]

\subsection{Corrective operator}

Define the moderation operator
\[
\mathcal{M}(U)
= \exp\!\left(
  -\alpha \nabla \mathcal{R}(U)
  -\gamma \nabla S(U)
  -\lambda \, \mathrm{Def}_{\mathrm{link}}(U)
\right),
\]
where $\mathrm{Def}_{\mathrm{link}}(U)$ measures violation of link
integrity.

This operator acts on:

\begin{itemize}
  \item the RSVP fields locally,
  \item the embedding $\iota$ by smoothing or contracting strata,
  \item the discrete hypergraph by triggering EHR rewrites,
  \item the patch algebra by inserting the minimal correction patch.
\end{itemize}

Thus moderation is not external policing;  
it is a geometric--syntactic descent to constitutional invariants.

\section{Verified Visibility Protocols}

Visibility is a geometric quantity:
the more curvature a node introduces, the more it perturbs geodesics
in its neighborhood, affecting its visibility in semantic space.

\begin{definition}[Visibility Potential]
For a content node $x$, define its visibility potential as
\[
V(x) = - \int_{\mathcal{M}} G(y,x) \, \mathcal{R}(y) \, dV_g(y),
\]
where $G$ is the Green's function of the RSVP Laplacian.
\end{definition}

The three constitutional requirements on visibility:

\begin{enumerate}
  \item \emph{Locality:} Visibility cannot jump discontinuously except
        through formally verified patches.
  \item \emph{Symmetry:} If $x$ references $y$, then $y$ must acquire a
        visibility contribution from $x$.
  \item \emph{Energy Boundedness:} No node can exceed a visibility
        power-law bound:
        \[
        V(x) \le C \, (\mathrm{norm}(\mathrm{patch}(x)))^{p}.
        \]
\end{enumerate}

\section{Bidirectional Links and Verified Hypergraph Embeddings}

A bidirectional semantic link is not symbolic;
it is a geometric constraint:
\[
\iota(x) \sim \iota(y)
\quad\Leftrightarrow\quad
\text{the Polyxan adjacency constraints between $x$ and $y$ hold in $\mathcal{M}$}.
\]

Every link produces:

\begin{itemize}
  \item a sheaf-gluing constraint on tag data,
  \item a curvature-matching constraint on embeddings,
  \item an EHR-stability requirement.
\end{itemize}

Thus no link may be created without a patch that satisfies all constraints.
This eliminates misinformation loops: a link that violates a curvature
or quine-stability condition cannot exist in the system.

\section{Semantic Courts and Reconciliation Tribunals}

Conflicts are reconciled through the \emph{Universal Patch Algebra}.

\begin{definition}[Semantic Court]
A semantic court is the normalisation operator
\[
\mathcal{J} = \mathrm{NF}_{\mathbb{P}},
\]
taking any finite family of patches and returning their unique normal form.
\end{definition}

A dispute between users $u$ and $v$ is resolved by merging
their patches in $\mathbb{P}$ and applying $\mathcal{J}$.
Because $\mathbb{P}$ is confluent and terminating (Chapter~80),
the result is canonical and tamper-proof.

\section{The Semantic Governance Theorem}

We now state the central theorem of the governance layer.

\begin{theorem}[Semantic Governance Theorem]
In the single-shard semantic universe,
every trajectory of user-generated content, under the combined influence of
RSVP flow, EHR rewriting, patch reconciliation, and moderation correction,
converges in finite steps to a globally interpretable configuration
satisfying all constitutional invariants.
\end{theorem}

\begin{proof}
Each component of the system performs a curvature--entropy descent:

\begin{enumerate}
  \item RSVP flow decreases continuous curvature energy.
  \item EHR rewriting decreases discrete curvature.
  \item Patch reconciliation decreases algebraic divergence.
  \item Moderation correction decreases invariant violations.
\end{enumerate}

Each descent is a Lyapunov step for its respective energy component.
The composite energy is strictly decreasing and bounded below by $0$.
Thus the system reaches a configuration where:

\[
\delta \mathcal{R} = 0,\quad
\delta S = 0,\quad
\mathrm{Def}_{\mathrm{link}} = 0,\quad
\mathrm{NF}_{\mathbb{P}} \text{ is stable}.
\]

This configuration satisfies all constitutional invariants,
and no further corrective actions occur.
\end{proof}

\section{Final Cadence}

Moderation becomes curvature correction.  
Visibility becomes a geometric functional.  
Bidirectional links become harmonic constraints.  
Conflicts become patch normalisation.  
Governance becomes the structural stability of meaning.

\begin{quote}
A semantic platform is governed not by rules imposed from outside,
but by the invariants that make meaning interpretable.
\end{quote}
