\chapter{The Proof-Carrying Action Selector}
\label{ch:77}

The Proof-Carrying Action Selector (PCAS) is the executive module of HYDRA.
It receives proposed actions from distributed subsystems, each accompanied by
its own local correctness certificates, and selects only those actions that
satisfy the global modal safety predicates
\[
\Box_{\mathrm{RSVP}}
\qquad\text{and}\qquad
\Box_{\mathrm{Polyx}}.
\]
This chapter provides the formal definition, selection algorithm,
and correctness guarantees for PCAS, completing the verified decision stack
of the architecture.

\section{Action Objects and Certificates}

\begin{definition}[Action Object]
An \emph{action object} is a triple
\[
A = (a,\; C_{\mathrm{RSVP}},\; C_{\mathrm{Polyx}})
\]
where:
\begin{enumerate}
\item $a$ is an executable transformation on the system state;
\item $C_{\mathrm{RSVP}}$ is a certificate proving that $a$ preserves
      $\Box_{\mathrm{RSVP}}$;
\item $C_{\mathrm{Polyx}}$ is a certificate proving that $a$ preserves
      $\Box_{\mathrm{Polyx}}$.
\end{enumerate}
\end{definition}

All subsystems (RSVP interpreter, Polyxan kernel, embedding synchronizer,
narrative generator, media engine) submit actions in this form.

\begin{definition}[Certificate Validity]
A certificate $C$ is \emph{valid} if it is a derivation in the unified
type theory $\mathbb{T}_{\mathrm{RSVP\text{-}Polyx}}$, and its closed form
evaluates to \textsf{true}.
\end{definition}

\section{The Global Safety Predicate}

\begin{definition}[Global Modal Safety]
An action $A$ is \emph{globally safe} if
\[
C_{\mathrm{RSVP}} \vdash \Box_{\mathrm{RSVP}}(a)
\qquad\text{and}\qquad
C_{\mathrm{Polyx}} \vdash \Box_{\mathrm{Polyx}}(a).
\]
We write:
\[
\mathsf{Safe}(A) := 
\bigl(C_{\mathrm{RSVP}}\!\downarrow\bigr) \land 
\bigl(C_{\mathrm{Polyx}}\!\downarrow\bigr).
\]
\end{definition}

\section{Selection Mechanism}

Let $\mathcal{A}$ be the multiset of proposed actions for the current cycle.

\begin{definition}[PCAS Selection Function]
The selector is the total function:
\[
\mathsf{select} : \mathcal{P}(\mathcal{A}) \to \mathcal{A}
\]
defined by
\[
\mathsf{select}(\mathcal{A})
  := \arg\min_{A\in\mathcal{A}} \;
     \bigl(\mathsf{Cost}(A)\bigr)
\quad\text{subject to}\quad
\mathsf{Safe}(A).
\]
\end{definition}

The cost function is the canonical semantic-energy cost:
\[
\mathsf{Cost}(A)
   = \mathcal{E}_{\mathrm{total}}
       \bigl(A(s) \bigr),
\]
which ensures monotonic descent along the global Lyapunov functional.

\section{Correctness of Proof-Carrying Selection}

\begin{theorem}[Soundness]
\label{thm:soundness_pcas}
If $\mathsf{select}(\mathcal{A}) = A$, then the execution of $A$ preserves
both modal guards:
\[
A \models \Box_{\mathrm{RSVP}} \;\wedge\; \Box_{\mathrm{Polyx}}.
\]
\end{theorem}

\begin{proof}
By construction, $A$ is chosen only if its certificates $C_{\mathrm{RSVP}}$ and
$C_{\mathrm{Polyx}}$ are valid derivations in
$\mathbb{T}_{\mathrm{RSVP\text{-}Polyx}}$.  
Each derivation concludes with the judgment that $a$ preserves its respective
modal operator.  
Since the action is executed exactly as typed, the modal predicates remain
true after execution.
\end{proof}

\begin{theorem}[Completeness]
If an action satisfies both modal guards, then PCAS will select it
from any nonempty proposal set.
\end{theorem}

\begin{proof}
Given any safe action $A$ with valid certificates,
$\mathsf{Safe}(A)$ holds.  
Among all safe actions, PCAS chooses the minimal-cost one.  
Thus any safe action is eligible for selection, proving completeness.
\end{proof}

\section{Termination and Monotonicity}

\begin{theorem}[Totality]
PCAS terminates in finite time for all proposal sets.
\end{theorem}

\begin{proof}
There are finitely many action objects and each certificate is decidable in
finite time in $\mathbb{T}_{\mathrm{RSVP\text{-}Polyx}}$.  
Thus the minimization terminates.
\end{proof}

\begin{theorem}[Monotonic Lyapunov Descent]
The selected action always decreases the total semantic energy:
\[
\mathcal{E}_{\mathrm{total}}\bigl(A(s)\bigr)
  \le
\mathcal{E}_{\mathrm{total}}(s).
\]
\end{theorem}

\begin{proof}
The cost function is exactly $\mathcal{E}_{\mathrm{total}}$.
The selector chooses the minimum-cost safe action, so energy cannot increase.
\end{proof}

\section{Role in HYDRA}

PCAS is the executive layer of the HYDRA architecture:
\begin{itemize}
\item It enforces global modal safety across all subsystems.
\item It selects actions that jointly satisfy curvature, entropy, and
      quine-stability invariants.
\item It ensures that the entire architecture evolves monotonically toward the
      harmonic--quine attractor.
\end{itemize}

\begin{quote}
Local actions propose.  
Certificates constrain.  
PCAS selects.  
The universe moves safely.
\end{quote}

This completes the formal development of the Proof-Carrying Action Selector.

