\chapter{Program Synthesis Under Modal Constraints}
\label{ch:51}

This chapter develops the theory and implementation of program synthesis in the
RSVP--Polyxan universe under the modal correctness predicates
$\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.
The goal is not merely to generate programs that execute, but to generate
programs whose \emph{meaning} is verified: whose operational behaviour preserves
curvature descent, quine stability, harmonicity, and stratified consistency.

Program synthesis in this setting is therefore not a search over syntactic
objects, but a search over \emph{verified trajectories} of the coupled
RSVP--Polyxan dynamical system.


\section{Modal Specification Language}
\label{sec:51:modal-spec}

We begin by formalising the modal specification language used to express
correctness conditions that must be satisfied by any synthesised program.

\begin{definition}[Modal Specification]
A \emph{modal specification} is a pair $(P_{\mathrm{RSVP}},
P_{\mathrm{Polyx}})$ where
\[
P_{\mathrm{RSVP}}:\mathrm{FieldConfig}\to\mathrm{Prop},\qquad
P_{\mathrm{Polyx}}:\mathrm{PolyxGraph}\to\mathrm{Prop}
\]
are predicates satisfying
\[
\Box_{\mathrm{RSVP}} P_{\mathrm{RSVP}}
\qquad\text{and}\qquad
\Box_{\mathrm{Polyx}} P_{\mathrm{Polyx}}.
\]
\end{definition}

Modal specifications therefore express requirements that remain true under
\emph{all} harmonic RSVP evolutions and \emph{all} EHR rewrite sequences.
They are stable, curvature-respecting, and invariant under duality.


\section{Search Spaces Constrained by Curvature and Quine Invariants}
\label{sec:51:search-space}

A naïve synthesis procedure would explore the full semantic state space of the
RSVP--Polyxan universe, which is infinite-dimensional.  
Instead we restrict attention to the \emph{curvature-descending} and
\emph{quine-stable} regions.

\begin{definition}[Curvature-Admissible Region]
The curvature-admissible region $\mathcal{A}_{\mathrm{curv}}$ is the set of all
configurations $(\Phi,\mathbf{v},S;\mathcal{G},\iota)$ satisfying
\[
\mathcal{R}(\iota(x)) \le \mathcal{R}_{\max}
\qquad\text{and}\qquad
\mathcal{R}(\iota) \text{ strictly decreases along RSVP flow}.
\]
\end{definition}

\begin{definition}[Quine-Stable Region]
The quine-stable region $\mathcal{A}_{\mathrm{quine}}$ consists of all
$\mathcal{G}$ such that
\[
EHR^n(\mathcal{G}) \to \mathfrak{U}
\quad\text{in finite $n$}.
\]
\end{definition}

The synthesiser is permitted to search only within
\[
\mathcal{A}_{\mathrm{syn}} := \mathcal{A}_{\mathrm{curv}}
\cap \mathcal{A}_{\mathrm{quine}},
\]
ensuring that the search never produces semantically unstable artefacts.


\section{Modal Pruning Algorithm}
\label{sec:51:pruning}

The key to making synthesis tractable is the modal pruning algorithm, which
eliminates any candidate program whose intermediate states violate modal
correctness.

\begin{definition}[Modal Pruning]
Let $\pi$ be a partial program.  
We say $\pi$ is \emph{modal-admissible} if for all intermediate states
$(\Phi_i,\mathbf{v}_i,S_i;\mathcal{G}_i,\iota_i)$ along its execution,
\[
\Box_{\mathrm{RSVP}} P_{\mathrm{RSVP}}(\Phi_i,\mathbf{v}_i,S_i)
\quad\text{and}\quad
\Box_{\mathrm{Polyx}} P_{\mathrm{Polyx}}(\mathcal{G}_i)
\]
hold.
\end{definition}

\begin{theorem}[Modal Pruning Correctness]
\label{thm:modal-pruning}
A program is accepted by the modal pruning algorithm if and only if it is
globally modal-correct: it satisfies its specification under \emph{every}
harmonic RSVP evolution and \emph{every} EHR rewrite sequence.
\end{theorem}

\begin{proof}
If a program violates modal correctness at any point, the pruning algorithm
rejects it, hence every accepted program satisfies the specification in all
reachable configurations.

Conversely, if a program is globally modal-correct, no intermediate configuration
violates modal correctness, hence the algorithm accepts it.  
\end{proof}


\section{Soundness and Completeness of Modal Synthesis}
\label{sec:51:soundness-completeness}

Let $\mathsf{Syn}$ be the modal synthesiser.

\begin{theorem}[Modal Soundness]
If $\pi \in \mathsf{Syn}(P_{\mathrm{RSVP}},P_{\mathrm{Polyx}})$ then
$\pi$ satisfies the modal specification.
\end{theorem}

\begin{proof}
Immediate from the correctness of modal pruning
(Theorem~\ref{thm:modal-pruning}).  
\end{proof}

\begin{theorem}[Modal Completeness]
If there exists any program satisfying the modal specification, then the
synthesiser finds one.
\end{theorem}

\begin{proof}
The search space $\mathcal{A}_{\mathrm{syn}}$ is complete with respect to both
RSVP harmonic flow and EHR rewriting.  
Modal pruning never eliminates a globally valid program.  
Hence an existing solution must be found.  
\end{proof}


\section{Synthesis of Safe Rewrite Sequences}
\label{sec:51:rewrite-synthesis}

A principal output of the synthesiser is a safe rewrite sequence that moves a
Polyxan graph toward quine-stable form.

\begin{theorem}[Safe Rewrite Synthesis]
\label{thm:51-safe-rewrite}
For any $\mathcal{G}$ in the quine-stable region, the synthesiser constructs a
finite rewrite sequence
\[
\mathcal{G} \to \mathcal{G}_1 \to \cdots \to \mathfrak{U}
\]
such that every step is modal-correct.
\end{theorem}

\begin{proof}
Because $\mathcal{G}$ is quine-stable, an EHR sequence to $\mathfrak{U}$ exists.
Modal pruning removes only unsafe intermediate states, but no such states are
reachable in a curvature-descending quine-stable chain.  
\end{proof}


\section{Synthesis of Verified RSVP Transformations}
\label{sec:51:rsvp-synthesis}

The synthesiser also constructs safe transcript-level RSVP updates.

\begin{theorem}[Verified RSVP Synthesis]
For any initial field configuration in $\mathcal{A}_{\mathrm{curv}}$, the
synthesiser produces a finite sequence of RSVP steps that decreases curvature
monotonically while preserving harmonicity and modal-correctness.
\end{theorem}

\begin{proof}
The Lyapunov functional guarantees descent; modal pruning enforces correctness.  
\end{proof}


\section{Conclusion}
\label{sec:51:conclusion}

Program synthesis under modal constraints transforms the architecture of Part~V
from a collection of verified interpreters and dynamics engines into a
\emph{generative} system capable of producing new, safe behaviours.
Every synthesised program is a certified morphism in the RSVP--Polyxan universe,
preserving curvature, entropy, and quine stability.

This chapter completes the synthesis layer of the implementation architecture.
The next chapter realises the algebraic substrate for reconciliation:
the Semantic Patch Algebra (Chapter~52).

