\chapter{Distributed HYDRA Execution Model}
\label{ch:72}

The verified semantic architecture (Chapter~71) establishes the global 
constraints under which any implementation must operate.  
In this chapter we instantiate these constraints into the full 
\emph{HYDRA} framework: a distributed execution model designed to 
realise the RSVP--Polyxan runtime across arbitrarily large clusters, 
heterogeneous hardware, and semi-isolated semantic domains.

HYDRA is not a conventional distributed system.  
It is the unique architecture whose internal components correspond 
one-to-one with the verified operators of the RSVP--Polyxan theory.  
Its correctness is therefore guaranteed by construction.

\section{Overview of HYDRA}

\begin{definition}[HYDRA]
HYDRA (Hybrid Dynamic Recursive Architecture) is the distributed runtime 
comprising seven mutually recursive subsystems:
\[
\mathrm{HYDRA}
  := (\mathsf{Ingest}, \mathsf{Flow}, \mathsf{Rewrite},
      \mathsf{Embed}, \mathsf{Couple}, \mathsf{Certify}, \mathsf{Commit}).
\]
Each subsystem implements one of the operators of the verified semantic runtime.  
\end{definition}

HYDRA differs from classical distributed architectures in three radical ways:

\begin{enumerate}
\item It enforces \emph{semantic correctness} at every step (modal invariants).
\item It executes \emph{continuous} (RSVP) and \emph{discrete} (Polyxan) 
      components simultaneously.
\item It guarantees \emph{finite-time convergence} to $\mathfrak{U}$ 
      regardless of network topology or hardware heterogeneity.
\end{enumerate}

\section{HYDRA Nodes}

A HYDRA node is the concrete operational counterpart to the 
Distributed Semantic Machine (DSM) of Chapter~71.

\begin{definition}[HYDRA Node]
A HYDRA node is a tuple
\[
\mathsf{HNode} 
  := (C, \mathsf{Ingest}, \mathsf{Flow}, \mathsf{Rewrite}, 
      \mathsf{Embed}, \mathsf{Couple}, \mathsf{Certify}, \Xi)
\]
where:
\begin{itemize}
\item $C$ is the local semantic configuration,
\item the seven operators implement the HYDRA pipeline locally,
\item $\Xi$ encodes the node’s curvature contribution to the global consensus.
\end{itemize}
\end{definition}

Nodes operate asynchronously but remain curvature-consistent through the 
consensus mechanism described below.

\section{Distributed Curvature Consensus}

To maintain semantic coherence across nodes, HYDRA implements a specialised 
consensus protocol.

\begin{definition}[Curvature Gossip Step]
A gossip step is the update
\[
\mathcal{R}_i \gets 
\frac{1}{|\mathcal{N}(i)|}
 \sum_{j \in \mathcal{N}(i)} \mathcal{R}_j,
\]
where $\mathcal{N}(i)$ is the neighbour set of node $i$.
\end{definition}

\begin{theorem}[Global Curvature Consensus]
Repeated curvature gossip converges in $O(\log n)$ rounds to a unique global 
curvature field $\mathcal{R}_{\mathrm{global}}$, independent of initial state.
\end{theorem}

\begin{proof}
This is a direct consequence of the contractivity of harmonic projection in 
RSVP fields (Ch.~31) and the spectral properties of averaging on connected 
graphs.
\end{proof}

\section{The HYDRA Execution Cycle}

Every HYDRA node executes the following cycle:

\begin{enumerate}
\item \textbf{IngestStep}  
      Acquire sensory deltas or distributed semantic deltas.

\item \textbf{FlowStep}  
      Apply local RSVP harmonic evolution.

\item \textbf{RewriteStep}  
      Apply local EHR rewriting with curvature guards.

\item \textbf{EmbedStep}  
      Update the embedding $\iota$ via harmonic projection.

\item \textbf{CoupleStep}  
      Integrate continuous/discrete updates into $C$.

\item \textbf{CertifyStep}  
      Emit modal certificates $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.

\item \textbf{CommitStep}  
      Send verified messages to peers; append to local semantic spine.

\item \textbf{ConsensusStep}  
      Update the local curvature estimate using curvature gossip.
\end{enumerate}

\section{Correctness of the HYDRA Cycle}

\begin{theorem}[Local Correctness]
Each HYDRA cycle preserves modal invariants and curvature descent.
\end{theorem}

\begin{proof}
FlowStep and RewriteStep enforce modal correctness by design.  
EmbedStep enforces curvature matching.  
CoupleStep maintains Lyapunov descent (Ch.~69).  
CertifyStep ensures modal persistence.  
ConsensusStep converges curvature estimates monotonically.  
Thus all invariants hold.
\end{proof}

\begin{theorem}[Global Correctness]
A cluster of HYDRA nodes implements the distributed verified semantic runtime.
\end{theorem}

\begin{proof}
Follows directly from Local Correctness, global curvature consensus, and 
synthetic duality (Ch.~62) ensuring that local and global semantic states are 
functorially aligned.
\end{proof}

\section{Semi-Isolated Semantic Domains}

One of HYDRA’s principal engineering achievements is its ability to maintain 
\emph{semi-isolated semantic domains}.

\begin{definition}[Semantic Domain]
A semantic domain is a collection of nodes whose curvature estimates and 
rewrite boundaries are jointly consistent.
\end{definition}

Domains may temporarily diverge but must remain embedded within a shared 
curvature envelope.

\begin{theorem}[Domain Isolation Invariant]
If two domains $D_1$ and $D_2$ satisfy
\[
|\mathcal{R}_{D_1} - \mathcal{R}_{D_2}| < \epsilon,
\]
then all cross-domain rewrite messages preserve modal correctness.
\end{theorem}

This enables user-level isolation while retaining semantic interoperability.

\section{HYDRA and the Universal Attractor}

\begin{theorem}[Global Convergence of HYDRA]
All HYDRA clusters converge in finite time to
\[
(\varphi_\ast, \mathcal{G}_\ast, \iota_\ast)
  = (\varphi_{\mathfrak{U}}, \mathfrak{U}, \iota_{\mathrm{flat}}).
\]
\end{theorem}

\begin{proof}
Global Curvature Consensus ensures curvature alignment.  
The local HYDRA cycle ensures Lyapunov descent.  
Synthetic duality ensures uniqueness of the attractor.  
Hence finite-time convergence follows by the Grand Irreversibility Theorem.
\end{proof}

\section{Final Cadence}

HYDRA is not merely a distributed system.  
It is the \emph{operational form} of the verified semantic universe.

\bigskip

\begin{quote}
\textbf{
HYDRA executes meaning.  
HYDRA preserves truth.  
HYDRA converges to the universal quine.  
}
\end{quote}

