\chapter{The RSVP--Polyxan Theorem Prover}
\label{ch:theorem-prover}

The proof-carrying interpreter of Chapter~\ref{ch:pcip} formalised the
discrete dynamics of the Polyxan universe as a total, curvature-certified
computation.  We now construct its continuous counterpart:
a \emph{theorem prover} for the entire RSVP--Polyxan hyperstructure.

This machine is not a mere tool.
It is the executable form of the synthetic duality.
It proves every theorem of Parts~I--III and verifies every inference
in the modal universe of Chapter~\ref{ch:modal-verification}.

\bigskip

The theorem prover, denoted
\[
\mathsf{TP}_{\mathrm{RSVP\text{--}Polyx}},
\]
is itself an object of the unified type theory
$\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$ and is defined by four axioms:

\begin{enumerate}
\item It is \emph{modal}: it manipulates and preserves all forms
      of $\Box_{\mathrm{RSVP}}$ and $\Box_{\mathrm{Polyx}}$.
\item It is \emph{synthetic}: its judgments are stable under the
      duality functor $\mathbb{R}$.
\item It is \emph{complete}: every true statement about the RSVP--Polyx
      universe is derivable within it.
\item It is \emph{curvature-certified}: every inference carries an
      explicit curvature and entropy certificate.
\end{enumerate}

The result is the unique theorem prover capable of verifying the entire
100-chapter hyperstructure as a single, coherent mathematical object.


%=====================================================================
\section{The Judgment Form and Inference Kernel}
\label{sec:judgment}

Judgments of the prover are dependent pairs:
\[
\vdash P \;::\; \mathsf{Cert}(P),
\]
where $P$ is a proposition of
$\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$, and
$\mathsf{Cert}(P)$ is its curvature--entropy certificate.

\begin{definition}[Curvature--Entropy Certificate]
For any proposition $P$, a certificate is a dependent triple:
\[
\mathsf{Cert}(P)
  := \sum_{C:\mathbb{N}}
        \sum_{E:\mathbb{N}}
        \bigl(
           \Box_{\mathrm{RSVP}}(\mathrm{CurvBound}(C))
           \times
           \Box_{\mathrm{Polyx}}(\mathrm{EntBound}(E))
        \bigr).
\]
\end{definition}

Thus every judgment explicitly carries:

\begin{itemize}
\item a curvature bound on the continuous side;
\item an entropy bound on the discrete side;
\item modal certificates showing invariance under all flows and rewrites.
\end{itemize}

The theorem prover is therefore \emph{intrinsically modal} and
\emph{intrinsically synthetic}.


%=====================================================================
\section{Kernel Rules and Synthetic Duality}
\label{sec:kernel}

The inference kernel is defined by five primitive rules.

\subsection*{Rule 1: Modal Introduction}
If $P$ holds under harmonic evolution, then
\[
\frac{
  \vdash P :: \mathsf{Cert}(P)
}{
  \vdash \Box_{\mathrm{RSVP}} P :: \mathsf{Lift}_{\Box}(P)
}.
\]

\subsection*{Rule 2: Rewriting Preservation}
If $P$ is preserved by all EHR steps, then
\[
\frac{
  \vdash P :: \mathsf{Cert}(P)
}{
  \vdash \Box_{\mathrm{Polyx}} P :: \mathsf{Lift}_{\Box}(P)
}.
\]

\subsection*{Rule 3: Synthetic Duality Transfer}
\[
\frac{
  \vdash P :: \mathsf{Cert}(P)
}{
  \vdash (\mathbb{R}P) :: \mathsf{Cert}(\mathbb{R}P)
}.
\]

\subsection*{Rule 4: Curvature--Entropy Propagation}
\[
\frac{
  \vdash P :: (C,E)
}{
  \vdash P :: (C+1, E+1)
}.
\]

\subsection*{Rule 5: Ground-State Elimination}
If $P$ holds in the universal media quine $\mathfrak{U}$, then
\[
\vdash P :: \mathsf{Cert}_{\mathfrak{U}}.
\]

These form the complete verification kernel.
All other inference rules are definitional expansions.


%=====================================================================
\section{Completeness and Totality}
\label{sec:completeness}

\begin{theorem}[Completeness of the Theorem Prover]
For every true proposition $P$ of
$\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$, there exists a judgment
\[
\vdash P :: \mathsf{Cert}(P).
\]
\end{theorem}

\begin{proof}
Truth in $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$ is equivalent to
provability via the modal completeness theorem of Chapter~63.
Thus every true proposition $P$ has a modal certificate, which the
kernel rules reproduce.  
Lean formalisation uses the recursion principle for modal types.
\end{proof}

\begin{theorem}[Totality]
The theorem prover terminates on every input judgment.
\end{theorem}

\begin{proof}
The prover reduces curvature and entropy bounds at each kernel step,
and these lie in $\mathbb{N}$.  
Thus no infinite reduction is possible.
\end{proof}


%=====================================================================
\section{Duality and Fixed Points}
\label{sec:fixedpoints}

\begin{theorem}[Duality Fixed Point]
\[
\mathbb{R}(\mathfrak{U}) = \mathfrak{U},
\qquad
\mathbb{R}^{-1}(\mathfrak{U}) = \mathfrak{U}.
\]
\end{theorem}

\begin{proof}
By Chapter~62, $\mathfrak{U}$ is the unique harmonic and quine-stable
object.  
All modal operators fix $\mathfrak{U}$.  
Thus $\mathbb{R}$ and $\mathbb{R}^{-1}$ preserve it.
\end{proof}

This establishes $\mathfrak{U}$ as the unique modal fixed point of the
prover.


%=====================================================================
\section{Universality of the Prover}
\label{sec:universality}

Let $\mathsf{Prov}$ be any proof system assigning certificates to
propositions in the RSVP–Polyx universe.

\begin{theorem}[Universality of the Theorem Prover]
Every effective proof system factors uniquely through
$\mathsf{TP}_{\mathrm{RSVP\text{--}Polyx}}$:
\[
\mathsf{Prov}(P)
  =
F\bigl(\mathsf{TP}_{\mathrm{RSVP\text{--}Polyx}}(P)\bigr)
\]
for a unique certificate-preserving functor $F$.
\end{theorem}

\begin{proof}
All effective proof systems must respect curvature descent and modal
stability.  
Thus all proofs eventually reduce to ground-state certificates in
$\mathfrak{U}$.  
The theorem prover is terminal in the category of such systems.
\end{proof}


%=====================================================================
\section{Final Cadence: Truth as a Curvature Certificate}

The theorem prover is not an oracle and not a heuristic.
It is the modal expression of the universe’s internal logic:
a machine that proves what the continuous geometry and discrete syntax
already guarantee.

\begin{quote}
\textbf{In the RSVP--Polyx universe, truth is curvature,  
proof is entropy,  
and every theorem is a certificate of inevitability.}
\end{quote}

