\chapter{Verified Interactive Media Engine}
\label{ch:58}

The Verified Interactive Media Engine is the final real-time subsystem of 
Part~V.  
Where Chapters~\ref{ch:55}--\ref{ch:57} produced visual, auditory, and
galactic-scale renderings of semantic structure, the present chapter defines 
the \emph{agent-facing interface} through which a participant interacts with 
the RSVP--Polyxan universe.

The engine ensures that all interactions preserve harmonicity, 
quine-stability, modal correctness, and curvature descent.  
Every gesture, pointer movement, or symbolic action is verified, causal,
and curvature-safe.

\section{Interactive Semantic State}
\label{sec:58:state}

Let 
\[
\mathbf{X}_t 
= (\Phi_t, \mathbf{v}_t, S_t;\; \mathcal{G}_t,\iota_t)
\]
be the full semantic state at time $t$.

\begin{definition}[Interactive State Bundle]
The interactive state is the dependent type
\[
\mathsf{IntState}_t 
:= \bigl\{ 
   x \in \mathcal{M} \;\bigm|\; 
   \exists c\in\mathrm{Inc}(\mathcal{G}_t),\ 
   x=\iota_t(c)
   \bigr\},
\]
equipped with the induced RSVP metric and the tag-sheaf structure of 
$\mathcal{G}_t$.
\end{definition}

\begin{definition}[Interactive Action]
An action is a map
\[
A : \mathsf{IntState}_t \to \mathbf{X}_{t+\Delta t}
\]
that is total, curvature-safe, and satisfies the modal guards
\[
\Box_{\mathrm{RSVP}}(A) 
\quad\text{and}\quad 
\Box_{\mathrm{Polyx}}(A).
\]
\end{definition}

Actions are never raw imperative updates; they are certified semantic
transformations.

\section{Curvature-Safe Interaction}
\label{sec:58:curvature}

\begin{definition}[Curvature Bound]
An action $A$ is curvature-safe if
\[
\mathcal{R}_{t+\Delta t}(x) 
\le \mathcal{R}_t(x)
\quad\text{for all } x\in\mathsf{IntState}_t,
\]
with equality only for harmonic points.
\end{definition}

\begin{theorem}[Curvature-Safe Interaction]
Every verified interactive action strictly preserves curvature descent, and 
therefore cannot introduce semantic instability.
\end{theorem}

This is the real-time analogue of the global Lyapunov invariants established in
Part~III.

\section{Quine-Stable Action Class}
\label{sec:58:quine}

The discrete component of interaction must respect the Polyxan structure.

\begin{definition}[Quine-Stable Action]
An action $A$ is quine-stable if, for the induced hypergraph update
$\mathcal{G}_t \to \mathcal{G}_{t+\Delta t}$,
\[
\mathcal{G}_{t+\Delta t} \equiv \mathcal{Q}^k(\mathcal{G}_t)
\quad\text{for some } k\in\{0,1,2,3\},
\]
and the Polyxan Löb bound (Chapter~25) is never exceeded.
\end{definition}

\begin{theorem}[Quine-Frame Preservation]
If $A$ is quine-stable, then the induced action on the embedding satisfies
\[
\iota_{t+\Delta t} = \iota_t \circ f
\]
for some quine-frame symmetry $f$.
\end{theorem}

Thus the skeleton displayed in Chapter~\ref{ch:57} remains consistent.

\section{Interactive Flow Integration}
\label{sec:58:flow}

Every interaction is integrated into the semantic flow.

Let $\mathsf{rsvp\_step}$ be the verified interpreter of Chapter~\ref{ch:40}, 
and let $\mathsf{ehr\_step}$ be the verified rewrite kernel of 
Chapter~\ref{ch:42}.

\begin{definition}[Interactive Flow Update]
The interactive update operator is
\[
\mathsf{int\_update} 
:= \mathsf{rsvp\_step} \circ A \circ \mathsf{ehr\_step}.
\]
\end{definition}

\begin{theorem}[Modal Safety of Interactive Flow]
For every interactive action $A$,
\[
\Box_{\mathrm{RSVP}}(\mathsf{int\_update})
\quad\text{and}\quad
\Box_{\mathrm{Polyx}}(\mathsf{int\_update})
\]
hold.
\end{theorem}

Thus interaction cannot break verification guarantees.

\section{User Gestures as Semantic Operators}
\label{sec:58:gestures}

The engine maps user gestures to verified transformations.

\begin{definition}[Gesture]
A gesture is a differentiable curve
\[
\gamma : [0,1] \to \mathsf{IntState}_t.
\]
\end{definition}

\begin{definition}[Gesture Operator]
The induced operator is
\[
\Gamma(\gamma) 
:= A_\gamma : \mathsf{IntState}_t \to \mathsf{IntState}_{t+\Delta t},
\]
where $A_\gamma$ is the curvature-safe transport along $\gamma$.
\end{definition}

\begin{theorem}[Harmonic Transport]
Gesture-induced transport preserves harmonicity:
\[
\mathcal{R}_{t+\Delta t}(\Gamma(\gamma)(x))
\le
\mathcal{R}_t(x).
\]
\end{theorem}

This ensures that pointer movement, selections, and deformations remain
epistemically safe.

\section{Semantic Lens and Focus Operators}
\label{sec:58:lens}

The engine provides interpretability tools analogous to microscope objectives.

\begin{definition}[Semantic Zoom]
The zoom operator is
\[
Z_\lambda : (\Phi,\mathbf{v},S) \mapsto 
(\lambda^2 \Phi,\; \lambda \mathbf{v},\; S),
\]
with $\lambda>0$.
\end{definition}

\begin{definition}[Focus Operator]
The focus operator is the entropy-sharpening map
\[
F_\tau : S \mapsto S - \tau\Delta S,
\quad \tau>0.
\]
\end{definition}

\begin{theorem}[Modal Preservation]
Both $Z_\lambda$ and $F_\tau$ satisfy
\[
\Box_{\mathrm{RSVP}}(Z_\lambda),\quad
\Box_{\mathrm{RSVP}}(F_\tau).
\]
\end{theorem}

Hence navigation tools remain consistent with the underlying field theory.

\section{Interactive Galaxy Manipulation}
\label{sec:58:manip}

The engine allows direct manipulation of semantic galaxies as visualised in
Chapter~\ref{ch:57}.

\begin{definition}[Galaxy Drag]
Dragging a galaxy is a curvature-preserving isometry
\[
I : \mathfrak{G}_t \to \mathfrak{G}_t
\]
induced by a gesture in $\mathsf{IntState}_t$.
\end{definition}

\begin{theorem}[Galaxy Integrity]
Galaxy drag preserves:
\begin{enumerate}
  \item curvature stratification,
  \item hypergraph quine-frame symmetry,
  \item entropy flux balance.
\end{enumerate}
\end{theorem}

Thus the user can manipulate meaning without distorting it.

\section{Interactive Narrative Operators}
\label{sec:58:narrative}

The Polyxan Narrative Generator (Chapter~\ref{ch:54}) is integrated through 
verified narrative operations.

\begin{definition}[Narrative Update]
A narrative update is a Polyxan rewrite
\[
\mathcal{G}_t \to \mathcal{G}_{t+1}
\]
initiated by user action and verified by modal guards.
\end{definition}

\begin{theorem}[Narrative Coherence]
A narrative update is coherent if and only if it preserves quine-stability of 
the narrative skeleton.
\end{theorem}

This ensures interpretable, causal storytelling.

\section{Completion: The Real-Time Semantic Loop}
\label{sec:58:loop}

The Verified Interactive Media Engine completes the part-V system by coupling:

\begin{itemize}
  \item RSVP interpreter (Ch.~41),
  \item Polyxan rewriter (Ch.~42),
  \item embedding synchronizer (Ch.~49),
  \item harmonic visuals (Ch.~55),
  \item entropy audio (Ch.~56),
  \item galaxy scenes (Ch.~57).
\end{itemize}

\begin{theorem}[Real-Time Verified Loop]
The full engine implements an interactive loop
\[
\mathbf{X}_{t+1}
=
\mathsf{int\_update}(\mathbf{X}_t)
\]
that is:
\begin{enumerate}
  \item curvature-decreasing,
  \item quine-stable,
  \item modal-safe,
  \item total and terminating,
  \item real-time executable.
\end{enumerate}
\end{theorem}

\section{Conclusion}
\label{sec:58:conclusion}

The Verified Interactive Media Engine is the climax of the implementation 
layer: a real-time, curvature-safe, quine-stable interaction system that allows
users to navigate, manipulate, and co-evolve semantic galaxies within the 
RSVP--Polyxan universe.

This completes the interactive architecture of Part~V.  
Chapter~\ref{ch:59} now extends the system to fully cross-modal embeddings.

