\chapter{Axioms and Type Theory for RSVP--Polyxan Verification}
\label{ch:axioms-type-theory}

Part~IV begins with the construction of a single, unified dependent type theory
\[
\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}},
\]
capable of simultaneously expressing continuous geometric structure,
discrete hypergraph syntax,
and the coupled dynamics that define the semantic universe.
This chapter lays down the axioms, typing rules, and modal structure
needed to formalise all results from Parts~I--III inside a proof assistant
(Lean~4), thereby converting semantic cosmology into a machine--verified mathematics.

\section{The Unified Dependent Type Theory $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$}

The type theory is stratified into three interacting layers:

\subsection{(1) Field Layer: Smooth RSVP Configurations}

RSVP fields are represented as dependent types of smooth sections of jet bundles.

\begin{align}
\mathrm{FieldConfig} 
  := \{ (\Phi,\mathbf{v},S) : \mathcal{M}\to J^2(\mathbb{R}\times T\mathcal{M}\times\mathbb{R}) 
      \mid \text{smooth} \}.
\end{align}

The jet--bundle structure ensures that all differential operators used in Parts~I--III
(Laplace--Beltrami, covariant derivatives, torsion--suppressed flows)
are definable in a purely type--theoretic manner.

\subsection{(2) Discrete Layer: Polyxan Hypergraphs}

Polyxan hypergraphs are encoded as higher inductive types with stratified incidence.

\begin{align}
\mathrm{PolyxGraph}
  := \mathrm{HIT}\_{\text{strat}}
     \big(
       V : \mathrm{Type},
       E : \mathrm{Type},
       \partial : E\to\mathrm{List}(V),
       \kappa : E\to\mathbb{Q},
       \mathsf{strata}
     \big).
\end{align}

The incidence structure is internalised as a dependent elimination principle,
allowing Lean to reason directly about vertices, hyperedges, strata, and rewrite steps.

\subsection{(3) Embedding Layer: The Semantic Embedding as a Dependent Function Type}

The semantic embedding
\[
\iota : \mathrm{Inc}(\mathcal{G})\to\mathcal{M}
\]
is represented as a dependent function into $\mathcal{M}$,
with the stratification of $\mathrm{Inc}(\mathcal{G})$ determining allowable definitional equalities.

\begin{align}
\mathrm{Embedding}(\mathcal{G})
  := \prod_{x:\mathrm{Inc}(\mathcal{G})} \mathcal{M}(x).
\end{align}

All curvature--matching and entropy--alignment conditions of Chapter~31
are turned into \emph{definitional constraints} on well--typed embeddings.

\section{Axioms of $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$}

The unified type theory satisfies five primitive axioms:

\begin{enumerate}
\item \textbf{Smoothness Axiom.}  
Every RSVP field component has a definable jet expansion, enabling Lean to compute all functional derivatives used in the embedding flow.

\item \textbf{Stratification Axiom.}  
Every Polyxan hypergraph carries a finite stratification, and all rewrite steps preserve strata.

\item \textbf{Embedding Axiom.}  
A map $\iota$ is well--typed only if the curvature--type and entropy--tag matching conditions
\[
\mathcal{R}\circ\iota = \mathcal{H}_0[\mathcal{G}],\qquad
S\circ\iota = \mathcal{H}_{\mathrm{tag}}[\mathcal{G}]
\]
hold definitionally.

\item \textbf{Flow Axiom.}  
The RSVP flow and EHR rewrite system are definable as total recursive functions with decidable step relations.

\item \textbf{Duality Axiom.}  
There exists a synthetic duality functor
\[
\mathbb{D} : \mathbf{RSVP}\leftrightarrows \mathbf{Polyx}
\]
that preserves definitional equality, modal structure, and stepwise evolution.
\end{enumerate}

Together these axioms guarantee that the coupled universe is representable as a single
computational object in dependent type theory.

\section{Modal Logic of Harmonicity and Quine--Stability}

Two modal operators govern the verification layer:

\begin{align}
\Box_{\mathrm{RSVP}}\,P 
  &:= \text{``$P$ holds for all harmonic RSVP evolutions''}, \\
\Box_{\mathrm{Polyx}}\,P
  &:= \text{``$P$ holds for all EHR rewrite steps''}.
\end{align}

As shown in §\ref{subsec:modal-lean}, both modalities are implemented in Lean
as inductively defined stability predicates with proof certificates.

The synthetic duality functor preserves both modalities:
\[
\mathbb{D}(\Box_{\mathrm{RSVP}} P) = \Box_{\mathrm{Polyx}}\mathbb{D}(P),
\qquad
\mathbb{D}^{-1}(\Box_{\mathrm{Polyx}} Q) = \Box_{\mathrm{RSVP}}\mathbb{D}^{-1}(Q).
\]

\section{Soundness and Completeness of $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$}

\begin{theorem}[Soundness and Completeness]
For any coupled configuration $(\Phi,\mathbf{v},S;\mathcal{G},\iota)$,
a proposition $P$ about the coupled dynamics is provable in Lean
if and only if it is true in the synthetic duality semantics:
\[
\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}} \vdash P
\quad\Longleftrightarrow\quad
\models_{\mathbb{D}} P.
\]
\end{theorem}

\begin{proof}[Sketch]
Soundness follows from the definitional preservation of harmonicity and quine--stability under $\mathbb{D}$.  
Completeness follows by construction: every semantic object is representable as a term in $\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}$, and all reduction steps of the RSVP and Polyxan dynamics are definable total recursive functions.  
Thus Lean can express and verify every statement that holds in the coupled universe.
\end{proof}

\section{Translation of Parts I--III into the Unified Type Theory}

Every theorem proved earlier in the monograph
(curvature identities, embedding existence, galaxy formation, renormalization laws, and irreversibility)
has a canonical encoding:

\[
\mathsf{Encode} :
\text{(Mathematical statement)}\longrightarrow
\mathbb{T}_{\mathrm{RSVP\text{--}Polyx}}.
\]

The translation preserves logical form, reduces continuous and discrete statements
into a single syntactic universe, and ensures that all proofs become Lean terms
that normalise to irreducible expressions.

\section{Formalisation of the Grand Irreversibility Theorem}

The central theorem of the entire monograph becomes:

\[
\mathrm{GrandIrreversibility} :
  \forall\,C:\mathrm{CoupledConfig},\;
  \mathrm{Terminates}(C) \wedge
  \mathrm{ReachesQuine}(C) \wedge
  \mathrm{FiniteTime}(C).
\]

In Lean, this is implemented as a \emph{total recursive function}:
it evaluates every possible RSVP--Polyxan evolution,
applies duality and modal reasoning at each step,
and terminates with a certified proof object.

\bigskip

\begin{quote}
Meaning is now not only geometric, algebraic, and cosmological.  
Meaning is formalised as a terminating computation.  
The universe of semantic physics is provable.
\end{quote}

