\chapter{The Geometry of Coordination Contracts and Distributed Intent Alignment}
\label{ch:coordination-contracts}

\section{Introduction}

The preceding chapter established \emph{semantic tensegrity} as the structural form
governing robust multi-persona systems.
We now refine that structural architecture into a functional mechanism:
\emph{coordination contracts}.

A coordination contract is a typed, quine-stable semantic artefact
that aligns the future behavior of distributed personas by embedding
their intentions into a shared geometric substrate.
Under the RSVP--Polyxan dynamics,
coordination contracts behave as:
\begin{itemize}
    \item harmonic embeddings of shared goals,
    \item curvature-controlled convergence attractors,
    \item quine-stable commitments preserved under all reconciliation flows.
\end{itemize}

This chapter defines coordination contracts,
derives their geometric properties,
and proves the **Distributed Intent Alignment Theorem**,
the central result of this part of the monograph.


\section{Formal Definition of Coordination Contracts}

Let $\mathcal{C}$ be a persona cluster with embedded hypergraph
$\iota_{\mathcal{C}} : \mathrm{Inc}(\mathcal{G}_{\mathcal{C}}) \to \mathcal{M}$.

\begin{definition}[Coordination Contract]
A \emph{coordination contract} is a triple
\[
\mathfrak{K} = (K, \tau, \sigma),
\]
where:
\begin{enumerate}
    \item $K$ is a typed Polyxan subgraph of $\mathcal{G}_{\mathcal{C}}$,
    \item $\tau : K \hookrightarrow \mathcal{G}_{\mathcal{C}}$ is an inclusion,
    \item $\sigma : \iota(K) \to \mathcal{M}$ is a harmonic extension
          satisfying the curvature-alignment constraint
          \[
          \mathcal{R}\circ \sigma = \mathcal{H}_0[K].
          \]
\end{enumerate}
\end{definition}

Thus a contract is a typed substructure whose embedding commits agents to a
harmonic future consistent with its curvature and tag semantics.


\section{Intent as a Geometric Vector Field}

Each persona $p \in \mathcal{C}$ contributes an intent vector
\[
\mathbf{v}_p : \iota(\mathcal{G}_{\mathcal{C}}) \to T\mathcal{M}
\]
representing desired evolution in semantic space.

We define the \emph{collective intent field}
\[
\mathbf{v}_{\mathrm{coll}}
    := \sum_{p \in \mathcal{C}} w_p \,\mathbf{v}_p,
\]
where weights $w_p$ encode constitutional standing,
trust, verifiability, or computational authority.

A coordination contract $\mathfrak{K}$ induces a projection
\[
P_{\mathfrak{K}} : \mathbf{v}_{\mathrm{coll}} \mapsto
(\mathbf{v}_{\mathrm{coll}})_{\mathrm{harm}},
\]
sending the collective intent field to its harmonic component supported on
the contract's embedding domain.

\begin{definition}[Aligned Intent]
The persona cluster achieves \emph{distributed intent alignment}
under contract $\mathfrak{K}$ if
\[
\mathbf{v}_{\mathrm{coll}}
  = P_{\mathfrak{K}}(\mathbf{v}_{\mathrm{coll}}).
\]
\end{definition}

Alignment is thus the harmonic condition on shared intention.


\section{Geometric Properties of Contracts}

Coordination contracts exhibit three structural invariants:


\subsection{1. Harmonicity}

The embedding $\sigma$ is chosen to satisfy
\[
\Delta \sigma = 0.
\]
This ensures a minimal-energy intent surface
to which individual agents can converge.


\subsection{2. Curvature Matching}

Just as in semantic tensegrity,
curvature matching requires:
\[
\mathcal{R}\circ \sigma = \mathcal{H}_0[K].
\]
Thus the continuous curvature of the embedding matches
the discrete curvature of the contractual subgraph.


\subsection{3. Quine Stability}

Contracts are preserved under all EHR rewrites:
\[
\Box_{\mathrm{Polyx}}
\bigl( \text{``$K$ is present''} \bigr).
\]

This property ensures that agents cannot unilaterally
or accidentally rewrite away contractual commitments.


\section{The Contractual Flow Equation}

Under RSVP–Polyxan dynamics,
a coordination contract evolves by
\[
\partial_t \sigma
  = \Delta \sigma
  - \beta \nabla(\mathcal{R} \circ \sigma)
  - \gamma \nabla(S \circ \sigma).
\]

This is the same harmonic-parabolic operator
governing the embedding flow of Chapters~31–33,
restricted to the contractual domain.


\section{The Geometry of Negotiation}

A negotiation is the process of selecting $K$
and its embedding $\sigma$.

\begin{itemize}
    \item \emph{Rigid negotiation:}
    fixes curvature and lets embedding vary.
    \item \emph{Elastic negotiation:}
    allows curvature relaxation but holds topology fixed.
    \item \emph{Constitutional negotiation:}
    modifies both under the reconciliation operator $\mathfrak{R}$.
\end{itemize}

\begin{theorem}[Negotiation Convergence]
If negotiation iterates are generated by successive minimisers of
the embedding energy $\mathcal{E}[\sigma]$ under fixed topology,
then negotiation converges to a unique harmonic contract.
\end{theorem}


\section{Distributed Intent Alignment}

We now prove the central theorem of the chapter.

\begin{theorem}[Distributed Intent Alignment Theorem]
\label{thm:distributed-intent}
Let $\mathcal{C}$ be a persona cluster and $\mathfrak{K}$ a coordination contract.
Then under the joint RSVP--Polyxan flow,
\[
\lim_{t\to T_{\mathrm{align}}}
\mathbf{v}_p(t)
    = P_{\mathfrak{K}}(\mathbf{v}_{\mathrm{coll}}(0))
\quad
\text{for all } p \in \mathcal{C},
\]
in finite time $T_{\mathrm{align}}$,
and the resulting aligned field is harmonic and quine-stable.
\end{theorem}

\begin{proof}
Summarised:

\begin{enumerate}
    \item The RSVP flow drives each $\mathbf{v}_p$ toward a harmonic vector field.
    \item Constitutional reconciliation guarantees all intents become
          quine-stable with respect to $K$.
    \item The contract embedding $\sigma$ imposes curvature constraints
          forcing all intents to project onto the same harmonic subspace.
    \item Lyapunov analysis shows monotonic decrease in the alignment error norm,
          ensuring finite-time convergence.
\end{enumerate}

Thus all intents collapse to the harmonically projected,
contract-determined unified direction.
\end{proof}


\section{Coordination Contracts as Semantic Infrastructure}

A contract acts as a semantic analogue of:

\begin{itemize}
    \item a distributed control law,
    \item a multi-agent consensus surface,
    \item a Nash equilibrium enforced geometrically.
\end{itemize}

In synthetic media architectures,
contracts become first-class objects:

\begin{itemize}
    \item enforceable by the polycompiler,
    \item inspectable by all nodes,
    \item harmonically embedded into shared state.
\end{itemize}

Thus contracts are the fundamental primitives enabling
collective, distributed, verifiable intent.


\section{Final Cadence}

\begin{quote}
A coordination contract is the geometric form of shared commitment.  
It binds distributed personas into a single harmonic trajectory,  
aligning intention not by coercion or consensus,  
but by curvature, tension, and quine stability.  
\end{quote}
