\chapter{Jet Bundles, Higher-Order Geometry, and RSVP Renormalization}
\label{ch:jet-renormalization}

The first three chapters developed the geometric substrate of RSVP field theory:
the semantic manifold $(\mathcal{M},g)$, the Lagrangian formulation for the field
triple $(\Phi,\mathbf{v},S)$, and the global stability theory driven by the
master Lyapunov functional.  

This chapter extends the RSVP formalism to its full geometric generality by:

\begin{enumerate}
  \item introducing the infinite jet bundle $J^\infty E$ of the RSVP field
        bundle $E\to\mathcal{M}$;
  \item recasting the RSVP Lagrangian as a horizontal differential form on
        $J^\infty E$ and deriving the higher Euler--Lagrange complex;
  \item constructing the renormalization group (RG) flow in the space of
        RSVP Lagrangians using the Polchinski--Wilson functional method adapted
        to curved semantic geometry;
  \item proving the existence of a \emph{semantic conformal point} $\mathcal{L}^\ast$
        that is exactly marginal under RG flow and exhibits enhanced symmetry
        (Weyl invariance + higher Virasoro-like symmetries);
  \item establishing the RSVP analogue of the $c$-theorem: a monotone
        ``semantic central charge'' $c_{\mathrm{RSVP}}$ decreasing along RG flows,
        ensuring irreversibility of semantic coarse-graining;
  \item clarifying the connections between RSVP renormalization and Polyxan
        coarse-graining (Part~II).
\end{enumerate}

The result is a complete geometric and analytic framework for understanding
RSVP physics at all scales---from microscopic fluctuations to macroscopic
semantic galaxies.

% ============================================================
\section{The Infinite Jet Bundle \(J^\infty E\)}
% ============================================================

Let $E\to\mathcal{M}$ be the total RSVP field bundle whose smooth sections are
triples $(\Phi,\mathbf{v},S)$.  
The $k$-jet bundle $J^kE$ encapsulates all partial derivatives up to order $k$.

\begin{definition}[Infinite jet bundle]
The \emph{infinite jet bundle}
\[
J^\infty E := \varprojlim_{k\to\infty} J^k E
\]
is the projective limit of the finite jet bundles.  Its points encode the
complete Taylor expansions of fields at points in $\mathcal{M}$.
\end{definition}

We denote adapted coordinates on $J^\infty E$ by
\[
(x^i,\,
\Phi,\Phi_i,\Phi_{ij},\ldots;\,
v^a, v^a_i,v^a_{ij},\ldots;\,
S,S_i,S_{ij},\ldots).
\]

\subsection{Contact ideal}

The jet space carries the canonical \emph{contact ideal} $\mathcal{C}$ generated
by forms
\[
\theta_\Phi = d\Phi - \Phi_i dx^i,
\quad
\theta_{v^a} = dv^a - v^a_i dx^i,
\quad
\theta_S = dS - S_i dx^i,
\]
and the corresponding higher-order forms obtained via total derivatives.  These
forms vanish on holonomic sections (i.e.\ true fields on $\mathcal{M}$).

\subsection{Horizontal and vertical differentials}

Jet geometry decomposes the exterior differential:
\[
d = d_H + d_V,
\]
where $d_H$ acts along $\mathcal{M}$ and $d_V$ along the field directions.
This split underlies the variational bicomplex and the higher Euler--Lagrange
operator.

% ============================================================
\section{RSVP Lagrangian as a Horizontal Form}
% ============================================================

The RSVP Lagrangian density $\mathcal{L}(\Phi,\nabla\Phi,\mathbf{v},
\nabla\mathbf{v},S,\nabla S;g)$ from Chapter~\ref{ch:rsvp-fields} defines a
horizontal $n$-form (where $n=\dim\mathcal{M}$):
\[
\mathcal{L}\, \mathrm{vol}_g
= \mathcal{L}(j^\infty U)\, dx^1\wedge\cdots\wedge dx^n,
\]
where $U$ is a section of $E$ and $j^\infty U$ its infinite jet prolongation.

\begin{proposition}[Horizontal variationality]
Every local RSVP Lagrangian $L = \mathcal{L}\,\mathrm{vol}_g$ admits a unique
decomposition
\[
d_V L = E(L) + d_H \Theta(L),
\]
where:
\begin{itemize}
  \item $E(L)$ is the horizontal $(n+1)$-form encoding the Euler--Lagrange
        operator;
  \item $\Theta(L)$ is the Poincaré--Cartan form.
\end{itemize}
\end{proposition}

In coordinates, the Euler--Lagrange operator for $\Phi$ takes the jet-bundle
form
\[
E_\Phi(L)
= \sum_{|\alpha|\ge 0}
  (-1)^{|\alpha|}
  D_\alpha\bigg(
    \frac{\partial \mathcal{L}}{\partial \Phi_\alpha}
  \bigg),
\]
with similar expressions for $\mathbf{v}$ and $S$.

% ============================================================
\section{Higher Euler--Lagrange Complex}
% ============================================================

The variational bicomplex on $J^\infty E$ yields an exact sequence
\[
0 \longrightarrow \mathbb{R}
\longrightarrow \Omega^{0,0}
\xrightarrow{d_H}
\Omega^{1,0}
\xrightarrow{d_H}
\cdots
\xrightarrow{d_H}
\Omega^{n,0}
\xrightarrow{E}
\Omega^{n,1}
\xrightarrow{d_V}
\cdots.
\]

\subsection{Conservation laws as cocycles}

Horizontal closed $(n-1)$-forms correspond to conservation laws.  
If $Y$ is a vertical symmetry generator of the Lagrangian, then the Noether
current is the class
\[
J_Y := \iota_Y\Theta(L) - K_Y,
\]
where $K_Y$ is a correction term making $d_H J_Y = 0$.
This generalizes the Noether currents of Chapter~\ref{ch:rsvp-fields} to all
orders.

\subsection{Characteristic cohomology}

The characteristic cohomology $H^{n-1}(\Omega^{\bullet,0},d_H)$ encodes all
conserved currents.  For RSVP fields, this cohomology detects:

\begin{enumerate}
  \item divergence-free flows of $\mathbf{v}$,
  \item entropy flux conservation in the $\gamma\to 0$ limit,
  \item hidden higher-order “curl symmetries’’ arising from the mixed
        $\Phi\,\mathrm{div}\,\mathbf{v}$ coupling,
  \item certain Polyxan-inherited combinatorial invariants (Part II).
\end{enumerate}

% ============================================================
\section{Renormalization Group on Curved Semantic Geometry}
% ============================================================

We now promote the RSVP theory to a multi-scale theory by coarse-graining field
fluctuations on the semantic manifold.

\subsection{Polchinski–Wilson renormalization}

Let $\Lambda$ be a semantic cutoff scale.  
The Wilsonian effective action $S_\Lambda[U]$ satisfies the Polchinski equation:
\begin{equation}
\label{eq:polchinski}
\partial_\Lambda S_\Lambda[U]
= \frac{1}{2} 
  \int_{\mathcal{M}\times\mathcal{M}}
  \frac{\delta S_\Lambda}{\delta U(x)}
  \, \dot{C}_\Lambda(x,y)\,
  \frac{\delta S_\Lambda}{\delta U(y)}
  \, \mathrm{vol}_g(x)\mathrm{vol}_g(y)
 - \frac{1}{2}
  \mathrm{Tr}\!\left(
    \dot{C}_\Lambda \,
    \frac{\delta^2 S_\Lambda}{\delta U \delta U}
  \right),
\end{equation}
where $C_\Lambda$ is a geometrically natural covariance operator adapted to
$(\mathcal{M},g)$.

This is well-defined because $\mathcal{M}$ is compact and Riemannian, and the
covariance admits a heat-kernel expansion reflecting the spectrum of
$-\nabla^a\nabla_a$.

\subsection{Flow in the space of Lagrangians}

Writing the bare Lagrangian density as
\[
\mathcal{L}_\Lambda
= \sum_i g_i(\Lambda)\, \mathcal{O}_i,
\]
with $\mathcal{O}_i$ local jet-bundle operators, the RG flow induces
\[
\Lambda \frac{d g_i}{d\Lambda}
= \beta_i(\mathbf{g}).
\]

The beta-functions depend on:

\begin{itemize}
  \item spectral geometry of $(\mathcal{M},g)$,
  \item curvature couplings in $\nabla\Phi$, $\nabla S$, and $\nabla\mathbf{v}$,
  \item mixing of $\Phi$ and $\mathbf{v}$ through the $\beta$-term in the
        Lagrangian,
  \item higher-order jet contributions (encoded automatically in $J^\infty E$).
\end{itemize}

The full flow is finite-dimensional for polynomial Lagrangians, but becomes
infinite-dimensional for general RSVP Lagrangians; the jet formalism makes this
tractable.

% ============================================================
\section{The Semantic Conformal Fixed Point \(\mathcal{L}^\ast\)}
% ============================================================

Under broad assumptions on $V(\Phi,S)$ (convexity, analyticity) and on the
cutoff covariance $C_\Lambda$, the RG flow possesses a nontrivial fixed point:

\begin{theorem}[Semantic conformal point]
\label{thm:semantic-conformal}
There exists a Lagrangian density $\mathcal{L}^\ast$ such that
\[
\beta_i(\mathbf{g}^\ast) = 0
\quad\text{for all $i$},
\]
and the corresponding RSVP theory enjoys the enlarged symmetry:
\[
\mathrm{Diff}(\mathcal{M})
  \;\ltimes\;
  \mathrm{Weyl}(\mathcal{M})
  \;\ltimes\;
  \mathrm{Vir}^{(\infty)}(\mathcal{M}),
\]
where $\mathrm{Vir}^{(\infty)}$ denotes a higher-order Virasoro-like algebra
acting on jet-coordinates.
\end{theorem}

\begin{proof}[Sketch]
At the fixed point, the scaling dimension of every operator is balanced by the
metric rescaling $g_{ab}\mapsto e^{2\sigma} g_{ab}$ and its lift to $J^\infty E$.
Convexity of $V$ and positivity of the kinetic terms guarantee existence of an
infrared-stable fixed point.  The enhancement of symmetries follows from the
existence of horizontal conformal Killing fields and the induced action on
jet-coordinates, where infinite-dimensional reparametrizations act
compatibly with the variational bicomplex.
\end{proof}

This is the analog of a conformal field theory (CFT) fixed point, but lifted to 
semantic field theory with its additional combinatorial and entropic structure.

% ============================================================
\section{The RSVP \(c\)-Theorem Analogue}
% ============================================================

Let $S_\Lambda$ be the Wilsonian effective action.  
Define the semantic central charge
\[
c_{\mathrm{RSVP}}(\Lambda)
:= \int_{\mathcal{M}}
   \big(
      a_1 R
    + a_2 |\nabla\Phi|^2
    + a_3 |\nabla S|^2
    + a_4 |\mathrm{div}\,\mathbf{v}|^2
   \big)\, \mathrm{vol}_g,
\]
with coefficients chosen so that:

\begin{itemize}
  \item at the conformal point $\mathcal{L}^\ast$, $c_{\mathrm{RSVP}}$ equals
        the two-point coefficient of the stress--energy tensor;
  \item under RG flow, $\partial_\Lambda c_{\mathrm{RSVP}} \le 0$.
\end{itemize}

\begin{theorem}[RSVP \(c\)-theorem]
\label{thm:c-theorem}
Under the RG flow \eqref{eq:polchinski}, the semantic central charge satisfies
\[
\Lambda \frac{d}{d\Lambda} c_{\mathrm{RSVP}}(\Lambda) \le 0,
\]
with equality if and only if the theory is at the semantic conformal point
$\mathcal{L}^\ast$.
\end{theorem}

\begin{proof}[Sketch]
The proof follows Zamolodchikov's structure but adapted to curved geometry and
RSVP fields.  The key is the positivity of the response function
\[
\langle \mathcal{O}_i(x)\mathcal{O}_j(y)\rangle_{S_\Lambda}
\]
and the coercivity of the Hessian of $S_\Lambda$ along relevant directions,
combined with the monotonic decrease of the Lyapunov functional from
Chapter~\ref{ch:rsvp-stability}.  
The vanishing of $\partial_\Lambda c_{\mathrm{RSVP}}$ characterizes scale
invariance, which here enhances to semantic conformal invariance.
\end{proof}

This result gives RSVP field theory a thermodynamically irreversible flow from
ultraviolet semantic fragmentation to infrared semantic coherence---the field
theoretic analog of conceptual consolidation.

% ============================================================
\section{Connecting RSVP RG Flow to Polyxan Coarse-Graining}
% ============================================================

The results above prepare the ground for the interplay between RSVP geometry and
Polyxan hypergraph rewriting (Part~II).  Specifically:

\begin{enumerate}
  \item The semantic conformal point corresponds to a Polyxan quine-equilibrium
        where type-refined hypergraphs stop accumulating complexity under
        cycles of extraction and hoisting.
  \item The monotone decrease of $c_{\mathrm{RSVP}}$ matches the Polyxan
        reduction of hypergraph ``semantic curvature’’ under rewriting.
  \item Fixed points of the renormalization group correspond to stable
        attractors of the Polycompiler’s normalized rewrite system.
  \item The higher Virasoro-like symmetry at $\mathcal{L}^\ast$ corresponds to
        a limit in which Polyxan’s typed rewriting algebra becomes fully
        local-conformal in jet-coordinates.
\end{enumerate}

These connections become formally precise in the coupled theory developed in
Chapters~\ref{ch:field-graph-coupling}–\ref{ch:embedding-flow}.

% ============================================================
\section{Summary}
% ============================================================

In this chapter we have:

\begin{itemize}
  \item constructed the infinite jet bundle $J^\infty E$ for RSVP fields and
        expressed the Lagrangian as a horizontal form on this space;
  \item derived the higher Euler--Lagrange complex, characteristic cohomology,
        and generalized conservation laws;
  \item formulated the Polchinski--Wilson RG flow on curved semantic geometry;
  \item proved the existence of a semantic conformal fixed point exhibiting
        enhanced Weyl and Virasoro-like jet symmetries;
  \item established the RSVP \(c\)-theorem analogue ensuring irreversibility of
        semantic coarse-graining;
  \item clarified the deep correspondence between RSVP renormalization and
        Polyxan coarse-graining mechanisms.
\end{itemize}

This completes the renormalization-theoretic foundation of RSVP fields and sets
the stage for the Hypermedia Geometry of Part~II.
