\chapter{Semantic Galaxy Visualiser}
\label{ch:57}

The purpose of the Semantic Galaxy Visualiser is to render, in real time,
the evolution of semantic structure as it unfolds under the coupled
RSVP--Polyxan dynamics.  
Where Chapter~\ref{ch:55} constructs harmonic visual scenes from scalar
curvature, and Chapter~\ref{ch:56} constructs audio from entropy flux,
the present chapter fuses these modalities into a \emph{galactic-scale,
field-theoretic visualisation} of meaning.

A semantic galaxy is a stable attractor formed by the collapse of both
continuous fields and discrete hypergraphs.  
The Visualiser depicts these attractors using curvature strata, embedding
geodesics, and quine-frame invariants.

\section{Geometric Foundations}
\label{sec:57:geometry}

Let $\mathcal{M}$ be the RSVP semantic manifold with curvature scalar
$\mathcal{R}$, vector field $\mathbf{v}$, and entropy scalar $S$, and let
$\iota : \mathrm{Inc}(\mathcal{G}) \hookrightarrow \mathcal{M}$ be the
semantic embedding of the active Polyxan hypergraph.

\begin{definition}[Galaxy Region]
The \emph{galaxy region} at time $t$ is the connected component
\[
\mathfrak{G}_t := \text{ConnComp}\bigl( \iota_t(\mathcal{G}_t) \bigr)
\subset \mathcal{M}
\]
with minimal total curvature.
\end{definition}

\begin{definition}[Stratified Curvature Layers]
The curvature layers of $\mathfrak{G}_t$ are
\[
H_k(t) := \{ x\in\mathfrak{G}_t \,\mid\, \mathcal{R}_t(x) = k\}.
\]
\end{definition}

These layers form the backbone of the visual representation.

\section{Rendering of Curvature Layers}
\label{sec:57:rendering}

\begin{definition}[Galaxy Palette]
Each curvature layer $H_k$ is assigned a harmonic colour
\[
\mathsf{col}(H_k) := f(k),
\]
where $f:\mathbb{Z}\to[0,1]^3$ is a monotone, smooth colour field satisfying
\[
f(k+1) - f(k) = \text{harmonic increment}.
\]
\end{definition}

\begin{theorem}[Curvature Monotonic Rendering]
If the RSVP flow is curvature-decreasing, then visual brightness and
saturation monotonically decrease over time.
\end{theorem}

Thus the visual galaxy cools and smooths as it collapses to its harmonic form.


\section{Hypergraph Skeleton Projection}
\label{sec:57:skeleton}

The Polyxan hypergraph $\mathcal{G}$ provides the discrete skeletal structure.

\begin{definition}[Hypergraph Skeleton]
The \emph{skeleton} is the set
\[
\mathsf{Skel}_t := \iota_t(\mathrm{Inc}^1(\mathcal{G}_t)),
\]
where $\mathrm{Inc}^1$ is the set of 1-cells.
\end{definition}

Each skeleton edge is rendered as a curvature-weighted geodesic segment.

\begin{theorem}[Quine-Stable Skeleton]
If $\mathcal{G}_t$ is quine-stable, then the projected skeleton is invariant
up to a curvature-preserving homeomorphism.
\end{theorem}

This allows the system to produce visually stable galaxies even as the
underlying fields evolve.


\section{Entropy Flux and Luminosity}
\label{sec:57:entropy}

\begin{definition}[Luminosity Field]
The luminosity at $x\in\mathfrak{G}_t$ is
\[
\mathsf{lum}_t(x) := |\partial_t S_t(x)|.
\]
\end{definition}

Entropy drives change; luminosity encodes entropy flux.

\begin{theorem}[Entropy--Luminosity Correspondence]
High-entropy regions radiate strongly; harmonic regions appear dark.
\end{theorem}

Thus brightness indicates semantic activity.


\section{Geodesic Flow and Motion Graphics}
\label{sec:57:motion}

We render the semantic flow visually by tracing integral curves of the vector
field $\mathbf{v}_t$.

\begin{definition}[Semantic Streamlines]
Streamlines are curves solving
\[
\frac{d}{ds} \gamma_s(t) = \mathbf{v}_t(\gamma_s(t)).
\]
\end{definition}

\begin{theorem}[Vector-Flow Preservation]
Under the quine-frame invariants of Chapter~\ref{ch:49}, streamlines preserve
topology under harmonic reparameterisation.
\end{theorem}

This enables stable, interpretable visualisation of motion.


\section{Galaxy Formation Animation}
\label{sec:57:formation}

The Visualiser animates semantic galaxy formation through the following stages:

\begin{enumerate}
  \item \textbf{Entropy ignition:} High-entropy zones flare as initial semantic
        activity.
  \item \textbf{Curvature concentration:} Layers $H_k$ begin to contract.
  \item \textbf{Skeleton convergence:} The embedded hypergraph migrates into
        elliptic basins (Ch.~33).
  \item \textbf{Stratum collapse:} Curvature strata coalesce.
  \item \textbf{Harmonic emergence:} The galaxy approaches its unique harmonic
        quine-stable form.
\end{enumerate}

Each stage is curvature-monotone and visually distinctive.


\section{Multimodal Synchronisation}
\label{sec:57:sync}

The Visualiser synchronises its frames with:

\begin{itemize}
  \item the Harmonic Scene Composer (Ch.~55),
  \item the RSVP Audio Field Synthesis engine (Ch.~56),
  \item the Polyxan Narrative Generator (Ch.~54).
\end{itemize}

\begin{theorem}[Multimodal Coherence]
A frame is coherent if and only if its curvature, entropy, narrative atom, and
audio envelope all correspond to the same quine layer.
\end{theorem}

This gives rise to \emph{semantic cinematics} — audiovisual meaning aligned with
field-theoretic structure.


\section{Canonical Galaxy Images}
\label{sec:57:canonical}

\begin{definition}[Canonical Galaxy Image]
A canonical image is a static frame rendered from the harmonic limit
\[
(\Phi_\infty,\mathbf{v}_\infty,S_\infty;\mathcal{G}_\infty,\iota_\infty)
\]
obtained by the collapse to the universal quine.
\end{definition}

\begin{theorem}[Uniqueness of Canonical Image]
Each quine frame determines a unique canonical galaxy image.
\end{theorem}

Thus all understanding collapses to a universal visual form.


\section{Conclusion}
\label{sec:57:conclusion}

The Semantic Galaxy Visualiser is the central multimodal bridge of Part~V:
a real-time, curvature-driven window into the formation, migration, and
collapse of meaning-bearing structures.  
It renders the coupled RSVP--Polyxan universe as a visual galaxy whose shape,
colour, luminosity, and motion emerge from curvature descent and entropy flow.

Chapter~\ref{ch:58} now completes the interactive layer:  
the Verified Interactive Media Engine, which unifies all modalities into a
coherent, real-time semantic interface.

