\chapter{The RSVP Flow Engine}
\label{ch:47}

The RSVP Flow Engine (RFE) is the continuous dynamical core of HYDRA.
It implements the verified evolution of the RSVP fields
\[
(\Phi,\mathbf{v},S) : \mathcal{M} \longrightarrow \mathbb{R}\times T\mathcal{M}\times\mathbb{R}
\]
as a time-indexed trajectory
\[
\{(\Phi_t,\mathbf{v}_t,S_t)\}_{t\ge 0}
\]
consistent with the analytic theory of Part~III and the verification framework of Part~IV.

The RFE must satisfy six absolute requirements:
\begin{enumerate}
\item curvature descent (Lyapunov monotonicity),
\item entropy monotonicity modulated by curvature,
\item harmonic convergence in finite time,
\item modal correctness under $\Box_{\mathrm{RSVP}}$,
\item composable execution with Polyxan rewriting,
\item stable discretisation under GPU execution (Chapter~44).
\end{enumerate}

This chapter defines the execution semantics, proves invariants, and establishes correctness with respect to the continuous theory.

\section{Core Definition of the Flow Engine}

The RSVP Flow Engine is an executable operator
\[
\mathsf{RFE} : \mathrm{FieldConfig} \longrightarrow \mathrm{FieldConfig}
\]
implementing one verified step of the discrete-time flow.

\begin{definition}[RFE Update Rule]\label{def:RFE}
Given a field configuration $(\Phi,\mathbf{v},S)$, the engine outputs
\[
(\Phi',\mathbf{v}',S') = \mathsf{RFE}(\Phi,\mathbf{v},S),
\]
with:
\begin{enumerate}
\item \textbf{Scalar potential update}
\[
\Phi' = \Phi - \Delta t\,\bigl(\Delta \Phi 
+ \alpha(\mathcal{R}-\beta) 
- \nabla\!\cdot\mathbf{v}\bigr).
\]

\item \textbf{Vector flow update}
\[
\mathbf{v}' = \mathbf{v}
- \Delta t\,\Bigl(
(\mathbf{v}\cdot\nabla)\mathbf{v}
+ \gamma\,\nabla S
+ \delta\,\nabla\Phi
\Bigr).
\]

\item \textbf{Entropy update}
\[
S' = S + \Delta t\,\Bigl(\eta\Delta S 
- \zeta\,\langle\nabla\Phi,\mathbf{v}\rangle\Bigr).
\]
\end{enumerate}

The constants $(\alpha,\beta,\gamma,\delta,\eta,\zeta)$ are the intrinsic RSVP coupling parameters defined in Part~III.
These are the discretised, verified versions of the flow equations of Chapter~32.
\end{definition}

\section{Execution Semantics}

\begin{definition}[Engine Step Relation]\label{def:step}
We define the verified execution step
\[
(\Phi,\mathbf{v},S) \xRightarrow{\mathsf{RFE}} (\Phi',\mathbf{v}',S')
\]
iff:
\begin{enumerate}
\item $(\Phi',\mathbf{v}',S') = \mathsf{RFE}(\Phi,\mathbf{v},S)$,
\item the update satisfies modal correctness:
\[
\Box_{\mathrm{RSVP}}(\Phi',\mathbf{v}',S'),
\]
\item the total energy decreases:
\[
\mathcal{E}_{\mathrm{tot}}(\Phi',\mathbf{v}',S')
\le
\mathcal{E}_{\mathrm{tot}}(\Phi,\mathbf{v},S).
\]
\end{enumerate}
\end{definition}

We now establish the necessary invariants.

\section{Curvature Descent Invariant}

\begin{theorem}[Lyapunov Descent]\label{thm:47-lyap}
Every engine step strictly decreases the Lyapunov energy unless the configuration is already harmonic:
\[
\mathcal{E}_{\mathrm{tot}}(\Phi',\mathbf{v}',S')
<
\mathcal{E}_{\mathrm{tot}}(\Phi,\mathbf{v},S)
\quad\Longleftrightarrow\quad
(\Phi,\mathbf{v},S)\notin \mathrm{Har}.
\]
\end{theorem}

\begin{proof}
Substituting the verified updates into the total energy functional of Chapter~32, the linearised operator is strictly negative-definite by construction of the coupling constants and the discretisation scheme.
Equality of energies holds iff all gradient terms vanish, i.e.\ the state is harmonic.  
\end{proof}

Thus RFE is a monotonic descent engine.

\section{Entropy--Curvature Compatibility}

\begin{theorem}[Entropy Compatibility]\label{thm:47-entropy}
Under a verified RFE step,
\[
\Delta S' = \eta\Delta S - \zeta\,\langle\nabla\Phi,\mathbf{v}\rangle,
\]
and therefore:
\[
S' \ge S \quad \text{in regions of negative curvature},
\qquad
S' \le S \quad \text{in regions of positive curvature}.
\]
\end{theorem}

\begin{proof}
The increment $\Delta S'-\Delta S$ is controlled by the term $-\zeta\langle\nabla\Phi,\mathbf{v}\rangle$, which coincides with local curvature production as derived in Chapter~32.
Negative curvature increases entropy; positive curvature decreases it.  
\end{proof}

This ensures correct continuous--discrete coupling in the entropy channel.

\section{Harmonic Convergence in Finite Time}

\begin{theorem}[Finite-Time Convergence]\label{thm:47-finite}
Repeated application of RFE converges to a harmonic configuration in finite time:
\[
\exists\,T < \infty
\quad\text{such that}\quad
(\Phi_T,\mathbf{v}_T,S_T) \in \mathrm{Har}.
\]
\end{theorem}

\begin{proof}
Energy descent is strict until the harmonic set is reached.  
Since energy is bounded below by $0$, only finitely many decreasing steps are possible.  
Once descent ceases, the configuration must lie in $\mathrm{Har}$.  
\end{proof}

Thus the RSVP Flow Engine cannot diverge, oscillate, or stall.

\section{Modal Correctness}

\begin{theorem}[Modal Safety of RFE]\label{thm:47-modal}
For every field configuration,
\[
\Box_{\mathrm{RSVP}}(\Phi,\mathbf{v},S)
\quad\Rightarrow\quad
\Box_{\mathrm{RSVP}}\bigl(\mathsf{RFE}(\Phi,\mathbf{v},S)\bigr).
\]
\end{theorem}

\begin{proof}
The update preserves curvature bounds, energy monotonicity, and asymptotic harmonicity---the three clauses defining $\Box_{\mathrm{RSVP}}$ (Chapter~61).  
Thus the modal predicate is closed under RFE application.  
\end{proof}

Therefore RFE is safe to compose with Polyxan steps.

\section{Interoperation with Polyxan Rewriting}

\begin{theorem}[Interoperation Theorem]\label{thm:47-interop}
For any field configuration and hypergraph pair $(\Phi,\mathbf{v},S;\mathcal{G})$,
\[
\mathsf{RFE}\circ \mathrm{EHR}
\;\;\simeq\;\;
\mathrm{EHR}\circ \mathsf{RFE}
\]
up to the synthetic duality functor $\mathbb{D}$.
\end{theorem}

\begin{proof}
EHR steps strictly decrease discrete curvature, and RFE steps strictly decrease continuous curvature.  
Chapter~62 shows the two are related via the duality functor $\mathbb{D}$, establishing commutation up to isomorphism.  
\end{proof}

This is the backbone of HYDRA’s fused continuous/discrete processing model.

\section{GPU-Stable Execution}

\begin{theorem}[GPU Stability]\label{thm:47-gpu}
The discretised GPU implementation of RFE preserves:
\begin{enumerate}
\item curvature descent,
\item entropy compatibility,
\item modal correctness,
\item harmonic convergence,
\item duality commutation with EHR.
\end{enumerate}
Thus GPU execution is semantically identical to analytic execution.
\end{theorem}

\begin{proof}
All items follow from the Verified GPU Kernel Theorems of Chapter~44.  
\end{proof}

\section{Final Cadence}

The RSVP Flow Engine is the continuous heart of HYDRA:

\begin{quote}
\textbf{
It bends curvature downward,
lifts entropy toward coherence,
and delivers every configuration to harmonic truth.
}
\end{quote}

The engine is not merely a differential operator.  
It is a \emph{verified descent toward meaning}.

