\chapter{Verified Sensorial Ingestion Layer}
\label{ch:73}

HYDRA’s distributed execution framework (Chapter~72) begins, at every node, 
with the \emph{Sensorial Ingestion Layer}.  
This layer provides the only legitimate entry point by which external signals, 
human actions, machine actions, or semantic deltas may enter the verified 
RSVP--Polyxan runtime.

Because ingestion constitutes the boundary between the unverified world and the 
verified semantic universe, its design is subject to the most stringent set of 
invariants in the entire architecture.  
This chapter presents its construction, semantics, correctness guarantees, and 
its role in the global curvature envelope.

\section{Purpose and Scope}

The Sensorial Ingestion Layer (SIL) is responsible for the following tasks:
\begin{enumerate}
\item converting raw external signals into geometric field-compatible deltas,
\item ensuring curvature and entropy consistency of incoming information,
\item routing deltas into the RSVP field engine and Polyxan rewriting kernel,
\item emitting modal certificates establishing that the ingestion step is safe,
\item coordinating with the HYDRA consensus layer to maintain global coherence.
\end{enumerate}

No unverified information may bypass the SIL.  
All signals must pass through its verification pipeline.

\section{Types of Input}

Every external signal must be classified into exactly one of the following 
modal categories:

\begin{enumerate}
\item \emph{Geometric sensory deltas} 
      (continuous-valued; destined for RSVP fields),
\item \emph{Symbolic structural deltas} 
      (discrete; destined for Polyxan rewriting),
\item \emph{Hybrid deltas} 
      (containing both continuous and discrete semantic components),
\item \emph{Domain boundary deltas} 
      (cross-domain reconciliation messages under curvature guards).
\end{enumerate}

\begin{definition}[Ingestion Packet]
An ingestion packet is a tuple
\[
\mathsf{InPkt}
  := (\Delta_{\mathrm{raw}}, \tau, \sigma),
\]
where:
\begin{itemize}
\item $\Delta_{\mathrm{raw}}$ is the unprocessed raw signal,
\item $\tau$ is its modal classification,
\item $\sigma$ is a provenance signature used for verification.
\end{itemize}
\end{definition}

\section{Preprocessing and Normalisation}

Incoming deltas must be transformed into canonical forms compatible with the 
RSVP manifold structure.

\subsection{Geometric Preprocessing}

\begin{definition}[Curvature-Compatible Normalisation]
A geometric delta $\Delta$ is curvature-compatible if
\[
\|\nabla \Delta\| \le \kappa_{\max}
\qquad\text{and}\qquad
\Delta \in H^1(\mathcal{M}),
\]
for the node’s curvature bound $\kappa_{\max}$.
\end{definition}

Normalisation enforces these bounds by harmonic projection:
\[
\Delta_{\mathrm{norm}}
  = P_{\mathrm{harm}}(\Delta_{\mathrm{raw}}).
\]

\subsection{Symbolic Preprocessing}

\begin{definition}[Rewrite-Compatible Form]
A symbolic delta is in rewrite-compatible form if it is a finite collection of 
Polyxan cells with attributed curvature tags matching the embedding signature.
\end{definition}

Symbolic deltas are normalised via tag refinement and curvature guard insertion.

\subsection{Hybrid Preprocessing}

Hybrid inputs are decomposed:
\[
\Delta_{\mathrm{raw}} 
  \mapsto (\Delta_{\mathrm{cont}}, \Delta_{\mathrm{disc}})
\]
and normalised separately before recombination.

\section{Modal Verification of Inputs}

Each ingestion step must satisfy the modal correctness predicates:
\[
\Box_{\mathrm{RSVP}}(\Delta)
\qquad\text{and}\qquad
\Box_{\mathrm{Polyx}}(\Delta).
\]

\begin{definition}[Modal Verification]
A delta $\Delta$ is modally verified if:
\[
\begin{aligned}
\text{(i)} &\quad \Delta \text{ preserves curvature bounds},\\
\text{(ii)} &\quad \Delta \text{ does not increase entropy beyond admissible limits},\\
\text{(iii)} &\quad \Delta \text{ respects quine-frame stability},\\
\text{(iv)} &\quad \Delta \text{ is compatible with the local patch algebra}.
\end{aligned}
\]
\end{definition}

\begin{theorem}[Verified Ingestion]
Every modally verified delta may be safely forwarded to the RSVP engine 
and Polyxan kernel without violating global invariants.
\end{theorem}

\begin{proof}
Curvature constraints maintain geometric stability.  
Entropy constraints maintain thermodynamic coherence.  
Quine-frame stability ensures self-reference does not escape the permitted 
strata.  
Patch algebra compatibility ensures seamless integration with the reconciliation 
layer.  
Thus the delta is safe.
\end{proof}

\section{Integration into the Execution Pipeline}

After verification, deltas are routed to the appropriate HYDRA subsystems:

\begin{enumerate}
\item Geometric deltas $\to$ \textbf{FlowStep},  
      where they alter $\Phi$, $\mathbf{v}$, $S$ through harmonic evolution.
\item Symbolic deltas $\to$ \textbf{RewriteStep},  
      where they alter $\mathcal{G}$ through curvature-guarded EHR rewriting.
\item Hybrid deltas $\to$ \textbf{EmbedStep} and \textbf{CoupleStep},  
      where continuous and discrete parts are synchronised.
\end{enumerate}

The ingestion layer therefore acts as the semantic stabiliser between the 
external world and the verified semantic dynamics.

\section{Cross-Domain Input and Isolation Guarantees}

When deltas originate from other semantic domains, the SIL evaluates:
\[
|\mathcal{R}_{\mathrm{local}} - \mathcal{R}_{\mathrm{remote}}|
\le \epsilon.
\]

\begin{theorem}[Boundary Safety]
Cross-domain deltas remain safe if and only if curvature mismatch 
remains within the admissible envelope.
\end{theorem}

\begin{proof}
Curvature mismatch serves as the invariant that ensures foreign deltas can be 
embedded within the local manifold without violating harmonic conditions.  
If mismatch exceeds $\epsilon$, embedding would require strata shifts that 
break modal predicates.
\end{proof}

This mechanism enables semi-isolated domains while maintaining 
the integrity of global semantic coherence.

\section{Certification and Emission}

Each ingestion step produces a modal certificate:
\[
\mathsf{Cert}(\Delta)
  := (\Box_{\mathrm{RSVP}}(\Delta),\;\Box_{\mathrm{Polyx}}(\Delta)).
\]

These certificates are appended to the node’s semantic spine and propagated 
through HYDRA’s verification network.

\begin{theorem}[Soundness of Certification]
For every ingested delta, certification is sound:
\[
\mathsf{Cert}(\Delta) = \top
\quad\Longrightarrow\quad
\Delta \text{ is safe}.
\]
\end{theorem}

\begin{proof}
Direct from Verified Ingestion and the construction of the modal predicates.
\end{proof}

\section{Final Cadence}

The Sensorial Ingestion Layer is the narrowest point in the architecture:  
the membrane separating raw reality from verified meaning.  
Every delta that enters the verified universe must first survive this passage.

\bigskip

\begin{quote}
\textbf{
To ingest is to translate the unverified into the verified.  
To verify is to bind sensation to meaning.  
Through ingestion, the universe becomes legible.  
}
\end{quote}

