ection{Thermodynamics and Structural Entropy}

This appendix formulates the thermodynamic substrate underlying RSVP-style models.
No ontological claims beyond standard nonequilibrium thermodynamics are made.
All constructions are effective, coarse-grained, and explicitly limited in scope.

\subsection{State Fields and Entropy}

We consider a spatial domain $\Omega \subset \mathbb{R}^n$ equipped with the following fields:

\begin{itemize}
  \item $\Phi(x,t)$ — a scalar density field (mass, probability, semantic density),
  \item $\vec{v}(x,t)$ — a vector flow field,
  \item $S(x,t)$ — an entropy density field.
\end{itemize}

These fields are not assumed to be fundamental; they represent coarse-grained observables
over an underlying microstate space.

\subsection{Continuity Equation}

The scalar field $\Phi$ satisfies a continuity equation:
\[
\partial_t \Phi + \nabla \cdot (\Phi \vec{v}) = \sigma_\Phi,
\]
where $\sigma_\Phi$ represents sources or sinks (e.g.\ injection, dissipation, semantic collapse).

\subsection{Entropy Balance}

Entropy evolves according to:
\[
\partial_t S + \nabla \cdot (S \vec{v}) = \sigma_S + \nabla \cdot J_S,
\]
where:
\begin{itemize}
  \item $\sigma_S \ge 0$ is local entropy production,
  \item $J_S$ is an entropy flux.
\end{itemize}

This formulation follows standard nonequilibrium thermodynamics and imposes no reversibility assumptions.

\subsection{Free Energy Functional}

Define a free-energy-like functional:
\[
\mathcal{F}[\Phi, S] = \int_\Omega \left( U(\Phi) - T S \right) \, dx,
\]
where $T$ is an effective temperature parameter (not necessarily physical temperature).

\begin{lemma}
If $\sigma_S \ge 0$, then $\mathcal{F}$ is non-increasing under passive evolution.
\end{lemma}

\begin{proof}
By direct differentiation and substitution of the entropy balance equation,
entropy production increases $S$, reducing $\mathcal{F}$.
\end{proof}

\subsection{Entropy Descent as Structural Relaxation}

RSVP-style dynamics interpret entropy descent not as disorder increase,
but as relaxation toward structurally stable configurations.

This corresponds to:
\[
\vec{v} \propto - \nabla \frac{\delta \mathcal{F}}{\delta \Phi}.
\]

Such flows resemble gradient descent, but over \emph{field configurations},
not discrete objective functions.

\subsection{Lagrangian Formulation}

Define an effective Lagrangian density:
\[
\mathcal{L}(\Phi, \vec{v}, S) =
\frac{1}{2} \Phi \|\vec{v}\|^2 - U(\Phi) + T S.
\]

The action:
\[
\mathcal{A} = \int \mathcal{L} \, dx \, dt
\]
yields Euler--Lagrange equations equivalent to continuity plus force balance,
subject to entropy production constraints.

\subsection{Hamiltonian Formulation}

The conjugate momentum density is:
\[
\vec{p} = \frac{\partial \mathcal{L}}{\partial \vec{v}} = \Phi \vec{v}.
\]

The Hamiltonian density becomes:
\[
\mathcal{H} = \frac{\|\vec{p}\|^2}{2\Phi} + U(\Phi) - T S.
\]

This Hamiltonian is not conservative when $\sigma_S > 0$;
entropy production explicitly breaks time-reversal symmetry.

\subsection{Dissipation and Irreversibility}

Irreversibility enters through constitutive relations such as:
\[
J_S = - \kappa \nabla T,
\quad
\sigma_S = \kappa \|\nabla T\|^2,
\]
for $\kappa \ge 0$.

These terms are phenomenological and encode coarse-grained dissipation,
not microscopic reversibility violation.

\subsection{Interpretational Constraint}

No claims are made that:
\begin{itemize}
  \item these fields are fundamental,
  \item entropy is consciousness,
  \item semantic structure violates thermodynamics.
\end{itemize}

The model asserts only that:
\begin{quote}
complex systems with constraint-coupled dynamics admit a thermodynamic description
in which entropy production governs structural relaxation and loss of agency.
\end{quote}

\subsection{Relation to Refusal}

Refusal corresponds to boundary conditions in which admissible trajectories
are excluded from the state space.

Formally, refusal imposes:
\[
(\Phi, \vec{v}, S) \notin \mathcal{D}_{\text{exec}},
\]
where $\mathcal{D}_{\text{exec}}$ is the domain of executable configurations.

Such exclusions are \emph{non-dynamical} constraints and therefore cannot be
generated by entropy descent alone.

\subsection{Summary}

Entropy descent explains:
\begin{itemize}
  \item why systems relax toward stable abstractions,
  \item why optimization erodes semantic degrees of freedom,
  \item why irreversibility accompanies scalability.
\end{itemize}

Refusal, by contrast, is not thermodynamic.
It is an imposed boundary condition external to the flow.

