% ============================================================
% CHAPTER 6
% ============================================================

\chapter{The TARTAN--RSVP Framework}
\label{chap:tartan-rsvp}

\section{Motivation}

Traditional distributed computation concerns consistency and
synchronization across spatially separated agents. Conflict--free
replicated data types (CRDTs) guarantee eventual consistency under
concurrent updates without requiring global coordination. In the
lamphrodynamic setting, distributed geometric computation is realized as
a field theory of interacting configurations. The guiding question of
this chapter is:

\begin{quote}
Can CRDT semantics be recovered from lamphrodynamic evolution?
\end{quote}

We will show that specific CRDTs can be modeled as \emph{field
configurations} whose update rules correspond to lamphrodynamically
admissible transformations. Eventual consistency emerges from entropy
dissipation and geometric collapse.


\section{CRDTs as Field Configurations}

Let $\mathsf{CRDT}$ denote a category of CRDTs, with morphisms given by
admissible updates. We construct a functor
\[
  \CRDT \colon \mathsf{CRDT} \longrightarrow \dConfig,
\]
assigning to each CRDT a derived configuration object. Updates are
mapped to lamphrodynamic transformations respecting entropy admissibility.

\begin{definition}
A \emph{lamphrodynamic CRDT} is a CRDT whose state space is realized as
a subset of $\dConfig$, and whose update rules correspond to merge or
collapse operations.
\end{definition}


\section{Eventual Consistency}

CRDTs guarantee eventual consistency: given sufficiently many updates
and message deliveries, all replicas converge to the same state. In
lamphrodynamic terms, convergence arises from entropy minimization under
repeated merge and collapse operations.

\begin{proposition}
Let $X,Y\in\dConfig$ represent two CRDT replicas. If updates are
entropy--admissible and lamphrodynamically bounded, then repeated merge
and collapse operations converge to a common configuration
$Z\in\dConfig$.
\end{proposition}

\begin{proof}[Sketch]
Each merge reduces discrepancy while preserving entropy admissibility.
Collapse eliminates internal inconsistency while preserving boundary
conditions. Lamphrodynamic energy estimates guarantee convergence to a
fixed point.
\end{proof}


\section{Gray-Code Rewriting}

TARTAN incorporates Gray--code rewriting to minimize semantic conflict
and maintain lamphrodynamic coherence across distributed computational
paths. Let $G$ denote a Gray--code rewrite operator acting on
configurations.

\begin{definition}
A \emph{lamphrodynamic Gray--code path} is a sequence of configurations
connected by Gray--code rewrites which preserve entropy admissibility.
\end{definition}

\begin{proposition}
Gray--code lamphrodynamic rewrites minimize semantic distortion under
concurrent merge operations.
\end{proposition}

\begin{proof}[Idea]
Gray--code ordering ensures that successive rewrites differ in only one
bit of boundary entropy, minimizing lamphrodynamic discrepancy.
\end{proof}


\section{Formal Statement of Convergence}

\begin{theorem}[Lamphrodynamic Eventual Consistency]
\label{thm:eventual-consistency}
Let $\{X_i\}$ be a family of lamphrodynamic CRDT replicas. Assume update
rules are merge--collapse admissible, and communication is eventual.
Then there exists a limiting configuration $X_\infty\in\dConfig$ such
that $X_i \rightsquigarrow X_\infty$ for all $i$.
\end{theorem}

\begin{proof}[Sketch]
Repeated merge produces a decreasing lamphrodynamic discrepancy. Collapse
removes internal semantic conflict. Standard fixed--point techniques for
entropy dissipative systems yield a common limit.
\end{proof}


\section{Distributed SpherePop}

Distributed SpherePop generalizes merge--collapse computation to multiple
agents. Each agent holds a derived configuration; updates propagate via
entropy--admissible messages; merge and collapse maintain local
coherence.

\begin{remark}
Unlike traditional distributed computation, lamphrodynamic propagation
imparts geometric structure to communication and ensures physical
coherence across replicas.
\end{remark}


\section{Open Problems}

\begin{enumerate}
\item Characterize necessary and sufficient conditions for
lamphrodynamic convergence.
\item Develop complexity bounds for distributed SpherePop.
\item Explore applications to decentralized lamphrodynamic inference.
\end{enumerate}

\newpage
