% ============================================================
% CHAPTER 7
% ============================================================

\chapter{Ethical Constraints in Geometric Computation}
\label{chap:ethical-constraints}

\section{Motivation}

Geometric computation via merge and collapse lacks an inherent notion of
\emph{ethical admissibility}. Yet distributed computation, especially in
multi--agent systems, must incorporate ethical constraints that prevent
harmful or undesirable outcomes. The central hypothesis of this chapter
is that \emph{ethical norms} can be encoded as \emph{geometric
constraints} on lamphrodynamic evolution, inducing an admissible subset
of configurations.

We investigate the mathematical structure of such constraints and
explore philosophical implications for distributed geometric
computation.


\section{Ethical Admissibility}

Let $\Eth$ denote a collection of ethical conditions expressible as
constraints on lamphrodynamic fields. A configuration is
\emph{ethically admissible} if it satisfies all conditions in $\Eth$.

\begin{definition}
An ethical constraint is a lamphrodynamic inequality of the form
\[
  F(\PhiField,\vField,\SField) \ge 0,
\]
preserved under merge and collapse.
\end{definition}

\begin{remark}
Ethical admissibility resembles entropy admissibility, but may involve
additional state variables or derived invariants.
\end{remark}


\section{Collapse and Salience}

Collapse abstracts internal details. Ethical collapse assigns greater
weight to boundary conditions corresponding to ethically salient
features, effectively reweighing lamphrodynamic fields by ethical
importance.

\begin{proposition}
Under ethical reweighting, collapse preserves ethically salient boundary
conditions while eliminating ethically neutral internal variations.
\end{proposition}


\section{Distributed Ethics}

Distributed lamphrodynamic computation raises the question of whether
ethical admissibility can be guaranteed across multiple agents.
Consistency requires:
\begin{enumerate}
\item local ethical admissibility,
\item global lamphrodynamic compatibility,
\item admissibility under merge and collapse.
\end{enumerate}

\begin{theorem}[Ethical Consistency]
If each agent satisfies ethical admissibility locally, and updates are
lamphrodynamically admissible, then distributed merge--collapse evolution
preserves ethical admissibility.
\end{theorem}

\begin{proof}[Sketch]
Ethical admissibility is preserved by local updates and merge; collapse
removes ethically neutral detail; lamphrodynamic invariance ensures
closure.
\end{proof}


\section{Philosophical Implications}

Interpreting ethical norms as geometric constraints suggests that ethics
is not merely external to computation, but internal to lamphrodynamic
structure. Ethical admissibility becomes a physical constraint analogous
to conservation laws or entropy conditions.

\begin{remark}
This perspective hints at a lamphrodynamic foundation for ethical
agency, linking physical admissibility to normative structure.
\end{remark}


\section{Open Problems}

\begin{enumerate}
\item Characterize ethical invariants preserved by collapse.
\item Develop categorical semantics for ethical admissibility.
\item Investigate lamphrodynamic foundations of moral salience.
\end{enumerate}

\medskip
\noindent
This completes Part~III. We now turn to spectral geometry and cymatics.
\newpage

