\chapter*{Appendix W: Computational Architecture and Execution Semantics}
\addcontentsline{toc}{chapter}{Appendix W: Computational Architecture and Execution Semantics}

\section{Execution Model}

The computational model consists of three layers:

\begin{enumerate}
\item the field-evolution layer implementing RSVP dynamics,
\item the hypergraph-rewrite layer implementing Polyxan EHR rewriting,
\item the reconciliation layer ensuring consistency between the two.
\end{enumerate}

A complete execution state is a triple
\[
(\mathsf{Field},\mathsf{Graph},\mathsf{Embed}),
\]
where the embedding is constrained to remain curvature-matched at all steps.

\section{Intermediate Representations}

\subsection{Field IR}

A field configuration is represented in memory as a structure
\[
\mathsf{FIR} = \{ \Phi_i, v_i, S_i \}_{i = 1}^N,
\]
with each component stored in a contiguous block.  
Curvature, divergence, and gradient operators act via stencil kernels.

\subsection{Hypergraph IR}

The hypergraph representation is:
\[
\mathsf{GIR} = (V,E,\mathrm{tag}),
\]
where tags consist of integer curvature metadata and entropy counters.

EHR rules select subgraphs matching a pattern and replace them according to the reduction rule set.

\subsection{Reconciliation IR}

The representation of patches is a list of atomic modifications:
\[
\mathsf{PIR} = \{ (u \mapsto u')_k \},
\]
with reduction through normalisation.

\section{Scheduling Semantics}

A scheduler assigns time-slices to the three computational layers according to invariant preservation.

\subsection{Harmonic Scheduling}

A time step consists of:
\[
\mathsf{RSVP\_step} \;\rightarrow\; \mathsf{EHR\_step} \;\rightarrow\; \mathsf{Recon\_step}.
\]

The scheduler enforces curvature-decreasing order by requiring:
\[
\mathcal{E}_{\mathrm{RSVP}}(t{+}1) \le \mathcal{E}_{\mathrm{RSVP}}(t),
\qquad
\mathcal{E}_{\mathrm{Polyx}}(t{+}1) \le \mathcal{E}_{\mathrm{Polyx}}(t).
\]

\subsection{Parallel Execution}

Both RSVP and Polyxan layers may be parallelised under the condition that each kernel or rewrite rule is local and maintains modal invariants.  
Reconciliation is global and therefore serial.

\section{Memory Model}

\subsection{Field Memory}

Fields are stored in three aligned arrays to permit SIMD operations.  
All operators use bounded-access stencils to satisfy locality constraints.

\subsection{Hypergraph Memory}

Hypergraph nodes and edges are stored in adjacency lists with immutable tags.  
Rewrite rules produce new nodes but never mutate existing ones.  
Garbage collection is performed via reference-count collapse once a patch becomes idempotent.

\subsection{Patch Memory}

Patches are maintained in normalised lists.  
A normalised patch retains only those atomic edits not eliminated by confluence.

\section{Operational Semantics}

\subsection{Field Step}

The RSVP field update is defined by:
\[
\Phi^{t+1} = \Phi^t - \Delta \Phi^t,
\qquad
v^{t+1} = v^t - \nabla \times (\nabla \times v^t),
\qquad
S^{t+1} = S^t - \nabla\cdot(\nabla S^t).
\]

The step is accepted only if harmonic constraints remain satisfied.

\subsection{Rewrite Step}

For a hypergraph $G$:
\[
G^{t+1} = R(G^t),
\]
where $R$ is one EHR rule satisfying curvature-decrease.

\subsection{Reconciliation Step}

Given two updated structures, the reconciliation operator computes:
\[
(\mathsf{Graph},\mathsf{Field})^{t+1}
  = \mathsf{Recon}((\mathsf{Graph},\mathsf{Field})^{t+1}).
\]

The operator produces a normalised patch and applies it to both structures.

\section{Resource Semantics}

\subsection{Curvature Budget}

Each step is allocated a curvature budget $\mathcal{B}$:
\[
\Delta\mathcal{E}_{\mathrm{RSVP}} + 
\Delta\mathcal{E}_{\mathrm{Polyx}} \le \mathcal{B}.
\]

If the budget would be exceeded, the step is postponed.

\subsection{Entropy Budget}

A similar budget exists for entropy change:
\[
\Delta S_{\mathrm{total}} \le \mathcal{B}_{S}.
\]

\section{Compiler-Level Issues}

\subsection{Static Checks}

A static checker verifies:
\begin{itemize}
\item stencil safety on fields,
\item rewrite termination on hypergraphs,
\item patch confluence,
\item embedding curvature invariance.
\end{itemize}

\subsection{Code Generation}

The compiler produces kernels for:
\begin{itemize}
\item discrete Laplacian,
\item curl-curl operator,
\item entropy-gradient operator,
\item EHR match-and-rewrite rules,
\item reconciliation normalisation.
\end{itemize}

\section{Execution Guarantees}

Every execution of the system satisfies:

\begin{enumerate}
\item local harmonic invariants for fields,
\item local curvature-decrease on hypergraphs,
\item global consistency after reconciliation,
\item finite-time convergence under bounded energy and entropy.
\end{enumerate}

These guarantees are enforced through the invariant-preserving scheduler and the reconciliation operator.


