\chapter*{Appendix A: The Fundamental RSVP PDE System}
\addcontentsline{toc}{chapter}{Appendix A: The Fundamental RSVP PDE System}

\section*{A.1 \; Overview}

The Relativistic Scalar--Vector Plenum (RSVP) framework describes the
co-evolution of three interacting fields:
\begin{itemize}
  \item $\Phi(x,t)$: a scalar \emph{semantic field} measuring local meaning density,
  \item $\mathcal{v}(x,t)$: a vector \emph{flow field} representing baryonic motion,
        embodied agency, or distributed effort,
  \item $S(x,t)$: an \emph{entropy potential} encoding drives, preferences, and
        free-energy gradients.
\end{itemize}

These fields obey a coupled system of nonlinear partial differential equations
representing the minimal classical dynamics underlying all higher-level
RSVP formalisms (Lagrangian, Hamiltonian, categorical, AKSZ, BV,
derived-symplectic, and quine-based constructions).

The purpose of this appendix is to give a compact and self-contained overview
of the basic equations, independent of the advanced formalisms appearing in later
chapters.

\section*{A.2 \; Semantic Field Evolution}

The scalar field $\Phi$ satisfies an advective--diffusive equation with an
entropy-driven source term:
\begin{equation}
\label{eq:RSVP-Phi}
\partial_t \Phi
  = \kappa\,\Delta\Phi
    - \nabla\!\cdot(\Phi\,\mathcal{v})
    + \lambda\,S.
\end{equation}

The terms have the following interpretation:
\begin{itemize}
  \item $\kappa\,\Delta\Phi$ smooths incoherent or noisy semantic structure,
  \item $-\nabla\!\cdot(\Phi\,\mathcal{v})$ advects semantic density along the flow,
  \item $\lambda\,S$ injects semantic structure from entropy gradients.
\end{itemize}

Equation~\eqref{eq:RSVP-Phi} is the RSVP analogue of a Fokker--Planck equation
with meaningful transport and thermodynamic coupling.

\section*{A.3 \; Flow Field Dynamics}

The vector field $\mathcal{v}$ evolves according to a modified Navier--Stokes
equation:
\begin{equation}
\label{eq:RSVP-v}
\partial_t \mathcal{v}
  + (\mathcal{v}\!\cdot\nabla)\mathcal{v}
    = -\nabla S
      + \nu\,\Delta\mathcal{v}
      + \alpha\,\nabla\Phi.
\end{equation}

The terms represent:
\begin{itemize}
  \item $-(\nabla S)$: descent on entropy gradients, expressing drive reduction,
  \item $\nu\,\Delta\mathcal{v}$: viscous damping of overreaction and instability,
  \item $\alpha\,\nabla\Phi$: semantic forces, where meaning gradients attract or
        repel the flow.
\end{itemize}

This equation encodes the RSVP principle that motion is guided jointly by
semantic predictions and entropic demands.

\section*{A.4 \; Entropy Potential Evolution}

The entropy field $S$ satisfies a dissipative balance equation:
\begin{equation}
\label{eq:RSVP-S}
\partial_t S
  = D_S\,\Delta S
    - \beta\,\mathcal{v}\!\cdot\nabla S
    - \gamma\,(\Phi - \Phi^\ast).
\end{equation}

Here:
\begin{itemize}
  \item $D_S\,\Delta S$ allows entropy to diffuse spatially,
  \item $-\beta\,\mathcal{v}\!\cdot\nabla S$ transports entropy along the flow,
  \item $-\gamma(\Phi - \Phi^\ast)$ decreases entropy when predictions match
        expected structure $\Phi^\ast$.
\end{itemize}

Equation~\eqref{eq:RSVP-S} captures the RSVP claim that entropy decreases under
predictive alignment and increases under dissonance.

\section*{A.5 \; Coupled Structure of the System}

Equations \eqref{eq:RSVP-Phi}--\eqref{eq:RSVP-S} form a closed feedback loop:
\[
S \longrightarrow \mathcal{v}
  \longrightarrow \Phi
  \longrightarrow S.
\]

\begin{itemize}
  \item $S$ shapes motion via $\nabla S$,
  \item $\mathcal{v}$ shapes $\Phi$ through advection,
  \item $\Phi$ shapes $S$ through predictive error $(\Phi-\Phi^\ast)$.
\end{itemize}

This minimal loop produces semantic coherence, agency transport, and entropy
regulation without assuming any external controller or pre-existing hierarchy.

\section*{A.6 \; Energy, Stability, and Dissipation}

Each equation obeys a local balance law.  For example:
\[
\partial_t\!\left(\tfrac{1}{2}\Phi^2\right)
  + \nabla\!\cdot(\Phi^2\mathcal{v})
  = -\kappa\,|\nabla\Phi|^2 + \lambda\,S\,\Phi.
\]

Similar identities hold for $\mathcal{v}$ and $S$, implying:
\begin{itemize}
  \item diffusion terms always dissipate energy,
  \item semantic and entropic couplings can inject or remove energy,
  \item long-term behaviour is governed by curvature reduction and entropy flow.
\end{itemize}

These classical energy identities connect directly to the derived BV symplectic
structure and the curvature-minimising normal forms appearing in Chapters~79–80.

\section*{A.7 \; Relation to the Geometric and Derived Formulations}

The PDE system above is not the most fundamental structure of RSVP.
Rather:
\begin{itemize}
  \item it is the classical, continuum limit of the derived AKSZ/BV model,
  \item it is the shadow of the RSVP Lagrangian and Hamiltonian from Appendix~B,
  \item it is recovered from the RSVP quine monad and semantic patch algebra,
  \item it is the local coordinate expression of the categorical dynamics.
\end{itemize}

All higher-level constructions reduce, in the classical limit, to
\eqref{eq:RSVP-Phi}--\eqref{eq:RSVP-S}.  
Thus this appendix serves as the minimal reference point for readers before they
encounter the geometric, categorical, and derived-symplectic machinery of the
later appendices.

\section*{A.8 \; Summary}

The fundamental RSVP PDE system consists of:
\begin{align*}
\partial_t\Phi &= \kappa\,\Delta\Phi
                   - \nabla\!\cdot(\Phi\,\mathcal{v})
                   + \lambda S, \\
\partial_t\mathcal{v}
  &= -\nabla S
     + \nu\,\Delta\mathcal{v}
     + \alpha\,\nabla\Phi
     - (\mathcal{v}\!\cdot\nabla)\mathcal{v}, \\
\partial_t S
  &= D_S\,\Delta S
     - \beta\,\mathcal{v}\!\cdot\nabla S
     - \gamma(\Phi - \Phi^\ast).
\end{align*}

These equations encode:
\begin{itemize}
  \item semantic smoothing and transport,
  \item agency motion through meaning and value gradients,
  \item entropy regulation through predictive alignment.
\end{itemize}

They supply the intuitive physical picture that grounds the more abstract and
verified structures developed throughout the monograph.

