\chapter*{Appendix P: Functorial Semantics}
\addcontentsline{toc}{chapter}{Appendix P: Functorial Semantics}

\section{Categories of Syntactic and Semantic Structures}

\subsection{Polyxan Hypergraphs}

Let $\mathbf{Polyx}$ be the category whose objects are typed Polyxan hypergraphs and whose morphisms are structure-preserving homomorphisms:
\[
\mathbf{Polyx}
=
\left(
\mathrm{Obj}=\{\mathcal{G}\},\ 
\mathrm{Mor}=\{f:\mathcal{G}\to\mathcal{H}\}
\right).
\]
Composition is ordinary morphism composition; identities are the canonical inclusions.

\subsection{Semantic Manifolds}

Let $\mathbf{RSVP}$ be the category whose objects are curvature–entropy–vector-field semantic manifolds and whose morphisms are stratified harmonic maps:
\[
\mathbf{RSVP}
=
\left(
\mathrm{Obj}=\{\mathcal{M}\},\ 
\mathrm{Mor}=\{\varphi:\mathcal{M}\to\mathcal{N}\}
\right).
\]
Morphisms preserve stratification, curvature index, and harmonicity.

\section{Interpretation Functor}

\subsection{Definition}

Define the interpretation functor
\[
\llbracket - \rrbracket : \mathbf{Polyx} \longrightarrow \mathbf{RSVP}
\]
as follows:
\begin{itemize}
\item For each hypergraph $\mathcal{G}$,  
  \[
  \llbracket \mathcal{G} \rrbracket
  = 
  \mathcal{M}_{\mathcal{G}},
  \]
  the canonical semantic manifold equipped with the harmonic embedding of its incidence complex.
\item For each morphism $f:\mathcal{G}\to\mathcal{H}$,  
  \[
  \llbracket f \rrbracket:\mathcal{M}_{\mathcal{G}}\to\mathcal{M}_{\mathcal{H}}
  \]
  is the unique harmonic extension respecting curvature and type data.
\end{itemize}

\subsection{Functoriality}

For hypergraph morphisms $f:\mathcal{G}\to\mathcal{H}$ and $g:\mathcal{H}\to\mathcal{K}$,
\[
\llbracket g\circ f \rrbracket
=
\llbracket g \rrbracket \circ \llbracket f \rrbracket,
\qquad
\llbracket \mathrm{id}_{\mathcal{G}} \rrbracket
=
\mathrm{id}_{\mathcal{M}_{\mathcal{G}}}.
\]

\section{Satisfaction Relation}

\subsection{Definition}

For a semantic manifold $\mathcal{M}$ and a hypergraph $\mathcal{G}$, define
\[
(\mathcal{M},\iota) \models \mathcal{G}
\quad\Longleftrightarrow\quad
\iota : \mathrm{Inc}(\mathcal{G}) \hookrightarrow \mathcal{M}
\ \text{is a stratified harmonic embedding}.
\]

This relation is preserved under $\llbracket - \rrbracket$.

\section{Semantic Institution Structure}

\subsection{Components}

The quadruple
\[
\mathfrak{S}
=
(\mathbf{Polyx},\mathbf{RSVP},\llbracket - \rrbracket,\models)
\]
forms a semantic institution in the sense of category-theoretic model theory:
\begin{itemize}
\item $\mathbf{Polyx}$ provides the syntactic structures,
\item $\mathbf{RSVP}$ provides semantic structures,
\item $\llbracket - \rrbracket$ interprets syntax in semantics,
\item $\models$ relates semantic objects to syntactic ones.
\end{itemize}

\subsection{Satisfaction Condition}

For every morphism $f:\mathcal{G}\to\mathcal{H}$ and embedding $\iota:\mathcal{M}\to\mathcal{N}$,
\[
(\mathcal{M},\iota) \models \mathcal{G}
\quad\Longrightarrow\quad
(\mathcal{N},\iota\circ \llbracket f \rrbracket) \models \mathcal{H}.
\]

\section{Natural Transformations and Evolution}

\subsection{Semantic Evolution}

Let $E:\mathbf{RSVP}\to\mathbf{RSVP}$ denote the evolution endofunctor induced by harmonic flow.

A natural transformation
\[
\theta : \llbracket - \rrbracket \Longrightarrow E\circ \llbracket - \rrbracket
\]
assigns to every hypergraph $\mathcal{G}$ a flow-compatible embedding
\[
\theta_{\mathcal{G}} : \mathcal{M}_{\mathcal{G}} \to E(\mathcal{M}_{\mathcal{G}}).
\]

\subsection{Compatibility with Morphisms}

For every $f:\mathcal{G}\to\mathcal{H}$,
\[
E(\llbracket f \rrbracket)\circ \theta_{\mathcal{G}}
=
\theta_{\mathcal{H}} \circ \llbracket f \rrbracket.
\]

\section{Rewriting and Functoriality}

\subsection{EHR Rewriting as an Endofunctor}

Define the rewriting endofunctor
\[
\mathcal{R} : \mathbf{Polyx} \to \mathbf{Polyx}
\]
by sending each hypergraph to its curvature-reduced form and each morphism to the induced structural map.

\subsection{Interpretation Preservation}

\[
\llbracket \mathcal{R}(\mathcal{G}) \rrbracket
=
E(\llbracket \mathcal{G} \rrbracket),
\]
i.e., rewriting corresponds to semantic flow under interpretation.

\section{Adjunctions}

\subsection{Adjoint Pair}

The functors $\llbracket - \rrbracket$ and the synthetic duality functor $\mathbb{D}:\mathbf{RSVP}\to\mathbf{Polyx}$ form an adjunction:
\[
\llbracket - \rrbracket \dashv \mathbb{D}.
\]

\subsection{Unit and Counit}

The unit $\eta:\mathrm{Id}_{\mathbf{Polyx}}\Rightarrow \mathbb{D}\circ\llbracket - \rrbracket$  
and counit $\varepsilon:\llbracket - \rrbracket\circ \mathbb{D}\Rightarrow \mathrm{Id}_{\mathbf{RSVP}}$  
satisfy the triangle identities.

\section{Equivalence on Stable Objects}

\subsection{Stable Subcategories}

Let $\mathbf{Polyx}_{\mathrm{stable}}$ be the full subcategory of quine-stable hypergraphs and
$\mathbf{RSVP}_{\mathrm{harm}}$ the full subcategory of harmonic semantic manifolds.

\subsection{Equivalence}

\[
\llbracket - \rrbracket :
\mathbf{Polyx}_{\mathrm{stable}}
\stackrel{\simeq}{\longrightarrow}
\mathbf{RSVP}_{\mathrm{harm}}
\]
is an equivalence of categories.


