\documentclass[11pt]{article}

% --------------------------------------------------
% Packages
% --------------------------------------------------
\usepackage{geometry}
\usepackage{setspace}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{csquotes}
\usepackage{microtype}

\geometry{margin=1in}
\setstretch{1.15}

% --------------------------------------------------
% Theorem Environments
% --------------------------------------------------
\newtheorem{definition}{Definition}[section]
\newtheorem{axiom}{Axiom}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{lemma}{Lemma}[section]

% --------------------------------------------------
% Title
% --------------------------------------------------
\title{Spherepop OS:\\
A Deterministic Semantic Operating System\\
\large Formal Specification, Architectural Rationale, and Foundations of Pre-Linguistic Intelligence}
\author{Flyxion}
\date{\today}

\begin{document}
\maketitle

% ==================================================
\begin{abstract}
% ==================================================

Spherepop OS is a deterministic semantic operating system whose primary
abstraction is neither the process, nor the file, nor the thread, but an
append-only, causally ordered event substrate from which all semantic state
is derived. The system is designed to support collaborative, time-aware,
and introspectable computation by enforcing strict replayability, total
causal order, and a principled separation between authoritative state
transitions and non-authoritative views.

This paper presents a comprehensive formal specification of Spherepop OS
together with the architectural rationale underlying each design decision.
Beyond operating-system concerns, the paper situates Spherepop OS within a
broader theoretical context, arguing that its event-sourced semantics
instantiate foundational principles of intelligence articulated by Yi Ma
and others: parsimony, self-consistency, and abstraction grounded in
structure rather than language.

By distinguishing compression as reduction in causal construction from
abstraction as reduction in semantic representatives, Spherepop OS resolves
a longstanding ambiguity in the theory of intelligence. The resulting
architecture functions as a pre-linguistic semantic substrate, positioned
at the level of animal cognition rather than symbolic reasoning. In this
sense, Spherepop OS occupies an intermediate position between an operating
system kernel, a distributed database, and a field-theoretic model of
intelligence grounded in time, construction, and equivalence.
\end{abstract}

% ==================================================
\section{Motivation and Design Orientation}
% ==================================================

Modern operating systems are historically shaped by assumptions that no
longer hold in collaborative, distributed, or semantically rich
computational environments. The dominant abstractions of processes,
address spaces, and mutable global state emerged in an era where computation
was primarily local, sequential, and opaque. While these abstractions have
proven effective for managing hardware resources, they impose severe
limitations on reproducibility, introspection, and principled reasoning
about time.

In particular, traditional systems make it difficult to answer questions
such as: \emph{What exactly happened to produce the current state?} \emph{Can
that state be reconstructed independently?} \emph{Can multiple observers
reason about the same history without interfering with one another?} These
difficulties are not accidental but structural, arising from the use of
mutable state, implicit causality, and side-effect-laden control flow.

Spherepop OS is motivated by the conviction that these limitations can only
be resolved by elevating time and construction to first-class concepts.
Rather than treating history as an implementation detail, Spherepop OS
treats it as the sole source of authority. All semantic state exists only as
the deterministic consequence of replaying an immutable, totally ordered
event log.

This orientation aligns with a deeper theoretical perspective on
intelligence. As emphasized by Yi Ma, intelligent systems are governed by
two foundational principles: parsimony and self-consistency. Parsimony
demands that the system explain the world using as little structure as
necessary, while self-consistency demands that the resulting model be rich
enough to reliably recreate and simulate its environment. Spherepop OS
embeds these principles architecturally rather than heuristically: every
unnecessary construction leaves a permanent trace, and every inconsistency
is exposed through replay.

The design goals of Spherepop OS therefore follow not from performance
considerations but from epistemic ones. The system is designed to guarantee
deterministic replay, enforce a single global causal order, separate causes
from observations, privilege semantic relations over procedural artifacts,
and support collaborative introspection without destabilization. These
commitments collectively rule out large classes of conventional kernel,
database, and version-control architectures, motivating the log-centric
design developed in the sections that follow.

% ==================================================
\section{Core Ontology}
% ==================================================

The ontology of Spherepop OS is intentionally minimal. Rather than
introducing a rich taxonomy of object types, the system defines a small set
of primitives from which all higher-level structure is derived.

\begin{definition}[Semantic Object]
A semantic object is an abstract entity identified by a stable handle.
Semantic objects possess no intrinsic meaning, internal state, or behavior.
Their significance arises solely through participation in relations,
equivalence classes, and events.
\end{definition}

This definition reflects a deliberate refusal to ascribe semantics at the
object level. Objects are not records, files, or processes; they are
structural placeholders whose interpretation emerges only through context.

\begin{definition}[Relation]
A relation is a typed, directed association between two semantic objects.
Relations are first-class entities and may carry auxiliary metadata, but
their semantic force derives from their position within the relational
structure induced by the event log.
\end{definition}

\begin{definition}[Equivalence]
An equivalence relation over semantic objects identifies multiple handles as
representing a single canonical representative. Equivalence is induced
explicitly by merge events and maintained via a union--find structure during
replay.
\end{definition}

Crucially, none of these entities exist independently of the event log.
There is no mutable object table, no hidden registry of relations, and no
authoritative state outside replay. This ensures that the ontology of the
system is entirely historical: to exist is to have been constructed.

% ==================================================
\section{The Authoritative Event Log}
% ==================================================

\begin{axiom}[Append-Only Authority]
All authoritative state transitions in Spherepop OS are recorded as events
in a single append-only log. No other structure is permitted to introduce,
modify, or erase semantic state.
\end{axiom}

Each event is assigned a strictly increasing event identifier by a single
arbiter, establishing a total causal order over all state transitions. This
total order is not an emergent property but an enforced invariant.

\begin{proposition}[Deterministic State Reconstruction]
Given an initial empty state and any finite prefix of the event log, the
resulting semantic state is uniquely determined.
\end{proposition}

\begin{proof}
All kernel transitions are pure functions of prior state and the next event
in the log. Since the log is immutable and totally ordered, replay yields a
unique result.
\end{proof}

This proposition elevates the event log from a storage mechanism to the
primary semantic artifact of the system. The log is not merely a record of
what happened; it \emph{is} what happened.

In the context of intelligence theory, this construction has a significant
implication. Compression in Spherepop OS is not an abstract information
measure but a concrete reduction in causal construction. Abstraction, as
will be shown later, is not symbolic generalization but structural
equivalence over constructed entities. Both are directly visible in the
log, rendering the system introspectable in a way that purely statistical
models are not.

% ==================================================
\section{Kernel Semantics and Deterministic Execution}
% ==================================================

At the heart of Spherepop OS lies a deliberately minimal kernel. Unlike
traditional kernels, which manage processes, memory, and scheduling, the
Spherepop kernel performs a single task: it deterministically interprets the
authoritative event log. All higher-level phenomena—state, structure,
abstraction, and even time—emerge from this interpretation.

\begin{definition}[Kernel]
The kernel is the deterministic transition function that maps an event-log
prefix to a semantic state. It possesses no mutable internal state beyond
that which is derivable from replay, and it produces no side effects beyond
derived views.
\end{definition}

This definition is intentionally austere. The kernel is not a scheduler, a
memory manager, or a policy engine. It is a semantic interpreter whose sole
responsibility is to ensure that identical histories yield identical worlds.

This design enforces self-consistency in the strongest possible sense.
Because the kernel is pure, replayable, and free of hidden state, the system
is always capable of reconstructing its own present from first principles.
There is no distinction between \emph{how} the state was reached and
\emph{what} the state is; the former fully determines the latter.

\subsection{Derived State and Non-Authority}

Although the kernel maintains derived structures—such as equivalence
classes, relation tables, and metadata maps—these structures are explicitly
non-authoritative. They exist only as caches that accelerate observation.
At any time, they may be discarded and reconstructed by replay.

This distinction is essential. In conventional systems, caches frequently
become sources of truth through optimization or necessity. In Spherepop OS,
by contrast, authority is rigidly centralized in the log. The kernel is
therefore immune to a wide class of synchronization errors and semantic
drift.

\subsection{Kernel State Formalization}

To make these claims precise, we formalize the kernel state.

\begin{definition}[Kernel State]
A kernel state is a tuple
\[
\sigma = (O, U, R, M),
\]
where $O$ is a finite set of object identifiers, $U : O \to O$ is a
union--find parent map encoding equivalence, $R \subseteq O \times O \times
\mathcal{T}$ is a set of typed relations, and $M : O \to \mathcal{K}
\rightharpoonup \mathcal{V}$ is a partial metadata map.
\end{definition}

We write $\mathrm{rep}_\sigma(o)$ for the canonical representative of object
$o$ under the equivalence relation induced by $U$. Path compression is
applied deterministically during replay.

\subsection{Operational Semantics of Events}

Kernel execution is defined by a labeled transition relation
\[
\sigma \xrightarrow{e} \sigma',
\]
where $e$ is an event drawn from a fixed and closed set of event types. Each
event type corresponds to a semantic commitment that refines the structure
of the world.

\paragraph{Object Introduction (POP).}
A POP event introduces a new semantic object into existence. The object has
no properties beyond its identity and initial self-representation under
equivalence. This reflects the principle that objects are inert until
related.

Formally, if $o \notin O$, then
\[
(O, U, R, M) \xrightarrow{\mathrm{POP}(o)}
(O \cup \{o\},\; U[o \mapsto o],\; R,\; M).
\]

\paragraph{Equivalence Induction (MERGE).}
A MERGE event identifies two existing objects as equivalent. Importantly,
this operation does not destroy either object; instead, it collapses them
into a shared representative.

If $o_1, o_2 \in O$, then
\[
\sigma \xrightarrow{\mathrm{MERGE}(o_1,o_2)} \sigma',
\]
where the representative of $o_2$ is redirected to that of $o_1$.

This operation is idempotent and confluent. Applying MERGE to already
equivalent objects has no effect, and the final equivalence relation is
independent of merge order. These properties ensure that abstraction is
stable under replay and resistant to race conditions.

\begin{proposition}[Confluence of Equivalence]
The equivalence relation induced by a set of MERGE events depends only on the
set of merges performed, not on their order.
\end{proposition}

\begin{proof}
Union--find computes the transitive closure of the merge relation, which is
order-independent.
\end{proof}

\paragraph{Relation Construction (LINK and UNLINK).}
LINK events introduce directed, typed relations between canonical
representatives of objects. UNLINK events remove such relations. Both
operations are interpreted relative to current equivalence classes,
ensuring that abstraction propagates uniformly through relational structure.

\paragraph{Bulk Abstraction (COLLAPSE).}
The COLLAPSE event generalizes MERGE by inducing equivalence over an entire
region of objects. Conceptually, COLLAPSE corresponds to a deliberate act of
abstraction: many constructed entities are retrospectively identified as
instances of a single concept.

This operation is especially significant from the perspective of
intelligence theory. COLLAPSE does not reduce the number of events that
occurred; instead, it reduces the number of distinct representatives that
matter. As such, it exemplifies abstraction without compression, a
distinction that will become central later in the paper.

\paragraph{Metadata Attachment (SET\_META).}
SET\_META attaches auxiliary information to objects without affecting
semantic structure. Metadata is explicitly non-causal: it does not influence
equivalence, relations, or object existence.

\begin{axiom}[Metadata Orthogonality]
Metadata updates commute with all kernel transitions and have no semantic
effect.
\end{axiom}

\subsection{Determinism and Semantic Stability}

The operational semantics above satisfy strong determinism guarantees.

\begin{theorem}[Kernel Determinism]
For any kernel state $\sigma$ and event $e$, there exists at most one state
$\sigma'$ such that $\sigma \xrightarrow{e} \sigma'$.
\end{theorem}

This theorem underwrites every higher-level property of Spherepop OS. It
ensures that observers, regardless of vantage point or reconstruction
method, converge on the same semantic world when given the same history.

From the perspective articulated by Yi Ma, this determinism is not merely a
technical convenience but a prerequisite for intelligence. A system that
cannot reconstruct its own world consistently cannot form reliable models
of that world. By embedding determinism into its core, Spherepop OS enforces
self-consistency as an architectural invariant rather than a learned
behavior.

% ==================================================
\section{Observation, Diffs, and Non-Interference}
% ==================================================

While the authoritative event log determines all semantic state, a practical
system must permit observation without compromising determinism. Spherepop
OS resolves this tension by introducing a strict separation between
authoritative causation and non-authoritative observation.

\begin{definition}[Diff]
A diff is a derived, non-authoritative description of the semantic change
induced by applying a single event to a kernel state.
\end{definition}

Diffs are emitted by the kernel as observational artifacts. They are never
themselves logged, never replayed, and never consulted during future kernel
execution. Their sole purpose is to permit observers to maintain an
incremental view of semantic state without replaying the entire log.

This separation is enforced by construction rather than convention. The
kernel computes diffs as a pure function of state transition, but diffs have
no causal power. They may be dropped, delayed, reordered, or ignored by
observers without affecting the correctness of the system as a whole.

\begin{axiom}[Observational Non-Interference]
Diffs have no effect on authoritative state and do not influence future
kernel execution.
\end{axiom}

This axiom eliminates an entire class of feedback-induced inconsistencies
that plague conventional systems. Observation is rendered epistemic rather
than ontological: it reveals what has occurred without altering what can
occur.

\subsection{Incremental Reconstruction}

Observers may reconstruct semantic state incrementally by applying diffs in
sequence order. This reconstruction is semantically equivalent to full
replay.

\begin{proposition}[Incremental Equivalence]
Let $\sigma_n$ be the kernel state obtained by replaying events
$e_1,\dots,e_n$ from the empty state. Let $\hat{\sigma}_n$ be the observer
state obtained by applying diffs $d_1,\dots,d_n$. Then $\sigma_n$ and
$\hat{\sigma}_n$ are semantically equivalent.
\end{proposition}

\begin{proof}
Each diff $d_i$ is deterministically derived from the kernel transition
$\sigma_{i-1} \xrightarrow{e_i} \sigma_i$. Because kernel transitions are
pure and diffs preserve all semantic information, incremental application
yields a state equivalent to replay.
\end{proof}

This proposition ensures that no privileged reconstruction path exists.
Replay-based and diff-based observers converge on the same semantic world.

% ==================================================
\section{Snapshots as Derived Views}
% ==================================================

To support late-joining observers and historical inspection, Spherepop OS
permits the construction of snapshots.

\begin{definition}[Snapshot]
A snapshot is a complete serialization of kernel state derived by replaying
a prefix of the event log.
\end{definition}

Snapshots are derived artifacts, not authoritative records. They contain no
information that cannot be reconstructed from the corresponding log prefix.

\begin{axiom}[Snapshot Non-Authority]
Snapshots introduce no new semantic information and have no effect on future
kernel execution.
\end{axiom}

\begin{proposition}[Snapshot Purity]
For any snapshot $S_k$ derived from events $[1..k]$, there exists a unique
kernel state $\sigma_k$ such that $S_k$ is a faithful serialization of
$\sigma_k$.
\end{proposition}

Snapshots therefore serve as epistemic conveniences rather than ontological
commitments. They accelerate access but do not alter meaning.

\subsection{Replay Equivalence}

The relationship between replay, diffs, and snapshots is captured by a
single unifying result.

\begin{theorem}[Replay Equivalence]
Let $\ell = e_1,\dots,e_n$ be an event sequence. Let $\sigma_{\mathrm{replay}}$
be the state obtained by replaying $\ell$ from the empty state. Let
$\sigma_{\mathrm{incr}}$ be the state obtained by applying diffs
$d_1,\dots,d_n$. Then $\sigma_{\mathrm{replay}} = \sigma_{\mathrm{incr}}$ up
to semantic equivalence.
\end{theorem}

\begin{proof}
Both constructions apply the same deterministic kernel transitions in the
same order. Since diffs are derived views and introduce no additional
authority, the resulting states coincide.
\end{proof}

This theorem guarantees global coherence across observers and reconstruction
strategies.

% ==================================================
\section{Seekable Time and Historical Reconstruction}
% ==================================================

A distinctive feature of Spherepop OS is explicit support for temporal
navigation. Rather than treating history as opaque or irreversible, the
system exposes time as a first-class query dimension.

\begin{definition}[Seek Operation]
A seek operation $\mathrm{seek}(k)$ constructs an ephemeral kernel state by
replaying the event log prefix $[1..k]$.
\end{definition}

Seek operations instantiate temporary kernel instances that are isolated
from the live system. They permit historical inspection, debugging, and
counterfactual reasoning without side effects.

\begin{axiom}[Temporal Isolation]
Seek operations produce no modifications to authoritative state and emit no
diffs.
\end{axiom}

\begin{corollary}
Historical inspection cannot influence present or future semantic state.
\end{corollary}

\subsection{Temporal Monotonicity}

The authoritative event sequence is strictly monotone. No operation may
decrease, rewrite, or fork the authoritative timeline.

\begin{proposition}[Temporal Monotonicity]
If $i < j$, then $\mathrm{eid}(e_i) < \mathrm{eid}(e_j)$.
\end{proposition}

This enforces a strong notion of causality: the past is immutable, and the
future is always constructed relative to a fixed history.

% ==================================================
\section{Self-Consistency and World Simulation}
% ==================================================

The architectural commitments described above jointly enforce a strong form
of self-consistency. At any moment, the system can reconstruct its entire
semantic state from first principles by replaying the event log.

\begin{theorem}[Strong Self-Consistency]
For any time $t$, the semantic state of Spherepop OS is uniquely determined
by the event log prefix $[1..t]$.
\end{theorem}

This property directly instantiates the second foundational principle of
intelligence articulated by Yi Ma: a world model must be consistent enough
to support accurate simulation. In Spherepop OS, simulation is not an
approximation layered atop mutable state; it is an intrinsic consequence of
replay.

Because seek operations permit reconstruction of past states, and
speculative branches (introduced in the next section) permit exploration of
possible futures, the system possesses an internal model of its own
temporal evolution. This model operates below language and symbolism. It is
not a narrative or a description, but a constructive history that can be
replayed, inspected, and extended.

In this sense, Spherepop OS implements self-consistency not as a learning
objective but as a structural invariant. Any deviation from consistency
would manifest immediately as replay failure or divergence, rendering the
system epistemically unstable. By design, such instability is impossible.

% ==================================================
\section{Speculative Branches and Counterfactual Evaluation}
% ==================================================

An intelligent system must be capable not only of reconstructing its past
but of reasoning about possible futures. Spherepop OS addresses this
requirement through the introduction of speculative branches, which permit
counterfactual exploration without compromising causal integrity.

\begin{definition}[Speculative Branch]
A speculative branch is defined by a base event identifier $k$ together with
a client-local sequence of hypothetical events that are evaluated against
the authoritative history but not committed to it.
\end{definition}

Speculative branches are evaluated by replaying the authoritative log prefix
$[1..k]$ and then applying the speculative events in sequence. Crucially,
this process uses the same deterministic kernel as authoritative execution.
Speculation is therefore not a heuristic approximation but a faithful
simulation of what would occur if the proposed events were to be accepted.

\begin{axiom}[Speculative Isolation]
Speculative branches have no effect on authoritative state unless their
events are explicitly validated and sequenced by the arbiter.
\end{axiom}

This axiom ensures that counterfactual reasoning remains epistemic rather
than ontological. Speculation explores possibilities without altering
reality.

\begin{proposition}[Counterfactual Soundness]
Speculative evaluation preserves semantic consistency with the base history.
\end{proposition}

\begin{proof}
Speculative execution reuses the same deterministic transition rules as
authoritative replay. Since speculative events are isolated, they cannot
introduce contradictions into the authoritative timeline.
\end{proof}

From the standpoint of intelligence theory, speculative branching plays a
role analogous to mental simulation in biological systems. The system can
evaluate the consequences of potential actions without enacting them, a
capacity widely regarded as central to adaptive behavior.

% ==================================================
\section{Compression and Abstraction as Distinct Structural Operations}
% ==================================================

A recurring theme in contemporary discussions of artificial intelligence is
the ambiguity between compression and abstraction. Yi Ma has emphasized
this ambiguity as a central open problem, asking what distinguishes mere
data compression from genuine understanding.

Spherepop OS provides a structural resolution to this question by making
both operations explicit and orthogonal within the event-derived semantic
substrate.

\subsection{Compression as Reduction in Causal Construction}

In Spherepop OS, compression is defined operationally rather than
statistically.

\begin{definition}[Compression]
Compression is the reduction in the number of events required to construct a
given semantic state.
\end{definition}

Compression is therefore visible directly in the event log. Two histories
that yield identical semantic states but differ in event count exhibit
different degrees of compression. Reuse of existing objects, avoidance of
redundant constructions, and deliberate minimization of causal steps all
manifest as shorter logs.

This notion of compression aligns with parsimony as articulated by Yi Ma.
The system is forced to confront the cost of every construction: unnecessary
operations leave permanent traces in the log. There is no opportunity to
retroactively erase inefficiency.

\subsection{Abstraction as Reduction in Semantic Representatives}

Abstraction, by contrast, operates on a different structural dimension.

\begin{definition}[Abstraction]
Abstraction is the reduction in the number of distinct semantic
representatives required to produce a given semantic effect.
\end{definition}

Abstraction is realized in Spherepop OS through equivalence operations.
MERGE and COLLAPSE events identify multiple constructed objects as
representing the same underlying entity. Importantly, abstraction does not
remove events from history; it reinterprets their consequences by reducing
the number of distinct representatives that matter going forward.

\begin{proposition}[Orthogonality of Compression and Abstraction]
Compression and abstraction are independent structural operations. One may
increase without increasing the other.
\end{proposition}

\begin{proof}
A system may reuse a single object many times, achieving high compression
with minimal abstraction. Conversely, a system may construct many distinct
objects and later collapse them into a single representative, achieving
high abstraction without reducing event count. The operations act on
different aspects of the event-derived structure.
\end{proof}

This distinction resolves Yi Ma’s question without appealing to semantics or
behavior. Compression concerns \emph{how much happened}; abstraction
concerns \emph{how many distinct things matter}. Both are visible in the
log, and neither subsumes the other.

% ==================================================
\section{Avoiding the Language Compression Trap}
% ==================================================

Yi Ma has argued that contemporary large language models operate primarily
by compressing already-compressed signals. Natural language itself is the
product of extensive abstraction and parsimony, encoding generations of
world modeling into symbolic form. Models trained exclusively on text
therefore inherit abstraction without participating in its formation.

Spherepop OS avoids this trap by operating below language. Its primitives
precede symbols, syntax, and narration. Objects are not words, relations are
not predicates, and equivalence is not semantic synonymy but structural
identification.

\begin{theorem}[Pre-Linguistic Abstraction]
All abstraction in Spherepop OS arises from explicit equivalence operations
over constructed entities rather than from symbolic compression.
\end{theorem}

\begin{proof}
Equivalence is induced only by MERGE and COLLAPSE events applied to objects
that have been constructed through prior events. No symbolic encoding or
linguistic representation is required.
\end{proof}

This positions Spherepop OS at the level of intelligence Yi Ma associates
with animals and humans prior to language: a world model built through
interaction, regularity, and structural consolidation rather than textual
description.

% ==================================================
\section{Parsimony and Self-Consistency as Architectural Invariants}
% ==================================================

Yi Ma identifies parsimony and self-consistency as the governing principles
of intelligence. In Spherepop OS, these principles are not optimization
objectives or emergent behaviors; they are architectural invariants.

Parsimony is enforced by the append-only log. Every construction is
accounted for, and inefficiency cannot be hidden. Self-consistency is
enforced by deterministic replay. Any inconsistency would manifest as
divergence between replayed and observed state, an outcome that is
structurally prohibited.

\begin{theorem}[Invariant Satisfaction]
Spherepop OS satisfies parsimony and self-consistency by construction.
\end{theorem}

\begin{proof}
Parsimony follows from the irreversibility of event construction, while
self-consistency follows from deterministic replay over an immutable log.
\end{proof}

Together, these invariants ensure that the system’s internal model of the
world remains both minimal and reliable, capable of supporting simulation,
counterfactual reasoning, and abstraction without collapse.

% ==================================================
\section{Geometry, Layout, and Gauge Freedom}
% ==================================================

Spherepop OS enforces a strict conceptual separation between semantic
structure and geometric presentation. This separation is not a matter of
implementation convenience but a foundational design principle intended to
prevent accidental entanglement between meaning and visualization.

Geometric information—such as spatial layout, clustering, or visual
proximity—is treated as non-causal metadata. Observers may attach, modify,
or discard such information freely without affecting semantic state.

\begin{definition}[Layout Hint]
A layout hint is non-authoritative metadata associated with a semantic
object that proposes a geometric or visual interpretation of that object.
\end{definition}

Layout hints are never consulted by the kernel during event interpretation.
They do not influence equivalence, relations, or object existence. As such,
they commute with all kernel operations.

\begin{axiom}[Geometric Gauge Freedom]
For any two assignments of layout hints, replaying the same event log yields
identical semantic state.
\end{axiom}

This axiom parallels gauge freedom in physics: multiple geometric
representations may correspond to the same underlying structure. The
semantic content of the system is invariant under changes of layout gauge.

This distinction is essential for collaborative environments. Different
observers may adopt radically different visualizations suited to their
tasks, cognitive preferences, or display constraints, while remaining in
perfect semantic agreement.

% ==================================================
\section{Observers and Causal Synchronization}
% ==================================================

Observers in Spherepop OS are epistemic agents rather than causal actors.
They receive information about state transitions but do not participate in
authoritative execution.

\begin{definition}[Observer]
An observer is an external process that receives diffs derived from kernel
state transitions and maintains a local semantic view.
\end{definition}

Observers are passive with respect to causation. They cannot modify kernel
state directly and cannot introduce events into the authoritative log.

All diffs are delivered to observers in strictly increasing event order.
This delivery discipline ensures that every observer witnesses a causally
consistent history.

\begin{proposition}[Observer Convergence]
Any two observers that process the same prefix of diffs converge to
semantically equivalent states.
\end{proposition}

\begin{proof}
Since diffs are deterministically derived from kernel transitions and are
applied in the same order, the resulting observer states differ only in
non-authoritative metadata.
\end{proof}

Late-joining observers may be bootstrapped using snapshots followed by
diffs, achieving eventual convergence without special handling.

% ==================================================
\section{Arbiter and Collaborative Commit Model}
% ==================================================

While observation is decentralized, authority in Spherepop OS is centralized
by design.

\begin{definition}[Arbiter]
The arbiter is the sole entity authorized to append events to the
authoritative log.
\end{definition}

Clients submit proposals to the arbiter, which validates them against the
current kernel state and, if accepted, assigns them sequence identifiers and
commits them to the log.

\begin{axiom}[Single Sequencer]
There exists exactly one arbiter per authoritative log.
\end{axiom}

This axiom enforces total causal order. Unlike distributed consensus systems
that permit partial ordering or eventual convergence, Spherepop OS commits
to a single, linear history.

\begin{proposition}[Proposal Determinism]
Given identical kernel state and proposal content, the arbiter produces the
same canonical event or rejection.
\end{proposition}

This property ensures fairness, reproducibility, and auditability. The
arbiter is not a policy engine but a deterministic mediator of semantic
commitment.

% ==================================================
\section{Computational Complexity and Scalability}
% ==================================================

The computational properties of Spherepop OS follow directly from its
kernel design. Because all operations are defined in terms of union--find
and relation table updates, complexity bounds are well understood.

\begin{proposition}[Kernel Operation Complexity]
Object introduction is constant time. Equivalence induction via MERGE
operates in inverse-Ackermann time amortized. Relation creation and removal
scale logarithmically with the number of relations. Bulk collapse scales
linearly with the size of the collapsed region, up to inverse-Ackermann
factors.
\end{proposition}

\begin{proposition}[Replay Complexity]
Replaying an event log of length $\ell$ requires time linear in $\ell$ up to
inverse-Ackermann factors determined by equivalence maintenance.
\end{proposition}

These bounds ensure that long-lived semantic histories remain tractable,
even as abstraction accumulates. Importantly, abstraction tends to reduce
effective state complexity by consolidating representatives, mitigating the
cost of replay over time.

% ==================================================
\section{Rationale and Comparative Perspective}
% ==================================================

Spherepop OS occupies a position that is difficult to classify within
existing computational paradigms. It is not a conventional operating system
in the POSIX sense, as it does not manage hardware resources or schedule
processes. It is not a database, as it does not support arbitrary mutation
or query-driven updates. It is not a version control system, as speculation
is continuous rather than episodic.

Instead, Spherepop OS is best understood as an operating system for semantic
time. It manages the construction, identification, and stabilization of
meaning over time through explicit events, equivalence, and replay.

In contrast to language-centric artificial intelligence systems, Spherepop
OS does not begin with symbols. It begins with construction. Meaning emerges
only after structure has been assembled, reused, and abstracted. This
architectural choice aligns Spherepop OS with theories of intelligence that
locate understanding in interaction and structure rather than in linguistic
compression.



ection{Accumulation Regimes and the Stratification of Intelligence}
% ==================================================

A central contribution of Spherepop OS is that it renders visible, within a
single formal substrate, multiple distinct regimes of knowledge
accumulation that are often conflated in both artificial intelligence and
cognitive science. These regimes correspond closely to the stratification
identified by Yi Ma—phylogenetic, ontogenetic, social, and abstract—but are
expressed here in strictly operational and architectural terms.

The key observation is that accumulation is not a unitary process. Different
forms of intelligence arise from different mechanisms of compression,
stabilization, and reuse, each operating on different timescales and with
different degrees of internal autonomy. Spherepop OS provides a framework in
which these differences are not merely conceptual but structurally enforced
and empirically observable.

\subsection{Phylogenetic Accumulation as Log Externalization}

Phylogenetic accumulation corresponds to the slow incorporation of
environmental regularities into the structure of an organism through
evolutionary processes. In informational terms, this is a brutal and highly
lossy form of compression, mediated by selection rather than inference.

Within the Spherepop OS framework, phylogenetic accumulation is best
understood as \emph{log externalization}. Structural commitments are fixed
outside the active semantic substrate and appear to the system as immutable
primitives. The system does not record the evolutionary history that produced
these structures; it merely inherits them as given.

This regime is characterized by the absence of introspectable construction
history. The reasons for a structure’s existence are not available within
the system itself. As a result, phylogenetic knowledge is maximally stable
but minimally revisable. It supports survival but not explanation.

Spherepop OS deliberately does not model this regime internally. Instead, it
assumes a fixed kernel and primitive operation set, mirroring the way
biological organisms inherit neural architectures without access to their
evolutionary derivation.

\subsection{Ontogenetic Accumulation as Event-Sourced Construction}

Ontogenetic accumulation corresponds to individual learning through
interaction with the environment. This is the regime in which Yi Ma locates
animal and human intelligence at its core.

Spherepop OS is explicitly designed to operate at this level. Ontogenetic
learning is modeled as the accumulation of events in an authoritative log,
with semantic objects, relations, and equivalence classes emerging through
replay. Crucially, this accumulation is internally driven. While external
inputs may trigger events, the structural consequences of those events are
determined by the system’s existing semantic organization.

Compression at this level is not mere reduction of data volume. It is the
progressive elimination of redundant constructions through equivalence,
reuse, and abstraction. The system learns not by memorizing observations but
by stabilizing invariants across events.

Because the full construction history remains available, ontogenetic
knowledge in Spherepop OS is introspectable, revisable, and replayable. This
property sharply distinguishes it from both evolutionary inheritance and
static database storage.

\subsection{Social Accumulation as Shared Log Alignment}

Social accumulation arises when multiple systems align their semantic
substrates through shared symbols, conventions, and artifacts. Language,
writing, and institutions belong to this regime.

From the perspective of Spherepop OS, social knowledge corresponds to partial
log alignment across agents. Shared abstractions emerge when distinct event
histories induce compatible equivalence classes and relations. Importantly,
this does not require identical histories, only compatible semantic
commitments.

This framing reveals why social knowledge is powerful but fragile. Because
it operates above individual ontogenetic logs, it depends on continuous
maintenance, interpretation, and repair. Misalignment does not appear as a
logical contradiction but as semantic drift.

Spherepop OS supports this regime by design through its observer model.
Multiple observers may maintain divergent views, layouts, and speculative
branches while remaining anchored to a common authoritative core.

\subsection{Abstraction and the Transition to Formal Knowledge}

The highest regime identified by Yi Ma is the transition from empirical
compression to abstract, deductive knowledge—science and mathematics. This
transition represents a qualitative shift rather than a quantitative
extension.

In Spherepop OS terms, this corresponds to the emergence of abstraction
layers that operate primarily on representatives rather than on raw events.
Once equivalence classes stabilize, reasoning can proceed by manipulating
canonical objects without revisiting their construction histories.

This transition is visible in the log as a change in event character. Early
logs are dominated by object creation and relation formation. Mature logs
exhibit a preponderance of merges, collapses, and high-level relational
rewriting. The system increasingly reasons about its own abstractions.

Crucially, this phase transition does not eliminate the underlying event
substrate. Formal reasoning remains grounded in replayable construction,
preventing abstraction from drifting into incoherence.

\subsection{Why This Architecture Resolves the Compression--Abstraction
Problem}

Yi Ma poses a fundamental open question: what distinguishes compression from
abstraction? Spherepop OS resolves this question structurally rather than
definitionally.

Compression corresponds to a reduction in the number of events required to
construct a semantic effect. Abstraction corresponds to a reduction in the
number of distinct representatives required to account for those effects.
Both processes are visible, measurable, and enforceable within the log.

This distinction explains why large language models, which operate primarily
on already-compressed linguistic artifacts, struggle to achieve genuine
understanding. They compress signals that have already undergone social and
ontogenetic abstraction, without access to the underlying construction
history.

Spherepop OS, by contrast, operates below language. It records the formation
of meaning itself, rather than its symbolic residue. In doing so, it occupies
the pre-linguistic substrate of intelligence that Yi Ma associates with
animal cognition and world modeling.

\subsection{Implications for Artificial Intelligence}

By grounding intelligence in event-sourced semantic construction rather than
symbolic manipulation, Spherepop OS avoids the trap of conflating memorization
with understanding. Intelligence emerges not from scale alone, but from the
capacity to accumulate, stabilize, and reorganize structure over time.

This perspective suggests that progress toward artificial general
intelligence will require systems that expose and preserve their
construction histories, rather than systems that merely optimize predictive
performance on compressed corpora.

Spherepop OS offers a concrete architectural instantiation of this idea,
providing a substrate on which learning, abstraction, and reasoning can be
studied as unified phenomena.



ection{A Field-Theoretic Interpretation of Semantic Construction}
% ==================================================

The architecture of Spherepop OS admits a natural interpretation in
field-theoretic terms. This interpretation is not metaphorical but
structural: the semantic state induced by the event log behaves as a
dynamical field whose configuration evolves through localized
interactions, stabilizes into attractors, and exhibits gauge freedom in
its observational representations.

This perspective provides a unifying language for understanding learning,
abstraction, and coherence within the system, and connects Spherepop OS to
a broader tradition of field-based theories in physics, cognition, and
cybernetics.

\subsection{Semantic State as a Field Configuration}

At any sequence index $k$, the kernel state $\sigma_k = (O_k, U_k, R_k, M_k)$
induces a semantic configuration over the space of object handles. Rather
than viewing this state as a static data structure, it is more appropriate
to regard it as a field assigning semantic potentials to relations between
objects.

Relations act as couplings, equivalence classes as condensations, and
metadata as gauge-dependent annotations. The replay operation corresponds
to evolving this field forward in discrete time according to deterministic
local update rules.

Importantly, the field is not defined over physical space but over a
semantic manifold whose coordinates are given by object identity and
relational context. The geometry of this manifold is emergent rather than
prescribed.

\subsection{Events as Local Field Perturbations}

Each event in the log introduces a localized perturbation to the semantic
field. Object creation introduces new degrees of freedom; relation
creation alters coupling terms; merge and collapse events reduce
dimensionality by identifying degrees of freedom.

These perturbations propagate through the field according to the kernel’s
deterministic transition rules. Because all events are replayed in a total
order, the resulting field evolution is well-defined and path-independent.

Learning, in this view, corresponds to the progressive reshaping of the
field so that subsequent perturbations produce smaller global effects.
The system becomes increasingly insensitive to noise and increasingly
responsive to structurally meaningful changes.

\subsection{Attractors and Semantic Stability}

As event accumulation proceeds, the semantic field develops attractors:
regions of configuration space toward which diverse event sequences
converge. These attractors correspond to stable abstractions—equivalence
classes and relational motifs that persist across contexts.

The existence of attractors explains how abstraction emerges without
explicit symbolic reasoning. Abstractions are not imposed; they are the
fixed points of semantic dynamics under repeated construction.

This view aligns with empirical observations in biological cognition, where
concepts appear as stable patterns of activation rather than as discrete
symbols. Spherepop OS provides an explicit operational substrate in which
such patterns can be defined, observed, and replayed.

\subsection{Gauge Freedom and Observational Equivalence}

A central principle of field theories is gauge freedom: multiple
representations may correspond to the same physical state. Spherepop OS
exhibits an analogous property with respect to layout, geometry, and other
non-causal metadata.

Different observers may assign different geometric embeddings or visual
layouts to the same semantic field configuration without altering its
authoritative content. These choices correspond to gauge transformations.

This interpretation clarifies why visualization, speculation, and
exploration can be safely decoupled from semantic authority. The kernel
operates on gauge-invariant quantities; observers are free to choose gauges
appropriate to their tasks.

\subsection{Relation to Cybernetics and Predictive Models}

The field-theoretic view also connects Spherepop OS to cybernetic theories
of intelligence. Feedback, error correction, and prediction correspond to
the system’s capacity to compare expected semantic effects with observed
perturbations and to reorganize its field accordingly.

However, unlike classical cybernetic models, Spherepop OS makes the
construction history explicit. Prediction is not merely statistical but
structural, grounded in replayable semantic commitments.

This distinction is crucial. It explains how intelligence can arise from
compression without collapsing into mere memorization, and why abstraction
requires access to generative history rather than surface regularities.


% ==================================================
\section{Conclusion}
% ==================================================

Spherepop OS presents a deterministic semantic operating system grounded in
causal clarity, replayability, and structural abstraction. By elevating the
event log to the sole source of authority and rigorously separating causes
from observations, the system supports collaboration, introspection, and
counterfactual reasoning without sacrificing consistency.

More importantly, Spherepop OS provides a concrete architectural answer to
foundational questions in the theory of intelligence. Compression is revealed
as reduction in causal construction; abstraction as reduction in semantic
representatives. Both are enforced structurally and rendered visible in the
system’s history.

By operating below language and symbolism, Spherepop OS occupies a
pre-linguistic stratum of intelligence, corresponding to the level at which
world models are formed through interaction and consolidation rather than
description. In this sense, the system is not merely an operating system but
a semantic substrate upon which more familiar cognitive and computational
structures may be built.

% ==================================================
% Appendices
% ==================================================

\appendix

\section{Determinism and Confluence}

\begin{theorem}
For any finite prefix of the authoritative event log, the resulting semantic
state is unique.
\end{theorem}

\begin{proof}
By induction on the length of the log prefix, using kernel determinism.
\end{proof}

\section{Compression--Abstraction Duality}

The dual roles of compression and abstraction are independent and jointly
necessary. Compression governs the efficiency of construction; abstraction
governs the stability of meaning. Neither alone suffices to produce robust
intelligence.

\section{Relation to Foundational Theories of Intelligence}

Yi Ma’s principles of parsimony and self-consistency are instantiated
architecturally within Spherepop OS. Norbert Wiener’s cybernetic emphasis
on feedback and control is extended through explicit temporal replay and
counterfactual branching.



%ppendix
\section{Formal Distinction Between Compression and Abstraction}
% ==================================================

This appendix formalizes a distinction that is central to the theoretical
claims of the paper: the difference between compression and abstraction.
While these terms are often used interchangeably in the literature, they
correspond to distinct structural transformations in Spherepop OS.

\subsection{Preliminaries}

Let $\ell = (e_1, \ldots, e_n)$ be a prefix of the authoritative event log,
and let $\sigma(\ell)$ denote the kernel state obtained by replaying $\ell$
from the empty initial state.

Let $|\ell|$ denote the length of the log prefix, and let $|O(\ell)|$ denote
the number of object identifiers present in $\sigma(\ell)$. Let
$|U(\ell)|$ denote the number of equivalence classes induced by the union--
find structure.

\subsection{Compression as Event Reduction}

\begin{definition}[Semantic Effect]
A semantic effect is an equivalence class of kernel states under
observational equivalence, where two states are observationally equivalent
if they induce the same canonical representatives and relations.
\end{definition}

\begin{definition}[Compression]
Compression is the process of reducing the minimal log length required to
produce a given semantic effect.
\end{definition}

Formally, compression occurs when there exists a log prefix $\ell'$ such
that $\sigma(\ell')$ is observationally equivalent to $\sigma(\ell)$ and
$|\ell'| < |\ell|$.

Compression therefore operates on the temporal dimension of construction.
It removes redundant or unnecessary events while preserving semantic
outcomes.

\subsection{Abstraction as Representative Reduction}

\begin{definition}[Abstraction]
Abstraction is the process of reducing the number of distinct canonical
representatives required to account for a semantic effect.
\end{definition}

Formally, abstraction occurs when, for a fixed semantic effect,
$|U(\ell')| < |U(\ell)|$, typically as the result of merge or collapse
events.

Abstraction operates on the structural dimension of the semantic field. It
does not shorten the log but simplifies the ontology induced by the log.

\subsection{Independence of Compression and Abstraction}

\begin{proposition}
Compression and abstraction are logically independent operations.
\end{proposition}

\begin{proof}
Compression may occur without abstraction by removing redundant events that
introduce objects which are later unused, without altering equivalence
classes. Conversely, abstraction may occur without compression when merge
events increase log length while reducing the number of representatives.

Therefore, neither operation implies the other.
\end{proof}

\subsection{Visibility in the Log}

\begin{proposition}
The distinction between compression and abstraction is observable in the
event log and kernel state.
\end{proposition}

\begin{proof}
Compression is observable as a reduction in minimal log prefixes producing
a given semantic effect. Abstraction is observable as a reduction in the
cardinality of equivalence classes in $\sigma(\ell)$. Both quantities are
explicitly computable from the log and kernel state.
\end{proof}

\subsection{Implications for Learning Systems}

This formal distinction explains why systems that optimize for compression
alone may fail to achieve abstraction. Statistical models that reduce data
volume without altering underlying representational commitments compress
signals without restructuring semantics.

By contrast, abstraction requires the system to identify and stabilize
equivalences across contexts, an operation that cannot be performed solely
at the level of surface compression.

Spherepop OS enforces this distinction architecturally, ensuring that
abstraction remains grounded in replayable construction rather than in
opaque parameter adjustment.



ection{Category-Theoretic Structure of Spherepop Semantics}
% ==================================================

This appendix provides a categorical interpretation of the semantic
machinery of Spherepop OS. The goal is not abstraction for its own sake,
but to clarify which aspects of the system are structural invariants and
which are implementation artifacts.

\subsection{Objects, Morphisms, and Construction Histories}

Let $\mathcal{C}$ be a category whose objects correspond to semantic object
handles and whose morphisms correspond to construction relations induced
by events. Each event introduces either a new object, a new morphism, or a
constraint identifying existing objects.

The event log $\ell = (e_1, \ldots, e_n)$ induces a diagram $D_\ell$ in
$\mathcal{C}$ whose shape is determined by the dependency structure of
events. Replay corresponds to the progressive extension of this diagram.

Importantly, the category $\mathcal{C}$ is not assumed to be small or
finitely generated a priori; its finiteness at any given time is an
emergent property of the log prefix.

\subsection{POP as Object Introduction}

A POP event introduces a new object $o$ into the category, corresponding to
the free adjunction of an object. Categorically, this is a left adjoint to
the forgetful functor that removes isolated objects.

This interpretation clarifies why object creation is irreversible in the
authoritative log: left adjoints do not, in general, admit inverses.

\subsection{LINK as Morphism Introduction}

A LINK event introduces a morphism between two objects, enriching the
diagram without collapsing structure. Morphisms are directed and typed,
corresponding to labeled arrows in $\mathcal{C}$.

LINK events preserve object identity and do not alter equivalence classes.
They increase relational density without reducing dimensionality.

\subsection{MERGE and COLLAPSE as Coequalizers}

MERGE and COLLAPSE events impose equivalence relations between objects.
Categorically, these operations correspond to the formation of
coequalizers.

Given objects $o_1, o_2$ with parallel morphisms or semantic roles, a MERGE
event enforces their identification by introducing a coequalizer object
representing their quotient. COLLAPSE generalizes this to finite families.

The resulting quotient object is universal with respect to the imposed
identification, explaining the confluence and order-independence of merge
operations.

\subsection{Semantic State as a Colimit}

For any log prefix $\ell$, the kernel state $\sigma(\ell)$ corresponds to
the colimit of the diagram $D_\ell$. Replay computes this colimit
incrementally.

This perspective yields a concise explanation of determinism: colimits, if
they exist, are unique up to isomorphism. Different replay strategies
compute the same semantic object.

\subsection{Views as Functorial Projections}

Derived views, snapshots, and diffs correspond to functors from
$\mathcal{C}$ into categories of representations, such as graphs,
geometric spaces, or visual layouts.

These functors need not be faithful. Lossy projections are permitted,
provided they preserve semantic equivalence. This formalizes the principle
that views are observational and non-authoritative.

\subsection{Speculation as Diagram Extension}

Speculative branches correspond to local extensions of the diagram
$D_\ell$ without committing to the colimit. Such extensions exist in a
slice category over the base diagram and may be discarded without
affecting the canonical colimit.

This explains why speculation is structurally safe: it never alters the
universal property defining semantic state.

% ==================================================
\section{Thermodynamic Constraints on Semantic Construction}
% ==================================================

This appendix situates Spherepop OS within the framework of
non-equilibrium thermodynamics. The purpose is to clarify the physical
interpretation of irreversibility, determinism, and abstraction in the
system.

\subsection{Event Logs as Irreversible Processes}

The append-only nature of the authoritative log defines an arrow of time.
Each committed event increases the length of the log and cannot be
removed without violating determinism.

This irreversibility mirrors the second law of thermodynamics. The log
plays the role of an entropy ledger, recording irreversible semantic work
performed by the system.

\subsection{Replay as Reversible Computation}

While log growth is irreversible, replay itself is reversible in the
computational sense. Given a log prefix, the kernel can reconstruct past
states exactly.

This separation mirrors the distinction between microscopic reversibility
and macroscopic irreversibility in statistical mechanics. The kernel is a
reversible interpreter acting on an irreversible history.

\subsection{Abstraction as Entropy Reduction}

Merge and collapse events reduce the number of effective degrees of
freedom in the semantic state. From an informational perspective, these
operations correspond to entropy reduction within the semantic field.

However, this reduction is not free. It requires work in the form of
explicit merge events, which increase log length. Thus, abstraction trades
temporal entropy (longer history) for structural simplicity (fewer
representatives).

This tradeoff mirrors Landauer’s principle: erasure of distinctions incurs
a thermodynamic cost.

\subsection{Self-Consistency and Free Energy}

A semantic state is self-consistent if replay of new events produces
localized effects rather than global reorganization. Such states minimize
semantic surprise and stabilize abstractions.

This condition parallels the minimization of free energy in predictive
systems. The system maintains a structured internal model that absorbs
perturbations efficiently.

Spherepop OS enforces this condition structurally rather than statistically.
Self-consistency is not inferred but constructed through replayable
commitments.

\subsection{Why Statistical Compression Is Insufficient}

Purely statistical compression reduces description length but does not
alter the generative structure responsible for producing observations.
Such compression operates on already-formed signals.

By contrast, semantic abstraction in Spherepop OS operates at construction
time, before signals are rendered. This explains why compression at the
level of natural language or sensory data cannot substitute for
pre-linguistic semantic construction.

\subsection{Non-Equilibrium Steady States}

Over long histories, systems operating under Spherepop semantics tend
toward non-equilibrium steady states characterized by stable abstractions
and bounded perturbation propagation.

These states are neither static nor maximally entropic. They are sustained
by continual semantic work, analogous to dissipative structures in
thermodynamics.


% ==================================================
% Bibliography
% ==================================================

\begin{thebibliography}{9}

\bibitem{ma2023}
Y.~Ma,
\emph{Learning Deep Representations of Data Distributions},
MIT Press, 2023.

\bibitem{wiener1948}
N.~Wiener,
\emph{Cybernetics: Or Control and Communication in the Animal and the Machine},
MIT Press, 1948.

\end{thebibliography}

\end{document}

